# Causal Structure: Drawing the Graph

## Welcome {.unnumbered}

Back in Chapter 6 we fitted a model of sales on customer rating, read
the slope, and then added a caveat that we didn't do anything about:

> One more caveat worth saying out loud, because a break-even
> calculation quietly assumes it: the slope describes an
> **association** between rating and sales among current salespeople.
> Treating it as what *would* happen if you intervened to raise ratings
> is a causal claim the data alone doesn't support.

That was honest, and it was also a dead end. It told you what you
*couldn't* say without telling you what it would take to say it. This
chapter is the missing half.

The good news is that it takes less machinery than you'd expect. You
don't need an experiment, and you don't need a new modelling framework —
the `brm()` skeleton is the same one you've been using since Chapter 5.
What you need is a way to write down what you believe about how the
world works, and a rule for reading off which variables belong in the
model. That's it. The hard part isn't the statistics.

::: callout-important
The core idea in one sentence: **which variables you control for is a
question about causes, not about model fit — and no amount of data will
answer it for you.**
:::

## What you'll be able to do by the end {.unnumbered}

1.  Draw a DAG — a diagram of what you believe causes what — and read it
    as a set of falsifiable claims rather than a picture
2.  Recognise the three structures that generate every confounding
    problem you'll meet: the fork, the chain and the collider
3.  Explain why "control for everything you have" is wrong, and name the
    two distinct ways it goes wrong
4.  Use `dagitty` to derive the set of variables you must adjust for
5.  Fit that adjusted model and state what the posterior means — and
    what it still assumes
6.  Recognise that **every analysis of employee data is already
    conditioned on a collider**, and say something sensible about it

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)
library(ggdag)
library(dagitty)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na()
```

Two new packages. `ggdag` draws causal diagrams and `dagitty` does the
graph theory underneath — works out, from the diagram, which variables
you need to adjust for. Neither fits models; they operate purely on your
assumptions, before any data is touched. That separation is the whole
point.

------------------------------------------------------------------------

## Why "control for everything" is wrong

Ask most analysts how to decide what goes in a model and you'll get one
of two answers. Either *put in everything you've got and let the model
sort it out*, or *put in whatever improves the fit statistics*. Both are
reasonable-sounding, and both are wrong in ways that matter.

The fit-statistics answer fails because Chapter 7's LOO comparison
answers a different question than the one being asked. LOO tells you
which model predicts held-out data better. A model can predict beautifully
and give you a completely wrong answer about what would happen if you
intervened — those are different jobs, and nothing in the model output
distinguishes them.

The everything-you've-got answer fails for a subtler reason: **some
variables make your estimate worse when you adjust for them.** Not
"add noise" worse. Systematically, confidently wrong worse. And they
look exactly like the helpful ones from inside the regression output.

Three small simulations make this concrete. In each one we know the true
answer because we generated the data, which is the only situation where
you can actually check whether a method works.

------------------------------------------------------------------------

## The three structures

### The fork: a common cause

The familiar one. Something causes both your predictor and your outcome,
manufacturing an association between them that isn't a causal path.

```{r sim-fork}
n <- 2000

tenure <- rnorm(n)                                                  # <1>
rating <- 0.6 * tenure + rnorm(n)                                   # <2>
sales  <- 0.8 * tenure + rnorm(n)                                   # <3>

fork <- tibble(tenure, rating, sales)
```

1.  Standardised, so every coefficient below is on the same scale and
    the numbers are directly comparable.
2.  Longer-tenured salespeople get better customer ratings.
3.  Longer-tenured salespeople also sell more — and note what is *not*
    in this line. `rating` does not appear. **The true causal effect of
    rating on sales is exactly zero.**

```{r slope-helper}
#| include: false
# Pull one named coefficient out of an lm() into a one-row tibble, so the
# comparisons below read as a table rather than two blocks of model output.
slope_of <- function(fit, term, label) {
  s <- summary(fit)$coefficients[term, ]
  tibble(model = label, estimate = s[["Estimate"]], se = s[["Std. Error"]])
}
```

```{r sim-fork-fits}
bind_rows(
  slope_of(lm(sales ~ rating, data = fork),          "rating", "rating alone"),
  slope_of(lm(sales ~ rating + tenure, data = fork), "rating", "rating + tenure")
)
```

The first model reports a solid positive relationship between rating and
sales. It is not a small effect, it is not marginal, and if you saw it in
a real dataset you would believe it. It is also entirely an artefact of
tenure. Adjusting for tenure returns the estimate to approximately zero,
which is the truth.

This is the case everyone knows, and it's the one that motivates the
instinct to control for everything. So far so good.

### The chain: a mediator

Now the same picture with the arrows rearranged.

```{r sim-chain}
manager_quality <- rnorm(n)
engagement      <- 0.7 * manager_quality + rnorm(n)                 # <1>
attrition_risk  <- -0.6 * engagement + rnorm(n)                     # <2>

chain <- tibble(manager_quality, engagement, attrition_risk)
```

1.  Good managers raise engagement.
2.  Engaged people are less likely to leave. Again, note what's absent:
    `manager_quality` doesn't appear here. Managers affect attrition
    **only** through engagement — every bit of their effect flows down
    this path.

```{r sim-chain-fits}
bind_rows(
  slope_of(lm(attrition_risk ~ manager_quality, data = chain),
           "manager_quality", "manager_quality alone"),
  slope_of(lm(attrition_risk ~ manager_quality + engagement, data = chain),
           "manager_quality", "manager_quality + engagement")
)
```

Here the *first* model is right and the second is wrong. Manager quality
genuinely reduces attrition risk — that's how we built the data — and
the first model finds it. Adjusting for engagement makes managers look
irrelevant, because we've held constant the very thing through which
they act.

::: callout-important
Same regression output, opposite conclusion about which model to trust.
Nothing inside the model tells you which situation you're in. Tenure was
a confounder to be adjusted for; engagement is a mediator to be left
alone. They are statistically indistinguishable and causally opposite.
:::

This one causes problems constantly in People Analytics, because the variables
sitting in an HRIS are mostly *downstream* of the thing being studied.
Engagement, performance rating, promotion, internal mobility — these are
consequences at least as often as they are causes. Throwing them into a
model as "controls" is not neutral.

### The collider: a common effect

The third structure is the one almost nobody has an intuition for, and
it's the most important one in this field.

```{r sim-collider}
aptitude      <- rnorm(n)                                           # <1>
interpersonal <- rnorm(n)

screen_score <- aptitude + interpersonal + rnorm(n, sd = 0.5)       # <2>
hired        <- screen_score > quantile(screen_score, 0.8)          # <3>

collider <- tibble(aptitude, interpersonal, hired)
```

1.  In the applicant pool, aptitude and interpersonal skill are
    generated independently — knowing one tells you nothing about the
    other.
2.  The hiring process weighs both.
3.  The top 20% get offers.

```{r sim-collider-cors}
tibble(
  population = cor(collider$aptitude, collider$interpersonal),
  among_hires = with(filter(collider, hired), cor(aptitude, interpersonal))
)
```

In the applicant pool the correlation is essentially zero. Among the
people who were hired it is solidly negative — and nothing about anyone
changed. We didn't condition on aptitude or on interpersonal skill. We
conditioned on *being hired*, which both of them cause.

The mechanism, once you see it, is obvious: to clear a bar that sums two
things, a candidate who is weak on one must be strong on the other. The
weak-on-both candidates were filtered out, and they were the ones holding
the lower-left corner of the scatterplot.

```{r collider-plot}
#| code-fold: true
collider |>
  mutate(status = if_else(hired, "Hired", "Not hired")) |>
  ggplot(aes(aptitude, interpersonal, colour = status)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(data = ~filter(.x, status == "Hired"),
              method = "lm", se = FALSE, colour = red) +
  scale_colour_manual(values = c("Hired" = navy, "Not hired" = "grey80")) +
  labs(
    title = "A relationship created entirely by hiring",
    subtitle = "Independent in the applicant pool; negatively related among hires",
    x = "Aptitude", y = "Interpersonal skill", colour = NULL
  )
```

::: callout-important
## The uncomfortable general case

You can only analyse people your organisation hired and who haven't yet
left. Both of those are caused by many of the variables you care about.
**Every dataset in this book, and every dataset on your laptop, is
already conditioned on at least one collider.**

That isn't a reason to stop working. It is a reason to stop being
surprised when a validated selection test shows no relationship with
performance among your employees, or when two things that obviously ought
to go together don't. Range restriction, the "why does our assessment
have zero validity" problem, and the four-stage selection story that
distorts observed performance distributions are all this structure,
described in different vocabularies.
:::

------------------------------------------------------------------------

## A DAG is a set of claims

Now the notation. A DAG — directed acyclic graph — is those pictures,
written down formally enough that software can reason about them.

```{r first-dag}
sales_dag <- dagify(
  sales  ~ rating + tenure,
  rating ~ tenure,
  exposure = "rating",
  outcome  = "sales"
)

sales_dag |>
  tidy_dagitty(seed = 2026) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_colour = "grey60") +
  geom_dag_point(colour = navy, size = 18) +
  geom_dag_text(colour = "white", size = 3.2) +
  theme_dag()
```

Three rules for reading it, in order of how often they're forgotten:

1.  **An arrow is a claim that one thing causes another.** Not
    "correlates with", not "is associated with". Causes.
2.  **A missing arrow is a stronger claim than a present one.** Drawing
    an arrow says "there might be an effect here." Leaving one out says
    "there is *no* direct effect here, and I'm willing to be held to
    that." Most of a DAG's content is in what you didn't draw.
3.  **Acyclic means no loops.** You cannot have A causing B causing A.
    Real organisations are full of feedback loops, and handling them
    properly means adding time — engagement in Q1 affecting performance
    in Q2 affecting engagement in Q3 — so the arrows always point
    forward.

::: callout-note
## A different lens: two dialects of the same language

If you came to this from econometrics, DAGs may feel like an unfamiliar
route to familiar destinations. The econometrics tradition typically
works from a structural equation and an identification argument stated in
prose — omitted variable bias, exogeneity, exclusion restrictions. The
graphical tradition, associated with Judea Pearl, works from the picture
and derives the same conclusions as graph theory.

These are dialects, not rival theories. Omitted variable bias is the
fork. Bad control is the chain and the collider. Sample selection bias —
Heckman's problem — is conditioning on a collider. The graphical version
buys you two things: a diagram a non-technical stakeholder can argue with,
and software that finds the adjustment set for you rather than requiring
you to reason it out by hand each time. If you already think fluently in
the algebraic dialect, treat this chapter as translation rather than
instruction.
:::

::: callout-tip
## For the ML/DS crowd

Feature selection and adjustment-set selection are different operations
with different objectives, and using one for the other is the single most
common way a good predictive modeller produces a bad causal estimate.

Feature selection asks: which inputs improve held-out prediction?
Adjustment-set selection asks: which inputs, if held constant, isolate
the causal path I care about? A collider will often *improve* predictive
accuracy — it carries information — while destroying your ability to
interpret the coefficient of interest. Regularisation doesn't help;
shrinking a bad control toward zero shrinks the bias too, but doesn't
remove it. Neither does more data: bias from adjusting on a collider
doesn't shrink with $n$, it just gets measured more precisely.

The practical tell: if you cannot state, for each variable in your model,
*why it is there in causal terms*, you have a prediction model and should
not be reading its coefficients.
:::

------------------------------------------------------------------------

## Letting the graph choose the model

The payoff of writing assumptions down formally is that the adjustment
set becomes a computation instead of an argument.

```{r adjustment-sets}
adjustmentSets(sales_dag, exposure = "rating", outcome = "sales")
```

Given that graph, adjusting for tenure is sufficient to identify the
effect of rating on sales. Change the graph and the answer changes —
which is the point. The graph is doing the arguing, out in the open,
where a stakeholder can disagree with a specific arrow rather than with
your model as a whole.

Now a graph for the actual data. We have `sales`, `customer_rate` and
`performance` in `salespeople`, and no tenure variable. Suppose we
believe that a salesperson's underlying capability — captured imperfectly
by their performance rating — drives both the ratings customers give
them and the sales they make:

```{r salespeople-dag}
sp_dag <- dagify(
  sales         ~ customer_rate + performance,
  customer_rate ~ performance,
  exposure = "customer_rate",
  outcome  = "sales",
  labels = c(
    sales         = "Sales",
    customer_rate = "Customer\nrating",
    performance   = "Performance"
  )
)

sp_dag |>
  tidy_dagitty(seed = 2026) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_colour = "grey60") +
  geom_dag_point(colour = navy, size = 20) +
  geom_dag_text(aes(label = label), colour = "white", size = 2.8) +
  theme_dag()

adjustmentSets(sp_dag, exposure = "customer_rate", outcome = "sales")
```

::: callout-important
This graph is a **hypothesis**, and a contestable one. You could argue
the arrow between performance and customer rating runs the other way, or
that both are caused by something we haven't measured. Those are real
objections and the DAG doesn't settle them — it *states* the position
clearly enough to be objected to. That is a large improvement on the
usual alternative, which is an unstated position embedded in a variable
list.
:::

### Look at the arrows before you fit anything

A fork only distorts your estimate if **both** of its arrows are
substantial. A common cause that barely moves the predictor, or barely
moves the outcome, has little to bias away. Worth two lines before
committing to a model:

```{r check-the-arrows}
salespeople |>
  summarise(
    `performance -> customer_rate` = cor(performance, customer_rate),
    `performance -> sales`         = cor(performance, sales)
  )
```

Keep those two numbers in mind — they predict how much the adjustment
below can possibly change.

### Fitting it

Same skeleton as every model since Chapter 5, one extra term. We fit
both versions here so the difference is visible in one place rather than
requiring you to flip back:

```{r fit-adjusted}
salespeople <- salespeople |>
  mutate(customer_rate_c = customer_rate - mean(customer_rate))     # <1>

priors <- c(
  prior(normal(400, 200),   class = Intercept),                     # <2>
  prior(normal(0, 100),     class = b),
  prior(exponential(0.005), class = sigma)
)

fit_unadjusted <- brm(
  sales ~ customer_rate_c,
  data = salespeople, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 20, refresh = 0
)

fit_adjusted <- brm(
  sales ~ customer_rate_c + performance,                            # <3>
  data = salespeople, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 20, refresh = 0
)
```

1.  Centred exactly as in Chapter 6, so the intercept stays
    interpretable.
2.  The same priors as Chapter 6, deliberately. The only thing that
    differs between these two models is the variable list, which is the
    whole argument of the chapter.
3.  `performance` is a 1–4 rating and we're treating it as continuous
    here, which is a simplification — Chapter 18 does it properly. It
    doesn't change the point being made.

Steps 5, 7 and 8 of the workflow are deliberately quiet here, and that is
worth stating rather than leaving as a gap. The **priors are Chapter 6's,
unchanged and already visualised there** — the annotation above says so,
because the whole argument depends on the two models differing in
nothing but their variable list. **Diagnostics and the posterior
predictive check** apply exactly as they did in Chapter 6, and for the
same reason: this is the same Gaussian model with one extra term.

What this chapter adds is a step the earlier chapters could not reach.
Step 2 — deciding what you are claiming — is the subject of everything
above, and step 10 follows immediately below.

```{r compare-adjusted}
adjustment_comparison <- bind_rows(
  fit_unadjusted |> spread_draws(b_customer_rate_c) |>
    median_qi(.width = 0.95) |> mutate(model = "rating alone (Chapter 6)"),
  fit_adjusted |> spread_draws(b_customer_rate_c) |>
    median_qi(.width = 0.95) |> mutate(model = "adjusted for performance")
) |>
  mutate(model = factor(model, levels = c("rating alone (Chapter 6)",
                                          "adjusted for performance")))

adjustment_comparison |>
  select(model, estimate = b_customer_rate_c, .lower, .upper)
```

```{r compare-adjusted-plot}
#| code-fold: true
ggplot(adjustment_comparison, aes(x = b_customer_rate_c, y = fct_rev(model))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  geom_linerange(aes(xmin = .lower, xmax = .upper),
                 colour = navy_light, linewidth = 3) +
  geom_point(colour = navy, size = 3) +
  labs(
    title = "The same slope, before and after adjusting for performance",
    subtitle = "Point = posterior median; bar = 95% credible interval. Red dashed = no effect.",
    x = "Sales per point of customer rating", y = NULL
  )
```

### Both models fit, and that is exactly the problem

Chapter 10's step 7 asks whether the model reproduces the data. Run it on
both:

```{r ppc-adjusted}
#| code-fold: true
pp_check(fit_adjusted, ndraws = 50) +
  labs(
    title = "Posterior predictive check: the adjusted model",
    subtitle = "Thin lines = datasets the model would generate; thick line = the real data",
    x = "Sales", y = "Density"
  )
```

Swap `fit_adjusted` for `fit_unadjusted` and run it again. The chart is
essentially identical — which is the entire point of this section.

::: callout-important
## The most important negative result in the chapter

The posterior predictive check passes. It also passed for the unadjusted
model. It would have passed for a model adjusted for a **collider** too —
adjusting for one of the simulations at the top of this chapter would
likely have *improved* the fit while making the causal estimate
systematically wrong.

That is worth sitting with, because it contradicts an instinct the previous
ten chapters have carefully built. Everything up to Chapter 20 has treated
"does the model fit the data?" as the question that decides whether to trust
a result. Here it decides nothing. Both specifications describe the data
they were given perfectly adequately, and one of them answers the question
that was asked.

So the check still belongs in the workflow — it catches a badly specified
likelihood, and that failure is real. It just cannot be asked to do a job
it has no access to. **No diagnostic computed from your data can tell you
whether your adjustment set is right, because the data doesn't contain that
information.** The DAG does, and the DAG came from a conversation.
:::

### When the number doesn't move

If you were expecting the adjustment to overturn Chapter 6's answer, it
didn't. The two estimates are close and their intervals overlap almost
entirely — which the correlations above should have led you to expect, and
which the chart makes obvious in a way the two rows of numbers don't.

That overlap is the finding, and it is worth being precise about how to
show it. Two intervals sitting on top of each other is a picture anyone can
read; two pairs of decimal numbers invites a stakeholder to seize on the
difference between them, which here is noise. Plot the comparison and the
right conclusion is the one that's easiest to reach.

That is a genuinely useful outcome, and it is worth being careful about
what it does and doesn't mean.

::: callout-important
## "So the adjustment was pointless?"

No — and this is the most important paragraph in the chapter.

Before the adjustment, we had a number nobody could interpret. Chapter 6
said so explicitly and left it there: the slope described an association,
and whether it was also an intervention effect was unknown and
unknowable from the model alone.

After the adjustment, we have a number that means something specific:
*given this graph*, it is the effect of raising a salesperson's customer
rating while holding their performance fixed. The number is similar. The
**claim** is completely different, and only one of the two versions can
be put in front of someone about to spend money.

You could not have known in advance that it wouldn't move. That's the
point of doing it. An analysis that changes the estimate and an analysis
that confirms it are the same analysis — you just don't get to skip it on
the grounds that you expect the second outcome.
:::

::: callout-note
## What to say when it *does* move

It often will. When it does, the temptation is to present the adjusted
model as the better one, and that framing invites an argument you don't
want to have.

Both models are correct answers to different questions. The unadjusted
number answers *among our current salespeople, how much more do the
highly-rated ones sell?* — the right number for a forecast, or for
deciding who to talk to. The adjusted number answers *if we raised one
person's rating while their underlying performance stayed the same, how
much more would they sell?* — the right number for deciding whether to
fund a customer-experience programme.

Say which question you answered. That sentence prevents more
misinterpretation than any diagnostic in this book.
:::

------------------------------------------------------------------------

## What this still assumes

The posterior above is a causal effect **conditional on the graph being
right**. It carries every assumption in the diagram, including all the
arrows we didn't draw. That's not a weakness of the method; it's the
method being explicit about something every regression assumes silently.

The assumption most likely to fail is that there is no unmeasured common
cause. In People Analytics there usually is one — motivation, manager
support, the state of the territory — and it is usually unmeasured
precisely because it's hard to measure.

Two honest responses, neither of which is "add more controls":

**Say what would have to be true.** Rather than claiming no confounding,
state how strong an unmeasured confounder would need to be to overturn
your conclusion. If the answer is "it would need to be more strongly
related to sales than performance is," most stakeholders can judge that
for themselves. Searching for *sensitivity analysis for unmeasured
confounding*, *E-values* and *tipping-point analysis* will find the
formal versions.

**Put a prior on it.** This is where the Bayesian framing earns its keep:
an unmeasured confounder can be included in the model as a parameter with
an informative prior representing your beliefs about its strength, and
the posterior for your effect then integrates over that uncertainty
instead of ignoring it. Chapter 23's elicitation protocol is the natural
way to get that prior, and Chapter 14's `prior_strength` discussion is the
same conversation about how much a belief should count.

------------------------------------------------------------------------

## The graph as an elicitation artefact

Worth pausing on, because it's the reason this chapter sits where it
does. A DAG is not really a statistical object. It's a record of what a
group of people believe about how their organisation works — and drawing
one with your stakeholders does something that no amount of modelling
does afterwards.

It surfaces disagreement early. Two HRBPs who both nod along to
"engagement drives retention" will draw different arrows when asked
whether manager quality acts *through* engagement or alongside it, and
that disagreement is worth more than a fit statistic. It also makes the
control-variable decision a business conversation rather than a technical
one, which is exactly where it belongs — it is a question about causes,
and they know more about the causes than you do.

This is the same move as Chapter 23's prior elicitation, applied to
structure rather than to magnitude. Same workshop, same people, often the
same hour. One produces the priors; the other produces the model
specification.

::: callout-caution
## Try it yourself

1.  Redraw `sp_dag` with the arrow between `performance` and
    `customer_rate` reversed, so customer ratings feed the performance
    rating rather than the other way round. Rerun `adjustmentSets()`.
    What changes, and what does that imply about the model you should
    fit?
2.  Add an unmeasured node `capability` that causes all three observed
    variables (`latent = "capability"` in `dagify()`). Does an
    adjustment set still exist?
3.  Take the collider simulation and vary the selection rate from 0.8
    down to 0.2. How does the induced correlation among hires change as
    the process gets more selective?
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Most of the questions People Analytics gets asked are causal, and almost
none of them are phrased that way. *Does our manager training work? Is
flexible working hurting promotion? Would raising pay reduce attrition?*
Each one is an intervention question dressed as a data question.

The value of this chapter isn't that it lets you answer them
definitively — often the honest answer is that the data can't. It's that
it lets you say precisely **what you'd need in order to answer**, which
is a far better deliverable than a regression coefficient with a caveat
attached. "We can estimate this if we can measure territory quality;
without it, the estimate is biased upward by an unknown amount" is a
sentence that gets budget for measuring territory quality.

It also gives you a principled defence for a variable list, which you
will need the first time someone senior asks why you didn't control for
performance rating.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Which variables to control for is a **causal** question. Fit
    statistics can't answer it, and neither can more data.
2.  A **fork** (common cause) creates bias you fix by adjusting.
3.  A **chain** (mediator) creates bias you fix by *not* adjusting —
    controlling for a mediator hides the effect you're looking for.
4.  A **collider** (common effect) creates bias when you adjust for it,
    including when the adjustment is invisible because it happened in
    the selection of your sample.
5.  Every employee dataset is conditioned on at least one collider,
    because you only observe people who were hired and haven't left.
6.  A DAG makes those assumptions explicit and contestable, and
    `dagitty::adjustmentSets()` turns them into a variable list.
7.  The resulting posterior is a causal effect **given the graph** —
    state the graph, and state what would have to be true for it to be
    wrong.
:::

## Next chapter {.unnumbered}

**Causal designs: finding variation you didn't create** — the DAG tells
you what you'd need to adjust for. Sometimes you can't, because the
confounder is unmeasured or unmeasurable. The next chapter covers what
to do instead: difference-in-differences, event studies and the family of
designs that exploit variation the organisation generated for its own
reasons — policy changes, staggered rollouts, thresholds — rather than
variation you had to assume away.
