# Causal Designs: Finding Variation You Didn't Create

## Welcome back {.unnumbered}

Chapter 21 ended on an assumption we couldn't check. The adjustment set
identifies a causal effect *given the graph*, and the graph asserts there
is no unmeasured common cause. In People Analytics there almost always
is one — motivation, manager support, whatever made someone volunteer for
the programme — and it's unmeasured precisely because it's hard to
measure.

Adjusting your way out of that is not available. You cannot control for a
variable you don't have.

What *is* available, more often than people expect, is variation the
organisation created for its own reasons. A policy that started on a
particular date. A rollout that reached the Northern region in March and
the Southern in September. A bonus threshold that applies at exactly 24
months' service. None of these were designed as experiments, and all of
them produce comparisons where the thing you care about moved for a
reason that has nothing to do with the people it moved for.

This idea has a name and a long history. In *Mostly Harmless
Econometrics*, Angrist and Pischke put **natural experiments** near the
front of the book, and anyone who reads economics papers will recognise
the pattern: a striking number of them open not with a hypothesis but
with a discovery — the authors explaining what changed, for whom, and
when, before any model appears. Finding the natural experiment *is* the
contribution. The estimation afterwards is comparatively routine.

Two habits sit behind that, and both are rarer outside economics than
they should be:

-   **An insistence on identifying causality.** "Correlation doesn't
    imply causation" is treated as the beginning of the problem rather
    than the end of the discussion. The interesting question is what
    would have to be true for a causal reading to be earned, and whether
    anything in the data can get you there.
-   **A willingness to work with found data** — data that nobody designed
    an experiment to produce, generated by an organisation going about
    its business.

People Analytics has the second in abundance and has borrowed very little
of the first. Almost everything we work with is found data, and the field
has largely inherited the caveat without the toolkit that was built to
get past it. Which is an odd place to end up, because the caveat is the
discouraging half.

That's the trick of this chapter. Stop trying to statistically remove
confounding, and start looking for places the organisation removed it for
you by accident.

::: callout-important
The core idea in one sentence: **if you can find a moment when the
treatment changed but the people didn't, the comparison does the work
that adjustment couldn't.**
:::

## What you'll be able to do by the end {.unnumbered}

1.  Recognise the three kinds of accidental variation worth looking
    for — a date, an order, a threshold
2.  Estimate a **difference-in-differences** model in `brms` and say
    exactly what it assumes
3.  Build an **event study** to interrogate the parallel-trends
    assumption instead of asserting it
4.  Know why **staggered rollouts** break the obvious model, and what to
    search for
5.  Recognise when an instrument or a threshold gives you a design, and
    when it doesn't
6.  Run a **simulation-based design analysis** before collecting data —
    and say, in advance, whether the study you're being asked to run can
    answer the question at all

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(21)

# Shared colour tokens (matching theme/academicdesign*.scss). The navy ramp
# carries the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_mid   <- "#3d68a8"
navy_light <- "#8fabd0"
red        <- "#d32f2f"
```

No `peopleanalyticsdata` this time. Every dataset in that package is a
cross-section — one row per person, no time dimension and no policy
change — and the designs in this chapter need both. So we simulate.

That's not a compromise. It's the same discipline as Chapter 21's three
structures: when we generate the data, we know the true effect, and we
can check whether a method recovers it. You will never have that luxury
with real data, which is exactly why it's worth spending time with
simulated data while you're learning what these methods do.

------------------------------------------------------------------------

## Three kinds of accidental variation

Before any modelling, the thing to develop is an eye for the
opportunity. Three patterns account for most of what you'll find in an
organisation:

**A date.** Something changed for some people at a known moment. A
return-to-office mandate, a new parental leave policy, a change to the
bonus formula, a reorganisation that moved one function under new
leadership. The comparison is before-versus-after, for the affected group
against an unaffected one. This is difference-in-differences, and it's
most of the chapter.

**An order.** The same change reached different groups at different
times — a manager training programme rolled out region by region over
eighteen months, a new HRIS deployed in waves. Early groups act as a
comparison for later ones. This is a staggered rollout, and it is more
delicate than it looks.

::: callout-tip
## The cheapest win available to you

Rollouts are usually sequenced by operational convenience — whichever
region is ready first, whichever director shouted loudest. Almost nobody
asks what sequence would make the result *measurable*.

If you have the standing to be in that conversation, this is the highest
return-per-minute intervention in the whole book. You are not asking for
a budget, a headcount or a delay. You are asking that the order be
decided a little more deliberately, and ideally that one group goes last
rather than never — which costs nothing, because they were always going
to be last anyway.

Do it once and the organisation gets an answer it would otherwise never
have had. It is also a rare chance to make the function meaningfully more
evidence-driven through a single decision, taken before anything happens,
rather than through an analysis produced after the fact that nobody is
in a position to act on.
:::

**A threshold.** A rule applies at a sharp cutoff. Share options vest at
exactly two years. Employees below a rating of 3 go onto a performance
plan. A bonus multiplier kicks in at 110% of target. People just either
side of the line are near-identical, and one group got the treatment.
This is a regression discontinuity.

::: callout-tip
The practical version of this chapter is a habit rather than a technique:
**when someone asks a causal question, ask what changed and when.** Most
of the time the answer is "nothing" and you're back to Chapter 21. But it
is asked far too rarely, and organisations change things constantly.
:::

------------------------------------------------------------------------

## Difference-in-differences

### The situation

A company mandates three days a week in the office. It applies to the
sales and marketing functions from July; engineering is exempt and
carries on as before. Leadership wants to know what the mandate did to
engagement.

The obvious analysis — compare engagement in sales before and after — is
confounded by everything else that happened in July. The other obvious
analysis — compare sales to engineering after July — is confounded by
every way those two functions differ. Difference-in-differences uses both
comparisons to cancel each other out.

```{r simulate-panel}
n_offices <- 40
months    <- 1:12
policy_month <- 7
true_effect  <- -4                                                  # <1>

offices <- tibble(
  office      = 1:n_offices,
  treated     = as.integer(office <= n_offices / 2),               # <2>
  office_base = rnorm(n_offices, mean = 70, sd = 6)                # <3>
)

panel <- expand_grid(offices, month = months) |>
  mutate(
    post   = as.integer(month >= policy_month),
    trend  = 0.4 * month,                                          # <4>
    engagement = office_base + trend +
      true_effect * treated * post +                               # <5>
      rnorm(n(), sd = 3)
  )
```

1.  The mandate costs four engagement points. We know this because we're
    writing it; the whole exercise is checking whether the method finds
    it.
2.  Half the offices are in the affected functions.
3.  Offices differ from each other, permanently and substantially — a
    six-point spread against a four-point effect. This is the confounding
    that a simple after-only comparison would fall into.
4.  Engagement is drifting upward over the year for everyone, treated or
    not. This is the confounding that a simple before-and-after
    comparison would fall into.
5.  The effect only exists for treated offices, only after the policy.

### The 2×2, before any model

Difference-in-differences is worth meeting as four numbers before it's a
regression, because the regression obscures how simple it is:

```{r the-2x2}
cell_means <- panel |>
  group_by(treated, post) |>
  summarise(engagement = mean(engagement), .groups = "drop")

cell_means |>
  pivot_wider(names_from = post, values_from = engagement,
              names_prefix = "period_") |>
  mutate(change = period_1 - period_0)
```

Each group's change over time is the first difference. The difference
*between* those changes is the estimate — hence the name:

```{r the-did-by-hand}
changes <- cell_means |>
  pivot_wider(names_from = post, values_from = engagement,
              names_prefix = "period_") |>
  mutate(change = period_1 - period_0)

diff(changes$change)                                               # <1>
```

1.  Treated group's change minus control group's change. Compare it to
    the `true_effect` of −4 we built in.

The control group's change absorbs everything that happened to everyone —
the upward drift, the season, whatever else July brought. What's left is
what happened to the treated group *and nobody else*.

### Look at it first

The 2×2 collapses twelve months into two numbers, which is the right
summary and the wrong first look. Before modelling, plot it — the same
habit as every other chapter, and doubly worth it here because the shape
over time is what the design's credibility rests on:

```{r look-at-the-panel}
group_means <- panel |>
  group_by(treated, month) |>
  summarise(engagement = mean(engagement), .groups = "drop") |>
  mutate(group = if_else(treated == 1,
                         "Mandated (sales & marketing)",
                         "Exempt (engineering)"))

ggplot(group_means, aes(month, engagement, colour = group)) +
  geom_vline(xintercept = policy_month - 0.5,
             linetype = "dashed", colour = red) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = months) +
  scale_colour_manual(values = c("Mandated (sales & marketing)" = navy,
                                 "Exempt (engineering)"         = navy_mid)) +
  labs(
    title = "Engagement by month, mandated vs exempt functions",
    subtitle = "Red dashed = the mandate. Both groups drift up; one of them steps down in July.",
    x = "Month", y = "Mean engagement", colour = NULL
  ) +
  theme(legend.position = "top")
```

Two features matter, and they map onto the two confounders we built in.
Both lines drift upward — that's the trend a naive before-and-after
comparison would have charged to the policy. And the lines sit at
different heights throughout — that's the permanent difference between
functions a naive after-only comparison would have charged to it. The
policy is the *step*, not the level and not the slope.

### Priors

Same discipline as everywhere else: look at them before fitting. The
parameter we care about is the difference-in-differences term, so that's
the prior worth seeing:

```{r plot-prior-did}
#| code-fold: true
tibble(effect = seq(-40, 40, length.out = 400)) |>
  mutate(density = dnorm(effect, mean = 0, sd = 10)) |>
  ggplot(aes(effect, density)) +
  geom_area(fill = navy, alpha = 0.30) +
  geom_line(colour = navy, linewidth = 1) +
  geom_vline(xintercept = true_effect, linetype = "dashed", colour = red) +
  labs(
    title = "The prior on the policy effect: Normal(0, 10)",
    subtitle = "Red dashed = the effect we simulated. Centred on no effect, comfortably wide enough to find one.",
    x = "Engagement points", y = "Prior density"
  )
```

Centred on zero, because we should not assume in advance that a mandate
helps or hurts. A standard deviation of 10 points on a 0–100 engagement
scale is deliberately generous — it would take a swing of roughly 20
points before the prior started pulling the estimate back, and no
plausible HR policy moves engagement that far. That's a weakly
informative prior doing its job: ruling out the absurd, staying out of
the way of the plausible.

Worth checking what the priors imply about engagement scores themselves,
because that's where a weak prior can quietly become a silly one:

```{r prior-predictive}
#| code-fold: true
n_sim <- 4000

tibble(
  intercept = rnorm(n_sim, mean = 70, sd = 20),
  sigma     = rexp(n_sim, rate = 0.2),
  simulated = rnorm(n_sim, mean = intercept, sd = sigma)
) |>
  ggplot(aes(simulated)) +
  geom_histogram(bins = 50, fill = navy, alpha = 0.45) +
  geom_vline(xintercept = c(0, 100), linetype = "dashed", colour = red) +
  labs(
    title = "What the priors think an engagement score could be",
    subtitle = "Red dashed = the ends of the scale. Some mass falls outside it.",
    x = "Simulated engagement score", y = "Simulated draws"
  )
```

The prior puts some weight below 0 and above 100, which are impossible
scores. That's a mild flaw and not worth fixing here: with 480
observations the likelihood overwhelms it immediately, and the
alternative — a prior that respects the bounds — buys nothing for a
teaching example. It's flagged rather than hidden because the habit of
checking is what matters, and because on a smaller dataset it would
matter.

### The model

The same quantity, fitted properly so it comes with uncertainty:

```{r fit-did}
panel <- panel |>
  mutate(treated_post = treated * post)                            # <1>

fit_did <- brm(
  engagement ~ treated + post + treated_post + (1 | office),       # <2>
  data = panel, family = gaussian(),
  prior = c(
    prior(normal(70, 20),  class = Intercept),
    prior(normal(0, 10),   class = b),
    prior(exponential(0.2), class = sd),
    prior(exponential(0.2), class = sigma)
  ),
  chains = 4, iter = 2000, seed = 21, refresh = 0
)
```

1.  Built as its own column deliberately. `treated * post` in the formula
    is the idiomatic shorthand and gives an identical answer, but names
    the parameter `b_treated:post`, and a colon in a parameter name is a
    small nuisance every time you extract it. Being explicit costs one
    line and makes the estimate's meaning legible: *treated, and after*.
2.  `(1 | office)` because each office appears twelve times — the same
    repeated-measures structure as Chapter 8, and ignoring it would
    understate the uncertainty considerably.

```{r read-did}
fit_did |>
  spread_draws(b_treated_post) |>
  median_qi(.width = c(0.95, 0.5))
```

```{r did-posterior-plot}
#| code-fold: true
fit_did |>
  spread_draws(b_treated_post) |>
  ggplot(aes(x = b_treated_post)) +
  stat_halfeye(fill = navy_light, .width = c(0.5, 0.95)) +
  geom_vline(xintercept = true_effect, linetype = "dashed", colour = red) +
  labs(
    title = "The estimated effect of the mandate",
    subtitle = "Red dashed = the true effect we simulated",
    x = "Engagement points", y = NULL
  )
```

### Does it fit?

The prior predictive check above flagged that these priors will entertain
impossible engagement scores. The posterior predictive check is where you
find out whether that mattered once the data had its say:

```{r ppc-did}
pp_check(fit_did, ndraws = 50) +
  geom_vline(xintercept = c(0, 100), linetype = "dashed", colour = red) +
  labs(title = "Posterior predictive check: simulated vs actual engagement",
       subtitle = "Thin lines = datasets the model would generate; thick line = the real data. Red dashed = the ends of the scale.",
       x = "Engagement score", y = "Density")
```

::: callout-note
The simulated distributions should now sit comfortably inside the scale —
480 observations were more than enough to pull the model away from the
prior's impossible tails, which is what the earlier flag predicted.

Worth noting what this check can and can't tell you here. It confirms the
model reproduces the *distribution* of engagement scores, which is a real
check on the Normal likelihood. It says nothing whatever about whether
parallel trends holds — that's an assumption about a counterfactual, and no
posterior predictive check can interrogate it. The event study in the next
section is the tool for that, and the distinction is worth keeping straight:
**a model can fit its data beautifully and still be identifying the wrong
quantity.**
:::

::: callout-important
## What this buys you, and what it costs

The estimate recovers the true effect, and it does so **without
adjusting for anything about the people**. We never measured motivation,
manager quality or team composition. Those all still differ between
functions — they just cancel, because they're differences that were there
before the policy and are still there after.

That's the whole appeal, and it's why this design is worth actively
hunting for. It handles the confounders you couldn't have measured, which
is the exact failure mode Chapter 21 had to leave open.
:::

### The assumption, stated honestly

Difference-in-differences replaces "no unmeasured confounding" with a
different assumption, not with no assumption:

> **Parallel trends.** Absent the policy, the treated group's engagement
> would have moved the same way the control group's did.

This is a claim about something that didn't happen, so no test can
confirm it. What you can do is check whether it held in the period before
the policy, when both groups were untreated and you *can* observe both.
If the lines were parallel for six months and then diverged exactly when
the policy landed, the assumption is credible. If they were already
diverging, it isn't.

------------------------------------------------------------------------

## Event studies: interrogating the assumption

An event study estimates the gap between the groups **separately in each
period**, rather than collapsing everything into before and after. It
turns parallel trends from an assertion into a picture.

```{r fit-event}
panel <- panel |>
  mutate(month_f = factor(month))

fit_event <- brm(
  engagement ~ month_f + month_f:treated + (1 | office),           # <1>
  data = panel, family = gaussian(),
  prior = c(
    prior(normal(70, 20),  class = Intercept),
    prior(normal(0, 10),   class = b),
    prior(exponential(0.2), class = sd),
    prior(exponential(0.2), class = sigma)
  ),
  chains = 4, iter = 4000, seed = 21, refresh = 0                  # <2>
)
```

1.  Note what's missing: a `treated` term on its own. The obvious
    formula here is `treated * month_f`, and it causes a real problem —
    see the box below.
2.  Twice the usual iterations, for the same reason.

::: callout-note
## A model that strains, and why

Written the obvious way — `engagement ~ treated * month_f + (1 | office)`
— this model samples badly and `brms` says so:

```
Warning: Bulk Effective Samples Size (ESS) is too low, indicating
posterior means and medians may be unreliable.
```

Chapter 11's advice applies: don't ignore it, and don't just add
iterations until the warning goes away without understanding it. Here the
cause is structural rather than bad luck.

Treatment is a property of the **office**, and it never changes — office
7 is in a mandated function in every one of its twelve rows. So a single
`treated` coefficient and the office-level intercepts `(1 | office)` are
trying to explain the same thing: why some offices sit higher than
others. The sampler can raise one and lower the other and land in almost
exactly the same place, so it wanders along a ridge instead of exploring
a peak. That's what low bulk ESS looks like.

The fix is to stop asking for the redundant parameter. `month_f:treated`
without a `treated` main effect gives one gap per month — which is what
an event study wants anyway — and no global treated term for the office
intercepts to fight with. The extra iterations are belt and braces.

This situation is common enough to be worth recognising on sight:
**whenever a predictor is constant within a grouping factor you're also
modelling, expect the two to compete.** It shows up in any panel design,
and it's a case where the econometric convention — absorbing units with
fixed effects, so unit-level predictors drop out entirely — is dodging
the same problem by a different route.
:::

Rather than extracting a dozen interaction coefficients by name, ask the
model for the fitted mean of each group in each month and subtract:

```{r event-gaps}
event_gaps <- panel |>
  distinct(month, month_f, treated) |>
  add_epred_draws(fit_event, re_formula = NA) |>                   # <1>
  ungroup() |>
  select(month, treated, .draw, .epred) |>
  pivot_wider(names_from = treated, values_from = .epred,
              names_prefix = "group_") |>
  mutate(gap = group_1 - group_0) |>
  group_by(month) |>
  median_qi(gap, .width = 0.95)

event_gaps
```

1.  `re_formula = NA` marginalises over the office-level intercepts — we
    want the average treated-versus-control gap, not a particular
    office's.

```{r event-study-plot}
ggplot(event_gaps, aes(x = month, y = gap)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_vline(xintercept = policy_month - 0.5, colour = red, linetype = "dashed") +
  geom_linerange(aes(ymin = .lower, ymax = .upper), colour = navy_light) +
  geom_point(colour = navy, size = 2.5) +
  scale_x_continuous(breaks = months) +
  labs(
    title = "The gap between treated and control offices, month by month",
    subtitle = "Red dashed = the policy. Flat before, stepped after, is what you want to see.",
    x = "Month", y = "Treated − control (engagement points)"
  )
```

This chart is the deliverable, more than the coefficient is. It shows a
stakeholder three things at once: the groups were tracking each other
before the policy, they separated when it landed, and by how much.

::: callout-important
## Reading the pre-period is the point

The months **before** the red line are the evidence. They're periods
where nothing had happened yet, so the gap ought to be flat. If it is,
parallel trends looks reasonable and the headline number is credible.

Note that "flat" is the test, not "zero". The two groups can sit at
different levels throughout for all sorts of permanent reasons —
different functions, different pay bands, different work — and the design
doesn't care, because a constant difference cancels. The dotted line at
zero is a reading aid, not a target.

If the gap is already sloping before the policy — the treated group
drifting away for reasons of its own — then the difference-in-differences
estimate is picking up that drift and attributing it to the mandate. The
honest response is to say so, not to report the number with a caveat.

Watch also for a gap that starts moving *just* before the intervention.
That usually means people anticipated it, which is its own finding and
badly breaks the design.
:::

------------------------------------------------------------------------

## When the rollout is staggered

Our policy landed everywhere at once. Real rollouts usually don't — the
training reaches the North in March, the South in September, the
international offices next year.

The instinct is to keep the same model, replacing `post` with "has this
office been treated yet". This is the two-way fixed effects estimator,
and for two decades it was standard practice.

::: callout-important
## The obvious model is biased here, and badly

A substantial recent econometrics literature has established that with
staggered timing **and** effects that differ between groups or grow over
time, two-way fixed effects doesn't estimate the average treatment effect.
It estimates a weighted average of comparisons that includes
already-treated groups being used as controls for later-treated ones —
and some of those weights are negative. The estimate can come out with the
wrong sign while every underlying effect points the same way.

Both conditions are the norm in People Analytics. Effects almost always
vary between regions, and they almost always grow or fade over time.

The reason this box has no worked example is that the fix is a live
research area with several competing estimators, and picking one for you
would date badly. What matters is that you recognise the situation and
don't reach for the model in the previous section. The search terms are
**Goodman-Bacon decomposition** (why it breaks), and **Callaway &
Sant'Anna**, **Sun & Abraham** and **Borusyak et al.** (what to use
instead).
:::

That recognition is most of the value. A staggered rollout looks like an
easier version of the same problem and is a harder one.

------------------------------------------------------------------------

## Thresholds and instruments, briefly

Two more designs worth being able to recognise, treated at the same depth
as the "names worth knowing" section in Chapter 14 — enough to spot the
opportunity and know what to search for.

**Regression discontinuity** exploits a sharp rule. Options vest at
exactly 24 months; the bonus multiplier changes at 110% of target;
performance plans trigger below a rating of 3. Someone at 23 months and
someone at 25 months are near-identical people, and one of them got the
treatment. Comparing narrowly across the threshold gives you something
close to a randomised comparison. It's underused in People Analytics
despite HR being unusually full of sharp rules. The catch: it only tells
you about people *near the cutoff*, which may not be the population you
care about. Search for `rdrobust`, and for **bunching** — if people can
manipulate which side of the line they land on, the design collapses.

**Instrumental variables** need something that shifts the treatment
without affecting the outcome any other way. Distance from an office as
an instrument for actually attending; a randomly-assigned recruiter as an
instrument for time-to-hire; assignment to a training cohort as an
instrument for completing it. Genuinely good instruments in organisational
data are rare, and a weak or invalid one is worse than no analysis. The
econometrics literature here is deep and worth reading before use rather
than after.

::: callout-note
## A different lens: this chapter is econometrics' home ground

Nothing in this chapter is new to economics. Difference-in-differences,
event studies, RDD and IV are the core toolkit of applied
microeconometrics and have been for decades, and the personnel economics
literature has been applying them to workforce questions for nearly as
long.

What's new is the audience. These designs are largely absent from People
Analytics practice, which has tended to inherit its methods from
psychometrics and I/O psychology — traditions strong on measurement and
comparatively quiet on identification. That's the gap this chapter and
Chapter 21 exist to close, and the reading list is already written: the
econometrics textbooks cover this far more thoroughly than a chapter can.

The Bayesian framing adds two things rather than replacing anything. You
get a posterior for the effect rather than a point estimate and a
standard error, which makes "how likely is it that this cost us more than
two engagement points?" directly answerable. And the design analysis
below is much more natural as a simulation than as a formula.
:::

::: callout-tip
## For the ML/DS crowd

The closest familiar structure is a holdout, with the crucial difference
that you didn't choose it and it isn't random.

The temptation, when told an effect is confounded, is to reach for a
model that adjusts harder — more features, more capacity, propensity
scores, doubly-robust estimation. All of those help only with confounders
you *measured*. None of them help with the ones you didn't, and no amount
of model capacity changes that: a flexible model fitted to confounded data
gives you a precisely-estimated wrong answer.

The designs in this chapter are the alternative, and they're a change of
data rather than a change of model. The model at the end is a linear
regression with an interaction term. All the work is in finding the
comparison.
:::

------------------------------------------------------------------------

## Designing the study before you run it

Everything so far has been retrospective — something happened, and we
analysed it. The other half of this chapter is the question that arrives
before any data exists: *we're planning a pilot in one region; is it
worth running?*

The classical answer is a power calculation. The Bayesian answer is
better, and it's a simulation: **generate the study you're proposing,
fit the model you'd fit, and look at what you'd get.** Repeat a few
hundred times.

This is the same machinery as Chapter 5's prior predictive checks, doing
a different job. There we simulated from the prior to see whether our
assumptions were sane. Here we simulate the whole study to see whether
the design can answer the question.

```{r design-function}
simulate_one <- function(n_offices, n_months = 12, effect = -4,      # <1>
                         office_sd = 6, noise_sd = 3) {
  offices <- tibble(
    office      = 1:n_offices,
    treated     = as.integer(office <= n_offices / 2),
    office_base = rnorm(n_offices, 70, office_sd)
  )

  d <- expand_grid(offices, month = 1:n_months) |>
    mutate(
      post = as.integer(month >= (n_months / 2 + 1)),
      engagement = office_base + 0.4 * month +
        effect * treated * post + rnorm(n(), sd = noise_sd)
    )

  fit <- lm(engagement ~ treated + post + treated:post, data = d)    # <2>
  ci  <- confint(fit)["treated:post", ]

  tibble(
    estimate = coef(fit)[["treated:post"]],
    lower = ci[[1]], upper = ci[[2]],
    width = ci[[2]] - ci[[1]],
    excludes_zero = (ci[[1]] > 0) | (ci[[2]] < 0)
  )
}
```

1.  Every quantity you'd have to guess at is an argument, which is the
    discipline this exercise imposes: you cannot run it without
    committing to how big an effect would matter and how noisy your
    measure is. Those commitments are the valuable part, whatever the
    simulation returns.
2.  `lm()`, not `brm()`, and this is a deliberate shortcut. We're asking
    about *precision*, and with the weakly-informative priors this book
    uses the credible interval and the confidence interval are close
    enough for a design question. Fitting `brm()` a few hundred times
    would take hours and change no decision. Once you've settled on a
    design, refit that one with the model you'll actually use.

```{r run-design-sweep}
design <- expand_grid(n_offices = c(6, 10, 20, 40, 80), rep = 1:200) |>
  mutate(result = map(n_offices, simulate_one)) |>
  unnest(result)
```

Two things are worth reading off this, and only one of them is power.

```{r design-precision-plot}
#| code-fold: true
design |>
  group_by(n_offices) |>
  median_qi(width, .width = 0.95) |>
  ggplot(aes(n_offices, width)) +
  geom_linerange(aes(ymin = .lower, ymax = .upper), colour = navy_light) +
  geom_line(colour = navy, linewidth = 1) +
  geom_point(colour = navy, size = 2.5) +
  geom_hline(yintercept = abs(true_effect), linetype = "dashed", colour = red) +
  scale_x_log10(breaks = c(6, 10, 20, 40, 80)) +
  labs(
    title = "How precise an answer would each design give?",
    subtitle = "Red dashed = the size of the effect itself. Above the line, the interval is wider than the thing you're measuring.",
    x = "Number of offices (log scale)", y = "Width of the interval"
  )
```

The red line is the useful reference and it isn't a significance
threshold. If your interval is wider than the effect you're looking for,
the study cannot distinguish "this policy cost four points" from "this
policy did nothing" — regardless of what comes back significant.

```{r design-errors}
design |>
  group_by(n_offices) |>
  summarise(
    detected      = mean(excludes_zero),                             # <1>
    wrong_sign    = mean(excludes_zero & sign(estimate) != sign(true_effect)),
    exaggeration  = mean(abs(estimate[excludes_zero]) / abs(true_effect))
  )
```

1.  The proportion of studies returning an interval that excludes zero —
    the simulation's version of statistical power.

::: callout-important
## The two columns nobody reports

`wrong_sign` is the **Type S** error rate — how often a study that finds
something finds it *backwards*. `exaggeration` is the **Type M** error —
how much the studies that "find" an effect overstate its size.

At small sample sizes both get ugly, and they get ugly in a way a power
calculation never shows you. An underpowered study doesn't just fail to
find real effects. Conditional on finding something, it reports an effect
substantially larger than the truth, and occasionally with the wrong
sign — because the only estimates large enough to clear the threshold are
the ones that got lucky.

This is why "we ran it and found a significant 9-point improvement" from
a six-office pilot deserves scepticism rather than a rollout. The 9
points is not evidence of a big effect. It is what a small effect looks
like when the only way to detect it was to overestimate it.
:::

::: callout-tip
## The most valuable output is often "don't run it"

The honest conclusion from an exercise like this is frequently that the
proposed study cannot answer the question — that with six offices and a
noisy engagement measure, no result would be informative either way.

That is worth saying out loud, before the effort is spent, and it is a
far more valuable contribution than analysing the data afterwards and
hedging. It also tends to be persuasive in a way that methodological
objections after the fact never are: *"here's what this pilot would tell
us, at the size proposed"* is a chart, and it gets designs changed.
:::

::: callout-caution
## Try it yourself

1.  Rerun the design sweep with `effect = -1` instead of −4. How many
    offices would you need to detect a one-point change in engagement?
    Is that number achievable in your organisation?
2.  Break parallel trends deliberately: add a term to `simulate_one()`
    that gives treated offices their own upward drift. Refit the
    difference-in-differences model. How large does the drift have to be
    before the estimate is meaningfully wrong — and would the event-study
    plot have caught it?
3.  Change `office_sd` from 6 to 15 — offices that differ far more from
    each other. Does the difference-in-differences estimate get worse?
    Should it? Reason it through before running it.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Two habits, and they arrive at opposite ends of a project.

At the start: **ask what changed and when.** Organisations run natural
experiments constantly and never label them — policy changes, phased
rollouts, reorganisations, sharp thresholds in comp and benefits rules.
Almost nobody in People Analytics is looking for them, which means the
opportunities are largely unclaimed.

Before any pilot: **simulate it.** An afternoon spent generating the
study you're being asked to run is the cheapest thing in this book, and
it converts an argument about methodology into a chart about what the
proposed design can and cannot show. It also protects you from the much
worse conversation later, when the pilot has run, the result is
ambiguous, and you have to explain why after the money is spent.

And when the design genuinely isn't there — no date, no order, no
threshold — Chapter 21's answer still stands. Draw the graph, state the
assumption, and be explicit that the estimate rests on it.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  When a confounder is unmeasured, adjustment can't help — but
    variation the organisation created for its own reasons often can.
2.  Look for **a date**, **an order** or **a threshold**.
3.  **Difference-in-differences** cancels any confounder that was stable
    across the change, including ones you never measured.
4.  It assumes **parallel trends** — untestable directly, but
    interrogable in the pre-period.
5.  An **event study** turns that assumption into a chart, and the chart
    is usually the better deliverable.
6.  **Staggered rollouts** break the obvious model. Recognise them and
    reach for the modern estimators.
7.  **Simulation-based design analysis** tells you before you start
    whether a study can answer its question, and reports Type S and Type
    M errors that a power calculation hides.
:::

## Next chapter {.unnumbered}

**The elicitation workflow: building priors with your stakeholders** —
the closing chapter, and the one that ties Part V together. Chapter 21
made your causal assumptions explicit as a graph; this chapter made your
design assumptions explicit as a simulation. Both are things you should
be building *with* the people who know the organisation, not presenting
to them at the end. The elicitation workflow is how.

------------------------------------------------------------------------

<!-- NOTE FOR REVISION (not rendered)

## Note for revision

Two open questions for a later pass:

**A real dataset.** Everything here is simulated, which is defensible
(and Chapter 21 sets the precedent) but means the chapter never touches
organisational data. If a public panel with a policy change can be found —
or if an anonymised extract becomes available, as with `turnover` in
Chapter 17 — the difference-in-differences section would be stronger with
it. The simulation should stay regardless; it's what lets us check the
method against a known answer.

**Housekeeping item, now resolved.** The YAML header that used to sit at
the top of this file (`warning: false` / `message: false`) has been removed
— `_quarto.yml` sets both project-wide.

**A book-wide refactor, half done.** `_setup.R` defines the colour tokens
(`navy`, `red`, `navy_light`) and the shared `theme_set()`, but no chapter
sources it. Chapters 16–23 now define the tokens in their own setup chunk
and reference them by name instead of repeating hex literals inline, which
gets most of the benefit — the palette is legible and consistent, and each
chapter still renders standalone from `drafts/`. Chapters 1–15 still have
the hex values inline and should be brought into line.

The fully tidy arrangement is still the standard Quarto book pattern: one
`source("_setup.R")` line per chapter, which works because `_quarto.yml`
sets `execute-dir: project`. Two things to sort out before doing it:
`_setup.R` currently loads `skimr` and `rstanarm`, which most chapters
don't need and which would make every chapter fail if either package is
missing; and sourcing it breaks standalone rendering from `drafts/`. Trim
the library list first, then it's genuinely a one-line change per chapter.

**Where the design section belongs.** It sits here because experiments
only entered scope with this chapter, and because it shares the
simulation machinery. But it's equally at home in Chapter 20 (A/B
testing), which is where a reader planning a people experiment would
naturally look. Worth revisiting once both chapters are integrated —
splitting it would duplicate the machinery, so it's one or the other.

-->
