# Meeting Your Data

## Welcome {.unnumbered}

### The data a people analyst actually works with {.unnumbered}

Almost nothing arrives in one piece. A People Analytics team's data is
assembled from wherever it happens to live:

-   The **HRIS or HRMS** — Workday, SuccessFactors, BambooHR — holding
    the core record: who works here, in what role, reporting to whom,
    since when, on what salary.
-   **Survey platforms** — engagement, pulse, onboarding, exit — each
    with its own scale and its own idea of an employee identifier.
-   **Other HR systems** — the ATS for recruitment, the LMS for
    learning, performance management, absence, payroll. Rarely
    designed to talk to each other.
-   And, always, **spreadsheets**. Someone's headcount tracker.
    Somebody's list of high-potential employees. The org design work
    that never made it into a system.

Merging these is a large part of the job, and it's where most of the
errors that reach a boardroom are born. This book won't teach you
that plumbing — but everything below assumes you're sceptical about
what came out of the other end of it.

### Most of the job is presenting it {.unnumbered}

Be honest about how HR analysis actually gets spent. Think of the
dashboards and monthly packs that fill HR departments: headcount by
division, attrition against last year, time-to-hire by recruiter,
engagement scores sliced eleven ways. Producing them consumes an
enormous share of an analyst's week, and most of it is description —
counting, averaging, and drawing bars.

That work is worth doing well. It's also worth doing *thoughtfully*,
which is the difference this chapter is really about.

### Understand it before you analyse it {.unnumbered}

Whatever you've been asked for, the first move is the same: find out
what you actually have. This is **exploratory data analysis** (EDA) —
a term coined by John Tukey to describe looking at data with an open
mind before testing anything against it. In practice it's a short list
of habits:

-   Check the **structure** — how many rows, how many columns, what
    type is each one, and what does a single row represent?
-   Check the **gaps** — which values are missing, how many, and is
    the missingness patterned?
-   Look at **each variable on its own** — its centre, its spread, its
    shape.
-   Hunt for the **impossible and the implausible** — negative tenure,
    a salary of 1, an engagement score of 9 on a 4-point scale, six
    people reporting to themselves.
-   Look at **variables in pairs** — does anything move together in a
    way you didn't expect?
-   **Sense-check against the business.** You know roughly how many
    people work here. Does the data agree?

None of that is glamorous, and all of it is faster than explaining a
wrong number afterwards.

### Always ask why they're asking {.unnumbered}

Even when the request is purely descriptive — *"can you send me
attrition by department for the last six quarters?"* — the great
analyst does one more thing before opening the data: works out **what
the person is going to do with it**.

Nobody wants a table. They want to decide something, or defend
something, or reassure someone. Once you know which, you can often
offer an approach that serves them better than the one they asked for
— and that's the difference between an analyst who fields requests
and one who gets invited to the conversation earlier.

::: callout-note
This matters more than it sounds, and it's not a soft skill. We come
back to it at the end of this chapter, because there's a specific,
predictable thing people are usually doing with historical data — and
they're usually doing it badly.
:::

### How to read this chapter {.unnumbered}

As we talk through each idea, run the code chunk next to it, change a
number, and see what happens. This chapter follows the same rhythm
every chapter in this book will: a little theory, a lot of code you can
poke at.

Today is deliberately gentle: **no models, no probability, no maths**.
We meet the two datasets we'll use throughout the book, and learn to
*describe* data.

## What you'll be able to do by the end {.unnumbered}

1.  Load a dataset into R and inspect its structure
2.  Tell **categorical** from **numeric** variables — and why it matters
3.  Describe a numeric variable with **centre** and **spread**
4.  Describe a categorical variable with **counts** and **proportions**
5.  Read histograms, boxplots, violin plots and bar charts — and know
    which to put in front of a manager
6.  Explain why a request for historical data is usually a question
    about the future

## Why start here? {.unnumbered}

::: callout-important
Every analysis begins by describing the data. Before anyone believes
your model, they want to know you **understand what you're working
with**. Most analytical mistakes are made before any model is fitted,
by someone who didn't look at their data first.
:::

------------------------------------------------------------------------

## Setup

We load the tidyverse, set a consistent plot theme, and load our two
running datasets from the `peopleanalyticsdata` package (created by
Keith McNulty — see the [Welcome](../index.qmd) chapter for the full
credit). **This same setup block appears, in spirit, at the top of
every chapter in this book.**

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(skimr)

theme_set(theme_minimal(base_size = 13))

data("salespeople", package = "peopleanalyticsdata")
data("employee_survey", package = "peopleanalyticsdata")
```

::: callout-tip
## If you get a "package not found" error

Run this **once** in the console (not in a code chunk):

```{r}
#| eval: false
install.packages(c("tidyverse", "peopleanalyticsdata", "skimr"))
```
:::

------------------------------------------------------------------------

## What *is* a dataset?

### Rows and columns

A dataset is just a table:

-   Each **row** is one *observation* — one thing you measured
-   Each **column** is one *variable* — one thing you measured *about*
    it
-   Each **cell** is one *value*

That three-part description has a name: **tidy data**. It's the
convention the whole tidyverse — `dplyr`, `ggplot2`, and everything
else we use — is built around, and it's why those tools compose so
neatly. Give a tidy table to `group_by()`, `summarise()` or `ggplot()`
and it will do the obvious thing without argument.

::: callout-note
## Tidy for analysis, not necessarily for reading

Conveniently, tidy is roughly how HR systems already store data. An
HRIS keeps one row per employee, or one row per absence event, or one
row per survey response — because that's the sane way to store
records.

Inconveniently, it's often the *worst* way to present them. Nobody
wants a thousand-row table. A manager wants absence rates as a small
grid of departments down the side and months across the top — which
is emphatically not tidy, because the column headings have become
data.

Both shapes are correct for their purpose. Analyse long, present wide,
and use `pivot_longer()` and `pivot_wider()` to move between them.
:::

::: callout-important
## The sentence you must always be able to finish

**"One row is one \_\_\_."**

Get this wrong and every number that follows is wrong. In
`salespeople`, one row is one **salesperson**.
:::

### Our two datasets

We'll use these two throughout the book, so it's worth learning them
properly.

-   **`salespeople`** — one row per salesperson. Sales figures,
    customer ratings, performance, and whether they were promoted.
    Used to ask *"what predicts who gets promoted?"*
-   **`employee_survey`** — one row per **survey response** to an
    engagement survey, on a four-point scale. This is *survey data* —
    exactly the shape most People Analytics teams collect constantly.

### A first look

`glimpse()` shows the structure: every column, its type, and the first
few values.

```{r glimpse-salespeople}
glimpse(salespeople)
```

So we have `r nrow(salespeople)` salespeople described by
`r ncol(salespeople)` variables.

`glimpse()` answers *what columns do I have*. It deliberately does not
tell you what is **in** them — that comes shortly, with `skim()`, which is
the tool we use for the fuller look. Two functions, two jobs, and it is
worth keeping them separate in your head.

### Two kinds of variable

This distinction drives every choice you make later.

| | **Numeric** | **Categorical** |
|---|---|---|
| Meaning | arithmetic makes sense | a label for a group |
| Examples | `sales`, `customer_rate` | `promoted`, `performance` |
| Describe with | centre and spread | counts |
| Picture | histogram, boxplot | bar chart |

Averaging `performance` (a 1–4 rating) is defensible only if you're
comfortable treating it as a genuine number rather than an ordered
label — we'll come back to exactly that question, properly, in Chapter 18. For now, treat it as categorical.

### The other dataset

```{r glimpse-survey}
glimpse(employee_survey)
```

### Your turn {.unnumbered}

Look at that output and answer three questions before reading on. What
does **one row** represent — and how do you know? Why are there so many
more rows here than in `salespeople`? And are these columns numeric or
categorical, given every one of them holds a whole number from 1 to 4?

The third question is genuinely hard, and it's the one Chapter 18 is
built around.

------------------------------------------------------------------------

## Describing a number

### Two questions, always

For any numeric variable, ask:

1.  **Where is the middle?** — the typical value
2.  **How spread out is it?** — how much values vary

Answer both and you've described it. Everything else is detail.

### Mean vs median

Two ways to say "the middle":

-   The **mean** adds everything and divides — every value affects it,
    including extreme ones
-   The **median** is the middle value when sorted — it ignores *how*
    extreme the extremes are

When someone says "average" they almost always mean the **mean** —
it's what `AVERAGE()` computes in Excel, and it's what a stakeholder
pictures when they ask for one. That's fine when the data is roughly
symmetric, and misleading when it isn't.

### Which one, in HR {.unnumbered}

The choice is usually decided by whether a long tail exists:

-   **Salary → median.** Pay distributions are right-skewed: a handful
    of senior people pull the mean above what almost anyone earns. It
    is not an accident that UK gender pay gap reporting requires
    organisations to publish **both** the mean and the median gap —
    the pair tells you something neither does alone, and a large
    difference between them is itself a finding about how pay is
    distributed.
-   **Tenure → median.** A few twenty-year veterans will drag the mean
    somewhere no current employee actually sits.
-   **Time to hire → median.** One requisition open for eleven months
    can add a fortnight to the mean across a whole quarter.
-   **Absence days → median**, usually, and often better still as a
    distribution: most people take very few days, a small number take
    many.
-   **Engagement scores → mean is fine.** The scale is bounded 1 to 5,
    so there's no tail to distort anything, and means are easier to
    compare across teams.

::: callout-important
If you're not sure, compute both. If they agree, report the mean and
move on. If they don't, you've learned something about the shape of
your data that you needed to know anyway.
:::

::: callout-note
## Why the difference matters

Nine salespeople sell \$40,000 each and one sells \$2,000,000. The
**mean** is a figure nobody near actually sold. The **median** is
\$40,000.

Which better describes a typical salesperson?
:::

### The numbers for sales

`glimpse()` told us what the columns *are*. To see what is *in* them,
the one-line answer is `skim()`, from the `skimr` package:

```{r sales-skim}
skim(salespeople)
```

That is a lot of table at once, so read it in this order.

**`n_missing` first.** It counts the empty cells in each column, and it
is the reason for the `na.rm = TRUE` arguments you are about to meet.
We come back to it properly at the end of the chapter. Getting this on
screen in the first minute, rather than discovering it later, is the
main reason this book uses `skim()` rather than the shorter
`summary()`.

**Then the middle.** `mean` is the mean and `p50` is the median. When
those two disagree, the column is skewed — the distinction we were just
drawing, now visible for every column at once.

**Then the two ends.** `p0` and `p100` are the smallest and largest
values, and this is where impossible numbers show themselves: a negative
tenure, a salary of 1, an engagement score of 9 on a four-point scale.

`skim()` also groups its output by variable type, giving character
columns and numeric columns separate tables with different summaries.
Here there is only one table, because every column in `salespeople` is
stored as a number. That includes `performance`, which is really a 1--4
rating: a category stored as a number. Neither `skim()` nor R can tell
the difference, which is exactly why the numeric-versus-categorical
judgement earlier in this chapter is yours to make rather than the
computer's.

The tidyverse version is more typing but composes with everything else:

```{r sales-centre}
salespeople |>
  summarise(
    mean_sales   = mean(sales, na.rm = TRUE),
    median_sales = median(sales, na.rm = TRUE),
    sd_sales     = sd(sales, na.rm = TRUE)
  )
```

::: callout-tip
## What `na.rm = TRUE` is doing there

`sales` has a missing value in it, and R's summary functions **refuse
to guess**. Without `na.rm = TRUE` they hand back `NA` instead of a
number — or, in `quantile()`'s case, an outright error.

Treat `na.rm = TRUE` as a habit worth having by default whenever you
summarise real data. The last section of this chapter explains why
it's a habit rather than a fix.
:::

The **mean** is `r round(mean(salespeople$sales, na.rm = TRUE), 1)` and
the **median** is `r round(median(salespeople$sales, na.rm = TRUE), 1)`
(both in thousands of dollars).

### See it: the histogram

```{r sales-histogram}
ggplot(salespeople, aes(x = sales)) +
  geom_histogram(bins = 30, fill = "#122a52", alpha = 0.45, colour = "white") +
  geom_vline(xintercept = mean(salespeople$sales, na.rm = TRUE),
             colour = "#8a5a00", linewidth = 1) +                    # <1>
  geom_vline(xintercept = median(salespeople$sales, na.rm = TRUE),
             colour = "#8a5a00", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Distribution of sales",
    subtitle = "Solid line = mean, dashed line = median",
    x = "Sales (thousands of dollars)", y = "Number of salespeople"
  )
```

1.  Amber, and the same amber for both. The two lines are the *point* of
    this chart, so they need to read clearly against the navy bars —
    which a soft grey does not. **Linetype**, not colour, separates the
    mean from the median, so that the pair still reads as one idea.

### What the shape tells you

Most salespeople cluster in the middle, with a **tail of high
performers** stretching to the right — that tail is exactly why the
mean sits *above* the median.

::: callout-important
When the mean and median disagree, the data is **skewed** — and you
should report both.
:::

### Spread: two measures

-   **Standard deviation (SD)** — roughly the typical distance from the
    mean. It pairs with the mean, and like the mean it is pulled by
    every extreme value.
-   **Interquartile range (IQR)** — the range covering the middle 50%
    of the data, from the 25th to the 75th percentile. It pairs with
    the median, and like the median it is barely affected by extremes.

```{r sales-spread}
salespeople |>
  summarise(
    sd  = sd(sales, na.rm = TRUE),
    q25 = quantile(sales, probs = 0.25, na.rm = TRUE),
    q75 = quantile(sales, probs = 0.75, na.rm = TRUE),
    iqr = IQR(sales, na.rm = TRUE)
  )
```

Note the pairing. If you report a **mean**, report an **SD** next to
it; if you report a **median**, report the **IQR**. Mixing them —
a median with a standard deviation — is a small tell that someone
reached for whichever number came to hand.

### The boxplot

```{r sales-boxplot}
ggplot(salespeople, aes(y = sales)) +
  geom_boxplot(fill = "#122a52", alpha = 0.8) +
  labs(title = "Sales — boxplot", y = "Sales (thousands of dollars)", x = NULL) +
  theme(axis.text.x = element_blank())
```

### Reading it, and when not to use it

A boxplot is a **five-number summary** drawn to scale. The box runs
from the 25th to the 75th percentile — so the box *is* the IQR, and
the middle half of your salespeople sit inside it. The line across the
box is the median. The whiskers reach out to the furthest points
within 1.5 × IQR of the box, and anything beyond that is drawn as an
individual dot.

Analysts love these, for good reasons: they're compact, they encode
five numbers plus outliers in almost no ink, and — the real
attraction — you can line up eight of them side by side and compare
groups at a glance.

::: callout-important
## Why they often fail in front of a manager

Three problems, and they compound:

1.  **They need explaining.** If you have to say "the whiskers are 1.5
    times the interquartile range" out loud in a meeting, you've spent
    your audience's attention on the chart instead of the finding.
2.  **They hide the shape.** A boxplot of a two-humped distribution —
    say, a team split between long-tenured staff and a recent intake —
    looks identical to a boxplot of a single broad group. That's a
    genuinely important difference to conceal.
3.  **They hide the sample size.** A box drawn from six people looks
    exactly as solid as one drawn from six hundred. In People
    Analytics, where team sizes vary wildly, this is the dangerous
    one — and it's the problem Chapters 8, 14 and 15 exist to solve
    properly.
:::

### The violin plot

A **violin plot** solves the first two by drawing the actual shape of
the distribution — a smoothed histogram, mirrored to make it
symmetric. Wide means many people at that value; narrow means few.
Most people read one correctly without instruction, because width
maps onto "how many" the way intuition expects.

You don't have to choose. The usual recommendation is both — the
violin for the shape, a slim boxplot inside it for the summary
statistics:

```{r sales-violin}
ggplot(salespeople, aes(x = "", y = sales)) +
  geom_violin(fill = "#122a52", alpha = 0.30, colour = "#122a52") +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +   # <1>
  labs(title = "Sales — violin plot with boxplot inside",
       subtitle = "The shape shows where salespeople actually pile up",
       y = "Sales (thousands of dollars)", x = NULL)
```

1.  `outlier.shape = NA` suppresses the boxplot's outlier dots, since
    the violin already shows the tails. Without it you'd draw the same
    extreme salespeople twice.

Compare the two figures. The boxplot says sales are right-skewed. The
violin shows you *how* — a dense cluster of salespeople around 400,
thinning steadily upward, with no second group hiding anywhere.

::: callout-tip
Whichever you use, add the group size to the axis label — `Sales team
(n = 42)` — or plot the individual points on top when there are few
enough. It costs nothing and it stops a six-person team being read
with the same confidence as a six-hundred-person one.
:::

------------------------------------------------------------------------

## Describing a category

### Count, don't average

There's no "average performance rating" that means much on its own.
Instead we ask: **how many** fall into each group, and **what share**
is that?

```{r performance-counts}
salespeople |>
  count(performance) |>
  mutate(proportion = round(n / sum(n), 3))
```

### The bar chart

```{r performance-bar}
salespeople |>
  count(performance) |>
  ggplot(aes(x = factor(performance), y = n)) +
  geom_col(fill = "#122a52") +
  labs(title = "Performance ratings across the sales team",
       x = "Performance rating (1 = lowest, 4 = highest)", y = "Number of salespeople")
```

### Bars vs histograms

::: callout-tip
## Don't mix them up

-   A **bar chart** shows counts for a handful of **labels**. Gaps
    between bars are meaningful.
-   A **histogram** shows the shape of a **numeric** variable. Bars
    touch because the underlying scale is continuous.
:::

### Your turn {.unnumbered}

Make the same count-and-bar-chart for `promoted`. What share of the
sales team was promoted?

```{r promoted-your-turn}
# Your code here

```

------------------------------------------------------------------------

## The survey data

### Meet the engagement survey

Each row is one employee's response, with an overall `Happiness` score
(1–4) and several sub-scale items on the same scale: benefits (`Ben1-3`),
work environment (`Work1-3`), management (`Man1-3`), and career
prospects (`Car1-4`).

::: callout-note
This is exactly the shape of an engagement survey you'll meet in real
People Analytics work — which is why we'll keep coming back to it.
:::

```{r survey-overall}
employee_survey |>
  summarise(
    mean_happiness   = mean(Happiness, na.rm = TRUE),
    median_happiness = median(Happiness, na.rm = TRUE),
    sd_happiness     = sd(Happiness, na.rm = TRUE)
  )
```

### How employees rate their happiness

```{r survey-bar}
ggplot(employee_survey, aes(x = factor(Happiness))) +
  geom_bar(fill = "#122a52") +
  labs(title = "Overall happiness (1–4)",
       x = "Happiness rating", y = "Number of responses")
```

### Why that shape matters

Survey scores are **bounded** (they can't go below 1 or above 4) and
often **pile up toward one end**.

::: callout-important
Keep that shape in mind. It's extremely common in engagement data, it
breaks some standard modelling assumptions, and it's exactly why
Chapter 18 exists.
:::

------------------------------------------------------------------------

## A first look at relationships

Description gets interesting when we ask whether one variable *differs
across* another. We're not modelling yet — just looking.

### Does sales differ by promotion outcome?

```{r sales-by-promoted}
salespeople |>
  group_by(promoted) |>
  summarise(
    mean_sales = mean(sales, na.rm = TRUE),
    median_sales = median(sales, na.rm = TRUE),
    n = n()
  )
```

```{r sales-by-promoted-plot}
ggplot(salespeople, aes(x = factor(promoted), y = sales, fill = factor(promoted))) +
  geom_boxplot(alpha = 0.85, show.legend = FALSE) +
  scale_fill_manual(values = c("#122a52", "#3d68a8")) +
  labs(title = "Sales by promotion outcome",
       subtitle = "Promoted salespeople sell more — but is that the whole story?",
       x = "Promoted (1 = yes, 0 = no)", y = "Sales (thousands of dollars)")
```

That question — *is that the whole story?* — is what the rest of this
book teaches you to answer properly.

### A trick worth knowing

Because `promoted` is coded 0/1, its **mean is the promotion rate**.
We'll lean on this constantly.

```{r promotion-rate-by-performance}
salespeople |>
  group_by(performance) |>
  summarise(
    n = n(),
    promotion_rate = mean(promoted, na.rm = TRUE)
  )
```

```{r promotion-rate-plot}
salespeople |>
  group_by(performance) |>
  summarise(promotion_rate = mean(promoted, na.rm = TRUE)) |>
  ggplot(aes(x = factor(performance), y = promotion_rate)) +
  geom_col(fill = "#122a52") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Promotion rate by performance rating",
       x = "Performance rating", y = "Promotion rate")
```

------------------------------------------------------------------------

## Messy reality: missing values

Real organisational data is rarely as clean as it first looks — and,
in fact, neither is this teaching dataset. Let's check what we're
actually working with:

```{r missing-check}
colSums(is.na(salespeople))
```

One missing value in `sales`, one in `customer_rate`, one in
`performance`, none in `promoted`. Out of `r nrow(salespeople)` rows
that's almost
nothing — and it's still enough to turn a mean into `NA`, which is
exactly why the calculations earlier in this chapter used
`na.rm = TRUE`.

Notice too that the gaps aren't only in `sales`. Had we checked only
the column we happened to be summarising, we'd have missed two of
them.

```{r na-demo}
#| eval: false
mean(some_vector)                 # returns NA if anything is missing
mean(some_vector, na.rm = TRUE)   # ignores the missing values
quantile(some_vector, probs = 0.25)                 # errors outright if anything is missing
quantile(some_vector, probs = 0.25, na.rm = TRUE)   # works
```

::: callout-important
## Knowing your gaps

Notice the two different failure modes above: `mean()` fails
*quietly* — it hands you back `NA` and lets you not notice — while
`quantile()` fails *loudly*, with an error. The loud failure is the
friendlier one. The quiet one is how a wrong number ends up in a
report. Deciding what to do about missing data — ignore it, exclude
those rows, or fill it in — is a decision you must **make and
justify**, not something to discover by accident halfway through a
model.
:::

------------------------------------------------------------------------

## What they're really asking for

We promised at the start of the chapter to come back to this, and it's
a key idea of this chapter.

In my experience, when someone asks for historical data, **they very
rarely care about the past**. Attrition last year is over; nobody can
do anything about it. They want to look at the past because they are
trying to work out **what is going to happen next** — and whether they
should do something about it.

The request is descriptive. The question underneath it is predictive.
And we hand over the description and let them do the hard part in
their heads.

### People are not good at this

That last step — looking at history and inferring the future — is one
of the most thoroughly documented weaknesses in human judgement.
Decades of research on decision-making under uncertainty finds the
same failures, and every one of them turns up monthly in HR reporting:

-   **Small samples are trusted far too much.** Kahneman and Tversky
    called this belief in the "law of small numbers": people expect a
    small group to look like the population, so a 40% attrition figure
    from a five-person team is read with roughly the confidence of a
    figure from five hundred. Team-level dashboards are built almost
    entirely from small samples.
-   **Regression to the mean is mistaken for causation.** Extreme
    results tend to be followed by less extreme ones, for no reason
    other than arithmetic. So the worst-scoring team improves after
    the intervention, the best-performing manager slips the following
    year, and both movements get explained with a story. This is the
    single most common way HR credits itself with an effect it didn't
    have.
-   **Noise is read as trend.** Two points make a line. Attrition ticks
    from 11% to 13% and it becomes "a worrying upward trend" in the
    commentary, when it's well within the wobble you'd see from a
    stable process.
-   **Confidence intervals are far too narrow.** When people are asked
    for a range they're 90% sure about, the truth falls outside it
    much more often than 10% of the time. We are systematically more
    certain than we are right.
-   **The story arrives before the evidence.** Give someone a chart and
    a few seconds and they will produce a causal explanation for
    whatever it shows — including, in experiments, for patterns that
    were generated at random.

None of this is stupidity, and pointing it out to stakeholders doesn't
help. These are the default settings of human intuition working on
incomplete information, and analysts have them too.

::: callout-important
## Which is the actual argument for the rest of this book

If the inference is going to happen anyway, it's better done by a
method than by a hunch.

That's what the following chapters are for. Rather than hand over a
number and hope, you can present the data **and** a systematic
interpretation of it: an estimate, an honest range around that
estimate, and where appropriate a forecast that already accounts for
how little data a small team really provides. The uncertainty gets
quantified and shown rather than left to be misjudged.

Do that, and you haven't just answered the question that was asked.
You've answered the one that was meant — and given HR, and the
managers it serves, a genuinely better basis for a decision.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Every People Analytics deliverable opens with a **description of the
data**: how many people, what the key variables look like, and where
the gaps are. Before anyone believes a model, they want to know you
understand what you're working with — and a stakeholder who spots a
headcount that doesn't match theirs will not hear a word of the rest.

Two habits to take away today. **Check the gaps before you summarise
anything**, because a quiet `NA` becomes a wrong number in a board
pack. And **ask what the request is for**, because the answer changes
what you should send back.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  **Exploratory data analysis** comes before everything: structure,
    gaps, shapes, impossible values, and a sense-check against what
    you know of the business.
2.  A dataset is rows (observations) × columns (variables) with one
    value per cell — **tidy data**. Analyse long, present wide.
3.  Describe a **numeric** variable by its **centre** (mean, median)
    and **spread** (SD, IQR); a gap between mean and median signals
    skew. In HR that usually means **median for pay, tenure and
    time-to-hire**.
4.  Report **mean with SD**, or **median with IQR** — pair them
    properly.
5.  Describe a **categorical** variable with **counts** and
    **proportions**.
6.  **Histograms**, **boxplots** and **violin plots** picture numbers;
    **bar charts** picture categories. Boxplots are for analysts;
    violins travel better to managers, and neither shows the sample
    size unless you add it.
7.  **Check every column for missing values** before you summarise any
    of them, and decide what to do about the gaps deliberately.
8.  A request for **historical data is usually a question about the
    future** — and human intuition handles that step badly. Supplying
    the inference, with its uncertainty, is what the rest of this book
    is for.
9.  A 0/1 variable's **mean is a proportion** — our bridge to next
    chapter.
:::

## Next chapter {.unnumbered}

**Thinking in chances** — the first step towards supplying that
inference rather than leaving it to intuition. We turn the promotion
rate into a *probability*, and meet conditional probability: does the
chance of promotion change once we know something else about the
salesperson?
