# Thinking in Chances

## Welcome back {.unnumbered}

Last chapter we *described* data. Today we learn the language of
**uncertainty** — probability.

Still no heavy maths: we build every idea by **counting** in our data
and **simulating** on the computer.

## What you'll be able to do by the end {.unnumbered}

1.  Read a probability two ways — as a **long-run frequency** and as a
    **degree of belief** — and know when each applies
2.  State the **core rules** of probability in notation and in English
3.  Estimate probabilities directly from data by counting
4.  Combine events with the **addition** and **multiplication** rules,
    including the catches in both
5.  Compute and interpret **conditional probability**
6.  Use simulation to answer a real business question

## Why does a People Analytics book start here? {.unnumbered}

::: callout-important
Bayesian statistics is nothing more than a **disciplined way of
updating probabilities as evidence arrives**. If you're comfortable
with probability — especially *conditional* probability — the rest of
this book is mostly bookkeeping. So we'll take our time.
:::

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na(sales, customer_rate, performance, promoted)
```

------------------------------------------------------------------------

## What is a probability?

### The working definition

A probability is a number between **0** and **1** measuring how likely
something is. Zero means it cannot happen; one means it certainly
will; everything interesting lives in between.

The usual definition goes:

> the proportion of times it would happen if we could **repeat the
> situation many times**.

That's the **long-run frequency** reading, and it's the one most
people were taught. Hold onto it, but notice its limitation early: an
awful lot of the questions you'll be asked at work happen exactly
once. *Will this candidate accept the offer? Will this reorganisation
reduce attrition?* You cannot repeat next Tuesday two thousand times.

So there's a second reading, and this book will lean on it heavily:
a probability is a **degree of belief** — how confident you are,
expressed as a number, given what you currently know. The two
readings agree whenever both apply. The belief reading just also
works for the one-off questions, which is most of them.

::: callout-note
This is not a philosophical aside you can skip. The difference between
these two readings is, more or less, the difference between classical
and Bayesian statistics — and it's why a Bayesian can say "there's an
85% chance this team is genuinely underperforming" while a
frequentist, strictly, cannot.
:::

### You already reason this way

"There's a good chance she gets the promotion." · "Most applicants from
that channel accept the offer." · "About 1 in 10 new hires leave in the
first year."

Statistics just makes this precise enough to calculate with.

### A coin, and a magician

Here's a question everyone answers instantly. I toss a coin. What's
the probability of heads?

Fifty per cent, obviously. But pause on *why* you believe that. You
haven't tossed this coin. You've collected no data on it whatsoever.
You believe it because of what you know about coins in general —
they're roughly symmetric, they have two sides, nothing about the
situation favours one.

Now change one detail. **The coin was handed to you by a magician.**

Same coin, same physical symmetry, same argument as before — and
almost nobody still says 50%. Your belief moved sharply, and it moved
without a single new observation of the coin itself. What changed was
your **information about where the coin came from**.

::: callout-important
That's the whole idea in one example. A probability isn't only a
property of the object. It's a statement about the object **given
what you know** — and when what you know changes, the number should
change too.
:::

So how would you settle it? You'd toss the thing. And the more times
you tossed it, the more the running proportion of heads would tell
you what you're dealing with — the magician's coin separating itself
from an honest one as the evidence piles up.

### Watch a probability settle

Let's do exactly that, in the setting we care about. Imagine a
salesperson gets promoted with true probability 0.3 — a value we know
here only because we're inventing the data. We'll generate
salespeople one at a time and track the running proportion promoted,
just as you'd track the running proportion of heads.

```{r law-large-numbers}
set.seed(202)                                        # <1>

true_p <- 0.3
n      <- 2000

draws <- rbinom(n, size = 1, prob = true_p)          # <2>

running <- tibble(
  person       = 1:n,
  running_rate = cumsum(draws) / person              # <3>
)

ggplot(running, aes(x = person, y = running_rate)) +
  geom_line(colour = "#122a52") +
  geom_hline(yintercept = true_p, colour = "#d32f2f", linetype = "dashed") +
  labs(title = "The running promotion rate settles down",
       subtitle = "Red dashed line = the true probability (0.3)",
       x = "Number of salespeople simulated", y = "Proportion promoted")
```

1.  Seeded immediately before the simulation, so this figure is the
    same every time the book is built and every time you re-run the
    chunk.
2.  `rbinom()` with `size = 1` is a coin toss: 2,000 independent
    yes/no outcomes, each with a 30% chance of a 1. Here 1 means
    promoted.
3.  `cumsum()` gives the running total of promotions; dividing by the
    person number turns it into the running *rate*. This is the whole
    trick behind the plot.

::: callout-important
## Three lessons

1.  **Small samples are unreliable.** Early on the running rate swings
    wildly — at 20 people it could plausibly read 0.15 or 0.45. This
    is the single most under-appreciated fact in HR reporting.
2.  **It does settle.** Given enough data the estimate converges on
    the truth. That's the *law of large numbers*, and it's why more
    data helps.
3.  **Probability says nothing about the next person.** It describes
    the pattern over many, not the individual in front of you.
:::

### The rules of probability

Four rules cover almost everything. Each one is stated in notation and
in English, and none of them is harder than the English version.

**Rule 1 — probabilities live between 0 and 1.**

$$0 \le P(A) \le 1$$

*The chance of anything is somewhere between impossible and certain.*
If a calculation ever hands you 1.4, you have made an error, and it is
usually the one in Rule 3.

**Rule 2 — the possibilities add up to 1.**

$$P(A_1) + P(A_2) + \dots + P(A_k) = 1$$

*If you list every outcome that could happen, and no two of them can
happen together, their probabilities must total exactly 1.* Something
has to happen.

This rule has a consequence people genuinely resist:

::: callout-important
## When a new possibility appears, everything else must get less likely

Three internal candidates are shortlisted for a director role, and you
judge their chances at 50%, 30% and 20%. That's a complete list, and
it sums to 1.

A fourth candidate now enters the process — strong, credible, say a
30% chance in their own right. The other three probabilities **must
come down**, because the four numbers still have to total 1. Nobody's
CV changed. Nobody interviewed worse. Their chances fell anyway.

This feels wrong to people, and it's the source of a lot of muddled
thinking about hiring odds, promotion odds and forecasting. A
probability is never a property of one candidate alone — it's a share
of a fixed total, and the total is always 1.
:::

**Rule 3 — the complement.**

$$P(\text{not } A) = 1 - P(A)$$

*The chance something doesn't happen is one minus the chance it does.*
Obvious written down, and constantly useful: "at least one" questions
are almost always easier to answer as "one minus the chance of none".
If 30% of a bench gets promoted, 70% doesn't.

**Rule 4 — conditions only ever narrow things.**

$$\text{if } B \text{ implies } A, \text{ then } P(B) \le P(A)$$

*A more specific claim can never be more likely than the general one
it sits inside.* We'll come back to this one in a moment, because it's
the rule people break most often.

### Probability straight from the data

Because `promoted` is coded 0/1, the **probability someone is promoted
is just its mean**:

```{r marginal-prob}
salespeople |>
  summarise(
    n = n(),
    n_promoted = sum(promoted),
    p_promoted = mean(promoted)
  )
```

This is a **marginal probability**: the overall chance, ignoring
everything else.

------------------------------------------------------------------------

## Combining events

### Two questions, two rules

-   **OR** — what's the chance of *this or that*? → **add**
-   **AND** — what's the chance of *both together*? → **multiply**

Both have a catch, and almost everybody gets the second one wrong.

### The addition rule (OR)

$$P(A \text{ or } B) = P(A) + P(B) \quad \text{— if they can't both happen}$$

Performance is exactly one of four ratings — nobody is rated both 3
and 4 — so the chance of *either* of two ratings is simply the sum of
their proportions:

```{r addition-rule}
salespeople |>
  count(performance) |>
  mutate(prob = n / sum(n))
```

The chance of a salesperson being rated 3 **or** 4 is those two
numbers added together.

::: callout-important
## The catch: only add when they can't overlap

$$P(A \text{ or } B) = P(A) + P(B) - P(A \text{ and } B)$$

*Add the two, then subtract the overlap, or you'll count it twice.*

Say 40% of your employees completed the leadership programme and 30%
received a top performance rating. The share who did **either** is not
70%. Plenty of people did both, and adding naively counts them once as
a graduate and again as a top performer. Subtract the overlap and the
answer might be 55%.

The tell is easy: **if two things can be true of the same person, you
cannot just add.** Performance ratings are safe because they're
mutually exclusive by construction. Almost nothing else in HR is.
:::

### The multiplication rule (AND)

$$P(A \text{ and } B) = P(A) \times P(B) \quad \text{— if they're independent}$$

*Independent* means knowing one tells you nothing about the other.

Suppose roughly 25% of salespeople are top performers (`performance ==
4`) and about 30% get promoted overall. *If* those were unrelated, the
chance of being both would be roughly 0.25 × 0.30 ≈ 0.075. Let's check
reality:

```{r independence-check}
p_top      <- mean(salespeople$performance == 4)
p_promoted <- mean(salespeople$promoted)

predicted_if_independent <- p_top * p_promoted
actual_both <- mean(salespeople$performance == 4 & salespeople$promoted == 1)

tibble(predicted_if_independent, actual_both)
```

### What that mismatch means

The actual joint probability is well above the "independent"
prediction.

::: callout-important
So performance and promotion are **not independent**: knowing one
*changes* the chance of the other — which is exactly what we'd hope
a performance rating would do.

When events aren't independent, the general rule takes over:

$$P(A \text{ and } B) = P(A) \times P(B \mid A)$$

*The chance of both is the chance of the first, times the chance of
the second **given** the first has happened.* That second term is a
conditional probability, and it's the whole of the next section.
:::

### The rule almost everyone breaks

Look at what multiplication does. Both probabilities are at most 1, so
multiplying them together can only ever make the number **smaller**.
That's Rule 4 from earlier, and it holds without exception:

$$P(A \text{ and } B) \le P(A) \quad\text{and}\quad P(A \text{ and } B) \le P(B)$$

*Adding a condition can never make something more likely.*

Stated like that it's obvious. In practice people violate it constantly,
because a more detailed story sounds more plausible than a vague one:

::: callout-note
## Two ways to see it go wrong

**The forecast.** "There's a 20% chance of rain tomorrow." "And a 40%
chance of rain between 2pm and 3pm." These cannot both be true.
Raining in that hour *is* raining tomorrow — it's a subset of it — so
its probability cannot be larger. Narrowing the window can only lower
the number.

**The candidate.** "What's the chance this candidate accepts?" — 70%.
"What's the chance they accept **and** are still here in two years?"
The second must be **at most** 70%, because every person in the second
group is already in the first. Yet ask the question in that order and
people routinely give a higher figure for the richer story, because it
paints a more convincing picture.

Psychologists call this the **conjunction fallacy** — Tversky and
Kahneman's famous demonstration had people rate "bank teller and
active in the feminist movement" as more likely than "bank teller",
which is impossible.
:::

::: callout-important
The practical version, worth remembering next time you're asked for a
forecast: **every extra condition you attach to a prediction makes it
less likely to come true, not more.** Detail feels like confidence.
It's the opposite.
:::

::: callout-tip
## For the ML/DS crowd

The multiplication rule for independent events is the entire
mathematical foundation of a **naive Bayes classifier** — it
multiplies together the probability of each feature *as if* every
feature were independent given the class, which is almost never
strictly true (hence "naive") but works surprisingly well in practice.
The independence check above is exactly the kind of test you'd want to
run before trusting that assumption for a real feature set.
:::

------------------------------------------------------------------------

## Conditional probability

### The idea

**Conditional probability** asks: once I know a salesperson is a top
performer, what's the chance they're promoted? We write it **P(promoted
\| performance = 4)** — the "\|" means *given*.

::: callout-note
`performance` in this dataset is an integer from 1 to 4, where **4 is
the top rating** — so `performance == 4` is simply how a high
performer is coded here. Every organisation codes its ratings
differently, and getting this backwards is a common and embarrassing
way to invert a finding. Check the coding before you condition on it.
:::

To compute it from data you don't need a formula. Just **restrict to
the group** and take the mean.

```{r conditional-performance}
salespeople |>
  group_by(performance) |>
  summarise(
    n = n(),
    p_promoted_given_performance = mean(promoted)
  )
```

### Knowing something changes everything

::: callout-note
Same company, same promotion process — but a very different, and far
more useful, probability once you condition on performance. Good
prediction is mostly about finding the right things to condition on.
:::

### Your turn {.unnumbered}

Compute P(promoted \| a customer rating above the median) — split
`customer_rate` at its median and compute the promotion rate for each
half. Does a higher customer rating change the odds?

```{r your-turn-conditional}
# Your code here

```

### Where we're heading

::: callout-important
## The flip

Today we computed **P(promoted \| performance)** — easy, because we
could count.

From Chapter 4 we want the harder direction: **P(the true rate \| the
data we saw)**. Bayes' theorem is the bridge between those two.
Everything you did today is the foundation for it.
:::

------------------------------------------------------------------------

## Simulation for a real decision

### The succession-pipeline problem

Suppose your organisation has **10 regional manager roles** opening up
next year, filled from a bench of "ready now" candidates. Historically,
about 30% of bench candidates in any given cycle end up actually being
promoted into a role (the rest aren't ready when the role opens, decide
to leave, or the timing doesn't line up).

-   Keep exactly 10 candidates on the bench → you'll almost certainly
    come up **short**
-   Keep far too many → you're investing development budget in people
    who'll never get the call

How big should the bench be? There's no formula on the wall — but we
can *simulate* it.

### Simulate a cycle

Suppose you keep **28 candidates** on the bench, each independently
promoted with probability equal to our observed rate.

```{r bench-sim}
set.seed(203)

p_promoted <- mean(salespeople$promoted)
roles_open <- 10
bench_size <- 28
n_sims     <- 10000

promoted_count <- rbinom(n_sims, size = bench_size, prob = p_promoted)

cat("P(fewer than 10 promoted — roles left unfilled):",
    round(mean(promoted_count < roles_open), 3), "\n")
cat("P(5+ more candidates promoted than roles available):",
    round(mean(promoted_count > roles_open + 5), 3), "\n")
```

```{r bench-plot}
tibble(promoted_count) |>
  ggplot(aes(x = promoted_count)) +
  geom_histogram(binwidth = 1, fill = "#122a52", colour = "white") +
  geom_vline(xintercept = roles_open, colour = "#d32f2f", linewidth = 1) +
  labs(title = "Simulated number promoted from a 28-person bench",
       subtitle = "Bars left of the red line = roles left unfilled",
       x = "Candidates actually promoted", y = "Number of simulated cycles")
```

### The trade-off

There is no "right" answer — only a choice between the risk of unfilled
roles and the cost of an oversized bench.

::: callout-important
Statistics gives you the risks. **Choosing** between them is a
management decision — one you can now make with numbers rather than
instinct.
:::

### Your turn {.unnumbered}

Change `bench_size` to 20, then to 35. What size would you recommend,
and why?

```{r your-turn-bench}
# Your code here

```

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Almost every People Analytics question is a conditional-probability
question in disguise: *does the chance of Y change when X is
different?* "Are internal candidates more likely to succeed than
external hires?" "Does completing the training programme change
retention?" Learning to phrase questions this way — and estimate them
by conditioning — is the foundation of every model in this book.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  A **probability** is both a long-run frequency and a degree of
    belief. The belief reading is the one that works for the one-off
    questions you're usually asked — and it's the Bayesian one.
2.  Probabilities **sum to 1**, so a new possibility lowers everything
    else. Nobody's CV changed; their odds fell anyway.
3.  Estimates **settle as data grows**, and swing wildly before they
    do.
4.  For a 0/1 variable, the probability of the event is its **mean**.
5.  **Add** for OR — but subtract the overlap unless the events truly
    can't co-occur.
6.  **Multiply** for AND when independent, otherwise
    P(A) × P(B \| A). Naive Bayes classifiers lean on exactly that
    first case.
7.  **Every extra condition makes something less likely.** A more
    detailed forecast is a less probable one, however convincing it
    sounds.
8.  **Conditional probability** P(A \| B) is computed by restricting
    to group B — the seed of Bayes' theorem.
9.  **Simulation** turns a probability into a business risk you can
    act on.
:::

## Next chapter {.unnumbered}

**Distributions, sampling and the idea of a model.** We meet the
Normal, Binomial and Poisson distributions, and learn why two samples
from the same population never give quite the same answer.
