# Distributions, Sampling & the Idea of a Model

## Welcome back {.unnumbered}

This is the **bridge** chapter. Today we meet the handful of
**distributions** that describe most People Analytics data, and we
discover the single most important fact in statistics: **every sample
is a little different from the last.**

### The argument you've probably already had {.unnumbered}

You run a pulse survey in the Manchester office. Fifty responses,
average happiness **3.2** out of 4. You send it round.

A colleague runs the same survey the following week, same office,
fifty different people. She gets **2.9**.

Now watch what happens in the meeting. Somebody will want to know
which number is right. Somebody else will start looking for the
reason — *was the second week after the reorg announcement? Did your
sample catch more of the sales floor? Was the question worded the
same?* Within ten minutes there'll be a theory, and possibly an
action.

::: callout-important
Here's the possibility nobody in that room raises: **nothing
happened.** Two samples of fifty people from the same unchanged
office bounce around by a couple of tenths as a matter of course, for
no reason beyond which fifty people happened to reply.

Whether *this* gap is inside that range is a question with an actual
answer, and we'll work it out at the end of the chapter.

If you can't tell the difference between a real change and this — the
ordinary wobble of sampling — then every number you report is a
coin-flip away from a wrong conclusion, and no amount of careful
analysis afterwards will save it.
:::

By the end of this chapter you'll be able to say how big that wobble
should be, and therefore whether 3.2 versus 2.9 is worth a meeting.

## What you'll be able to do by the end {.unnumbered}

1.  Recognise the **Normal**, **Binomial** and **Poisson**
    distributions, and choose between them
2.  Explain the difference between a **population** and a **sample** —
    including why you still need statistics when you have everybody
3.  Describe **sampling variation** and the **sampling distribution**
4.  Use the **standard error** to say how much a number should be
    expected to move on its own
5.  Explain what it means to say a **model** is "signal + noise"

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na(sales, customer_rate, performance, promoted)

# Used once, for the Poisson section; we meet it properly in Chapter 8.
data("managers", package = "peopleanalyticsdata")
```

::: callout-tip
## A colour convention for this chapter

Most of the charts below compare **your data** against **a model of
it**. Throughout, navy is the data you actually have and red is the
model being held up next to it. When the two disagree, that's the
interesting bit.
:::

------------------------------------------------------------------------

## What a distribution tells you

A **distribution** describes which values are common, which are rare,
and how spread out they are — where the **middle** is, how **wide** the
spread is, whether it **leans**.

::: callout-note
Name the shape and you get a lot for free: a way to summarise it in a
couple of numbers, and a way to judge whether any observation is
surprising.
:::

Three shapes cover most People Analytics data, and choosing between
them is easier than it looks — you're not judging what the data
*looks* like, you're asking **what kind of thing you measured**.

| If one observation is… | The shape is | Example |
|-------------------|-------------------|----------------------------------|
| A **quantity** that could take any value | **Normal** | Sales, salary, time to fill a role |
| A **count of successes out of a fixed number of tries** | **Binomial** | 13 promotions out of 40 salespeople |
| A **count of events with no natural maximum** | **Poisson** | Transfer requests per team per year |

::: callout-important
Ask the question in that order — *what did I measure?* — and the
answer usually falls out. The common mistake is to reach for the
Normal by default because it's the one everybody remembers, then
apply it to a count that can't go below zero and doesn't come in
halves.
:::

------------------------------------------------------------------------

## The Normal distribution

### Why the bell curve appears *everywhere*

Because most things are pushed around by **many small, independent
influences that add up**. A salesperson's total sales are nudged by
territory, client mix, timing of deals, effort, luck… No single factor
dominates.

::: callout-important
Add lots of little nudges together and you almost always get a bell.
:::

You already meet Normals every day:

| **In daily life** | **In People Analytics** |
|---|---|
| Adult heights | Sales figures |
| Exam marks in a big cohort | Time to fill a role |
| Blood pressure readings | Daily applications received |
| Errors in careful measurement | Time to productivity for new hires |

### See it happen

Twelve small random nudges, added up:

```{r normal-emerges}
set.seed(301)

tibble(total = replicate(5000, sum(runif(12, min = -1, max = 1)))) |>   # <1>
  ggplot(aes(total)) +
  geom_histogram(bins = 45, fill = "#122a52", colour = "white") +
  labs(title = "A bell emerges from nothing",
       subtitle = "Nothing here is 'normal' by design — it simply appears",
       x = "Sum of 12 random nudges", y = "Count")
```

1.  Read this from the inside out. `runif(12, min = -1, max = 1)`
    draws twelve nudges, each equally likely to be anywhere between
    −1 and +1 — a perfectly flat distribution, nothing bell-shaped
    about it. `sum()` adds them into a single total. `replicate(5000,
    ...)` repeats that whole exercise 5,000 times. The bell is in
    none of the ingredients; it appears purely from the adding.

::: callout-note
This is the **central limit theorem**, and it's the reason the Normal
turns up so relentlessly. It isn't that nature likes bell curves. It's
that anything built by summing many small independent contributions
ends up bell-shaped almost regardless of what the contributions
looked like individually.
:::

### The anatomy of a Normal

A Normal distribution is fully described by just **two numbers**: its
mean (where the hump sits) and its SD (how wide it is). No other
shape parameters, no options. That economy is a large part of why
it's so widely used.

```{r normal-anatomy}
normal_widths <- bind_rows(
  tibble(x = seq(-4, 4, length.out = 400), width = "SD = 1") |>
    mutate(y = dnorm(x, mean = 0, sd = 1)),
  tibble(x = seq(-4, 4, length.out = 400), width = "SD = 1.8") |>
    mutate(y = dnorm(x, mean = 0, sd = 1.8))
)

ggplot(normal_widths, aes(x, y, colour = width)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("#122a52", "#3d68a8")) +
  labs(title = "Same centre, different width",
       subtitle = "The mean sets the centre; the SD sets the width. That's all.",
       x = NULL, y = "Density", colour = NULL)
```

### The 68–95 rule

Worth committing to memory, because it converts a standard deviation
into a sentence you can say out loud:

-   About **68%** of values fall within **1 SD** of the mean
-   About **95%** fall within **2 SD**

```{r rule-68-95}
normal_curve <- tibble(x = seq(-4, 4, length.out = 400)) |>
  mutate(y = dnorm(x, mean = 0, sd = 1))

ggplot(normal_curve, aes(x, y)) +
  geom_area(data = filter(normal_curve, x >= -2, x <= 2),
            fill = "#122a52", alpha = 0.20) +
  geom_area(data = filter(normal_curve, x >= -1, x <= 1),
            fill = "#122a52", alpha = 0.40) +
  geom_line(colour = "#122a52", linewidth = 1) +
  labs(title = "The 68–95 rule",
       subtitle = "Darker band ≈ 68% within 1 SD · both bands ≈ 95% within 2 SD",
       x = "Standard deviations from the mean", y = "Density")
```

::: callout-tip
Try it on something you know. If time-to-hire averages 42 days with an
SD of 10, then roughly two thirds of your roles fill between 32 and 52
days, and all but one in twenty between 22 and 62. A req that takes 70
days is genuinely unusual; one that takes 55 is not, whatever the
hiring manager says.
:::

### Now our own data

```{r sales-raw}
ggplot(salespeople, aes(x = sales)) +
  geom_histogram(bins = 30, fill = "#122a52", colour = "white") +
  labs(title = "Sales across the sales team",
       subtitle = "Does this look like the bell we've been drawing?",
       x = "Sales (thousands of dollars)", y = "Number of salespeople")
```

### Does a Normal actually fit?

```{r sales-normal-overlay}
mean_sales <- mean(salespeople$sales)
sd_sales   <- sd(salespeople$sales)

ggplot(salespeople, aes(x = sales)) +
  geom_histogram(aes(y = after_stat(density)),                        # <1>
                 bins = 30, fill = "#122a52", colour = "white", alpha = 0.85) +
  stat_function(fun = dnorm,                                          # <2>
                args = list(mean = mean_sales, sd = sd_sales),
                colour = "#d32f2f", linewidth = 1.1) +
  labs(title = "Sales with a Normal curve laid over it",
       subtitle = paste0("Normal with mean ", round(mean_sales),
                         " and SD ", round(sd_sales)),
       x = "Sales (thousands of dollars)", y = "Density")
```

1.  `after_stat(density)` rescales the histogram from counts to
    density, so its total area is 1. Without it the bars and the
    curve would be on completely different vertical scales and the
    comparison would be meaningless.
2.  The Normal we're testing against, given the mean and SD of the
    real data. We're not fitting anything — just asking whether the
    simplest possible summary of this variable does a decent job.

::: callout-note
The Normal describes the middle of the data reasonably well and may
miss a little in the tails — that's fine. A distribution is a **useful
simplification**, not a claim that reality is exactly this shape.
:::

------------------------------------------------------------------------

## The Binomial distribution

### When the outcome is yes or no

Some observations have only two outcomes, and we care **how many** went
one way:

-   10 job offers → how many accepted?
-   20 interviews → how many led to a hire?
-   **50 salespeople → how many get promoted?**

That count is **Binomial**, and it needs just two things:

-   **n** — how many opportunities
-   **p** — the chance each one is promoted

### Promotions out of 50 salespeople

```{r binomial}
p_promoted <- mean(salespeople$promoted)

tibble(promotions = 0:50) |>
  mutate(prob = dbinom(promotions, size = 50, prob = p_promoted)) |>
  ggplot(aes(x = promotions, y = prob)) +
  geom_col(fill = "#122a52", alpha = 0.85) +
  labs(title = "Binomial: promotions out of 50 salespeople",
       subtitle = paste0("Each salesperson is promoted with probability ",
                         round(p_promoted, 2)),
       x = "Number promoted", y = "Probability")
```

Bars, not a curve — you can promote 14 people or 15, never 14.5.

::: callout-tip
Read that chart as a range, not a point. With a 30% rate and 50
salespeople you'd *expect* 15 promotions, but anywhere from about 9 to
21 is unremarkable. So a division that promoted 11 this year and 18
last year has not necessarily changed anything at all — which is the
Manchester survey problem again, in a different form.
:::

::: callout-important
## Next chapter we run this backwards

Today: *given* a rate, how likely is each count?

Next chapter: we saw **13 promotions in 40 salespeople** — what is the
true **rate**? The Binomial is what lets us answer that.
:::

------------------------------------------------------------------------

## The Poisson distribution

### Counting events in a window

Sometimes there's no natural *n* — events just happen at some average
rate: complaints per month, transfer requests per team, applications
per posting. The Poisson needs only **one** number: the rate, **λ**
(lambda).

### Our data: transfer requests

We'll get properly acquainted with the `managers` dataset in Chapter 8
— for now, a first look at `transfers`, the number of transfer requests
filed by a manager's team:

```{r poisson-fit}
transfer_rate <- mean(managers$transfers, na.rm = TRUE)

managers |>
  count(transfers) |>
  mutate(observed = n / sum(n),
         poisson  = dpois(transfers, lambda = transfer_rate)) |>
  pivot_longer(c(observed, poisson), names_to = "source", values_to = "prob") |>
  ggplot(aes(x = transfers, y = prob, fill = source)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c("#122a52", "#d32f2f")) +
  labs(title = "Transfer requests vs a Poisson distribution",
       subtitle = paste0("Poisson rate (lambda) = ", round(transfer_rate, 2)),
       x = "Number of transfer requests", y = "Probability", fill = NULL)
```

Navy is what the managers actually filed; red is what a Poisson with
the same average rate would produce. They're close but not identical,
and the gaps are where the model's assumptions are showing.

::: callout-important
A distribution is a **model** — a useful approximation, never the
literal truth. The skill is knowing when the approximation is good
enough — and Chapter 15 comes back to this exact variable to show what
happens when you don't account for how much opportunity each manager's
team size gave for a transfer request to occur.
:::

------------------------------------------------------------------------

## Population vs sample

### You never see the whole picture

-   The **population** is everyone you care about — every employee,
    every candidate, ever
-   The **sample** is the handful you actually measured

::: callout-important
You compute everything from the sample, but want to say something
about the population. **That gap is where statistics lives.**
:::

### "But I have everyone" {.unnumbered}

This is the objection every People Analytics practitioner raises, and
it deserves a straight answer. You don't have a sample of employees.
You have the HRIS. Every single person, all 4,000 of them. Surely the
attrition rate you compute *is* the attrition rate, full stop — no
uncertainty, nothing to estimate?

It's a fair challenge, and the answer is that you're almost never
actually asking about those 4,000 people.

Think about what the question is for. "Our attrition was 14% last
year" is a complete and exact description of a year that has finished
and cannot be changed. Nobody needs it. What they want to know is
whether 14% is what this organisation *tends to produce* — and what it
will do next year, and whether the retention programme moved it.

Those are questions about the **process**, not the headcount. And of
that process, your 4,000 employees over one year are a sample: one
draw from all the ways last year could have gone. Twenty of those
leavers might easily not have left. A different competitor might have
opened an office nearby. Run the year again and you'd get 13%, or 16%.

::: callout-important
**You have the whole population of people, and a sample of one year
from the process that generated them.** The moment your question
concerns next year, or a comparison, or a cause, you are back to
inference — and the uncertainty is real even though the count is
exact.

This is also why small teams are so treacherous. A twelve-person team
with three leavers has 25% attrition, exactly and unarguably. As an
estimate of what that team *tends* to do, it's nearly worthless.
Chapters 8, 14 and 15 are about fixing precisely that.
:::

### Today we'll cheat

We'll treat all `r nrow(salespeople)` salespeople as if they were the
whole **population**, so we know the "true" average sales figure:

```{r population-mean}
population_mean_sales <- mean(salespeople$sales)
population_mean_sales
```

### Take a sample — then take another

```{r repeated-samples}
set.seed(302)

mean(slice_sample(salespeople, n = 50)$sales)
mean(slice_sample(salespeople, n = 50)$sales)
mean(slice_sample(salespeople, n = 50)$sales)
```

Three analysts, three answers, one unchanged company — and not one of
them has made a mistake.

::: callout-important
**This is the Manchester survey.** Nothing was different between those
three samples: not the week, not the reorg, not the question wording.
The only thing that changed was which fifty rows came out.

If you had received only the first number, you would have believed it.
:::

------------------------------------------------------------------------

## Sampling variation

### What if we did it a thousand times?

```{r sampling-distribution}
set.seed(303)

sample_means_50 <- replicate(                                    # <1>
  1000,
  mean(slice_sample(salespeople, n = 50, replace = TRUE)$sales)  # <2>
)

tibble(mean_sales = sample_means_50) |>
  ggplot(aes(x = mean_sales)) +
  geom_histogram(bins = 40, fill = "#122a52", colour = "white") +
  geom_vline(xintercept = population_mean_sales,
             colour = "#d32f2f", linewidth = 1) +
  labs(title = "Sampling distribution of the mean sales figure (n = 50)",
       subtitle = "Red = the truth. Estimates scatter around it, but centre on it.",
       x = "Sample mean sales", y = "Number of samples")
```

1.  `replicate()` runs the expression inside it 1,000 times and
    collects the answers. Here each run is a whole imaginary study:
    survey 50 people, compute their mean. We end up with 1,000
    studies that could have happened.
2.  `replace = TRUE` puts each salesperson back before the next is
    drawn, so every one of the 50 is an independent pick. We're
    treating our `r nrow(salespeople)` as a stand-in for a much larger
    workforce rather
    than a fixed pool being used up — which is what makes the
    arithmetic below come out cleanly.

::: callout-important
Look at the width of that histogram. Every bar is a mean that a
competent analyst could have reported, from the same company, on the
same day. **The spread of this histogram is the honest margin of error
on any one of them** — and it is the thing that almost never appears
on the slide.
:::

::: callout-note
## A different lens: this is a frequentist idea, on purpose

Repeating an imaginary study over and over and asking how estimates
would scatter is the **classical, frequentist** way of thinking about
uncertainty — and it's genuinely useful for building intuition, which
is why we've used it all chapter. Next chapter we introduce the
Bayesian alternative: instead of imagining repeated studies we haven't
run, we ask directly *"given the one sample I actually have, which
values are plausible?"* Both are legitimate; they're answering subtly
different questions, and it's worth noticing the switch when it
happens.
:::

### Bigger samples wobble less

```{r bigger-samples}
set.seed(304)

sample_means_200 <- replicate(
  1000,
  mean(slice_sample(salespeople, n = 200, replace = TRUE)$sales)
)

bind_rows(
  tibble(mean_sales = sample_means_50,  n = "n = 50"),
  tibble(mean_sales = sample_means_200, n = "n = 200")
) |>
  mutate(n = factor(n, levels = c("n = 50", "n = 200"))) |>      # <1>
  ggplot(aes(x = mean_sales, fill = n)) +
  geom_histogram(bins = 45, alpha = 0.65, position = "identity") +
  geom_vline(xintercept = population_mean_sales,
             colour = "#d32f2f", linewidth = 1) +
  scale_fill_manual(values = c("#8fabd0", "#122a52")) +
  labs(title = "Four times the data — about half the spread",
       x = "Sample mean sales", y = "Number of samples", fill = NULL)
```

1.  Setting the factor levels explicitly. Without this, ggplot orders
    the groups alphabetically — which puts "n = 200" before "n = 50"
    and quietly swaps the two colours relative to what you intended.
    A small thing that produces a confidently mislabelled chart.

::: callout-tip
## For the ML/DS crowd

Simulating "what if I drew a different sample" by resampling from data
you already have is exactly the **bootstrap** — a technique you've
likely used to put a confidence interval on a model metric. The
sampling-distribution plots above *are* a bootstrap, just aimed at
building intuition rather than producing a specific interval.
:::

### Putting numbers on it

That spread has a name — the **standard error** — and you don't have
to simulate it. There's a formula, and it's the most useful one in
this chapter:

$$\text{SE} = \frac{\sigma}{\sqrt{n}}$$

*How much your estimate wobbles equals how much the underlying data
varies, divided by the square root of how many people you measured.*

Two things follow immediately. Noisier data means a noisier estimate,
which is unsurprising. And precision improves with the **square root**
of sample size, which is where intuition fails.

```{r se-table}
set.seed(305)

sample_means_all <- replicate(
  1000,
  mean(slice_sample(salespeople, n = nrow(salespeople), replace = TRUE)$sales)
)

tibble(
  sample_size = c(50, 200, nrow(salespeople)),
  simulated   = c(sd(sample_means_50), sd(sample_means_200), sd(sample_means_all)),
  formula     = sd_sales / sqrt(sample_size)                       # <1>
)
```

1.  The theoretical standard error, straight from the formula above.
    Compare the two columns: the simulation and the arithmetic agree
    closely. That's worth seeing once — it means you can get this
    number without simulating anything.

::: callout-important
## The square root is why precision is expensive

To **halve** your uncertainty you need **four times** the data. To cut
it to a tenth, a hundred times.

Put that in survey terms. Going from 100 respondents to 400 is a hard
quarter's work and buys you a halving of the error bar. Going from 400
to 1,600 — a genuinely enormous undertaking in most organisations —
buys you another halving, and no more.

Two practical consequences. **Small samples improve cheaply**: if
you're at 30 responses, getting to 120 is transformative and probably
achievable. **Large samples improve expensively**: at 2,000 responses,
chasing another 500 will not visibly move anything, and that effort is
almost always better spent asking a better question of the people
you've already got.
:::

### Your turn {.unnumbered}

Repeat the sampling-distribution experiment for the **promotion rate**
(mean of `promoted`) instead of sales, using samples of 50. Is it
centred on the population rate?

```{r your-turn-sampling}
# Your code here

```

------------------------------------------------------------------------

## So: 3.2 versus 2.9?

We now have everything needed to answer the question the chapter
opened with, and it takes three lines.

Happiness sits on a 4-point scale, and scores like that typically have
a standard deviation somewhere around 0.8. With 50 respondents:

```{r manchester}
sd_happiness <- 0.8
n_responses  <- 50

se_one_survey <- sd_happiness / sqrt(n_responses)
se_difference <- se_one_survey * sqrt(2)                 # <1>

tibble(se_one_survey, se_difference,
       observed_gap = 3.2 - 2.9,
       gap_in_se    = 0.3 / se_difference)
```

1.  We're comparing **two** surveys, and both wobble. When you
    subtract two independent estimates their uncertainties combine —
    not by adding, but by adding the *squares* and taking the root,
    which for two equal-sized surveys works out as multiplying by
    √2.

So a single survey of 50 people carries a standard error of about
**0.11**, and the gap between two such surveys has a standard error of
about **0.16**.

::: callout-important
## The answer

The observed gap of 0.3 is a bit under **two standard errors**. That
puts it at the edge: larger than the everyday wobble, but a gap that
size will still turn up by chance roughly one time in twenty when
nothing whatsoever has changed.

So the honest answer in that meeting is neither "engagement has
dropped" nor "it's just noise". It's: *"that's on the boundary of what
two identical surveys would differ by anyway. It's worth a third
survey, or a bigger one. It is not yet worth a theory about the
reorg."*

Notice what the arithmetic bought you. Not certainty — it doesn't
exist here — but the ability to say **how surprised to be**, which is
the difference between a considered response and a room full of
speculation.
:::

::: callout-note
That answer still leans on frequentist machinery: it asks how often a
gap this big would appear if nothing had changed. From next chapter
we'll ask the question the room actually wanted answered — *what's the
probability engagement really did fall?* — which turns out to be a
different, and more useful, calculation.
:::

------------------------------------------------------------------------

## The idea of a model

### Signal and noise

Every observation can be thought of as:

$$
\text{observation} = \text{signal} + \text{noise}
$$

-   The **signal** is the systematic part we care about — the true
    average sales figure, the true promotion rate, the real effect
-   The **noise** is everything else — the random wobble that makes
    each salesperson, and each sample, different

::: callout-note
A **model** says: "the signal looks like this, the noise looks like
that." Every distribution today is a candidate description of the
**noise**.
:::

### Where this is heading

Frequentist statistics answers *"how much would my estimate wobble if
I repeated the study?"*

Bayesian statistics answers the more direct question:

> *"Given the one sample I actually have, which values of the signal
> are plausible — and how sure am I?"*

Next chapter we start answering it.

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Your engagement survey, your attrition figures, your last headcount
report — all of it is **one sample**. Everything you report is an
estimate that would have come out a little differently with a
different group of respondents or a different month. That's the
difference between "engagement is down" and the honest, defensible
version: *"engagement is down in this survey, by roughly this much,
give or take this much."*

The habit to build: **before you explain a movement, work out how big
a movement would have been unremarkable.** It takes a minute with the
standard error formula, and it will stop you building a narrative on
top of noise more often than any other single thing in this book.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Choose a distribution by asking **what you measured**, not what
    the histogram looks like: **Normal** for quantities, **Binomial**
    for successes out of a fixed number of tries, **Poisson** for
    counts with no natural maximum.
2.  The Normal turns up everywhere because **sums of many small
    influences are bell-shaped**, whatever the influences look like
    individually.
3.  A distribution is a **model** — a useful approximation, not the
    exact truth.
4.  A **sample** estimate is not the **population** value; it wobbles.
    Having every employee doesn't exempt you: your question is
    usually about the **process**, and of that you have one year's
    draw.
5.  The **sampling distribution** is centred on the truth and tighter
    with more data — a frequentist idea we've used to build intuition
    before switching lenses next chapter.
6.  **SE = σ / √n.** Precision improves with the *square root* of
    sample size, so halving your error bar costs four times the data.
    Small samples improve cheaply; large ones don't.
7.  Before explaining a movement, **work out how big a movement would
    have been unremarkable**.
8.  Data = **signal + noise**.
:::

## Next chapter {.unnumbered}

**Bayesian thinking — Bayes' theorem and your first inference.** We
finally put prior, likelihood and posterior together and estimate a
promotion rate the Bayesian way.
