# Bayesian Thinking: Bayes' Theorem & Your First Inference

## Welcome to the Bayesian part! {.unnumbered}

You now have everything you need: you can describe data (Chapter 1),
think in probabilities (Chapter 2), and you understand distributions
and sampling wobble (Chapter 3). Today those pieces click together.

::: callout-note
Nothing new and difficult arrives today — we just point what you
already know in a new direction.
:::

## What you'll be able to do by the end {.unnumbered}

1.  Turn a stakeholder's vague hunch into a **question data can
    answer**
2.  State Bayes' theorem: **prior × likelihood → posterior**
3.  Compute a posterior using **grid approximation** — and know why
    you won't use it for long
4.  See how the **prior** changes the answer — and where a defensible
    prior comes from
5.  Read a **Beta** and a **Binomial**, and use the Beta-Binomial
    shortcut
6.  Update **sequentially** and answer a decision question

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na(sales, customer_rate, performance, promoted)
```

------------------------------------------------------------------------

## The idea

### It starts with a hunch, not a question

Someone says something in a meeting. In my experience that is almost always how these types of things
begin.

> *"I don't think we're promoting enough of our salespeople. That's
> why people leave — they can't see a path."*

It's your VP of Sales, and she may well be right. She has run that
team for eleven years. But as it stands you can't do anything with
what she's said. It isn't a question, it's a feeling — an informed
one, formed over a long time, but not yet something data can answer.

### Turning it into something answerable

What happens next matters more than anything statistical you'll do
afterwards, and it isn't statistics at all. You have to negotiate that
hunch down into a question that has an answer.

"Are we promoting enough?" is really at least three separate questions
hidden inside one:

-   **What *is* our promotion rate?** A number nobody currently knows.
-   **Is it lower than it should be?** Needs a comparison — last year,
    another function, an industry benchmark.
-   **Is it why people leave?** A different analysis altogether, and a
    much harder one (Chapter 17).

You can't do all three at once, and the second and third both depend
on the first. So you start there. Even then it isn't precise enough:
promoted over what period, and out of whom? A short conversation
settles it:

> **Of the salespeople we hire, what proportion are promoted at least
> once?**

That's answerable. It has exactly one unknown number attached to it,
which is the shape of problem this chapter solves.

### What data do you actually have?

Rarely all of it. Promotions have been recorded properly since the
HRIS migration, which means you have reliable flags for **40
salespeople**. The rest are in the old system, in a field three
different people used three different ways. Those different ways, which nobody can really pinpoint, suggest that you should almost certainly start with the newer data.

Forty is not many. Your first instinct is to count the promotions,
divide by 40, and report a percentage — and that will certainly
produce a number. But Chapter 3 already warned you what happens to a
number built from 40 people: it wobbles. Draw a different 40 and
you'd get a different percentage. A single figure on a slide hides
that entirely.

It also throws away something valuable. Your VP has watched this
process for over a decade. She has a view about the promotion rate,
and it is not worthless just because it isn't a spreadsheet.

### Updating a belief

So you have two sources of information, and neither is good enough
alone. Experience without data is an anecdote. Forty rows without
context is noise.

::: callout-important
You don't throw away your colleague's opinion, and you don't ignore
the numbers. You **combine** them into a new, better-informed belief.

That combination is all Bayes' theorem does. The rest is arithmetic.
:::

### The three ingredients

$$
\underbrace{P(\theta \mid \text{data})}_{\text{posterior}} \;\propto\;
\underbrace{P(\text{data} \mid \theta)}_{\text{likelihood}} \;\times\;
\underbrace{P(\theta)}_{\text{prior}}
$$

-   **Prior** $P(\theta)$ — what you believed *before* seeing this data
-   **Likelihood** $P(\text{data} \mid \theta)$ — how well each possible
    answer *explains* the data you saw
-   **Posterior** $P(\theta \mid \text{data})$ — your updated belief

Our unknown $\theta$ is the **true promotion rate** — the rate the
process would produce in the long run, not the count in any
particular group of 40 people.

::: callout-note
## A different lens: you are not trying to find the true value

If you were trained the classical way, this is the moment to adjust
your grip. There, θ is a fixed unknown constant and the job is to
estimate it — to get as close as possible and quantify how far off
you might be. The answer is a number, with an error bar attached to
express regret about not knowing more.

Bayesian practice keeps the same θ and changes the job. **You will
never know θ exactly, and pinning it down isn't the goal.** The goal
is to hold an honest, well-calibrated description of what θ could be,
and to sharpen that description every single time new evidence
arrives.

So the posterior isn't a stepping stone towards a real answer. It *is*
the answer — and it's an answer built to be revised. This quarter's
posterior becomes next quarter's prior, as you'll see at the end of
this chapter. Learning doesn't finish; it just gets less uncertain.
:::

------------------------------------------------------------------------

## Our question

### What is the true promotion rate?

The 40 salespeople with clean promotion records are all we have to
learn from.

```{r the-data}
set.seed(404)                                       # <1>

sample40 <- salespeople |> slice_sample(n = 40)     # <2>

n_people   <- nrow(sample40)
k_promoted <- sum(sample40$promoted)                # <3>

cat("Salespeople:", n_people, " Promoted:", k_promoted,
    " Observed rate:", round(k_promoted / n_people, 3), "\n")
```

1.  Any chunk that draws a random sample needs a seed, or the numbers
    in the text change every time the book is rebuilt. Set it
    **immediately before** the sampling, not just once at the top of
    the chapter, so the result survives you re-running chunks out of
    order.
2.  A stand-in for the 40 clean records. The full dataset has
    `r nrow(salespeople)` salespeople once Chapter 1's missing values
    are dropped; we're deliberately handicapping ourselves to the
    situation you'll usually be in.
3.  `promoted` is coded 0/1, so summing it counts the promotions.

Our observed rate is `r round(k_promoted/n_people, 3)` — but with only
`r n_people` salespeople we know (Chapter 3!) that this estimate
wobbles.

::: callout-note
Rather than a single estimate, Bayes gives us a **whole distribution**
of plausible true rates, with more weight on the ones that explain the
data best. That distribution is the **posterior**.
:::

------------------------------------------------------------------------

## Grid approximation

### What it is

The promotion rate could be any number between 0 and 1. That's
infinitely many possibilities, and combining a prior and a likelihood
across all of them is, in general, a calculus problem.

**Grid approximation dodges the calculus by refusing to consider
infinitely many values.** Instead we line up a manageable number of
candidates — 0.00, 0.01, 0.02, all the way to 1.00 — and score each
one in turn. A hundred and one candidates is a poor substitute for a
continuum in principle, and a perfectly good one in practice.

### The strategy

1.  List many **candidate** values for the true rate
2.  Score each by its **prior** — how plausible was it before?
3.  Score each by its **likelihood** — how well does it explain the
    data?
4.  **Multiply**, then normalise so the scores sum to 1

That's it — no calculus required.

```{r grid}
p_grid <- seq(from = 0, to = 1, length.out = 101)
```

### Why we're doing it this way

Because you can watch it happen. Every step above is arithmetic you
could do in a spreadsheet, and by the end of this chapter you'll have
seen a posterior assembled from its parts with nothing hidden. That's
worth a great deal — when `brms` produces a posterior in the next
chapter, you'll know what it's producing, rather than trusting it.

::: callout-important
## And why you'll almost never use it again

Grid approximation is a teaching tool that runs out of road fast, and
it's worth knowing exactly where.

Here we have **one** unknown, so 101 candidates means 101
calculations. Add a second unknown and you need a grid across both:
101 × 101 = **10,201**. A third takes you past a million. A modest
model with five parameters — which is nothing, that's an intercept
and four predictors — needs 101⁵, about **ten billion**
evaluations, and it will still be a coarse grid.

This is the *curse of dimensionality*, and it is the reason Bayesian
statistics stayed impractical for two centuries. The fix is to stop
scoring every point and instead go wandering through the space,
spending time in proportion to how plausible each region is — which
is what Chapter 5's sampler does, and what Chapter 11 opens up if you
want to see inside it.
:::

### Step 1 — the prior, and look at it

We'll start with a **flat** prior: every rate equally plausible.

That is not what we believe. We opened this chapter by saying the VP's
eleven years of experience was worth keeping, and a flat prior throws
all of it away. We start here anyway, for one chapter only, because it
makes the machinery visible: with a flat prior the posterior is driven
entirely by the data, so anything that moves in the plots below moved
because of the 40 salespeople and nothing else. It's a controlled
baseline, not a recommendation.

Her real prior arrives a few sections from now, once you've seen what
the flat one does — and then you'll be able to see exactly how much
difference she makes.

```{r flat-prior}
prior_flat <- rep(1, length(p_grid))

tibble(p = p_grid, prior = prior_flat) |>
  ggplot(aes(p, prior)) +
  geom_area(fill = "#8fabd0", alpha = 0.30) +
  geom_line(colour = "#8fabd0", linewidth = 1) +
  labs(title = "A flat prior: all promotion rates equally likely",
       subtitle = "Deliberately uncommitted — we let the data do the talking",
       x = "Promotion rate", y = "Prior (relative)")
```

::: callout-important
**Always visualise your prior.** It's the assumption you're most often
asked to defend.
:::

### Step 2 — the likelihood

For each candidate rate, how probable is our data (`r k_promoted`
promotions in `r n_people` salespeople)? That's the Binomial from
last chapter:

```{r likelihood}
likelihood <- dbinom(k_promoted, size = n_people, prob = p_grid)  # <1>

tibble(p = p_grid, likelihood = likelihood) |>
  ggplot(aes(p, likelihood)) +
  geom_area(fill = "#d32f2f", alpha = 0.30) +
  geom_line(colour = "#d32f2f", linewidth = 1) +
  labs(title = "Likelihood: which rates explain our data best?",
       x = "Promotion rate", y = "Likelihood")
```

1.  Read this as a question asked 101 times over: *if the true rate
    were `prob`, how probable is it that we'd have seen exactly
    `k_promoted` promotions among `n_people`?* The data stays fixed
    and the candidate rate varies — which is the opposite of how you
    normally use `dbinom()`, and the single most common place people
    trip up.

### Step 3 — multiply and normalise

Bayes' theorem is applied to each candidate rate on its own: the prior
score for 0.30 multiplies the likelihood score for 0.30, and nothing
else. `likelihood * prior_flat` does all 101 of those in one line —
`*` in R pairs the two vectors up position by position and multiplies
each pair, returning a vector the same length again. (This is
element-wise multiplication, not matrix multiplication — `%*%` is the
operator for that, and it isn't what we want here.) Dividing by the
total then rescales the 101 scores so they sum to 1, which is what
makes them a distribution rather than a set of relative weights.

::: callout-tip
Worth knowing before it causes a problem: if the two vectors are *not* the
same length, R still hands you an answer — it recycles the shorter one
from the start and pairs it up again. If the longer length is an exact
multiple of the shorter it does this silently; otherwise you get a
warning, but a result all the same. Both of ours are 101 long, so the
pairing is the obvious one, but a length mismatch elsewhere can give you
a wrong answer rather than an error.
:::

```{r posterior}
posterior <- likelihood * prior_flat
posterior <- posterior / sum(posterior)

tibble(p = p_grid,
       prior = prior_flat / sum(prior_flat),
       likelihood = likelihood / sum(likelihood),
       posterior = posterior) |>
  pivot_longer(-p, names_to = "curve", values_to = "density") |>
  mutate(curve = factor(curve, c("prior", "likelihood", "posterior"))) |>
  ggplot(aes(p, density, colour = curve)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#8fabd0", "#d32f2f", "#122a52")) +
  labs(title = "Prior × Likelihood → Posterior",
       x = "Promotion rate", y = "Density (normalised)", colour = NULL)
```

### Summarising the posterior

```{r posterior-summary}
post_mean <- sum(p_grid * posterior)

cum <- cumsum(posterior)
ci_low  <- p_grid[which.min(abs(cum - 0.025))]
ci_high <- p_grid[which.min(abs(cum - 0.975))]

cat("Posterior mean:", round(post_mean, 3), "\n")
cat("95% credible interval: [", round(ci_low, 3), ",", round(ci_high, 3), "]\n")
```


### Say it out loud

> *"There is a **95% probability** that the true promotion rate lies
> between `r round(ci_low,2)` and `r round(ci_high,2)`."*

::: callout-note
## A different lens: this sentence isn't available to a frequentist interval

A classical 95% **confidence interval** is built the other way round:
it's a procedure that, *if you repeated the whole study many times*,
would produce an interval containing the true rate 95% of the time —
it is not, strictly, a 95% probability statement about this one
interval. A Bayesian **credible interval** is a direct probability
statement about the parameter, given the data you actually have. Both
are legitimate — they answer different questions — but the plain-English
reading most people reach for instinctively ("there's a 95% chance the
true value is in here") is only correct for the credible interval,
which is one reason Bayesian results are often easier to present to a
non-technical stakeholder without a caveat.
:::

------------------------------------------------------------------------

## Priors

### Where does a prior come from?

Four places, roughly in ascending order of effort:

-   **Physical and logical limits** — a rate can't be below 0 or above 1. Free, and more useful than it sounds.
-   **Your own data** — previous cohorts, last year's figures, the
    same measure in a neighbouring function.
-   **Domain knowledge** — "promotion rates in sales organisations
    tend to run 20–40% over a few years".
-   **Published research on the process itself** — the section below.

And if you want to go further, Chapter 23 covers a formal process for
eliciting a prior directly from an expert's judgement, rather than
inferring one from a simple rule.

### You already have a prior — the question is whether you'll admit it

An analyst who has worked in a domain for any length of time always
has some expectation of what they're about to see. You know roughly
what attrition looks like in your organisation. You'd be startled by
a 90% promotion rate and unsurprised by 25%.

That expectation exists whether or not you write it down. A Bayesian
prior isn't the introduction of bias into a neutral process — it's
the act of stating the expectation you were carrying anyway, in a
form that can be inspected, challenged and overruled by data.

But there's a higher standard than "what I'd expect". You should have
a view about **the process you're analysing**, not just about the
numbers it tends to produce. Why would a promotion rate be what it
is? What determines it — span of control, growth rate, tenure
distribution, how the ladder is designed? An analyst who can answer
that has a prior worth having. An analyst who can't is guessing with
extra steps.

### Theory as a starting point — and why desk research earns its keep

This is where **evidence-based management** has something concrete to
offer. The discipline argues that good decisions draw on several
sources of evidence at once: the scientific literature, your
organisation's own data, practitioner expertise, and the concerns of
the people affected. In practice the first of those is the one
analysts skip. There's a strong pull to open the data immediately and
find out what's happening, and reading around the problem first feels
like a detour.

It isn't. Someone has almost certainly studied promotion rates,
internal mobility, or whatever your process is — across many more
organisations than you have access to. That literature is a genuine
source of information about your problem, and Bayesian analysis gives
you somewhere to *put* it. Desk research stops being background
reading and becomes an input to the model.

It has a second use, further down the line. Chapter 21 asks you to draw
what you believe causes what before deciding which variables belong in
a model. The same literature that gives you a prior is usually the best
available evidence for that drawing — someone else has already tested
whether the arrow you are about to draw exists.

::: callout-important
## Theory in, scepticism intact

The point is emphatically **not** to adopt published findings and
assume they hold in your organisation. Research on other firms in
other sectors may or may not generalise to yours — often it doesn't.

What you're doing is:

1.  Take what the theory suggests, and encode it as a prior — a
    **starting position**, held with a strength that reflects how much
    you trust it.
2.  Confront it with your own data.
3.  Report where you ended up, and how far the data moved you.

That's a disciplined way to use the best available thinking while
staying sceptical of it. If your data disagrees with the literature,
a weakly-held prior will get overruled and you'll see it happen. If
the two agree, you've earned a sharper answer than either could give
alone. Either way the assumption was on the table from the start.

Compare that with the alternative on offer: a method that claims to
use no prior information at all, run by an analyst who has plenty and
no legitimate place to declare it.

In fairness, classical practice does let prior-like information in — it
just doesn't call it a prior. Ridge regression is a Gaussian prior on
the coefficients; the LASSO is a Laplace one; empirical Bayes (Chapter
14) estimates the prior from the data itself. That strengthens the
point rather than weakening it. Everyone has a prior. The question is
only whether it is written down somewhere you can argue with it.
:::

::: callout-note
A prior is an assumption you **state openly and can defend** — which
is more honest than the hidden assumptions in any other method.
:::

### An experienced VP of Sales' prior

Our VP believes the rate is usually around 0.3, and she's confident
enough to argue for it but not so confident she'd bet the department.
We encode that as Beta(9, 21) — a shape we'll unpack properly in a
moment:

Note how that happened: we listened to her, and chose two numbers.
It works, and you should be slightly uncomfortable with it — a
different analyst in the same conversation could easily have landed on
Beta(6, 14). Chapter 23 replaces this with a structured elicitation
process that asks her answerable questions and derives the shape from
her answers, so the prior belongs to her rather than to your
interpretation of her.

```{r informed-prior-plot}
tibble(p = p_grid, informed = dbeta(p_grid, shape1 = 9, shape2 = 21)) |>
  ggplot(aes(p, informed)) +
  geom_area(fill = "#8fabd0", alpha = 0.30) +
  geom_line(colour = "#8fabd0", linewidth = 1) +
  labs(title = "An informed prior: Beta(9, 21)",
       subtitle = "Most belief between about 0.15 and 0.45",
       x = "Promotion rate", y = "Prior density")
```

### What it does to the answer

```{r compare-posteriors}
post_informed <- likelihood * dbeta(p_grid, shape1 = 9, shape2 = 21)
post_informed <- post_informed / sum(post_informed)

tibble(p = p_grid, flat = posterior, informed = post_informed) |>
  pivot_longer(-p, names_to = "which", values_to = "density") |>
  ggplot(aes(p, density, colour = which)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = k_promoted / n_people, linetype = "dashed", alpha = 0.5) +
  scale_colour_manual(values = c("#122a52", "#d32f2f")) +
  labs(title = "The prior nudges the posterior",
       subtitle = "Dashed line = observed rate in our 40 salespeople",
       x = "Promotion rate", y = "Posterior density", colour = "prior")
```

::: callout-important
## The teaching point

With only 40 salespeople, the prior still has a visible say. Get to a
few hundred and the likelihood dominates — the prior's influence
fades away without you having to do anything about it.

Priors matter most exactly when data is **scarce**, which is often the
case in People Analytics — a new office, a new role, a pilot
programme.
:::

### Your turn {.unnumbered}

Try a **sceptical** prior centred on 0.5,
`dbeta(p_grid, shape1 = 20, shape2 = 20)`.
Compute and plot its posterior alongside the others. Where does it
land?

```{r your-turn-prior}
# Your code here

```

------------------------------------------------------------------------

## The Beta-Binomial shortcut

### Two distributions, two different jobs

We've now used both of these, so it's worth being precise about what
each one is for. They're easy to confuse because both involve
proportions — but they describe different things.

**The Binomial is a distribution over counts.** Fix a rate and a
number of people, and it tells you how many promotions to expect:

```{r binomial-shape}
tibble(promotions = 0:40,
       probability = dbinom(0:40, size = 40, prob = 0.3)) |>
  ggplot(aes(promotions, probability)) +
  geom_col(fill = "#d32f2f", alpha = 0.45, width = 0.7) +
  labs(title = "Binomial: if the true rate really were 0.30, what would we see?",
       subtitle = "40 salespeople — counts near 12 are common, but 8 or 17 wouldn't be strange",
       x = "Number promoted out of 40", y = "Probability")
```

Notice it's made of **bars**, not a curve. You can promote 11 people
or 12, never 11.5. This is the distribution that generates data, which
is why it plays the role of the likelihood.

**The Beta is a distribution over rates.** It lives strictly between 0
and 1 and describes how plausible each possible *rate* is — belief
about a parameter, not a count of people. That makes it the natural
shape for a prior or a posterior about a proportion.

### Reading a Beta

`Beta(a, b)` has two parameters and the useful way to read them is as
**imaginary observations**: roughly, `a − 1` successes and `b − 1`
failures.

The awkward `− 1` is bookkeeping, and it comes from where the counting
starts. The no-evidence Beta is `Beta(1, 1)`, not `Beta(0, 0)` — the
1s are already there before you've seen anything. Since each real
observation adds 1 to `a` or to `b` (that's the update rule two
sections below), reading the counts back out means subtracting the 1
you began with.

If you want the reason underneath: the Beta density is proportional to
$p^{a-1}(1-p)^{b-1}$, and a Binomial likelihood for `k` successes in
`n` trials is proportional to $p^{k}(1-p)^{n-k}$. Line those two up
and the exponents *are* the counts — so `a − 1 = k` and `b − 1 = n − k`.
The `− 1` isn't a convention someone picked; it's the offset between a
parameter and the exponent it sits in.

So:

-   `Beta(1, 1)` — nothing observed at all. Perfectly flat, which is
    the flat prior we started the chapter with.
-   `Beta(9, 21)` — as if you'd watched 8 promotions among 28
    salespeople. Centred near 0.3, and about as confident as 28
    people's worth of evidence would justify. That's our VP.

Start from flat and add real data, and you can watch belief take
shape:

```{r beta-panel}
beta_examples <- tribble(
  ~label,               ~k,  ~n,
  "1 promoted of 7",      1,   7,
  "4 promoted of 10",     4,  10,
  "12 promoted of 30",   12,  30,
  "40 promoted of 100",  40, 100
) |>
  mutate(label = fct_inorder(label)) |>
  expand_grid(p = p_grid) |>
  mutate(density = dbeta(p, shape1 = 1 + k, shape2 = 1 + n - k))  # <1>

ggplot(beta_examples, aes(p, density)) +
  geom_area(fill = "#8fabd0", alpha = 0.30) +
  geom_line(colour = "#8fabd0", linewidth = 1) +
  geom_vline(aes(xintercept = k / n),                             # <2>
             linetype = "dashed", colour = "#d32f2f") +
  facet_wrap(~ label) +
  labs(title = "The same evidence, in increasing quantity",
       subtitle = "Dashed line = the raw observed rate",
       x = "Promotion rate", y = "Density")
```

1.  Each panel starts from a flat `Beta(1, 1)` and adds that panel's
    data — `k` successes and `n - k` failures. This *is* the update
    rule stated below; the panel is just showing it four times.
2.  The raw rate, for comparison with where the distribution actually
    puts its weight.

::: callout-important
## What the panel is telling you

Three of those four panels describe the same observed rate — 4 in 10,
12 in 30 and 40 in 100 are all 0.4 — and they are emphatically not
the same conclusion.

**Where the distribution sits is set by the ratio. How narrow it is
comes from the count.** Ten observations leave you barely able to
rule anything out between 0.2 and 0.7; a hundred pins you to roughly
0.3–0.5. A percentage on a slide can't distinguish those two
situations. This can.

One detail worth noticing: at small counts the distribution's *centre
of mass* sits slightly inside the dashed line, pulled towards 0.5. The
peak itself stays right on the raw rate — it is the lopsidedness of the
curve that shifts the average, with more of the weight lying on the 0.5
side. That's the flat prior still exerting a little influence, and it
fades as the data grows — the same effect you saw with the VP's prior a
few sections ago.
:::

### The shortcut itself

Here's the payoff. When the prior is a **Beta** and the data is
**Binomial**, the posterior is *always* another Beta, and you can
write it down without any grid, any sampling, or any calculus:

> Prior Beta(a, b) + data (k promotions out of n) → Posterior
> **Beta(a + k, b + n − k)**

Add your successes to the first parameter and your failures to the
second. That's the whole update. Distribution pairs that behave this
neatly are called **conjugate**, and while there are only a handful of
them, this one covers a great deal of People Analytics.

```{r beta-binomial}
a_prior <- 1   # flat prior = Beta(1, 1)
b_prior <- 1

a_post <- a_prior + k_promoted
b_post <- b_prior + (n_people - k_promoted)

cat("Posterior: Beta(", a_post, ",", b_post, ")\n")
cat("Posterior mean:", round(a_post / (a_post + b_post), 3), "\n")
cat("95% credible interval: [",
    round(qbeta(0.025, shape1 = a_post, shape2 = b_post), 3), ",",
    round(qbeta(0.975, shape1 = a_post, shape2 = b_post), 3), "]\n")
```

Compare these to the grid results above — they match.

------------------------------------------------------------------------

## Sequential updating

### Today's posterior is tomorrow's prior

Suppose **60 more salespeople** join, and 18 are eventually promoted.

```{r sequential}
n_new <- 60
k_new <- 18

a_post2 <- a_post + k_new
b_post2 <- b_post + (n_new - k_new)

tibble(p = p_grid,
       `after 40 people`  = dbeta(p_grid, shape1 = a_post,  shape2 = b_post),
       `after 100 people` = dbeta(p_grid, shape1 = a_post2, shape2 = b_post2)) |>
  pivot_longer(-p, names_to = "stage", values_to = "density") |>
  ggplot(aes(p, density, colour = stage)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#8fabd0", "#122a52")) +
  labs(title = "The posterior sharpens as data accumulates",
       x = "Promotion rate", y = "Posterior density", colour = NULL)
```

::: callout-important
## Learning never restarts — provided the process doesn't either

You never have to go back to the raw data. Last quarter's conclusion
becomes this quarter's starting point — and analysing data in batches
gives exactly the same answer as analysing it all at once.

That equivalence rests on an assumption worth saying out loud: that
there is one true rate θ, and it isn't moving. Here we'll take that as
given. In a real organisation it often isn't — a reorganisation, a new
promotion policy or a change of leadership can genuinely shift the
rate, and then evidence from three years ago is describing a process
that no longer exists. Piling it in as though it were current makes
you confidently wrong, and the more history you have the more
confidently.

There are ways to handle it — dropping history that predates a known
change, or letting older observations count for less as they age. Both
are ways of asking *how much should the past still count?* instead of
assuming the answer is "fully", and Chapter 14 comes back to them once
you've seen where a prior's strength comes from. For now, assume θ is
stable — and keep an eye on whether that's still true.
:::

::: callout-tip
## For the ML/DS crowd

This is the same principle behind **online / incremental learning** —
updating a model as new data streams in without retraining from
scratch. Bayesian sequential updating is one of the cleanest
mathematical justifications for why that works: the posterior already
captures everything the model learned so far, so it's a legitimate
starting point for the next batch.
:::

------------------------------------------------------------------------

## Making a decision

Managers ask concrete questions. *"What's the probability the true
promotion rate exceeds 25%?"* With a posterior, that's one line:

```{r decision}
prob_over_25 <- 1 - pbeta(0.25, shape1 = a_post2, shape2 = b_post2)
cat("P(rate > 0.25) =", round(prob_over_25, 3), "\n")
```

::: callout-note
Try phrasing that from a confidence interval — you can't, at least not
without a caveat. Direct probability answers are the Bayesian
superpower.
:::

### Your turn {.unnumbered}

Using the same posterior, compute the probability the rate is **below
20%**. What would you tell a VP of Sales worried the pipeline isn't
converting well enough?

```{r your-turn-decision}
# Your code here

```

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Any "what proportion / what rate?" question — the share of new hires
who pass probation, the fraction of applicants who accept an offer,
the conversion rate of a hiring channel — is a Beta-Binomial problem.
You'll be able to report a full posterior and make direct probability
statements ("there's an 80% chance the acceptance rate is above 60%"),
which reads far more naturally to a business stakeholder than a
p-value.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  The hardest step is usually the first one: turning a hunch into a
    **question with one unknown number in it**.
2.  **Bayes' theorem**: posterior ∝ likelihood × prior.
3.  **Grid approximation** computes a posterior by scoring candidate
    values — perfect for seeing the mechanism, useless beyond two or
    three unknowns.
4.  A **prior** can come from limits, past data, domain knowledge or
    published research. You have one either way; stating it is the
    honest move. Always visualise it.
5.  The **Binomial** is a distribution over counts, the **Beta** a
    distribution over rates. Location comes from the ratio,
    confidence from the count.
6.  The **Beta-Binomial** shortcut gives the exact posterior for a
    proportion instantly: add successes to `a`, failures to `b`.
7.  Updating is **sequential**, and the posterior answers decision
    questions directly — a credible interval means what it sounds
    like, unlike a confidence interval.
:::

## Next chapter {.unnumbered}

**Posteriors, priors and credible intervals** — and your first `brms`
model, the tool you'll use for the rest of the book.
