# Posteriors, Priors & Credible Intervals

## Welcome back {.unnumbered}

Last chapter you built posteriors by hand. That was essential for
*seeing* how Bayes works — but nobody does it by hand for real models.

::: callout-note
Today we meet **`brms`**, the tool you'll use for the rest of this
book and, very likely, your own analyses. The logic is identical to
last chapter. Only the machinery changes.
:::

### A short detour: why nobody taught me this {.unnumbered}

When I was taught statistics and econometrics, we were taught to
*think* like Bayesians — prior belief, evidence, updating — and then
made to *analyse* like frequentists. At the time that felt like an
odd inconsistency nobody quite explained. It wasn't perversity. The
methods in this chapter barely existed outside a handful of research
papers, and the computers we had couldn't have run them if they had.

I was unusual at university in owning a computer at all: a Mac LC, with what felt at
the time like a positively extravagant 2MB of RAM and a 40MB hard
drive. I loved that computer. It would have spent the better part of
a day on the model we're about to fit in a few seconds — assuming the
software had existed, which it didn't.

The ideas were much older than the hardware. The core algorithm goes
back to **1953**, at Los Alamos, where Metropolis and colleagues
worked out how to explore a complicated distribution by taking a long
random walk through it; Hastings generalised it in 1970. But for
decades it stayed in physics. The moment it crossed into statistics
is usually dated to **1990**, when Gelfand and Smith showed that this
family of methods could compute Bayesian posteriors that had been
analytically hopeless.

That's the same year the Mac LC went on sale and I raced to get a Mac 
with a **colour** screen! The algorithm and I turned up at university 
at roughly the same moment, and only one of us was in any state to do 
the work.

What followed was thirty years of turning *possible* into
*practical*. **BUGS** arrived in the early 1990s and put Bayesian
modelling within reach of researchers who weren't also programmers,
followed by **JAGS**. Then **Hamiltonian Monte Carlo** — a much
smarter way of wandering through the distribution, borrowed from
physics again — and **Stan**, first released in **2012**, which made
it fast and reliable enough for everyday use. **`brms`** landed in
**2015**: you write an ordinary R formula, `brms` writes the Stan
program for you, and Stan compiles it down to C++. Three decades of
work, wrapped in one function call.

::: callout-note
## Why this book exists now

I can teach this material to complete statistics beginners — which I
do — because that stack of advances removed the hard part. And I'd
argue it's now *easier* than the frequentist route, not harder.

The classical path asks you to memorise a menu: a t-test here, a
chi-square there, Cox regression for survival data, proportional-odds
for a Likert scale, each with its own function and its own
assumptions. From this chapter to the end of the book, every single
analysis is the same `brm()` call with different arguments. You learn
one thing properly instead of twenty things shallowly.
:::

The machinery still gets easier every year. It's worth knowing that
you're standing on it, and that not so long ago it wasn't there.

## What you'll be able to do by the end {.unnumbered}

1.  Fit a Bayesian model with **`brms`**, and read every line of its
    output
2.  Choose **weakly-informative priors** and **visualise** what they
    imply
3.  Run a **prior predictive check** — by hand and with `pp_check()`
4.  Report a posterior as a **point estimate + credible interval**
5.  Get **any** summary you want out of the posterior — a different
    interval width, a median, a direct probability
6.  **Compare two posteriors** by subtracting their draws, and report
    the probability that one quantity is larger than the other
7.  Check whether your conclusion is **sensitive to the prior**
8.  Know what a **Bayes Factor** is, and why this book generally
    doesn't reach for one

## One warning about the first run {.unnumbered}

::: callout-important
The first time you fit a `brms` model it **compiles** — it writes and
builds a small C++ program behind the scenes. That can take a minute.
After that it's fast. Nothing is broken; it's just warming up.
:::

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)
library(ggdist)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Book colour tokens: light blue = prior, red = likelihood, navy = posterior
prior_blue <- "#8fabd0"
red        <- "#d32f2f"
navy       <- "#122a52"
grey_alt   <- "#8a8f98"   # a comparison specification, not one of the three

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na(sales, customer_rate, performance, promoted)
```

------------------------------------------------------------------------

## The model: average sales

### It starts, again, in a meeting

The promotion question from Chapter 4 bought you some credibility, and
credibility in this job is mostly paid out in further questions. This
one arrives from Finance, three weeks before the planning cycle:

> *"What does a salesperson actually bring in? We need a number to
> build next year's quota on."*

There is an obvious temptation here, and it's the same one as last
time. You have `r nrow(salespeople)` salespeople with recorded sales.
You could take the
mean, put it on a slide, and be finished before lunch.

Two things stop you. The first is that the question isn't quite a
question yet. *A* salesperson — which one? The team has people who
joined last month and people who have run the same territory for a
decade. A short conversation settles what Finance actually wants:
**the sales figure a typical salesperson produces in a year**, as a
basis for a target, not a description of anyone in particular.

The second is Chapter 3. The mean of `r nrow(salespeople)` people is a
sample mean, and
sample means wobble. Handing Finance a single number invites them to
treat it as exact, and quotas built on numbers that are treated as
exact have a way of becoming somebody's performance review.

### Our question

**What is the average sales figure?** — not the average *in our data*,
which we already know and could compute in one line, but the
**underlying** average the process produces, reported with honest
uncertainty.

::: callout-note
This is the same shape of problem as Chapter 4: one unknown number,
and a body of data that constrains it without pinning it down. What
changes is that the unknown is now a quantity on a continuous scale
rather than a rate between 0 and 1 — and that we'll stop doing the
arithmetic by hand.
:::

### Writing a model down

Before you can learn anything about that average, you have to say what
you think generates the data. Ours says each salesperson's sales
figure sits around an unknown mean, scattered by an unknown amount:

$$
\text{sales}_i \sim \text{Normal}(\mu,\ \sigma)
$$

Two unknowns to learn: **μ**, the average we were asked for, and
**σ**, how far individual salespeople sit from it.

### Why *that* model?

Nothing forced the Normal on us. It earns its place for three reasons,
in descending order of importance:

1.  **We looked.** Chapter 3 plotted this exact variable and laid a
    Normal curve over it — the histogram is a single hump, roughly
    symmetric, with no second peak and no long tail dragging off to
    one side. That's the shape a Normal describes.
2.  **The generating story fits.** A salesperson's annual figure is the
    sum of many bookings, none of which dominates. Chapter 3's
    demonstration of why sums like that tend towards a bell applies
    directly here.
3.  **It has exactly the two unknowns we care about.** A Normal is
    described entirely by a centre and a spread, which is convenient
    when the centre is the thing Finance asked for.

::: callout-important
A likelihood is a **modelling choice**, not a fact about the world, and
it's the choice that most repays being explicit about. Ours has one
known flaw: a Normal puts some probability below zero, and nobody
books negative sales. With this data the flaw is harmless — the mean
sits far enough above zero that the implied fraction is tiny, and
we'll measure it rather than assert it when we run the prior
predictive check. Chapter 12 covers what to do when it isn't harmless.
:::

### The same thing in `brms`

```{r}
#| eval: false
brm(sales ~ 1, data = salespeople, family = gaussian())
```

-   `sales` — the outcome we're modelling
-   `~ 1` — "just an intercept", i.e. estimate one overall mean
-   `family = gaussian()` — the Normal likelihood

From next chapter we'll put predictors on the right of the `~`.

------------------------------------------------------------------------

## Priors — and always look at them

### What makes a good prior

::: callout-important
A **weakly-informative** prior rules out the absurd without dictating
the answer. `sales` in this dataset is recorded **in thousands of
dollars**, so the numbers run in the hundreds. A prior that allows
negative sales, or sales a thousand times larger than anything a
salesperson has ever booked, is not "neutral" — it's just careless.
:::

### Our two priors

Our model has two unknowns, so we need two priors: one for the average
**μ**, one for the spread **σ**. Take them in turn, in plain language
first — the code that hands them to `brms` comes once we know what
we're asking for.

### The average: what's a plausible number?

`normal(400, 200)` on the **Intercept** says: *before looking at the
data, my best guess for average sales is around 400 thousand dollars,
and I'd be unsurprised by anything within a couple of hundred either
side of that.*

The 400 is the **centre** — where I think the answer probably sits.
The 200 is the **standard deviation**, a measure of how spread out
the guess is: roughly two thirds of the prior sits between 200 and
600, and about 95% between 0 and 800. It's a deliberately loose
statement. I'm claiming to know the *scale* of the answer — hundreds,
not tens, not tens of thousands — and very little else.

### The spread: how different are salespeople from each other?

`exponential(0.005)` on **sigma** is the one that needs unpacking,
because it looks arbitrary and isn't.

σ is the spread of *individual* salespeople around the average — the
difference between a company where everyone sells much the same
amount and one where the stars sell five times what the strugglers
do.

The obvious question, given that we've just used a Normal for the
average and a Normal likelihood for the data, is why σ doesn't get a
Normal too. Three facts drive the choice:

1.  **A spread can't be negative.** This is the one that rules the
    Normal out. A Normal prior is symmetric and unbounded, so it would
    happily propose σ = −80, which isn't an unlikely value — it's a
    meaningless one. The exponential only exists above zero, so the
    impossible half of the number line is never on the table.
2.  **Smaller is more plausible than larger.** The exponential is
    highest at zero and falls away smoothly. That matches how spread
    usually behaves: modest differences are common, enormous ones are
    rare — but nothing is banned outright, so the data can still
    argue for a big σ if that's what it sees.
3.  **One number sets the scale.** The exponential has a single
    parameter, the **rate**, and its average value is simply
    `1 / rate`. So `rate = 0.005` means *"I expect the spread to be
    about 1 / 0.005 = 200 thousand dollars."*


::: callout-tip
## A simple rule for a `sigma` prior

**rate = 1 ÷ your honest guess at the spread.**

Guessing 200 here isn't clairvoyance. Sales in this business run from
somewhere near 100 to somewhere near 900, and a rough rule is that a
range covers about four standard deviations — so (900 − 100) / 4 ≈
200. That's the whole calculation. Guess 100 instead and you'd write
`exponential(0.01)`; the posterior would barely notice.
:::

::: callout-note
An exponential isn't the only defensible answer. A **half-Normal** —
a Normal folded at zero — is equally common and behaves similarly: it
also lives only on the positive side, and it puts slightly less weight
on very small spreads. The requirement is the boundary at zero, not
the particular curve. What you must not do is reach for an ordinary
Normal out of habit.
:::

### Writing them into the model

Now that we know what we're claiming, it's two lines of code:

```{r define-priors}
priors <- c(
  prior(normal(400, 200), class = Intercept),  # <1>
  prior(exponential(0.005), class = sigma)     # <2>
)
```

1.  `class = Intercept` attaches the first prior to the unknown
    **average**.
2.  `class = sigma` attaches the second to the unknown **spread**.

::: callout-note
This is the one place the book breaks its own rule about naming
arguments in full. Inside `prior()` you write the distribution in
**Stan's** notation, not R's, and Stan's arguments are positional:
`normal(centre, standard deviation)` and `exponential(rate)`. There is
no `sd =` to write. That's precisely why the paragraphs above exist —
when the code can't say which number is which, the prose has to.
:::

### Visualise both priors

```{r plot-prior-mean}
tibble(mu = seq(-200, 1200, length.out = 400)) |>
  mutate(density = dnorm(mu, mean = 400, sd = 200)) |>
  ggplot(aes(mu, density)) +
  geom_area(fill = prior_blue, colour = navy, linewidth = 0.6, alpha = 0.7) +
  labs(title = "Our prior for average sales",
       subtitle = "Normal(400, 200): centred on 400, most mass between 0 and ~800",
       x = "Mean sales, thousands of dollars (mu)", y = "Prior density")
```

```{r plot-prior-sigma}
#| code-fold: true
#| code-summary: "Show the plotting code (same pattern as above)"
tibble(sigma = seq(0, 1000, length.out = 400)) |>
  mutate(density = dexp(sigma, rate = 0.005)) |>
  ggplot(aes(sigma, density)) +
  geom_area(fill = prior_blue, colour = navy, linewidth = 0.6, alpha = 0.7) +
  geom_vline(xintercept = 200, linetype = "dashed", colour = grey_alt) +
  labs(title = "Our prior for the spread between salespeople",
       subtitle = "Exponential(0.005): positive only, average 200 (dashed), long right tail",
       x = "Spread of individual sales, sigma", y = "Prior density")
```

Note the shape of the second plot: it never crosses into negative
territory, its most likely values are small, and the tail runs a long
way right. Under this prior there's still roughly a 5% chance that σ
exceeds 600 — we've expressed a preference, not a rule.

::: callout-important
**Never trust a prior you haven't plotted.** Two lines of code each,
and it's the difference between a prior you chose and a prior you
copied.
:::

### The prior predictive check

A prior isn't really about the parameter in isolation — it's about the
**data it implies**.

::: callout-important
## Plotting is how you audit yourself — and how you explain yourself

`normal(400, 200)` is a claim about the world written in a notation
almost nobody reads fluently, including, on a bad day, you. Turning it
into a picture does two jobs.

The first is a **sense check**. A number that looked reasonable in code
often looks obviously wrong once drawn — a prior that puts a fifth of
its weight below zero, or one so tight it has effectively answered the
question before the data arrived. You catch that in the picture in
seconds; you rarely catch it by rereading the arguments.

The second is **communication**. "I assumed a Normal(400, 200) prior on
the intercept" ends most stakeholder conversations badly. A plot with
the caption *"before looking at the data, I treated anything from
roughly 0 to 800 as plausible"* invites the useful reply — *"that's far
too wide, we've never had a rep under 150"* — which is your VP handing
you a better prior. The chart isn't decoration; it's the only form in
which a non-statistician can disagree with your assumptions, which is
exactly what you want them to be able to do.
:::

::: callout-note
A prior predictive check simulates whole datasets **from the prior
alone** (before seeing any real data) and asks: *are these numbers
roughly the right size?*
:::

### Doing it by hand first

There is no magic in a prior predictive check, so it's worth building
one yourself once. The recipe is three lines of thinking:

1.  Draw a **μ** from its prior.
2.  Draw a **σ** from its prior.
3.  Generate a fake team of salespeople from `Normal(μ, σ)`.

Repeat 50 times and you have 50 imaginary companies, each one
consistent with your prior beliefs and nothing else.

```{r prior-predictive-manual}
prior_sims <- tibble(sim = 1:50) |>                                   # <1>
  mutate(
    mu    = rnorm(n(), mean = 400, sd = 200),                         # <2>
    sigma = rexp(n(), rate = 0.005)                                   # <3>
  ) |>
  mutate(
    sales_sim = map2(mu, sigma, function(m, s) {                      # <4>
      rnorm(300, mean = m, sd = s)
    })
  ) |>
  unnest(sales_sim)                                                   # <5>

ggplot(prior_sims, aes(x = sales_sim, group = sim)) +
  geom_density(colour = prior_blue, linewidth = 0.3) +
  geom_density(data = salespeople, aes(x = sales), inherit.aes = FALSE,
               colour = navy, linewidth = 1.2) +
  coord_cartesian(xlim = c(-500, 1500), ylim = c(0, 0.006)) +         # <6>
  labs(title = "Prior predictive check",
       subtitle = "Light lines = imaginary companies implied by the prior; dark = real data",
       x = "Sales (thousands of dollars)", y = "Density")
```

1.  Fifty imaginary companies. Each row will become one complete
    alternative world.
2.  Draw that world's **average sales** from the Intercept prior. Note
    we're sampling *from* the prior, not evaluating it — this is the
    same `normal(400, 200)`, used as a random number generator.
3.  Draw that world's **spread between salespeople** from the sigma
    prior, independently of the average.
4.  Given that pair of numbers, generate 300 salespeople. This is the
    likelihood — `Normal(μ, σ)` — doing exactly what the model says
    it does.
5.  Unpack the 50 lists of 300 into one long tibble, one row per
    imaginary salesperson.
6.  Clip the axes. See the callout below for why this is necessary and
    not a fudge.

::: callout-important
## Why we clipped the axes — and why the spikes are real

Look at `coord_cartesian()`: without it this plot is unreadable, and
the reason is worth understanding rather than hiding.

The `exponential(0.005)` prior on σ has its **highest density near
zero**. So a few of our 50 draws come back with a very small σ — an
imaginary company where every salesperson sells almost exactly the
same amount. A density curve for such a company is a needle: extremely
tall and extremely narrow. One needle sets the y-axis for the whole
plot and squashes everything else flat against the bottom.

That is not a bug, and it isn't a reason to change the prior. It's a
reminder of what an exponential prior actually says: *small spreads
are perfectly possible, large ones are increasingly unlikely.* We clip
the axes so we can see the bulk of the curves; the needles are still
there, running off the top.
:::

### The same check, the `brms` way

Once you've seen it by hand, `brms` will do it for you. Setting
`sample_prior = "only"` tells it to ignore the outcome data entirely
and sample from the prior:

```{r prior-predictive}
fit_prior_only <- brm(
  sales ~ 1, data = salespeople, family = gaussian(),
  prior = priors, sample_prior = "only",
  chains = 2, iter = 1000, seed = 5, refresh = 0
)

pp_check(fit_prior_only, ndraws = 50) +
  coord_cartesian(xlim = c(-500, 1500), ylim = c(0, 0.006)) +
  labs(title = "Prior predictive check, via pp_check()",
       subtitle = "Same picture: light = prior-implied datasets, dark = real data")
```

::: callout-tip
`pp_check()` almost always needs a `coord_cartesian()` when you use it
on a *prior*-only model. Prior draws are deliberately wider than
reality, so the default axes stretch to fit the most extreme
simulation and the interesting region collapses to a sliver. Add the
limits; don't assume the plot is broken.
:::

### Reading the verdict

A picture is good, but a number is easier to defend. Two questions
settle it:

```{r prior-predictive-numbers}
prior_sims |>
  summarise(
    prop_negative  = mean(sales_sim < 0),
    prop_over_2000 = mean(sales_sim > 2000),
    median_sales   = median(sales_sim)
  )
```

::: callout-important
The prior-implied sales figures overlap the real data's range —
sensible. A small fraction of impossible values (negative sales) is
the price of using a Normal likelihood on a positive quantity; a
*large* fraction would tell us to think again. A prior centred on
4,000 would sit far to the right, a clear warning sign. **This is the
check that lets you defend a prior when someone asks where it came
from.**
:::

------------------------------------------------------------------------

## Fit the model

```{r fit}
fit_sales <- brm(
  sales ~ 1, data = salespeople, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 5, refresh = 0
)
```

### Every argument, explained

This is the call you'll be writing for the rest of the book, so it's
worth taking apart properly. Only the first three arguments describe
your *model*; the rest describe how the computer should go looking for
the answer.

| Argument | What it does |
|------------------|------------------------------------------------------|
| `sales ~ 1` | The **model formula**. Outcome on the left, predictors on the right; `1` means "intercept only". |
| `data = salespeople` | Where the variables live. |
| `family = gaussian()` | The **likelihood** — how individual observations scatter around the mean. |
| `prior = priors` | The priors we defined and plotted above. Leave this out and `brms` picks defaults for you, which is exactly the sort of thing you don't want happening silently. |
| `chains = 4` | Run the sampler **four separate times**, from four different random starting points. |
| `iter = 2000` | Each chain takes 2,000 steps. |
| `seed = 5` | Fixes the random numbers so you and I get the identical answer. |
| `refresh = 0` | Suppresses the progress messages. Use `refresh = 100` when you want to watch a slow model. |

::: callout-note
## Chains, iterations and warmup

Chapter 4 built a posterior by evaluating it on a grid. `brms` can't
do that — real models have too many parameters — so instead it sends a
**sampler** wandering through the space of possible parameter values,
spending more time where the posterior is dense. Each visited point is
one **draw**. Pile the draws up and their histogram *is* the
posterior.

A **chain** is one such wandering walk. We run four because a walker
that starts in a silly place takes a while to find the interesting
region — and if all four chains, starting from different silly places,
end up describing the same distribution, that's strong evidence the
sampler found the real posterior rather than getting stuck.

The first half of each chain is **warmup** (`warmup = iter / 2` by
default, so 1,000 steps here). Those draws are thrown away: they're
the walker finding its feet, plus Stan tuning its step size. The
remaining 1,000 draws per chain × 4 chains gives us **4,000 posterior
draws** to work with — plenty for the intervals we'll report.

If a model feels slow, add `cores = 4` to run the chains in parallel.
It doesn't change the answer, just the wait.
:::

### Reading the summary

```{r fit-summary}
summary(fit_sales)
```

The output comes in blocks:

-   The **header** repeats the family, the formula and the sampling
    settings — useful when you come back to a saved model in six
    months.
-   **Regression Coefficients** (older `brms` versions call this block
    *Population-Level Effects*) holds the parameters from the formula.
    Right now that's just `Intercept`, the posterior for mean sales.
-   **Further Distributional Parameters** (older versions: *Family
    Specific Parameters*) holds everything else the family needs —
    here `sigma`, the spread of individual salespeople around that
    mean. Note that σ is a genuine unknown with its own posterior, not
    a nuisance figure computed on the side.

And the columns:

| Column | What it means |
|------------------|------------------------------------------------------|
| `Estimate` | The **mean of the posterior draws**. A summary of the distribution, not "the answer". |
| `Est.Error` | The **standard deviation of the posterior draws** — how wide our uncertainty is. It plays the role a standard error plays elsewhere, but it's a direct statement about the parameter. |
| `l-95% CI`, `u-95% CI` | The 95% **credible interval**: 2.5% of draws fall below the first, 2.5% above the second. |
| `Rhat` | Do the four chains agree? **1.00 is what you want.** Anything from about 1.01 upwards means they're describing different distributions — don't report the model. |
| `Bulk_ESS` | **Effective sample size** for the centre of the distribution. Draws are correlated, so 4,000 draws aren't worth 4,000 independent ones; this is the honest count. Want it in the hundreds at minimum, ideally thousands. |
| `Tail_ESS` | The same idea for the tails — the part that determines your interval endpoints. |

::: callout-tip
Rhat and ESS are **diagnostics about the computation**, not about your
model's quality. A model can sample perfectly and still be the wrong
model. Chapter 11 (optional) opens the sampler up; Chapter 7 covers
judging whether the model is any good.
:::

------------------------------------------------------------------------

## Read the posterior

### The whole distribution

```{r posterior-plot}
draws <- fit_sales |> spread_draws(b_Intercept)   # <1>

draws |>
  ggplot(aes(x = b_Intercept)) +
  stat_halfeye(fill = navy, slab_colour = navy,   # <2>
               slab_linewidth = 0.6, alpha = 0.85) +
  labs(title = "Posterior for mean sales",
       subtitle = "Every plausible value for the average, with its weight",
       x = "Mean sales (thousands of dollars)", y = NULL)
```

1.  `spread_draws()` pulls the raw posterior draws out of the fitted
    model and into a tidy tibble — one row per draw. `b_Intercept` is
    `brms`'s internal name for the intercept (`b_` for the
    coefficients block). **This object is the one we'll keep reusing
    for the rest of the chapter**; everything below is an ordinary
    dplyr operation on it.
2.  `stat_halfeye()` draws the density with the median and the 66% and
    95% intervals underneath it, so the point estimate and the
    uncertainty appear in one image.

### Point estimate + credible interval

```{r credible-interval}
draws |> mean_qi(b_Intercept, .width = 0.95)
```

::: callout-important
## Say it in plain English

*"Our best estimate of average sales is about [the mean], and there is
a **95% probability** it lies between [lower] and [upper]."*
:::

------------------------------------------------------------------------

## The posterior is the real prize

### You have 4,000 numbers, not one answer

Everything we've just reported — the mean, the 95% interval — came out
of the same object: a vector of 4,000 draws. Nothing was decided at
fitting time. **If you can phrase a question about that vector, you
can answer it**, and you can do it after the fact without refitting
anything.

Want a different interval width? Take it.

```{r other-intervals}
draws |> median_qi(b_Intercept, .width = c(0.5, 0.89, 0.95))
```

::: callout-note
The 89% interval isn't a typo. Richard McElreath uses it throughout
*Statistical Rethinking*, partly to make the point that **95% is a
convention, not a law of nature** — it's inherited from a
frequentist testing tradition this framework doesn't need. Pick a
width, say why, and stick to it.
:::

Want a different point summary? Take that too — mean, median, or the
most probable value:

```{r other-summaries}
draws |>
  summarise(
    mean    = mean(b_Intercept),
    median  = median(b_Intercept),
    lower_q = quantile(b_Intercept, probs = 0.25),
    upper_q = quantile(b_Intercept, probs = 0.75)
  )
```

### Questions you can only ask this way

Because the posterior is a probability distribution over the
parameter, you can ask directly for the probability of a claim — and
get a number a stakeholder can use:

```{r posterior-probability}
draws |>
  summarise(
    p_above_500 = mean(b_Intercept > 500),
    p_above_520 = mean(b_Intercept > 520),
    p_above_540 = mean(b_Intercept > 540)
  )
```

::: callout-important
*"There is a `r round(100 * mean(draws$b_Intercept > 520))`%
probability that average sales exceed 520."* That is a
sentence you can say out loud in a business meeting and have it mean
what the listener thinks it means. A p-value cannot be translated that
way, however often it is.
:::

::: callout-tip
The risk that comes with this flexibility is the temptation to go looking —
trying widths and thresholds until one looks decisive. Decide what
you'll report **before** you look, and report it whichever way it
falls.
:::

### Comparing two posteriors

We just compared the posterior to a **fixed number** — 520. The same
move works when the thing you are comparing against is *also*
uncertain, and that covers most of the questions people actually ask.
Is the team we trained selling more than the team we didn't? Is
attrition really worse in one region than another? Is this year's
score genuinely above last year's?

We haven't met categorical predictors yet — that's Chapter 7 — so
we'll use the crudest version that works: fit the same one-line model
twice, once on each group. Here, salespeople who have been promoted
against those who haven't.

```{r two-group-fits}
promoted_only <- salespeople |> filter(promoted == 1)
not_promoted  <- salespeople |> filter(promoted == 0)

fit_promoted <- brm(
  sales ~ 1, data = promoted_only, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 5, refresh = 0
)

fit_not_promoted <- update(fit_promoted,                    # <1>
                           newdata = not_promoted,
                           seed = 5, refresh = 0)
```

1.  `update()` refits the *same* model on different data. It reuses
    the compiled program rather than building a new one, so you pay
    the C++ compile cost once instead of twice.

Two models, two posteriors, 4,000 draws each. Now the whole technique,
which is one line of arithmetic:

```{r subtract-draws}
draws_promoted     <- spread_draws(fit_promoted,     b_Intercept)
draws_not_promoted <- spread_draws(fit_not_promoted, b_Intercept)

diff_draws <- tibble(
  difference = draws_promoted$b_Intercept - draws_not_promoted$b_Intercept
)
```

Subtract one bag of numbers from the other, element by element, and
what comes out is not a difference — it's **a posterior for the
difference**. It has a shape, a middle and a spread, exactly like the
posteriors it was built from, and every question you could ask of
those you can now ask of this.

```{r plot-difference}
ggplot(diff_draws, aes(x = difference)) +
  stat_halfeye(fill = navy, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +  # <1>
  labs(title = "Posterior for the difference in average sales",
       subtitle = "Promoted minus not promoted. Red dashed = no difference.",
       x = "Difference in average sales (thousands of dollars)",
       y = NULL)
```

1.  Zero is the only reference line that matters here. How much of the
    distribution sits to the right of it is the answer to "is one
    group higher?", and how far the bulk sits *from* it is the answer
    to "by how much?"

```{r summarise-difference}
diff_draws |>
  summarise(
    p_promoted_higher = mean(difference > 0),                # <1>
    median            = median(difference),
    lower             = quantile(difference, 0.025),
    upper             = quantile(difference, 0.975)
  )
```

1.  The proportion of draws above zero. Because the draws *are* the
    posterior, counting them is the same as integrating it — which is
    why a question that looks like it needs calculus needs `mean()`
    instead.

::: callout-important
## Say it in plain English

*"Promoted salespeople sell about [the median] more on average, with a
95% probability the true gap lies between [lower] and [upper]. Given
this data, there is a
`r round(100 * mean(diff_draws$difference > 0))`% probability their
underlying average is genuinely the higher of the two."*

Read the interval **before** the probability. If the interval
comfortably straddles zero, the probability is only telling you which
side of the fence most of the posterior sits on — not that a
difference has been established.
:::

::: callout-note
## Names worth knowing

The technique doesn't have one agreed name, which is part of why it
gets rediscovered so often. The vocabulary you'll meet:

-   The object you built is the **posterior distribution of the
    difference**, or a **posterior contrast** — "contrast" being the
    word inherited from ANOVA for any comparison you construct between
    groups.
-   The single number `mean(difference > 0)` is the **posterior
    probability of superiority**. Commercial A/B testing tools usually
    label it *"probability to beat"* or *"chance to win"*.
-   The reason it works at all is that **posterior draws survive being
    transformed**. Apply any function to the draws and you have draws
    from that function's posterior — no new derivation required. This
    is standard in the simulation-based Bayesian literature; Gelman et
    al.'s *Bayesian Data Analysis* sets it out as inference for
    derived quantities.

Two sources worth crediting. **David Robinson's** *Introduction to
Empirical Bayes* (2017) has the clearest short treatment I know: he
asks whether one baseball player is truly a better batter than
another, lists four ways to answer it, and puts *simulation of
posterior draws* first precisely because it needs no mathematics. And
**John Kruschke's** "Bayesian Estimation Supersedes the t Test"
(*Journal of Experimental Psychology: General*, 2013) is the paper
that made putting a posterior on the difference between two groups a
standard alternative to the t-test.
:::

::: callout-warning
## The trap: subtract the draws, not the summaries

Subtract two *summary numbers* — 570 minus 490 — and you get a
difference with no uncertainty attached, which is worse than useless
because it looks precise. The subtraction has to happen **draw by
draw**, so that the uncertainty in both quantities is carried through
into the answer.

There is a second reason, which matters more later. Here the two
posteriors came from two separate models, so their draws are
independent and the pairing is arbitrary. When both quantities come
from the **same** model — as they will from Chapter 7 onwards — the
draws are correlated, and matching row *i* to row *i* is what carries
that correlation into the difference. Any other pairing throws it
away and gives you an interval of the wrong width.
:::

You'll see this again twice. Chapter 7 does the same subtraction
inside a single model, where `compare_levels()` handles the
bookkeeping for you, and Chapter 20 builds a whole A/B test on it.
Both are the technique you have just learnt; only the wrapping
changes.

### "Isn't this just the bootstrap?"

If you've used bootstrapping, the mechanics will feel familiar: both
hand you a bag of numbers you can push through any function you like.
The **operational** flexibility really is the same. What differs is
what the numbers *are*.

| | Bootstrap sample | Posterior draws |
|-------------------|---------------------------|---------------------------|
| Made by | Resampling **your data**, refitting each time | Sampling a **probability distribution over the parameter** |
| Represents | How your *estimate* would wobble across repeated studies | What's plausible for the *parameter*, given this data and your priors |
| Uses prior information | No | Yes, explicitly |
| A 95% interval means | 95% of such intervals would cover the truth in the long run | 95% probability the parameter is in *this* interval |

::: callout-important
The bootstrap is a clever way of approximating a **sampling
distribution** — a statement about the procedure. The posterior is a
statement about the **parameter**. They can produce similar-looking
intervals in large samples with flat priors, and that coincidence is
exactly why the distinction gets blurred. It matters most when data
are thin, when you have real prior knowledge, or the moment someone
asks "so what's the probability we're above target?"
:::

------------------------------------------------------------------------

## Does the prior change the answer?

### The fair worry

Sooner or later someone will say: *you told the model the answer
before it started — of course it agrees with you.* It's a fair
challenge, and answering it is called a **sensitivity analysis**.

Notice what actually happened here, though. We centred the prior on
**400**. The posterior came back **well over a hundred higher** (see
the `Intercept` row in the summary above) — the data pulled
the estimate well away from where we pointed it, and the credible
interval is a fraction of the prior's width. That's the likelihood
doing the work, exactly as it should with `r nrow(salespeople)`
salespeople.

::: callout-note
That reassurance is suggestive, not proof. A posterior that moves
away from the prior doesn't guarantee the prior was harmless, and one
that *doesn't* move isn't evidence that it did damage — the prior and
the data may simply agree. The only way to know is to refit with a
deliberately different prior and compare.
:::

```{r sensitivity}
priors_vague <- c(
  prior(normal(0, 5000), class = Intercept),
  prior(exponential(0.001), class = sigma)
)

fit_vague <- brm(
  sales ~ 1, data = salespeople, family = gaussian(),
  prior = priors_vague,
  chains = 4, iter = 2000, seed = 5, refresh = 0
)

bind_rows(
  spread_draws(fit_sales, b_Intercept) |> mutate(prior = "weakly-informative"),
  spread_draws(fit_vague, b_Intercept) |> mutate(prior = "very vague")
) |>
  ggplot(aes(x = b_Intercept, fill = prior)) +
  stat_halfeye(alpha = 0.6) +
  scale_fill_manual(values = c("very vague" = grey_alt,          # <1>
                               "weakly-informative" = navy)) +
  labs(title = "The posterior barely moves when we change the prior",
       subtitle = paste0("With ", nrow(salespeople),
                         " salespeople, the data dominates"),
       x = "Mean sales", y = NULL)
```

1.  Both distributions here are **posteriors**, so the one we're
    actually reporting keeps the book's navy. The alternative
    specification is drawn in neutral grey rather than red — red is
    reserved throughout for a *likelihood*, and this isn't one.

::: callout-important
With plenty of data, **reasonable priors barely matter** — the
likelihood dominates. But you should always check, and **report that
you did**. With *little* data — as in Chapter 4's 40 salespeople — the
prior matters much more.
:::

### Your turn {.unnumbered}

Fit the same intercept-only model for **`customer_rate`** instead of
`sales`. Choose a sensible prior (customer ratings are typically on a
1–5 scale), run a prior predictive check, and report the posterior
mean with its credible interval.

Then, without refitting anything, answer two more questions from the
same draws: what is the **89%** interval, and what is the probability
that the average customer rating is **above 4**?

```{r your-turn}
# Your code here

```

------------------------------------------------------------------------

## A note on Bayes Factors

Everything above reports an **estimate** — a posterior distribution
and a credible interval. There's a second Bayesian approach worth
knowing about, because you'll meet it elsewhere: **Bayesian hypothesis
testing using Bayes Factors**, covered in depth in Keith McNulty's
book (his Chapter 12).

::: callout-note
## What a Bayes Factor actually does

A Bayes Factor compares the evidence for two competing models — say,
"the mean sales figure is exactly 500" (a sharp null hypothesis)
against "the mean is some other value" (a diffuse alternative) — and
reports a ratio: how many times more likely the data are under one
model than the other. It's the Bayesian cousin of a p-value, built to
answer *"is there truly an effect, or not?"* rather than *"how big is
the effect?"*
:::

### Why this book mostly doesn't use them

This book leans on estimation — posteriors, credible intervals, and
(from Chapter 7) LOO cross-validation for comparing models — rather
than Bayes Factors, for three practical reasons:

1.  **The question People Analytics usually needs answered is "how big
    and how sure," not "is it exactly zero."** A credible interval
    answers that directly; a Bayes Factor doesn't, on its own.
2.  **Bayes Factors are sensitive to how you specify the alternative
    hypothesis** — a different plausible choice of alternative model
    can shift the answer noticeably, which is an extra judgment call
    estimation avoids.
3.  **They are a separate skill, not an extension of the one you are
    building.** Once you can read a posterior and a credible interval,
    that same skill covers every chapter in this book. Bayes Factors
    are a different framework layered on top of it.

None of this means Bayes Factors are wrong — they're the right tool
when the actual scientific question is "does this effect exist at
all," which comes up more in some research settings (and is exactly
where McNulty's treatment is worth reading). We flag the difference
again in Chapters 7 and 20, at the two points where a classically
trained reader would most reflexively reach for a significance test.

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Every model in this book needs priors, and a sceptical stakeholder may
ask you to justify them. The workflow you practised today —

> *state the prior → visualise it → run a prior predictive check → fit
> → show a sensitivity analysis*

— is exactly what a rigorous methods write-up contains. Do this and
your analysis is defensible.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  `brms` fits Bayesian models with a formula like `sales ~ 1`;
    `chains`, `iter` and `seed` control the **sampler**, not the
    model.
2.  Read the summary in blocks: coefficients, then distributional
    parameters. **Rhat ≈ 1.00** and healthy **ESS** say the
    computation worked — not that the model is right.
3.  **Weakly-informative priors** rule out the absurd; always
    **visualise** them.
4.  A **prior predictive check** shows the data your prior implies.
    Prior-only plots usually need their axes clipped — that's the
    prior being honestly wide, not a broken plot.
5.  Report a posterior as a **mean + 95% credible interval** with
    `mean_qi()`.
6.  The posterior is **4,000 draws you own outright**: any width, any
    summary, any probability statement, no refitting. Similar
    flexibility to the bootstrap, but a different object — a
    distribution over the *parameter*, not the *estimator*.
7.  Two posteriors can be compared by **subtracting the draws** —
    element by element, never summary from summary. What comes out is
    a posterior for the **difference**, from which you get both a
    credible interval and a direct probability that one is higher.
8.  A **sensitivity analysis** shows whether the prior drove the
    result.
9.  **Bayes Factors** are a legitimate alternative Bayesian tool for
    "does this effect exist" questions — this book prefers estimation,
    for reasons that are worth stating explicitly rather than assuming.
:::

## Next chapter {.unnumbered}

**Bayesian regression** — from estimating one number to explaining a
relationship: does a salesperson's customer rating predict how much
they sell?
