# Bayesian Regression: Explaining & Predicting

## Welcome back {.unnumbered}

So far we've estimated single numbers — a rate, an average. Today we
take the biggest step in the book so far: **explaining one variable
with another**.

### Why you'd want to {.unnumbered}

Almost every question a business asks about its people is really a
question about a relationship between two things:

-   Do people with higher engagement scores stay longer?
-   Does the hiring source a candidate came from affect how quickly
    they reach full productivity?
-   Do larger teams perform worse than small ones, and from what size
    does it start to matter?
-   Is the pay gap between two groups still there once you account for
    role and tenure?

None of those can be answered by a single average. Each one asks how
an outcome *moves* as something else changes — and that is exactly
what regression measures. It's the tool underneath most quantitative
People Analytics work, and once you have it, most of the rest of this
book is variations on it.

### Where this one comes from {.unnumbered}

Your VP of Sales has taken the promotion analysis well, and she's back
with something she wants to spend money on.

> *"The best-rated reps are our best sellers. I want to put the whole
> team through customer-service training."*

There's a proposal attached, and it isn't cheap: **$1,500 per person**,
for every salesperson on the team. So the question isn't really
whether better-rated salespeople sell more — she has watched them for
a decade and she's almost certainly right about the direction. The
question is **how much**, because a number is the only thing that can
be set against a cost.

> **For each extra point of customer rating, how much more does a
> salesperson sell — and how likely is it that the gain covers the
> $1,500?**

Notice the shape of that second half. It isn't a question about the
data, it's a question about a **threshold**: there is some slope below
which the training loses money, and what your VP actually wants to
know is the probability we're above it. Hold onto that — we'll come
back and answer it once we have a posterior, and it turns out to cost
one extra line of code.

::: callout-note
## A yes/no question is the wrong question

The instinctive framing — *"is there a relationship between rating and
sales?"* — is answerable with a word, and the word is almost always
"yes" if you collect enough data. It's also useless to someone holding
a budget.

This is the estimation-over-testing argument from Chapter 5, arriving
in the place you'll meet it most often. Regression's real output isn't
a verdict on whether an effect exists. It's a **magnitude with honest
uncertainty attached** — and it's the uncertainty that tells your VP
whether the business case survives its own worst-case scenario.
:::

## What you'll be able to do by the end {.unnumbered}

1.  Fit a **Bayesian linear regression** with `brms`
2.  Interpret a **slope** as a posterior distribution — a magnitude,
    not a verdict
3.  **Centre a predictor** so the intercept means something
4.  Turn a posterior into a **break-even probability** a budget-holder
    can act on
5.  Distinguish **parameter uncertainty** from **prediction
    uncertainty**
6.  Check the model with a **posterior predictive check**

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)
library(ggdist)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |> drop_na(sales, customer_rate, performance, promoted)

# Centred version of the predictor — see "Why we centre" below.
mean_rate <- mean(salespeople$customer_rate)

salespeople <- salespeople |>
  mutate(customer_rate_c = customer_rate - mean_rate)
```

------------------------------------------------------------------------

## Look before you model

The first step of any analysis is to look at the data. Not summarise
it — *look* at it. A scatter plot takes ten seconds and will tell you
things no table can: whether the relationship is roughly a straight
line or a curve, whether it's tight or loose, whether a handful of
extreme points are about to drag your answer around, and whether the
variable is even measured the way you assumed.

Fitting a model before plotting is how analysts end up confidently
describing a relationship that isn't the shape they think it is.

Each point below is one salesperson.

```{r scatter}
ggplot(salespeople, aes(x = customer_rate, y = sales)) +
  geom_point(alpha = 0.4, colour = "#122a52") +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "#d32f2f", se = FALSE) +          # <1>
  labs(title = "Sales vs customer satisfaction rating",
       subtitle = "A gentle upward trend — happier customers, more sales",
       x = "Customer rating", y = "Sales (thousands of dollars)")
```

1.  That red line is a **classical** least-squares fit — `lm()`,
    dropped in by `geom_smooth()`. Note `se = FALSE`: we've explicitly
    switched off even the uncertainty band it would have offered.

::: callout-note
A word on why `customer_rate` goes in as a plain number here, when the
book elsewhere is careful about ratings. It is an *averaged*
satisfaction score — several customers' ratings pooled into one figure,
with decimals — so a step from 3.2 to 3.4 means much the same as a step
from 4.2 to 4.4. A single ordinal tick, such as the 1-to-4 performance
rating, is a different animal: nothing guarantees the gaps between its
levels are equal. That is why the "Your turn" exercise flags performance
and not customer rating, and why Chapter 18 gives ordered ratings a
model of their own.
:::

### Why not just stop there?

The red line is a perfectly good answer to *"which way does this
go?"*, and it took no effort. Reaching for it early is a good habit,
not a guilty one — it's fast, it aids comprehension, and later it
gives you something to sanity-check your Bayesian result against.

But look at what it is: **one line**. A single best guess, drawn with
total confidence, through a cloud that clearly permits a range of
lines. Tilt it slightly and it would still pass respectably through
those points. That range of defensible tilts is precisely what your
VP needs, because the business case depends on whether the effect is
70 per point or 50.

::: callout-important
## Two honest observations

1.  There **is** an upward trend — the red line slopes up.
2.  The cloud is **noisy** — customer rating is clearly not the only
    thing driving sales.

Both are true at once. Regression's job is to put a **number** on the
trend and an honest measure of **uncertainty** around it. The rest of
this chapter is about replacing that one confident line with every
line the data will support, weighted by how well each one does.
:::

------------------------------------------------------------------------

## The model

### Fitting a line through the cloud

$$
\text{sales}_i \sim \text{Normal}(\mu_i,\ \sigma), \qquad
\mu_i = \alpha + \beta \times (\text{customer\_rate}_i - \overline{\text{customer\_rate}})
$$

-   **α** (intercept) — average sales for a salesperson with an
    **average** customer rating
-   **β** (slope) — how much sales change **per extra point** of
    customer rating
-   **σ** — how much individual salespeople scatter around the line

::: callout-note
**β** is the one we came for. It's the number in your VP's business
case.
:::

### Why we centre

That subtraction in the formula is doing real work, and it's a habit
worth forming now.

Written the obvious way — $\mu_i = \alpha + \beta \times
\text{customer\_rate}_i$ — the intercept α is "expected sales when
customer rating is **zero**". But ratings in this dataset run from 1
to 5. A rating of zero doesn't exist and never could. The model will
still report a number for it, obtained by running the line off the
left-hand edge of the data, and that number is meaningless.

Subtract the mean rating from every value first and the problem
disappears. Now the predictor is zero at the *average* rating, so α
becomes "what a typically-rated salesperson sells" — a quantity your
VP can recognise, argue with, and sanity-check. The slope β is
completely unaffected; centring shifts where the line is anchored, not
how steep it is.

::: callout-important
## Something `brms` does quietly, that centring makes visible

There's a second reason to do this explicitly, and it will save you a
confusing afternoon at some point.

`brms` centres your predictors internally anyway, for reasons of
sampling efficiency. One consequence catches people out: **a prior you
set on `class = Intercept` applies to the centred intercept** — the
expected outcome at the *mean* of the predictors — while the
`Intercept` reported in `summary()` is converted back to the raw
scale, i.e. the outcome at predictor zero.

On an uncentred model those are two different numbers, and you can
spend a while trying to work out why the prior you set at 400 produced
a posterior sitting near 270. Centring the predictor yourself collapses
the two into one, and what you set is what you read.

(If you ever need the raw-scale intercept to be the one with the
prior on it, the formula syntax is `y ~ 0 + Intercept + x`.)
:::

### In `brms`

```{r}
#| eval: false
brm(sales ~ 1, ...)                 # last chapter: just a mean
brm(sales ~ customer_rate_c, ...)   # this chapter: a mean that depends on rating
```

Everything you learned about priors, posteriors and intervals carries
straight over.

::: callout-tip
## Worth naming: this is the recipe for the rest of the book

A formula, a `family`, and some priors — that's the whole thing `brm()`
needs. Nothing about *how* you fit a Bayesian model is about to change,
no matter how different the outcome gets. A yes/no outcome, a count, a
time-to-event, a Likert scale — every one of those, later in this book,
is still just `brm(formula, family = ..., prior = ...)`. Only the
formula, the `family`, and the priors change; the workflow around them
never does. Worth noticing now, because it stops being obvious once
you've seen ten different-looking outcomes in a row.
:::

### Three priors now, not two

Chapter 5's model had two unknowns. This one has three, and the new
arrival — the slope — gets `class = b`, the class every predictor
coefficient belongs to.

```{r priors}
priors <- c(
  prior(normal(400, 200),   class = Intercept),  # <1>
  prior(normal(0, 100),     class = b),          # <2>
  prior(exponential(0.005), class = sigma)       # <3>
)
```

1.  Unchanged from Chapter 5, and now unambiguous: thanks to
    centring, this really is our belief about sales for an
    average-rated salesperson.
2.  The slope. Centred on **zero** — we're not telling the model which
    way the relationship runs — with a standard deviation of 100,
    which says a swing of more than a couple of hundred thousand
    dollars per rating point would be surprising.
3.  Unchanged: the spread of individual salespeople around the line.

```{r plot-prior-slope}
tibble(beta = seq(-300, 300, length.out = 400)) |>
  mutate(density = dnorm(beta, mean = 0, sd = 100)) |>
  ggplot(aes(beta, density)) +
  geom_area(fill = "#8fabd0", alpha = 0.30) +
  geom_line(colour = "#8fabd0", linewidth = 1) +
  geom_vline(xintercept = 0, colour = "#d32f2f", linetype = "dashed") +
  labs(title = "Prior for the slope: Normal(0, 100)",
       subtitle = "Symmetric around zero — the data must tell us the direction",
       x = "Slope (change in sales per point of customer rating)", y = "Prior density")
```

::: callout-note
## Zero-centred is not the same as no opinion

Putting the slope prior on zero looks like a strong claim that there's
no effect. It isn't. It's a claim about **direction** — that we're not
going to prejudge the sign — combined with a claim about **plausible
magnitude**, which is what the 100 controls.

A prior centred on zero is easily overruled by data that disagrees, as
you're about to see. What it does buy you is protection against the
model taking a small, noisy pattern and reporting an implausibly
enormous effect.
:::

### What do these priors imply?

Chapter 5 checked a prior by simulating datasets from it. For a
regression there's a more direct version: simulate **lines**.

This is the **prior predictive check** from Chapter 5, in its natural home: a
prior on a slope means very little until you see the relationships it
implies.


```{r prior-predictive}
set.seed(606)

prior_lines <- tibble(
  alpha = rnorm(100, mean = 400, sd = 200),   # <1>
  beta  = rnorm(100, mean = 0,   sd = 100)
)

ggplot(salespeople, aes(customer_rate, sales)) +
  geom_point(alpha = 0.2, colour = "grey60") +
  geom_abline(data = prior_lines,
              aes(intercept = alpha - beta * mean_rate,   # <2>
                  slope     = beta),
              colour = "#8fabd0", alpha = 0.25, linewidth = 0.4) +
  coord_cartesian(xlim = c(1, 5), ylim = c(-500, 1500)) +
  labs(title = "100 relationships our priors consider plausible",
       subtitle = "Before seeing any data — wide, both directions, but not absurd",
       x = "Customer rating", y = "Sales (thousands of dollars)")
```

1.  Drawing straight from the two priors, exactly as written above.
    `brm(..., sample_prior = "only")` would do the same thing through
    the model; by hand is faster and shows there's no trick to it.
2.  Converting back to the raw rating scale for plotting. Our line is
    $\alpha + \beta(x - \bar{x})$, which rearranges to an intercept of
    $\alpha - \beta\bar{x}$ against plain $x$.

::: callout-important
That fan of lines is what we're claiming to believe before the data
speaks. It's wide, it slopes both ways, and it comfortably contains
anything the data might reasonably show — but it doesn't contain
lines implying a rating point is worth half a billion dollars. That's
what weakly-informative looks like for a slope.
:::

------------------------------------------------------------------------

## Fit and interpret

```{r fit}
fit_reg <- brm(
  sales ~ customer_rate_c, data = salespeople, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 6, refresh = 0
)

summary(fit_reg)
```

Same call as Chapter 5, same argument order, one extra term in the
formula. That's the whole difference.

### The slope is a distribution

```{r slope-plot}
fit_reg |>
  spread_draws(b_customer_rate_c) |>
  ggplot(aes(x = b_customer_rate_c)) +
  stat_halfeye(fill = "#122a52", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#d32f2f") +
  labs(title = "Posterior for the customer-rating slope",
       subtitle = "Where does the mass sit relative to zero?",
       x = "Change in sales per point of customer rating", y = NULL)
```

```{r slope-interval}
fit_reg |>
  spread_draws(b_customer_rate_c) |>
  mean_qi(b_customer_rate_c, .width = 0.95)
```

That interval is the sentence your VP is waiting for: *each additional
point of customer rating is worth roughly this much in sales, and
we're 95% confident the true figure lies between these two numbers.*

### Does it pay for itself?

The interval is the honest answer to the analytical question. It is
not yet an answer to the *business* question, because the business
question has a price tag in it.

So do the arithmetic the other way round. Rather than asking what the
slope is and hoping it's big enough, ask **how big it would have to
be** for the training to wash its face — and then ask the posterior
how likely that is.

Three numbers turn the cost into a threshold on the slope:

```{r break-even}
cost_per_head <- 1.5     # <1>
rating_uplift <- 0.1     # <2>
margin        <- 0.25    # <3>

break_even <- cost_per_head / (rating_uplift * margin)   # <4>

fit_reg |>
  spread_draws(b_customer_rate_c) |>
  summarise(
    break_even_slope  = break_even,
    p_pays_for_itself = mean(b_customer_rate_c > break_even)   # <5>
  )
```

1.  $1,500 per salesperson, in the thousands-of-dollars units the
    model works in.
2.  How much the training is expected to lift a rep's customer rating.
    This is **not** in the data — it's an assumption, and we'll come
    back to it.
3.  Extra sales aren't extra profit. Only the contribution margin
    covers the cost.
4.  The slope at which benefit exactly equals cost. Below this, the
    programme destroys value.
5.  The proportion of posterior draws above the threshold — which,
    because the posterior is a probability distribution, *is* the
    probability the training pays for itself.

That last line is doing the same trick as `p_above_520` in Chapter 5,
pointed at a decision instead of a description. And the answer is a
sentence with no statistics in it at all: *"given what we know about
the relationship, there's roughly an 80% chance this training returns
more than it costs."*

::: callout-important
## Which number the decision actually rests on

Look again at where the uncertainty in that answer comes from. The
slope is estimated from `r nrow(salespeople)` salespeople and is
pinned down reasonably well. The **0.1-point uplift is a guess** — nobody has measured it,
and halving it doubles the break-even slope.

That's a useful thing to discover, because it tells you where to spend
your next hour: not on a better model, but on a better estimate of
what the training actually does. Chapter 23 is about getting numbers
like that out of the people who hold them.

One more caveat worth saying out loud, because a break-even
calculation quietly assumes it: the slope describes an
**association** between rating and sales among current salespeople.
Treating it as what *would* happen if you intervened to raise ratings
is a causal claim the data alone doesn't support. Worth stating
plainly in the recommendation rather than leaving the reader to
assume it.

Stating it plainly is the minimum. **Chapter 21 is about what it would
take to say more** — and, on this exact model, arrives at a different
number for a different question.
:::

One more decision worth noticing before we move on, because nothing in
the output will point at it. We fitted a **straight** line. That was a
choice, not a default, and for this data it is a reasonable one. It is
not always: the relationship between tenure and attrition rises, falls
and rises again, and a straight line through it produces a small average
slope that describes nobody. Chapter 13 is about how to tell, and what to
do instead.

::: callout-tip
## Where this goes next: Bayesian decision analysis

What we just did — take a posterior, apply a cost, read off a
probability — is the doorstep of a whole field. **Bayesian decision
analysis** carries on from here: attaching a value to each possible
outcome, computing the expected value of each option you could choose,
and putting a price on the information that would reduce your
uncertainty before you commit.

It's out of scope for this book, which stays on estimation rather than
optimisation. But it's a natural next step once you're comfortable
with posteriors, and it's a recurring topic in my newsletter,
*Working Ideas*, at
[andrewmarritt.substack.com](https://andrewmarritt.substack.com).
:::

### Every line the data will support

The single red `lm()` line from earlier was one answer. Here are a
hundred, drawn from the posterior:

```{r posterior-lines}
set.seed(607)

posterior_lines <- fit_reg |>
  spread_draws(b_Intercept, b_customer_rate_c) |>
  slice_sample(n = 100)                                    # <1>

ggplot(salespeople, aes(customer_rate, sales)) +
  geom_point(alpha = 0.2, colour = "grey60") +
  geom_abline(data = posterior_lines,
              aes(intercept = b_Intercept - b_customer_rate_c * mean_rate,
                  slope     = b_customer_rate_c),
              colour = "#122a52", alpha = 0.15, linewidth = 0.4) +   # <2>
  labs(title = "100 relationships the data supports",
       subtitle = "Compare with the prior fan above — the data has narrowed it dramatically",
       x = "Customer rating", y = "Sales (thousands of dollars)")
```

1.  100 draws is plenty for a readable plot. Each one is a complete,
    internally consistent pair of intercept and slope — a whole
    candidate line, not two numbers picked independently.
2.  Every line has the **same** low opacity, deliberately. Each
    posterior draw is equally valid, so it would be wrong to draw some
    more boldly than others. The darkening you see where lines
    concentrate is honest overplotting: it's density, and density is
    the thing you want to read.

::: callout-important
## This is the picture worth keeping

Put this beside the prior fan. Same axes, same everything — but the
data has taken a wide range of plausible relationships and narrowed
it to a tight bundle. That narrowing *is* the learning, and the width
of what's left is the honest uncertainty in the answer.

It's also what the red `lm()` line was hiding. One line implies one
answer. The bundle shows you the set of answers still standing.
:::

::: callout-important
## Why "credibly positive" matters

If the 95% credible interval sits entirely above zero, you can say
plainly: *"we are confident the effect is real and positive — higher
customer satisfaction goes with higher sales, and it's about this
big."* No p-values, no "fail to reject" — just a direct statement about
the size of the effect.
:::

::: callout-note
## A different lens: this is the Bayesian version of `lm()`

`brm(sales ~ customer_rate_c)` and classical `lm(sales ~ customer_rate_c)`
will usually give very similar point estimates for the slope — this
isn't a case where the two philosophies typically disagree on the
number. What you gain here: a full posterior distribution for the
slope (rather than a point estimate and a standard error built on
asymptotic theory), and a natural way to fold in prior knowledge when
you have it. For a quick, well-behaved regression with plenty of data,
reaching for `lm()` first to sanity-check your `brms` result is a
perfectly reasonable habit.
:::

------------------------------------------------------------------------

## Two kinds of uncertainty

### The distinction that trips people up

Two very different questions:

-   *"What is the **average** sales figure for a customer rating of
    4.5?"*
-   *"What will **this particular salesperson** sell?"*

-   **Parameter uncertainty** — how unsure we are about the *line
    itself*. Shrinks as data grows.
-   **Prediction uncertainty** — how unsure we are about *one new
    salesperson*. Stays wide, because individuals scatter no matter
    how much data we have.

### See both at once

The distinction lives entirely in **which of two `tidybayes` functions
you call**, so it's worth reading this chunk slowly.

```{r two-uncertainties}
rating_grid <- tibble(
  customer_rate_c = seq(from = min(salespeople$customer_rate_c),
                        to   = max(salespeople$customer_rate_c),
                        length.out = 40)
) |>
  mutate(customer_rate = customer_rate_c + mean_rate)          # <1>

epred <- rating_grid |>
  add_epred_draws(fit_reg) |>                                  # <2>
  mean_qi(.epred, .width = 0.95)

pred <- rating_grid |>
  add_predicted_draws(fit_reg) |>                              # <3>
  mean_qi(.prediction, .width = 0.95)

ggplot() +
  geom_point(data = salespeople, aes(customer_rate, sales),
             alpha = 0.3, colour = "grey50") +
  geom_ribbon(data = pred, aes(customer_rate, ymin = .lower, ymax = .upper),
              fill = "grey70", alpha = 0.35) +                 # <4>
  geom_ribbon(data = epred, aes(customer_rate, ymin = .lower, ymax = .upper),
              fill = "#122a52", alpha = 0.60) +
  geom_line(data = epred, aes(customer_rate, .epred),
            colour = "#122a52", linewidth = 1) +
  labs(title = "Two bands: the line vs a single salesperson",
       subtitle = "Navy = uncertainty in the average line; grey = range for one salesperson",
       x = "Customer rating", y = "Sales (thousands of dollars)")
```

1.  The model is fitted on the centred predictor, so the grid has to
    be in centred units — but we carry the raw rating along too, so
    the plot's x-axis stays in units a reader recognises.
2.  **`add_epred_draws()`** — "e" for *expectation*. For each rating,
    where does the **average** sit? This uses only α and β, so it
    inherits their uncertainty and nothing else. It is the navy band.
3.  **`add_predicted_draws()`** — where would an actual **individual
    salesperson** land? Same α and β, but now σ is applied on top to
    scatter one person around the line. It is the grey band. **This
    single line of difference is the whole point of the section.**
4.  Grey rather than red here: in this book red means *likelihood*
    (see the palette in Chapter 4), and this band isn't that.

::: callout-important
## Which band answers which question

-   *"If we lift average ratings by a point, what happens to average
    sales?"* → **navy**. You're asking about the line.
-   *"Priya's rating is 4.5 — what will she sell?"* → **grey**. You're
    asking about a person.

Quoting the narrow navy band when someone asked about one specific
salesperson is how forecasts end up embarrassingly overconfident.

And note what happens as you collect more data: the navy band keeps
shrinking, because more evidence pins the line down. The grey band
mostly doesn't. Individual salespeople vary for reasons this model
never observed, and no amount of extra data makes that go away. If
your prediction interval for one person still looks uncomfortably
wide, that is not a failure of the analysis — it's the analysis being
honest with you.
:::

------------------------------------------------------------------------

## Does the model fit?

Chapter 5 ran `pp_check()` on a *prior*-only model, to see what our
assumptions implied before the data arrived. The same function on a
fitted model asks the opposite and more important question: **if this
model were true, would it generate data that looks like what we
actually collected?**

```{r ppc}
pp_check(fit_reg, ndraws = 50) +
  labs(title = "Posterior predictive check",
       subtitle = "Simulated sales (light) vs observed sales (dark)")
```

### What you're looking for

The dark line is your real sales data. The light lines are datasets
the fitted model invents. You want the dark line to look like it
belongs to the family — not identical to any one light line, but not
obviously the odd one out either.

Common failures worth recognising:

-   **The dark line is shifted** left or right of the pack — the model
    is systematically over- or under-predicting.
-   **The dark line is much narrower or wider** — σ is wrong, so the
    model misjudges how much salespeople vary.
-   **Real data has two humps, simulations have one** — there's a
    group structure the model doesn't know about. That's Chapter 7.
-   **Simulations run negative** where the real data can't — a Normal
    likelihood on a strictly positive outcome. Tolerable if it's a
    sliver, a signal to change family if it's substantial.

::: callout-note
A model that passes this check isn't proven correct — it's just not
visibly contradicted by the data. That's a lower bar than it sounds,
and still worth clearing before you present anything.
:::

### Your turn {.unnumbered}

**1. A different predictor.** Fit a regression of `sales` on
`performance` (treated here as a number, for practice). Centre it
first. Report the slope with its credible interval — and note, for
later, that Chapter 18 explains why treating a rating scale as a plain
number deserves a second look.

**2. Does the prior matter?** Chapter 5 insisted on a sensitivity
analysis, and we haven't run one here. Refit the customer-rating model
with a much vaguer slope prior — `prior(normal(0, 5000), class = b)` —
and plot the two slope posteriors together. With
`r nrow(salespeople)` salespeople, predict what you'll see before you
run it.

**3. How fragile is the business case?** Recompute
`p_pays_for_itself` across a range of assumed uplifts — say 0.05 to
0.30 rating points — and plot the probability against the assumption.
At what uplift does the recommendation flip? That plot, not the slope,
is usually the one worth taking into the meeting.

```{r your-turn}
# Your code here

```

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Regression is very likely the single most useful tool in this book for
daily work. What sets a Bayesian analysis apart is honesty about
uncertainty: you report the slope *with a credible interval*, and you
separate "how sure am I about the effect" from "how well can I predict
one specific person." Being able to make — and explain — that
distinction to a business stakeholder is a mark of a mature analysis.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  `brms` fits regression with `outcome ~ predictor` — same call as
    last chapter, one more term.
2.  Ask **how much**, not **whether**. The magnitude and its
    uncertainty are what a decision needs.
3.  **Centre your predictors** so the intercept means something — and
    so the prior you set is the quantity you read back.
4.  A **slope** is a posterior distribution; report its **credible
    interval**, and plot the bundle of lines it implies.
5.  Turn a cost into a **break-even threshold** and count the draws
    above it — the probability a decision pays off, in one line.
6.  **Parameter uncertainty** (the line, `add_epred_draws()`) shrinks
    with more data; **prediction uncertainty** (one person,
    `add_predicted_draws()`) largely doesn't.
6.  A **posterior predictive check** shows whether the model
    reproduces the data — necessary, not sufficient.
7.  Classical `lm()` and Bayesian regression usually agree closely on
    the point estimate — the posterior is the added value.
:::

## Next chapter {.unnumbered}

**Groups and categories** — adding categorical predictors, comparing
groups, and choosing between competing models.
