# When Data Has Structure: Multilevel Models

## Welcome back {.unnumbered}

People Analytics data is almost always **grouped**: employees within
managers, managers within offices, survey responses within teams,
performance within business units.

::: callout-important
Observations in the same group resemble each other. Pretending they're
independent makes you **overconfident** — and a sceptical stakeholder
notices.
:::

### The league table {.unnumbered}

Here is a thing that happens in every organisation, probably this
quarter.

Someone runs the numbers on manager assessment scores by office, sorts
them descending, and puts the table on a slide. The top office gets
praised and asked to share what it's doing. The bottom office gets a
development plan, or a visit.

The table is arithmetically correct. Every average in it is computed
properly. And it is very likely to be close to meaningless, because
**a league table of group averages is partly a league table of group
sizes** — the smaller the office, the further from the middle its
average can wander for no reason at all.

You already know why. Chapter 3 showed that an estimate from a small
sample swings wildly, and Chapter 1 named the trap this creates: the
extreme office regresses to the mean next year, and whatever
intervention happened in between gets the credit.

::: callout-note
So the honest question isn't *"which office is best?"* It's **"how
much of this ranking is real, and how much is sample size?"** — and
that question has a proper answer.
:::

Today we handle grouped data with a **multilevel model**, and meet one
of the most elegant ideas in statistics: **partial pooling**. It's the
foundation for Chapters 14, 15 and 16, which are the chapters most
People Analytics teams end up needing.

We switch to the `managers` dataset: manager test scores, grouped by
city.

## What you'll be able to do by the end {.unnumbered}

1.  Recognise **hierarchically structured** data, and why ignoring the
    structure makes you overconfident
2.  Explain **complete pooling, no pooling, and partial pooling** —
    and why the third is almost always right
3.  Fit a **varying-intercepts** model with `brms`
4.  See **shrinkage** happen, and explain to a stakeholder why the
    small office moved
5.  Answer the "shouldn't you correct for **multiple comparisons**?"
    challenge, and know exactly how far the answer goes
6.  Read the **group-level** (hyper)parameters, including one that is
    a business finding in its own right
7.  Predict for a group you have **no data on at all**

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("managers", package = "peopleanalyticsdata")

# Drop rows missing either variable we're about to model, following the
# check-before-you-summarise habit from Chapter 1.
managers <- managers |> drop_na(test_score, city)
```

`managers` has 571 managers, one row each, with a `test_score` (a score
on a standard assessment given to all managers) and `city` (their
office location) — exactly the grouped shape we need.

### Look at the scale before you choose a prior {.unnumbered}

Chapter 5's rule was never to write a prior for a variable you haven't
looked at. `test_score` is on an unfamiliar scale, so:

```{r score-scale}
managers |>
  summarise(
    n      = n(),
    mean   = mean(test_score),
    sd     = sd(test_score),
    lowest = min(test_score),
    highest = max(test_score)
  )
```

Scores run in the hundreds, not the tens — worth thirty seconds of
checking, because a prior written for the wrong order of magnitude
will quietly fight the data and produce a badly-behaved model.

I always plot the variable as well as summarising it. Summary
statistics tell you where a variable sits; only the picture tells you
whether it sits there in the way you assumed. Five numbers are
compatible with a great many shapes, and some of those shapes would
change the model.

```{r score-distribution}
ggplot(managers, aes(x = test_score)) +
  geom_histogram(bins = 30, fill = "#122a52", alpha = 0.45, colour = "white") +
  geom_vline(xintercept = mean(managers$test_score),
             colour = "#8a5a00", linewidth = 1) +
  geom_vline(xintercept = median(managers$test_score),
             colour = "#8a5a00", linewidth = 1, linetype = "dashed") +
  labs(title = "Distribution of manager test scores",
       subtitle = "Solid line = mean, dashed line = median",
       x = "Test score", y = "Number of managers")
```

One broad hump, mean and median close together, no second cluster and
no wall of scores piled at a maximum. That's the shape `family =
gaussian()` assumes, so the default choice is defensible here rather
than merely conventional. Had the picture shown two humps, or a hard
ceiling, or a long tail on one side, the model below would have needed
to change — and no summary table would have told us.

```{r priors}
priors <- c(
  prior(normal(200, 100),  class = Intercept),   # <1>
  prior(exponential(0.05), class = sd),          # <2>
  prior(exponential(0.01), class = sigma)        # <3>
)
```

1.  The **grand mean** across all offices. Centred at 200 with plenty
    of room either side — we know the scale, and little else.
2.  How much **offices differ from each other**. Rate 0.05 means an
    expected between-office spread of about 20 points (remember from
    Chapter 5: the exponential's average is 1 ÷ rate). This is the new
    parameter class in this chapter — `sd` is always the group-level
    variation.
3.  How much **individual managers differ within an office**. Expected
    around 100, using the range ÷ 4 rule from Chapter 5.

------------------------------------------------------------------------

## The problem

### Look at our cities

```{r raw-means}
raw <- managers |>
  group_by(city) |>
  summarise(
    raw_mean = mean(test_score),
    n        = n(),
    se       = sd(test_score) / sqrt(n())    # <1>
  ) |>
  arrange(n)

raw
```

1.  The standard error from Chapter 3 — σ ÷ √n. This is the column the
    league table never has, and it's the one that decides whether the
    ranking means anything.

Read the `n` column before the `raw_mean` column. Orlando has 23
managers; Toronto has 213. That is a ninefold difference in how much
evidence sits behind each average, and nothing on the slide version of
this table would tell you so.

Now look at `se`. One standard error is about 5 points for Toronto and
about 15 for Houston — and because a 95% range runs roughly two standard
errors either side, that is ±11 against ±31. The small offices are not
slightly less reliable than the large ones — they are about three times
less reliable.

Put the two together and the ranking starts to fall apart. New York
tops the table at 253 and San Francisco follows at 244, a gap of nine
points — but San Francisco's own average carries a standard error of
13. The bottom three offices sit at 223, 221 and 213, separated by
gaps smaller than any of their standard errors. Most of the ordering in
this table is not a finding about offices. It is the arithmetic of
unequal sample sizes.

### The league table, with its error bars restored

```{r raw-with-error}
ggplot(raw, aes(x = raw_mean, y = fct_reorder(city, raw_mean))) +
  geom_vline(xintercept = mean(managers$test_score),
             linetype = "dashed", colour = "#d32f2f") +
  geom_errorbarh(aes(xmin = raw_mean - 2 * se,
                     xmax = raw_mean + 2 * se),
                 height = 0.2, colour = "#122a52") +
  geom_point(aes(size = n), colour = "#122a52") +
  labs(title = "The same ranking, showing what each average is worth",
       subtitle = "Bars ≈ 95% range for each office average; red dashed = overall mean",
       x = "Mean test score", y = NULL, size = "Managers")
```

::: callout-important
## What the error bars do to the story

The point estimates give you a clean, confident ordering. The bars
show most of that ordering is inside the noise — the small offices
have intervals wide enough to be almost anywhere, and several offices
that look separated on the slide have intervals that overlap heavily.

Two things follow, and they pull in the same direction:

1.  **A small office's average is a weak estimate**, not a strong
    result. A handful of unusually strong or weak test-takers moves it
    a long way.
2.  **Ranking amplifies exactly this.** Sorting descending puts
    whichever group got the luckiest draw at the top, and small groups
    get lucky more often than large ones. The top of a league table is
    therefore *systematically* enriched with small groups — and they
    are the ones most likely to look ordinary again next year.
:::

::: callout-note
## Why not just drop the small offices?

It's the obvious fix and it's a bad one. You'd be discarding real
information about real employees, and you'd still have to defend the
threshold — is 30 managers enough? 50? Whatever line you draw, an
office just above it gets treated as reliable and one just below gets
deleted from the analysis entirely.

Partial pooling does something better: it keeps every office and
weights each one by how much evidence it actually carries. Nothing is
thrown away, and no threshold has to be defended.
:::

------------------------------------------------------------------------

## Three ways to handle groups

**Pooling** means combining data across groups. The question in this
section is how much of it to do, and there are only three answers
available.

At one extreme you can treat the six offices as if they were one — pool
everything, and accept that you have thrown away the differences between
them. At the other you can treat each office as a separate world with
nothing to learn from the others — pool nothing, and accept that a
23-manager office gets the same trust as a 213-manager one.

Neither is right, and the useful answer is the one in between. The rest
of this chapter is about how a model finds it for you.

### Option 1 — complete pooling

One average across all responses, used for every office.

-   **Pro** — uses all the data, very stable
-   **Con** — pretends every office is identical, which is plainly
    false

### Option 2 — no pooling

A separate average for each office — the raw means above.

-   **Pro** — respects that offices differ
-   **Con** — over-trusts tiny samples; the small-n offices dominate
    the extremes

### Option 3 — partial pooling

::: callout-important
Each office gets its **own** estimate, but offices with little data are
**pulled toward the overall average** — by an amount **the data
itself decides**.
:::

Big offices barely move. Small ones move a lot. Nobody has to choose
the amount by hand.

### How much does each office move?

The mechanism is worth seeing, because "the model decides" sounds like
magic and isn't. Each office's estimate is a **weighted average** of
its own raw mean and the grand mean, and the weight on its own data is:

$$w = \frac{n}{n + \sigma^2 / \tau^2}$$

*How much you trust an office's own average depends on how many
managers it has, relative to how noisy individuals are compared with
how much offices genuinely differ.*

Read what that does at the extremes:

-   **Large *n*** → *w* approaches 1. The office keeps its own
    average. It has earned it.
-   **Small *n*** → *w* approaches 0. The office is pulled towards the
    grand mean, because its own average is mostly noise.
-   **Offices genuinely differ a lot** (large τ) → everyone keeps more
    of their own average, because being different is expected.
-   **Offices are all much the same** (small τ) → everyone is pulled
    hard towards the middle, because apparent differences are
    probably noise.

::: callout-note
That last pair is the clever bit. The model doesn't just ask how much
data each office has — it also **learns from the offices collectively
how different offices tend to be**, and uses that to judge how
seriously to take any individual one. No single office could work this
out alone.
:::

------------------------------------------------------------------------

## Fit a varying-intercepts model

The formula `test_score ~ 1 + (1 | city)` reads: an overall average
(`1`), plus a **city-specific adjustment** (`(1 | city)`) drawn from a
shared distribution across cities. Still the same `brm()`, same
`family = gaussian()` — the only thing that's new is what's inside the
formula.

This model introduces a genuinely new kind of prior: not just "where's
the average," but **"how much do groups typically vary"** — the τ in
the weighting formula above. That's the `class = sd` prior we set in
the setup. As always, look at both before fitting:

```{r plot-ml-priors}
#| code-fold: true
#| code-summary: "Show the plotting code"
bind_rows(
  tibble(x = seq(-100, 500, length.out = 400), parameter = "Intercept: Normal(200, 100)") |>
    mutate(density = dnorm(x, mean = 200, sd = 100)),
  tibble(x = seq(0, 100, length.out = 400), parameter = "Between-city sd: Exponential(0.05)") |>
    mutate(density = dexp(x, rate = 0.05))
) |>
  ggplot(aes(x, density)) +
  geom_area(fill = "#8fabd0", alpha = 0.30) +
  geom_line(colour = "#8fabd0", linewidth = 1) +
  facet_wrap(~ parameter, scales = "free") +
  labs(title = "Our two headline priors",
       subtitle = "Both on the scale the data actually lives on",
       x = NULL, y = "Prior density")
```

```{r fit-ml}
fit_ml <- brm(
  test_score ~ 1 + (1 | city),
  data = managers, family = gaussian(),
  prior = priors,
  chains = 4, iter = 4000, seed = 8, refresh = 0,     # <1>
  control = list(adapt_delta = 0.95)                  # <2>
)

summary(fit_ml)
```

1.  Double the usual iterations. At the standard 2,000 this model
    returns `Rhat = 1.01` on the intercept — right at the threshold the
    diagnostics box below tells you not to accept. Six groups is a hard
    posterior to explore, and the cheapest fix is to run longer.
2.  The first time we've needed an extra argument beyond the standard
    call. Multilevel models have an awkwardly shaped posterior — when
    a group-level `sd` could plausibly be near zero, the geometry
    narrows to a spike the sampler can overshoot, and you get
    *divergent transition* warnings. Raising `adapt_delta` makes the
    sampler take smaller, more careful steps. It's slower and it's the
    standard first response. Chapter 11 explains what's actually
    happening.

    The default is **0.8**. The number is the sampler's target
    acceptance rate: the higher it is, the smaller and more cautious
    each step becomes, and the less likely the sampler is to overshoot
    the narrow part of the posterior. Raising it costs time and buys
    reliability, so the convention is to escalate only as far as you
    need — 0.95 first, 0.99 if warnings persist. We start at 0.95 here
    because multilevel models with few groups warn at the default often
    enough that it is not worth running twice to find out.

### Two new numbers to read

The output has grown a section. Before reading any of it as a finding,
here is what each block is telling you, in order of how much it matters.

**`sigma`, at the bottom, is how much individual managers differ from
each other.** About 80 points. Two managers picked at random from the
same office typically score around 80 points apart.

**`sd(Intercept)`, at the top, is how much the offices differ.** About
12 points. That is the number the whole league table was implicitly
claiming to measure.

**`Intercept`, in the middle, is the average across offices** — about
236 — now estimated properly rather than by averaging six averages of
very different sizes.

::: callout-important
Put the first two side by side and the chapter's argument is finished
in one line: **differences between individual managers are roughly
seven times larger than differences between offices.**

Around 2% of the variation in test scores sits between offices. The
other 98% sits between people working in the same office. A league
table of office averages is a ranking of the 2%, presented as though it
were the whole story.
:::

That ratio has a name — the intraclass correlation — and Chapter 16
turns it into a finding in its own right rather than a caveat.

::: callout-note
## `sd(Intercept)` is a real finding

It answers a genuine business question: *"how much does manager
assessment performance actually vary between our offices?"* A small
value means your offices are consistent and the league table was
noise; a large one means office genuinely matters and is worth
investigating.

Report it with its interval, not as a point. Which brings us to an
honest caveat.
:::

### Six groups is not many

Look at the credible interval on `sd(Intercept)`. It runs from about
1.5 to about 30 — the estimate is 12, and the data cannot rule out
either "offices are essentially identical" or "offices differ by 30
points". That interval is enormous, and it will be, because we are
trying to estimate *how much offices vary* from only **six offices**.
Six numbers is thin evidence about a spread, in exactly the way that
Chapter 3 warned a small sample would be.

This is the most common honest limitation of multilevel models in
People Analytics, and it's worth stating out loud rather than hiding:
you often have plenty of *people* and very few *groups*. The model is
still the right choice — partial pooling with six groups beats both
alternatives — but the between-group variance is the parameter you
should be most cautious about quoting.

A rough guide: with fewer than about five groups the variance
component is barely identified and you're leaning on your prior; by
ten or fifteen it starts to firm up; with fifty you can take it
seriously. Chapter 15 partially pools across **571 individual
managers**, and the difference in precision is obvious when you get
there.

::: callout-tip
## Check the diagnostics here especially

Chapter 5 said Rhat should read 1.00 and ESS should run into the
thousands. Multilevel models are the first place in this book where
that isn't automatic — the awkward geometry mentioned above makes them
genuinely harder to sample.

If you see `Rhat` at 1.01 or above, low `Bulk_ESS`, or divergent
transition warnings, **do not report the model**. Raise `adapt_delta`
towards 0.99, increase `iter`, and check your priors are on the right
scale — an off-scale prior fighting the data is a common cause, and
it's why we measured `test_score` before writing one.
:::

------------------------------------------------------------------------

### Does it fit?

Diagnostics said the sampler behaved. That is step 7. Step 8 asks a
different question — not *did the machinery work* but *can this model
produce data that looks like ours*:

```{r ppc-ml}
pp_check(fit_ml, ndraws = 100) +
  labs(title = "Posterior predictive check",
       subtitle = "Light lines = datasets the model would generate; dark = the real data",
       x = "Test score", y = "Density")
```

Look for the dark line sitting inside the light ones. A multilevel model
can pass every convergence check and still be the wrong shape — if test
scores are bimodal, or bounded, or have a spike at zero, a Normal
likelihood will smooth straight over it and the shrinkage estimates
inherit the error.

::: callout-note
Two things this check will *not* tell you, and it is worth knowing the
boundary. It does not tell you whether six groups is enough to estimate
`sd(Intercept)` — nothing does, beyond reporting the interval, which we
did above. And it does not tell you whether the offices differ *because
of* something about the offices. That is a causal question, and Chapter 21 is where it gets answered.
:::

------------------------------------------------------------------------

## See the shrinkage

We're about to use `add_epred_draws()` again — the "epred" is
*expected prediction*, introduced in Chapter 6 alongside
`add_predicted_draws()`. The reminder, since it matters here: `epred`
gives you the posterior for the **average** in each group, carrying
only the uncertainty in the model's parameters. `predicted` would give
you the range for a single new *manager*, which is much wider and is
not the question a league table asks.

```{r shrinkage}
grand_mean <- mean(managers$test_score)

pooled <- managers |>
  distinct(city) |>
  add_epred_draws(fit_ml) |>
  mean_qi(.epred)

compare <- left_join(raw, pooled, by = "city")

ggplot(compare, aes(x = n)) +
  geom_hline(yintercept = grand_mean, linetype = "dashed",
             colour = "#d32f2f", linewidth = 1) +
  geom_segment(aes(xend = n, y = raw_mean, yend = .epred),
               colour = "grey80") +
  geom_point(aes(y = raw_mean), colour = "grey70", size = 2.5) +
  geom_point(aes(y = .epred), colour = "#122a52", size = 2.5) +
  scale_x_log10() +
  labs(title = "Shrinkage: partial pooling pulls small offices inward",
       subtitle = "Grey = raw average, navy = model estimate, red dashed = grand mean",
       x = "Number of managers (log scale)", y = "Test score")
```

### Read it left to right

-   **Left** (few managers) — long grey lines: extreme raw averages
    pulled hard toward the middle
-   **Right** (many managers) — barely any movement: the model trusts
    them

::: callout-important
The model worked out **how much to trust each office**, all by itself.
Nobody set a threshold, nobody dropped a small office, and nobody had
to argue about where the cut-off should sit.
:::

------------------------------------------------------------------------

## The other way people fix this: the funnel plot

Partial pooling is not the only answer to the unequal-groups problem,
and it is not the best known one. In healthcare and education, where
published league tables of hospitals and schools do real damage, the
standard tool is the **funnel plot**, popularised by David
Spiegelhalter. It is worth knowing, both because you will meet it and
because comparing the two makes clear what partial pooling actually
buys you.

The idea is simple. Plot each group's average against its sample size,
then draw the range within which an average would be expected to fall
*if every group were identical*. That range is wide for small groups
and narrow for large ones, so the boundaries form a funnel.

```{r funnel-plot}
sigma_within <- managers |>                                        # <1>
  group_by(city) |>
  summarise(s = sd(test_score), n = n(), .groups = "drop") |>
  summarise(pooled = sqrt(sum((n - 1) * s^2) / sum(n - 1))) |>
  pull(pooled)

funnel <- tibble(n = seq(15, 230, length.out = 300)) |>
  mutate(
    lower = grand_mean - 2 * sigma_within / sqrt(n),               # <2>
    upper = grand_mean + 2 * sigma_within / sqrt(n)
  )

ggplot() +
  geom_line(data = funnel, aes(n, lower), colour = "grey70", linetype = "dashed") +
  geom_line(data = funnel, aes(n, upper), colour = "grey70", linetype = "dashed") +
  geom_hline(yintercept = grand_mean, colour = "#d32f2f", linetype = "dashed") +
  geom_point(data = raw, aes(n, raw_mean), colour = "#122a52", size = 3) +
  geom_text(data = raw, aes(n, raw_mean, label = city),
            nudge_y = 9, size = 3, colour = "#122a52") +
  labs(
    title = "A funnel plot of the same six offices",
    subtitle = "Grey dashed = the range you would expect if every office were identical; red dashed = overall mean",
    x = "Number of managers", y = "Mean test score"
  )
```

1.  The pooled within-office standard deviation — how much individual
    managers differ, averaged properly across offices of different
    sizes.
2.  The expected range for a group of size `n`, two standard errors
    either side of the overall mean. This is the Chapter 3 formula
    again, drawn as a curve instead of evaluated at one point.

Read it by asking which points fall outside the funnel. Inside means
the office is no further from the average than sampling noise would
comfortably explain. The plot makes the small-office problem
impossible to miss: the funnel is so wide on the left that an office of
23 managers has to be extraordinary before it counts as different at
all.

### Which should you use?

They answer different questions, and the difference is worth being
precise about.

The funnel plot asks, of each office separately: **could this
difference be chance?** It needs no model, takes one line to explain,
and is very hard to argue with in a room. What it does not do is give
you a better number. It tells you which averages to ignore; it leaves
the averages themselves untouched.

Partial pooling asks: **what is my best estimate for each office, given
everything I know?** It answers for every office at once, produces a
revised number rather than a verdict, and carries its uncertainty with
it.

::: callout-important
Which means the answer to the stakeholder who wanted a ranking is not
"you can't have one". **You can have a league table — just not one
built from raw averages.** Rank the partially pooled estimates instead.
Small offices will sit closer to the middle than they used to, which is
the honest position for them to occupy, and the ordering that survives
is the ordering the evidence actually supports.

That is a much better conversation than refusing the request, and it is
why this chapter ends with a model rather than a warning.
:::

The two tools also work well together. The funnel plot is the better
picture for showing an audience *why* the raw table was misleading; the
pooled estimates are the better basis for what you do next. Chapter 15
takes the second half of that much further, ranking 571 individual
managers.

------------------------------------------------------------------------


::: callout-note
## What to say when someone asks why the number changed

You will be asked this, and "shrinkage" is not the answer to give.

The version that works: *"That office has eleven managers. If we
reported their raw average we'd be treating eleven people's scores as
though they were as reliable as Toronto's two hundred, and they
aren't — with that few, the average could easily land several points
either side by chance. So the estimate leans partly on what we know
about offices in general. As the office grows, it'll rely more on its
own numbers and less on everyone else's."*

That's an argument about **how much evidence there is**, which
stakeholders accept readily. It's the same argument you'd make against
declaring a trend from two data points.
:::

### The multiple comparisons problem, quietly solved

Somebody trained classically will now raise a hand. We have six
offices. Comparing all of them against each other is fifteen
comparisons, and everyone learns that testing fifteen things at the
5% level will hand you a "significant" result roughly half the time by
chance alone. The classical fix is a **correction** — Bonferroni,
Tukey, false discovery rate — which widens every interval to buy back
the error rate you spent.

We haven't applied one, and we aren't going to. Here's why.

::: callout-important
## Partial pooling has already done the work

The classical problem arises because **no pooling** lets each office's
estimate go wherever its own small sample points, and then asks
fifteen questions of estimates that were free to wander. The
corrections are applied afterwards, to compensate.

Partial pooling stops it happening in the first place. Every office's
estimate has already been pulled towards the grand mean by an amount
set by its sample size, so the extremes that generate spurious
"differences" have been damped *before* any comparison is made. The
group-level prior — `sd(Intercept)`, learned from the offices
collectively — is doing structurally what a multiplicity correction
does by decree, except that the data chose the amount rather than a
convention.
:::

Two things follow that are worth being precise about, because it is
easy to overclaim here.

1.  **This is a property of the multilevel model, not of Bayes on its
    own.** A Bayesian model with a flat prior and no pooling would
    reproduce the classical problem faithfully. It's the shared
    distribution across groups — Chapter 5's point that a prior is a
    constraint, applied here to a whole family of estimates at once —
    that provides the protection. Gelman, Hill and Yajima make this
    argument in full in *"Why we (usually) don't have to worry about
    multiple comparisons"* [@gelman2012multiple]; it is the standard
    reference if a stakeholder or a reviewer pushes.
2.  **A credible interval doesn't change when you look at another
    one.** This is the deeper reason the machinery differs. A
    confidence interval is a statement about a procedure repeated over
    hypothetical datasets, so how many intervals the procedure
    produces changes what it guarantees. A posterior is a statement
    about the parameters given the data you have — asking a second
    question of the same posterior doesn't alter the answer to the
    first. There is no error rate being spent, so there is nothing to
    correct.

::: callout-note
None of this licenses fishing. If you compute all fifteen contrasts
and report the one that looks best, you'll mislead people with or
without a correction — the honest move is to say how many you looked
at. The claim is narrower and more useful: with partial pooling doing
the regularising, you don't need to inflate your intervals to make the
comparison legitimate.
:::

### Your turn {.unnumbered}

**1.** Which office has the largest gap between its raw average and its
model estimate? Confirm it's one with few managers.

**2.** Refit with `prior(exponential(0.5), class = sd)` — a prior
insisting offices barely differ. Does shrinkage get stronger or
weaker, and does that match the weighting formula from earlier?

```{r your-turn}
# Hint for 1: add a column = abs(raw_mean - .epred) and arrange().

```

------------------------------------------------------------------------

## Predicting a brand-new office

Suppose the business is opening in a city you have no managers in yet,
and someone asks what to expect. Complete pooling would give you the
grand mean with no acknowledgement that offices differ. No pooling
can't answer at all — there's no group to look up.

```{r new-office}
tibble(city = "NEW_OFFICE") |>
  add_epred_draws(fit_ml, allow_new_levels = TRUE) |>    # <1>
  mean_qi(.epred, .width = 0.95)
```

1.  `allow_new_levels = TRUE` tells `brms` this group wasn't in the
    training data and that's deliberate. It then draws a brand-new
    office-level offset from the distribution it learned across the
    six offices it *has* seen.

The estimate sits near the grand mean — sensible, since we know
nothing specific — but the interval is **wider** than for any existing
office, because it now carries the uncertainty about how much offices
differ *on top of* the usual uncertainty.

::: callout-important
That's the payoff of modelling the group-level distribution rather
than just the groups. You get a defensible answer for a group you have
never observed, with honest uncertainty attached — and "we opened a
new office, what should we expect?" is a question People Analytics
teams get asked constantly.
:::

::: callout-note
## Where this goes next

Chapter 14 applies this exact idea in closed form, without fitting a
full model (empirical Bayes). Chapter 15 goes deeper on partial
pooling applied to *individual managers* rather than offices — the
question most People Analytics teams actually get asked — and directly
compares this model to the classical mixed-effects model
(`lme4::lmer()`) you may already know.

Two further descendants are worth knowing about now, so you recognise
them when they arrive. Chapter 16 turns `sd(Intercept)` from a caveat
into the headline finding, decomposing pay variation across levels of an
organisation. And Chapter 17 attaches this same grouping term to a
*survival* model — a **frailty** model, where each group gets its own
baseline risk of someone leaving. That last one is where this chapter's
argument pays off most directly: it answers "which managers lose people
faster than their teams explain?" without the league table that started
this chapter.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

If your data is grouped — employees within managers, responses within
teams, outcomes within regions — a multilevel model is usually the
**correct** analysis, and a sophisticated stakeholder increasingly
expects it. It stops you over-claiming from small subgroups, borrows
strength across groups, and even lets you say something sensible about
a group you've never seen before.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Grouped data needs **multilevel models**; ignoring structure makes
    you overconfident.
2.  A **league table of group averages is partly a league table of
    group sizes**. Put error bars on one and most of the ranking
    dissolves.
3.  **Partial pooling** sits between one-average-for-all and
    a-separate-average-each — and beats both. Don't drop small groups
    instead; you'd lose data and still have to defend the threshold.
4.  How far each group moves is a **weighted average**, set by its
    sample size relative to how much groups genuinely differ. The
    model learns the second part from the groups collectively.
5.  `(1 | group)` adds a **varying intercept** per group in `brms`.
6.  **Shrinkage** pulls small-sample groups toward the grand mean —
    and improves predictions. Explain it to stakeholders as an
    argument about *how much evidence there is*, never by using the
    word "shrinkage".
7.  Partial pooling means you don't apply a **multiple comparisons
    correction**. The pooling has already damped the extremes that
    the correction exists to punish — and a credible interval doesn't
    change because you looked at another one.
8.  `sd(Intercept)` estimates **between-group spread**, itself a
    finding — but it needs a decent **number of groups** to be
    estimated well, and six is not many.
9.  Multilevel models are the first place **sampling diagnostics stop
    being automatic**. Check Rhat and ESS, and reach for
    `adapt_delta`.
10. With a group-level distribution in hand you can predict for a
    group you've **never observed**.
:::

## Next chapter {.unnumbered}

**Logistic regression** for a yes/no outcome, which HR has a large number of.
