# Inside the Sampler: MCMC & Trustworthy Models

## Optional deep-dive {.unnumbered}

Everything so far assumed the model "just worked." Usually it does.
This chapter explains **how it works**, and — more practically — **how
to tell when it hasn't**.

::: callout-important
But a sceptical stakeholder — or your own good judgment — may ask:
**"how do you know your model converged?"** If you can't answer that,
you can't defend any of your results.
:::

This chapter shows you how to check — and what a broken fit looks like
so you'll recognise one.

## What you'll be able to do by the end {.unnumbered}

1.  Explain in plain terms what **MCMC** does
2.  Read **trace plots**, **R̂ (Rhat)** and **effective sample size
    (ESS)**
3.  Recognise a **badly-behaved** fit and fix it

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(bayesplot)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("salespeople", package = "peopleanalyticsdata")

# Chapter 1 found a single missing value in each of `sales`,
# `customer_rate` and `performance`; dropping those rows here keeps
# the rest of this chapter's code simple.
salespeople <- salespeople |>
  drop_na(sales, customer_rate, performance, promoted) |>
  mutate(performance_f = factor(performance))
```

------------------------------------------------------------------------

## What is MCMC, really?

### The problem

In Chapter 4 we computed a posterior two ways: a grid, and the
Beta-Binomial shortcut.

::: callout-note
Both only work for the simplest models. Add a few predictors and there
is **no formula** and no grid fine enough — the space is too big.
:::

### The solution: explore it

Instead of solving for the posterior, `brms` (via Stan) **walks around**
it. Imagine the posterior as a landscape: every possible combination of
parameter values is a location, and the height at that location is how
plausible that combination is, given your data and priors. You cannot
see the landscape — it has too many dimensions to draw and no formula to
evaluate — but you can stand somewhere on it and feel how high you are.

So you walk. The whole algorithm, written out as if you were doing it on
foot in fog:

```
stand somewhere — anywhere
repeat, thousands of times:
    pick a nearby spot at random
    measure how high it is compared with where you are now

    if the new spot is higher:
        move there

    if the new spot is lower:
        move there anyway — but only sometimes,
        with a chance equal to how much lower it is
        (a little lower: usually move; far lower: rarely move)

    write down where you are standing
```

At the end you have a long list of locations. That list is the answer.

### Why it agrees to walk downhill

The step that looks like a mistake is the one that makes the whole thing
work, and it is worth slowing down on.

Suppose the rule were simply *"move if it's higher, otherwise stay"*.
The walk would climb the nearest hill, arrive at the top, and stop —
because every direction from a summit is down. You would learn where one
peak is and nothing else. Worse, if the landscape has several peaks, you
would find whichever one happened to be nearest your starting point and
never discover the others. That trap has a name you may know from
optimisation: a **local maximum**.

Sometimes accepting a downhill step is what lets the walker leave a peak,
cross a valley, and find out what else is out there. It is deliberate,
controlled randomness — noise added on purpose, as insurance against
confidently exploring only one small region.

::: callout-important
The downhill rule does something else, and it is the real reason for it.
Because the walker moves to lower ground *in proportion to how much
lower it is*, it ends up spending time in each region **in proportion to
how plausible that region is**.

That is the whole trick. It means you can simply count how often the
walk visited each area, and those counts reproduce the posterior — no
formula required.
:::

### What Stan actually does

The version above is the **Metropolis algorithm**, from 1953, and it is
the honest core of the idea. Stan uses a considerably smarter descendant
called **Hamiltonian Monte Carlo** — the "much smarter way of wandering"
mentioned in Chapter 5 — which uses the *slope* of the landscape to
propose steps that are long, well-aimed and rarely wasted, instead of
guessing at random and often being rejected.

The improvement is in how the next spot is chosen. The accept-or-reject
logic, and the reason for it, are exactly as above. Everything in the
rest of this chapter applies to both.

### The draws *are* the answer

The collection of places the walk visits — the **draws** — *is* our
approximation of the posterior. Every posterior plot and credible
interval you've made so far was computed from those draws.

::: callout-important
So if the walk went badly, **everything downstream is wrong** —
quietly, and without an error message.
:::

::: callout-tip
## For the ML/DS crowd

If you've trained models with gradient descent, MCMC will look
superficially similar — both are iterative — but they're solving
different problems. Gradient descent hunts for a **single best point**
(the minimum of a loss function, equivalent to the posterior's mode).
MCMC instead tries to visit **every plausible region in proportion to
how plausible it is**, so that the full spread of the draws describes
the whole posterior, not just its peak. That's why a converged MCMC
chain gives you a credible interval for free, where a converged
gradient descent run gives you one number and no direct measure of
uncertainty.
:::

### Two things must go right

-   **Convergence** — several independent walks (chains) must end up
    exploring the *same* region.
-   **Enough exploration** — the walk must move around freely, giving
    us plenty of effectively-independent draws.

Three diagnostics check exactly this.

------------------------------------------------------------------------

## A healthy model

```{r good-fit}
# Same priors as Chapter 7 - already visualised there. This chapter's
# focus is the sampler's behaviour, not the prior, so we reuse them as-is.
fit_good <- brm(
  sales ~ customer_rate + performance_f, data = salespeople, family = gaussian(),
  prior = c(prior(normal(400, 200), class = Intercept),
            prior(normal(0, 200),   class = b),
            prior(exponential(0.005), class = sigma)),
  chains = 4, iter = 2000, seed = 10, refresh = 0
)
```

### Diagnostic 1 — the trace plot

```{r good-trace}
mcmc_trace(fit_good, regex_pars = "b_")
```

::: callout-note
## Look for the fuzzy caterpillar

**Healthy** chains: **overlap** each other completely, look **flat and
stationary** — no drift or trend, and are **dense and fuzzy**, not slow
smooth waves. If one chain wanders off alone, or they drift upward
together, the sampler hasn't settled.
:::

### Diagnostic 2 — R̂ (R-hat)

Convergence is not a property of the model as a whole. It is checked
**one parameter at a time**, so the plot has one point per parameter —
every coefficient, plus `sigma`, plus a couple of internal quantities
Stan tracks (`lprior` and `lp__`). A model can settle down nicely for
most of its parameters and still be unreliable for one, which is
precisely what you are looking for here.

`bayesplot` hides the parameter names by default, on the grounds that
with a large model there would be too many to read. Ours is small, so we
switch them back on:

```{r good-rhat}
mcmc_rhat(brms::rhat(fit_good)) +
  yaxis_text(hjust = 1) +                                   # <1>
  labs(title = "R-hat (want all ≈ 1.00)")
```

1.  Turns the parameter labels back on and right-aligns them. Drop this
    line on a model with fifty parameters, where the pattern matters more
    than any individual name.

::: callout-tip
## A package-naming trap worth knowing about

Both `brms` and `bayesplot` export a function called `rhat()` (and
another called `neff_ratio()`) — whichever package you loaded *more
recently* wins when you call the bare function name, and the other
package's version usually can't make sense of a `brmsfit` object.
Writing `brms::rhat(fit)` explicitly, as above, sidesteps the ambiguity
entirely rather than hoping your `library()` calls are in the right
order. This kind of name collision is common enough across the R
ecosystem that reaching for `package::function()` whenever two loaded
packages might share a name is a good habit generally, not just here.
:::

::: callout-important
A simple rule: anything above **1.01** deserves attention.
:::

### Diagnostic 3 — effective sample size

```{r good-ess}
mcmc_neff(brms::neff_ratio(fit_good)) +
  yaxis_text(hjust = 1) +
  labs(title = "ESS ratio (higher is better)")
```

::: callout-note
**ESS** tells you how many genuinely independent draws your sample is
worth. You want it in the **hundreds or thousands**.
:::

All three diagnostics look healthy — we can trust `fit_good`.

------------------------------------------------------------------------

## A broken model (on purpose)

```{r bad-fit, warning=TRUE}
fit_bad <- brm(
  sales ~ customer_rate + performance_f, data = salespeople, family = gaussian(),
  prior = c(prior(normal(400, 200), class = Intercept),
            prior(normal(0, 200),   class = b),
            prior(exponential(0.005), class = sigma)),
  chains = 3, iter = 60, warmup = 25, seed = 10, refresh = 0
)
```

```{r bad-trace}
mcmc_trace(fit_bad, regex_pars = "b_")
```

```{r bad-rhat}
mcmc_rhat(brms::rhat(fit_bad)) + labs(title = "R-hat above 1.00 — a red flag")
```

### The danger

```{r compare-summaries}
summary(fit_bad)$fixed[, c("Estimate", "Rhat", "Bulk_ESS")]
```

::: callout-important
Notice: the model **still printed estimates**. Nothing crashed. If you
hadn't checked the diagnostics, you'd have reported these numbers — and
they're unreliable.
:::

------------------------------------------------------------------------

## Fixing convergence problems

-   **Run longer** — more `iter` (with adequate `warmup`). Fixes most
    cases.
-   **Better priors** — vague priors on awkward scales make the
    posterior hard to explore; weakly-informative priors help.
-   **Standardise predictors** — puts everything on a comparable
    scale.
-   **Handle divergences** — if `brms` warns about "divergent
    transitions," raise `control = list(adapt_delta = 0.95)` (or
    higher), or reparameterise.

```{r fixed}
fit_fixed <- update(fit_bad, iter = 2000, warmup = 1000, refresh = 0)
mcmc_rhat(brms::rhat(fit_fixed)) +
  labs(title = "R-hat back to ≈ 1.00 after running longer")
```

Nothing about the model changed — only how long we let it explore.

### Your turn {.unnumbered}

Fit a model with `chains = 2, iter = 40, warmup = 15` and inspect its
trace plot and R̂. Then fix it by running longer and confirm R̂ returns
to 1.00.

```{r your-turn}
# Your code here

```

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## A paragraph that signals competence

*"All R̂ values were 1.00, effective sample sizes were in the
thousands, and trace plots showed good mixing across four chains."*

One sentence in a methods appendix, and an obvious question from a
technical reviewer is answered before they ask it.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  **MCMC** approximates the posterior by exploring it; the **draws
    are the answer** — a different job from what gradient descent does.
2.  **Trace plots** should look like overlapping fuzzy caterpillars.
3.  **R̂ ≈ 1.00** and healthy **ESS** signal a trustworthy fit.
4.  A broken fit **still prints numbers** — always check.
5.  Most problems are fixed by **running longer** or **better priors**.
:::

## Next chapter {.unnumbered}

**Richer models** — varying slopes and count outcomes, for when your
own work needs more than the core toolkit.
