# Richer Models: When the Standard Recipe Runs Out

## Optional deep-dive {.unnumbered}

Four extensions covering most of what real People Analytics work needs
beyond the standard recipe: letting a *relationship* differ across groups
(**varying slopes**), modelling **counts** properly, handling outcomes
with **several unordered categories**, and dealing with data where
observations belong to **more than one kind of group** at once.

### Where each one shows up at work {.unnumbered}

These are easier to recognise from a situation than from a definition,
so here are two of them — the other two are introduced where they arise.

**"It worked. Well — it worked somewhere."** Your L&D team rolls out a
new onboarding programme across every region and asks whether it
improved time-to-productivity. You fit a model, find a small average
effect, and report it. Six months later a regional director tells you it
transformed her region and another says it changed nothing, and they are
both right. The average was real and it described nobody.

That is the varying-slopes problem. It appears whenever you have an
effect and more than one group: does the training work as well
everywhere, does an extra year of tenure buy the same thing in every
office, does the pay rise reduce attrition equally at every level? The
honest answer is usually no, and a single number cannot say so. **Where
an effect differs is often more actionable than the effect itself** —
you cannot copy what the strong region did if your model has averaged it
away.

**"Nobody has minus two sick days."** Your HR director wants to know
what drives absence. You have sick days per employee last year: mostly
0, 1 or 2, a few people at 15, nobody below zero and nobody at 3.5.
Fit an ordinary regression and it will happily predict −0.4 days for
some employees and 2.7 for others.

That is the count problem, and it covers a lot of HR data: sick days,
transfer requests, applications per vacancy, grievances per site,
support tickets per team, training courses completed. Anything you get
by *counting things that happened*. These need a model that knows the
answer must be a whole number and cannot be negative.

The chapter takes them in turn, using the `managers` dataset from
Chapter 8 throughout — test scores and tenure for the first, transfer
requests for the second, performance groups for the third.

## What you'll be able to do by the end {.unnumbered}

1.  Fit a **varying-slopes** multilevel model
2.  Interpret how an effect **differs across groups**
3.  Fit a **Poisson** regression for a count outcome
4.  Read coefficients as **rate ratios** and know when to reach for
    negative-binomial
5.  Recognise an outcome with **several unordered categories**, and know
    why its coefficients should never be presented
6.  Spot **crossed** grouping structure — the same person rated many
    times — and know what ignoring it costs you

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)
library(ggdist)
library(loo)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("managers", package = "peopleanalyticsdata")

# Guard against missing values before grouping and modelling (Chapter 1
# found gaps in a similar teaching dataset).
managers <- managers |>
  drop_na(test_score, yrs_employed, city, transfers, group_size,
          performance_group)
```

------------------------------------------------------------------------

## Part 1 — Varying slopes

### Where we left multilevel models

In Chapter 8 each city got its own **intercept** — its own average
test score.

::: callout-note
That says offices **start** in different places. But it assumes any
*effect* we add works **identically** everywhere.
:::

### The next question

> Does the effect of **years employed** on test score differ from city
> to city?

Plausibly yes: tenure might sharpen test performance strongly in some
offices (more mentoring, more internal knowledge-sharing) and barely
matter in others.

::: callout-important
If that's true, a single overall slope would **average away** the very
thing you want to report.
:::

### The formula

```{r}
#| eval: false
test_score ~ yrs_employed + (1 | city)                   # Chapter 8
test_score ~ yrs_employed + (1 + yrs_employed | city)    # today
```

`(1 + yrs_employed | city)` gives each city its **own intercept and its
own slope**.

### Fit it

The intercept, `sd`, and `sigma` priors are the same ones from Chapter 8. The one genuinely new number is the slope prior — how many points of
test score a year of tenure could plausibly move things:
```{r plot-prior-slope}
tibble(b = seq(-20, 20, length.out = 400)) |>
  mutate(density = dnorm(b, mean = 0, sd = 5)) |>
  ggplot(aes(b, density)) +
  geom_area(fill = "#122a52", alpha = 0.30) +
  geom_line(colour = "#122a52", linewidth = 1) +
  geom_vline(xintercept = 0, colour = "#d32f2f", linetype = "dashed") +
  labs(title = "Prior for the tenure slope: Normal(0, 5)",
       subtitle = "Most plausible values within about ±10 points per year",
       x = "Change in test score per year employed", y = "Prior density")
```

```{r fit-vs}
fit_vs <- brm(
  test_score ~ yrs_employed + (1 + yrs_employed | city),
  data = managers, family = gaussian(),
  prior = c(prior(normal(75, 15),   class = Intercept),
            prior(normal(0, 5),     class = b),
            prior(exponential(1),   class = sd),
            prior(exponential(0.5), class = sigma),
            prior(lkj(2),           class = cor)),
  chains = 4, iter = 4000, seed = 11, refresh = 0     # <1>
)

summary(fit_vs)
```

1.  Double the usual iterations. The correlation parameter introduced
    below is the hardest thing in this model to estimate, and at 2,000
    it comes back with a warning-level `Rhat`. Chapter 11's rule
    applies: run longer rather than report it.
### Reading the output, line by line

This is the most crowded output in the book, and it is worth going
through slowly. There are now three blocks instead of two, and the new
one is at the top.

**Regression Coefficients — the average city.** `Intercept` is the
expected test score for a manager with zero years of tenure, averaged
across cities. `yrs_employed` is the average slope: what one more year
of tenure does to the test score, *on average across all six offices*.
Note the sign here — the average effect of tenure is **negative**, which
is already worth a conversation. Hold that thought.

**Multilevel Hyperparameters — how much cities differ.** Three numbers,
and each answers a different question:

- `sd(Intercept)` — how much cities differ in their **starting point**.
  Around 1 point, with an interval that nearly touches zero. Cities
  start in essentially the same place.
- `sd(yrs_employed)` — how much cities differ in their **slope**. Around
  12 points per year, with an interval nowhere near zero. Cities differ
  enormously in what tenure does.
- `cor(Intercept, yrs_employed)` — whether a city's starting point and
  its slope are **related to each other**. This one gets its own section
  below.

**Further Distributional Parameters.** `sigma`, as always: how much
individual managers differ from the model's prediction for them.

::: callout-important
Put the average slope next to its variation and the finding is not the
one the average suggests. The average tenure effect is about −10 points
per year. The variation between cities is about 12.

A spread that large around an average that small means the effect is
**negative in some offices and positive in others**. "Tenure slightly
hurts test scores" is not the finding. "Tenure does completely different
things in different offices, and we should ask why" is.
:::

::: callout-caution
## Try it yourself

Before reading on: `sd(Intercept)` came out near zero and
`sd(yrs_employed)` came out large. In plain words, what does that
combination say about these six offices? Write one sentence you could
say to an HR director, then compare it with the paragraph above.
:::

::: callout-note
## Nobody reads this output fluently at first

Multilevel output is genuinely hard to read, and the difficulty is not a
reflection on you. It is dense, the labels are terse, and the same word
means different things in different blocks. Students struggle with it.
So do people who teach it. I still slow down and check myself against
the formula.

One practical suggestion, since it is available and it works: paste the
`summary()` output into an LLM and ask it what each line means, then
argue with the answer. Not because the machine knows better than the
book — it will occasionally be confidently wrong — but because being
told the same thing in different words is often the nudge that makes it
land. Use it to get unstuck, then come back and check the explanation
here agrees.
:::

### The correlation: what a city's starting point tells you about its slope

The third hyperparameter is the one most likely to be skipped, and it is
the most interesting idea in the chapter — partly because it cuts
against how most of us were taught to think about regression. In a
classical course, the intercept and the slope are two separate numbers
you estimate and then move on from. Here they can be **related across
groups**, and that relationship is itself a finding.

An example from my hospitality teaching makes it obvious. Imagine measuring the
queue at a chain of coffee shops, first thing in the morning and again
mid-afternoon:

- **Popular cafés** have a long morning queue — a high intercept — *and*
  the biggest drop by the afternoon: a steep slope.
- **Quiet cafés** have a short morning queue — a low intercept — and
  barely change all day: a flat slope.

Intercept and slope travel together. Knowing where a café starts tells
you something about how much it will move. A model that estimated the
two independently would be throwing that information away — and, more
practically, would predict badly for a new café you have only seen once,
in the morning.

The same shapes turn up constantly in People Analytics, and the **sign**
of the correlation tells you which story you are in:

- **Positive** — offices that start high also improve fastest. The gap
  between your best and worst sites is widening on its own.
- **Negative** — offices that start low improve fastest, because they
  have the most room. The gap is closing, and some of what looks like a
  successful intervention may be nothing more than that.

Those are opposite conclusions with opposite consequences for where you
put your money, and a varying-intercepts-only model cannot distinguish
them.

::: callout-important
Now look at what our model actually says. The estimate is near zero and
the interval runs from about −0.8 to +0.8 — which is very nearly the
entire range a correlation can take. **The data tells us essentially
nothing about this parameter.**

That is not a failure of the method, and it is worth reporting rather
than quietly dropping. With six cities you have six intercepts and six
slopes, and you are asking how two sets of six numbers move together.
Chapter 8 said six groups is thin evidence about a spread; it is thinner
still for a correlation. You would want dozens of groups before this
number said anything, and it is the parameter you should be most
reluctant to quote.
:::

The `lkj(2)` prior in the model is what keeps this well-behaved. It says,
before seeing data, that extreme correlations are less likely than mild
ones — a gentle scepticism that stops the model chasing a ±1 correlation
it cannot support from six points. Without it, this parameter would
happily wander to the edges.

### The number that answers the original question

::: callout-important
Under **Multilevel Hyperparameters**, `sd(yrs_employed)` is how much the
tenure *slope* varies across cities. It is clearly above zero, so the
relationship genuinely differs by office — and a single overall slope
would have hidden that completely.
:::

### See the varying slopes

```{r plot-slopes}
city_slopes <- coef(fit_vs)$city[, , "yrs_employed"] |>
  as_tibble(rownames = "city")

city_slopes |>
  ggplot(aes(x = Estimate, y = fct_reorder(city, Estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#d32f2f", linewidth = 1) +
  geom_pointrange(aes(xmin = Q2.5, xmax = Q97.5), colour = "#122a52") +
  labs(title = "Each city's years-employed slope",
       subtitle = "Red dashed = no effect. The spread across offices is the finding, not the average.",
       x = "Slope: test score change per extra year employed", y = NULL)
```

### Heterogeneity *is* a finding

::: callout-important
"Tenure predicts test performance strongly in some offices and barely
at all in others" is a **more useful** finding than a single average
slope near zero. An overall model would have reported "a modest
effect" — and been misleading about where it actually shows up.
:::

### Your turn {.unnumbered}

Compare `fit_vs` to the varying-intercepts-only model from Chapter 8
using LOO. Does allowing slopes to vary improve prediction here?

```{r your-turn-vs}
# Your code here

```

------------------------------------------------------------------------

## Part 2 — Count outcomes

### A third kind of outcome

Counts are neither continuous nor yes/no: transfer requests per team,
open reqs per recruiter, sick days per employee. These are
**non-negative whole numbers**: 0, 1, 2, 3… and often mostly small.

### Why ordinary regression fails

::: callout-important
1.  It can predict **negative** counts
2.  It can predict **fractional** counts (2.7 transfer requests)
3.  Counts are **skewed** with unequal spread — the Normal assumes
    neither

Using ordinary regression on counts is one of the most common
avoidable mistakes in applied People Analytics work.
:::

### The Poisson model

A third kind of outcome, and — as ever — still just `brm()` with a
formula and priors; only `family = poisson()` is new. Poisson works on
the **log** scale, so — same habit as the log-odds prior back in
Chapter 9 — translate before trusting it:

```{r plot-prior-pois}
set.seed(112)

tibble(log_intercept = rnorm(4000, 0, 1)) |>
  mutate(rate = exp(log_intercept)) |>
  ggplot(aes(rate)) +
  geom_histogram(bins = 40, fill = "#122a52", alpha = 0.6) +
  labs(title = "What Normal(0, 1) on the log scale implies",
       subtitle = "Simulated draws from the intercept prior, translated to a baseline transfer rate",
       x = "Implied baseline transfers per manager", y = "Simulated draws")
```

::: callout-note
Centred on 1 transfer, comfortably covering anything from a fraction
of one up to several — sensible for a rate we genuinely don't know yet,
and nowhere near ruling out either a quiet team or a high-churn one.
:::
::: callout-note
## Why no `scale()` this time?

Chapter 9 standardised its predictor so that `normal(0, 1)` on the
log-odds scale meant something sensible. Here we have left
`yrs_employed` and `group_size` in their raw units, and the reason is
the same reason, pointing the other way.

Standardising is a device for making a prior interpretable when the raw
unit is not. "One standard deviation of customer rating" is meaningless
to a reader, but it was the only way to say "a plausible-sized change"
for a variable whose scale we had no feel for.

Here the raw units are already the right size and already meaningful:
**one more year of tenure**, **one more direct report**. A prior of
`normal(0, 0.5)` on the log scale says one extra year multiplies the
expected count by something between about 0.6 and 1.6 — a statement you
can check against your own judgement in a way that "per standard
deviation of tenure" is not.

The rule that falls out: **standardise when the unit is arbitrary; keep
the raw scale when the unit is something a person could picture.** Your
coefficients are easier to explain either way, and one fewer translation
step is one fewer place to make a mistake.
:::

```{r fit-pois}
fit_pois <- brm(
  transfers ~ yrs_employed + group_size,
  data = managers, family = poisson(),
  prior = c(prior(normal(0, 1),   class = Intercept),
            prior(normal(0, 0.5), class = b)),
  chains = 4, iter = 2000, seed = 112, refresh = 0
)

summary(fit_pois)
```
Two things to notice before interpreting anything.

**There is no `sigma`.** Just as with the Bernoulli model in Chapter 9,
the Poisson has no separate spread parameter — the mean determines the
variance completely. That is a convenience now and an assumption we come
back to at the end of the chapter.

**The coefficients are on the log scale**, so they cannot be read
directly. An estimate of 0.07 for `group_size` does not mean "0.07 more
transfer requests per extra report". As with the log-odds in Chapter 9,
the number has to be translated before it means anything, which is the
next step.

### Interpret: rate ratios

```{r rate-ratios}
fixef(fit_pois) |>
  as_tibble(rownames = "term") |>
  mutate(rate_ratio = exp(Estimate)) |>
  select(term, Estimate, rate_ratio)
```

Exponentiating turns each coefficient into a **rate ratio** — a
multiplier rather than an addition. A rate ratio of 1.08 for
`group_size` means each extra direct report multiplies the expected
number of transfer requests by 1.08: an 8% increase per additional
report, compounding as the team grows. A ratio of 1.0 means the
predictor does nothing; below 1.0 means it reduces the count.

Multiplication rather than addition is the thing to hold on to. It is
why a Poisson model can never predict a negative count: whatever you
multiply a positive rate by, the answer stays positive.

::: callout-note
Chapter 15 revisits this exact variable, `transfers`, and shows why
comparing raw counts across managers with very different `group_size` is
misleading unless you account for it — there, as an offset defining a
*rate*, rather than as an ordinary predictor as we have done here.
:::

### Does it fit?

```{r ppc-pois}
pp_check(fit_pois, type = "bars", ndraws = 100) +
  labs(title = "Posterior predictive check for counts",
       subtitle = "Bars = what the model predicts; points = what actually happened",
       x = "Number of transfer requests", y = "Number of managers")
```

Read it one bar at a time, left to right. Each position on the x-axis is
a possible count — how many managers had 0 transfer requests, how many
had 1, how many had 2. The bar and its interval are what the model
expects; the point is what the data actually contains.

You want every point sitting inside its interval. Two failures are worth
being able to recognise on sight:

- **The point for zero sits above its bar.** The data has more managers
  with no transfer requests than the model can account for. Real HR
  counts do this often, and it has a name — *zero inflation* — usually
  because two different processes are mixed together: managers who could
  have had a transfer request and didn't, and managers for whom it was
  never a possibility at all.
- **The points in the right-hand tail sit above their bars.** The data
  has more high counts than the model expects. That is the failure the
  next section is about.

### When Poisson isn't enough

::: callout-important
## Overdispersion

The Poisson makes a strong assumption: the **mean equals the variance**.
If managers average 2 transfer requests, Poisson insists the variance is
also 2. Real counts are often **overdispersed** — more variable than
that, because something the model hasn't got hold of is pushing some
teams high and others low.

You will see it in the posterior predictive check: the model under-covers
the tails.
:::

You can see the assumption directly, without any modelling at all:

```{r check-dispersion}
managers |>
  summarise(
    mean_transfers = mean(transfers),
    var_transfers  = var(transfers),
    ratio          = var_transfers / mean_transfers      # <1>
  )
```

1.  Poisson requires this ratio to be about 1. Meaningfully above 1 is
    overdispersion; the further above, the worse the fit will be.

The fix is one word in the `family` argument:

```{r fit-nb}
fit_nb <- brm(
  transfers ~ yrs_employed + group_size,
  data = managers, family = negbinomial(),
  prior = c(prior(normal(0, 1),   class = Intercept),
            prior(normal(0, 0.5), class = b)),
  chains = 4, iter = 2000, seed = 112, refresh = 0
)
```

And the two are compared exactly as in Chapter 7:

```{r compare-pois-nb}
loo_compare(loo(fit_pois), loo(fit_nb))
```

Read the first column, `elpd_diff`. The better model sits on the top row
with a difference of 0, and every other row shows how much worse it is,
alongside the standard error of that difference. The rule of thumb: if
the gap is more than about twice its standard error, the difference is
real rather than noise.

Here it is not close. The negative binomial wins by around 45, with a
standard error of about 9 — roughly five standard errors, which is
about as decisive as this comparison gets. The variance-to-mean ratio
warned us, and the model comparison confirms it: **for this data, the
Poisson was the wrong family.**

### So what happens to the answer?

This is the honest consequence, and it is worth following through rather
than quietly moving on. We interpreted the Poisson rate ratios a few
sections ago. Those came from a model we have now shown to be
inadequate. Does the conclusion change?

```{r pois-vs-nb}
bind_rows(
  fixef(fit_pois)["group_size", ] |> as_tibble_row(),
  fixef(fit_nb)["group_size", ]   |> as_tibble_row()
) |>
  mutate(
    model      = c("Poisson", "Negative binomial"),
    rate_ratio = exp(Estimate),                          # <1>
    lower      = exp(Q2.5),
    upper      = exp(Q97.5)
  ) |>
  select(model, rate_ratio, lower, upper)
```

1.  Exponentiating the interval bounds as well, so the whole row is on
    the rate-ratio scale a reader can act on.

The **point estimate barely moves**. Both models agree that a bigger
team means more transfer requests, and by roughly the same multiplier.
What changes is the **interval**, which widens under the negative
binomial.

That is exactly what overdispersion does, and it is why it matters. The
Poisson did not get the effect wrong; it got the *confidence* wrong. It
reported a narrower range than the data supports, because it was
obliged to assume a variability it could see was too small.

::: callout-important
Which is the general shape of this kind of mistake, and worth carrying
beyond this chapter. **Choosing the wrong family will usually mislead
you about precision before it misleads you about direction.**

A model with the wrong family often still points the right way, which is
precisely why the error survives review. Nobody notices a conclusion
that is correct. They notice it much later, when the confident number
turns out to have been less certain than advertised.
:::

::: callout-warning
## When you can skip this

Checking for overdispersion costs one line — the variance-to-mean ratio
above — and you should always spend that line. Fitting the second model
is a different question.

**Skip it when the ratio is close to 1.** Poisson is doing its job and
the negative binomial has nothing to add.

**Skip it when you only need direction.** If the finding is "bigger teams
generate more transfer requests" and nobody is going to act on the exact
multiplier, overdispersion will widen your intervals without changing the
conclusion.

**Don't skip it when the interval is the point.** Overdispersion makes
Poisson intervals too narrow, so a model used to say "we expect between
X and Y grievances next quarter" will sound more certain than it has any
right to be. That is the case where the wrong family produces a
confidently wrong number rather than a slightly imprecise one.
:::

The negative-binomial adds a spare **dispersion** parameter, letting
variance exceed the mean. Compare the two with **LOO** — if
negbinomial wins clearly, you had overdispersion.

------------------------------------------------------------------------

## Part 3 — Outcomes with more than two categories

### The gap in the toolkit

Chapter 9 handled a **yes/no** outcome. Chapter 18 handles an **ordered**
one — a 1-to-5 rating, where the categories have a natural sequence.
Between them sits a shape neither covers: an outcome with several
categories that have **no order at all**.

People Analytics is full of them:

- **Reason for leaving** — better offer, relocation, management,
  retirement, dismissal.
- **Next move** — promoted, moved sideways, left, stayed put.
- **Benefit election** — plan A, plan B, plan C.
- **Offer declined because** — pay, location, timing, counter-offer.

"Better offer" is not more or less than "relocation", so the ordinal
machinery of Chapter 18 does not apply. And you cannot collapse them
into yes/no without throwing away the thing you were asked about.

### Priors need one extra step here

A multinomial model fits a **separate equation for every category**, so
there is no single coefficient called `b` to put a prior on — there is
one set per category. Ask `brms` what it expects:

```{r multinomial-default-prior}
managers <- managers |>
  mutate(performance_group = factor(performance_group))   # <1>

mn_prior <- default_prior(                                # <2>
  performance_group ~ yrs_employed,
  data = managers, family = categorical()
)

mn_prior
```

1.  Unordered labels, exactly as in Chapter 7 — `factor()` says these
    are names, not numbers.
2.  `default_prior()` lists every parameter the model will have, and what
    `brms` would use if you said nothing. Worth running on *any*
    unfamiliar model, not only this one: it is the fastest way to find
    out what you are being asked to have an opinion about.

Look at the `dpar` column. Each row belongs to one category's equation —
`mu` followed by the category name. Rather than typing those names out,
which depend on your data and are easy to get wrong, edit the table
`brms` has just handed you:

```{r fit-multinomial}
mn_prior$prior[mn_prior$class == "b" & mn_prior$coef == ""] <-
  "normal(0, 1)"                                          # <1>
mn_prior$prior[mn_prior$class == "Intercept"] <- "normal(0, 2)"

fit_multi <- brm(
  performance_group ~ yrs_employed,                       # <2>
  data = managers, family = categorical(),
  prior = mn_prior,
  chains = 4, iter = 2000, seed = 12, refresh = 0
)

summary(fit_multi)
```

1.  One statement setting every category's slopes at once. The
    `coef == ""` condition picks the rows that stand for *all*
    coefficients in an equation rather than one named coefficient.
2.  Does tenure predict which performance group a manager falls into?


### Reading it: one model per category, minus one

The output is longer than you might expect, and the reason is the whole
idea of the model.

A multinomial model picks one category as the **reference** — the same
device as the dummy variables in Chapter 7 — and then fits a *separate
set of coefficients for every other category*, each describing that
category **against the reference**. With three groups you get two sets;
with five you get four.

So a coefficient here never means "the effect of tenure on performance
group". It means "the effect of tenure on the log-odds of being in *this*
group **rather than the reference group**". Change the reference and
every number changes, while the model's predictions stay identical.

::: callout-important
Which makes the coefficients close to unreadable, and points at the same
answer Chapter 9 reached with odds ratios: **do not present these
numbers.** Present predicted probabilities instead — for each level of
your predictor, the probability of landing in each category. Those add up
to 1, they need no reference group, and a stakeholder can read them
directly.

`add_epred_draws()` handles this, returning one row per category per
draw.
:::

::: callout-warning
## When you can skip this

**Skip it when one category is the question.** If leadership only cares
about who left for a better offer, a yes/no model for that single
category is simpler, easier to explain, and answers what was asked.

**Skip it when the categories really are ordered.** Chapter 18's
proportional-odds model uses that order and gets more out of the same
data. Multinomial throws the ordering away.

**Don't skip it when the categories compete.** The reason multinomial
exists is that the probabilities must sum to 1 — if more people leave for
pay, fewer leave for everything else, by arithmetic. Fitting four
separate yes/no models ignores that constraint and can produce
probabilities that add to more than 1, which is difficult to defend in a
room.
:::

------------------------------------------------------------------------

## Part 4 — When observations come in more than one kind of group

### The structure Chapter 8 did not cover

Every multilevel model so far has had **one** grouping factor: managers
within cities, employees within offices. Each row belonged to exactly one
group, and the groups nested neatly.

A lot of People Analytics data does not look like that:

- **Pulse surveys** — one row per employee per wave. Rows group by
  *employee* and by *wave*, and neither is inside the other.
- **360 feedback** — one row per rating. Rows group by *ratee* and by
  *rater*.
- **Training evaluation** — one row per person per course. Rows group by
  *participant* and by *course*.

These are **crossed** rather than nested: an employee appears in every
wave, and every wave contains every employee. Chapter 8's `(1 | city)`
assumes each row sits in one place in one hierarchy, and that assumption
is simply false here.

### The formula, and why ignoring it is dangerous

The fix is smaller than the problem sounds — add a second term:

```{r}
#| eval: false
# One row per participant per course
rating ~ course_length + (1 | participant) + (1 | course)
```

Each participant gets their own intercept, each course gets its own, and
the model separates the two.

::: callout-important
The reason this matters is not precision, it is **honesty about how much
data you have**.

Five hundred ratings from fifty participants is not five hundred
independent pieces of evidence. Some people rate everything highly and
some rate everything harshly, so their ten ratings are ten variations on
one opinion. Ignore the participant grouping and the model counts them
all as independent, and reports an interval far narrower than the
evidence supports.

This is the same failure as Chapter 8's league table, arriving from a
different direction: treating repeated measurements of the same thing as
though they were measurements of different things.
:::

The `learning` dataset in `peopleanalyticsdata` — 4,974 feedback
responses from 326 participants across a range of programmes — is built
for exactly this and makes a good exercise. Chapter 19 then adds the
other half of the problem: those ratings are measured with error too.

------------------------------------------------------------------------

### Your turn {.unnumbered}

Refit the count model with `family = negbinomial()` and compare it to
the Poisson version with LOO. Is there evidence of overdispersion?

```{r your-turn-pois}
# Your code here

```

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Two common situations live here. If an effect plausibly **differs
across your groups** (offices, functions, cohorts), varying slopes let
you estimate and report that variation instead of averaging it away —
often the more actionable finding. If your outcome is a **count**,
Poisson or negative-binomial regression is the correct tool — using
ordinary regression on counts is a common and avoidable mistake.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  `(1 + x | group)` lets both **intercept and slope** vary by group.
2.  `sd(x)` at the group level measures how much the **effect**
    varies — itself a finding.
3.  **Poisson** regression (`family = poisson()`) models counts;
    interpret `exp(coef)` as **rate ratios**.
4.  Overdispersed counts call for **negative-binomial**.
:::

## Next chapter {.unnumbered}

**Model building as craft: the decisions software won't make for you** —
that completes the technical toolkit, and the next chapter is about
everything the toolkit does not decide for you. Which variables can be
changed by anybody in your organisation, what shape a relationship should
take, whether one effect depends on another, and what you would do
differently depending on the answer. None of it produces an error
message, and all of it changes what the model means.

Part V then turns to problems built specifically for the People Analytics
community: **empirical Bayes**, **Bayesian shrinkage** for ranking
managers and teams fairly, **hierarchical models for pay and variance**,
**survival analysis** for the employee lifecycle, **Likert and survey
data**, **measurement error**, **Bayesian A/B testing**, **causal
inference**, and a closing chapter on **building priors with your
stakeholders**.
