# Empirical Bayes: Estimating Rates You Don't Have Enough Data For

## Welcome {.unnumbered}

Every People Analytics team eventually hits the same wall: you want to
compare a rate across a set of groups — offices, cohorts, teams,
recruiting channels — and some of those groups are tiny. A city office
with four managers, a recruiting channel that hired six people last
quarter, a team of three. The raw rate for a small group swings wildly
on almost no evidence, and yet it's often the *extreme* small groups
that get the attention: the "worst" office, the "best" channel.

This chapter introduces **empirical Bayes** — a fast, elegant way to
stabilise exactly these estimates, without needing to fit a full
Bayesian model. It's the technique behind a genuinely famous piece of
applied statistics (estimating true batting averages in baseball from a
single season), and it maps almost exactly onto People Analytics
problems.

::: callout-important
The core idea in one sentence: **use the data itself to work out how
much you should trust any one group's raw number, then pull the
noisiest estimates back toward the average by exactly that much.**
:::

## What you'll be able to do by the end {.unnumbered}

1.  Explain why raw rates for small groups are systematically misleading
2.  Estimate a prior distribution *from your own data* (the "empirical"
    in empirical Bayes)
3.  Combine that prior with each group's own data to get a stabilised
    estimate
4.  Know when to reach for empirical Bayes versus a full multilevel
    model (Chapter 15)
5.  Use an **external benchmark** as a prior when you don't have enough
    of your own groups to estimate one reliably
6.  Build a prior from **your own history**, and decide how much of that
    history should still count

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

data("managers", package = "peopleanalyticsdata")

# Guard against missing values before grouping and estimating rates
# (Chapter 1 found gaps in a similar teaching dataset).
managers <- managers |> drop_na(concern_flag, city, high_hours_flag)
```

We'll use the `managers` dataset: 571 managers, one row each, with a
`concern_flag` (whether the manager has been the subject of a
complaint), their `city`, and a handful of other characteristics. This
data — and every other dataset used in Part V of this book — comes from
Keith McNulty's `peopleanalyticsdata` package; see the
[Welcome](../index.qmd) chapter for the full credit.

------------------------------------------------------------------------

## The problem

### A league table of cities

Suppose you want to know: in which cities are managers more likely to
have a complaint filed against them? A natural first move — group by
city, compute the raw rate:

```{r raw-rates}
city_rates <- managers |>
  mutate(concern = if_else(concern_flag == "Y", 1, 0)) |>
  group_by(city) |>
  summarise(
    n_managers = n(),
    n_concerns = sum(concern),
    raw_rate   = n_concerns / n_managers
  ) |>
  arrange(desc(raw_rate))

city_rates
```

```{r raw-rates-plot}
ggplot(city_rates, aes(x = n_managers, y = raw_rate)) +
  geom_point(size = 2.5, colour = "#122a52") +
  geom_hline(yintercept = mean(city_rates$n_concerns) / mean(city_rates$n_managers),
             linetype = "dashed", colour = "#d32f2f") +
  labs(
    title = "Raw complaint rate by city",
    subtitle = "Dashed line = overall rate. Small offices swing to the extremes.",
    x = "Number of managers in the city", y = "Raw complaint rate"
  )
```

### Why this table is dangerous to publish as-is

Look at where the highest and lowest rates sit: overwhelmingly among
the cities with the *fewest* managers. That's not a finding about those
cities — it's arithmetic. With four managers, the only possible rates
are 0%, 25%, 50%, 75% or 100%. A single complaint moves a small city
from "fine" to "an outlier" instantly. A city with sixty managers
barely moves at all for the same one complaint.

::: callout-important
If you rank cities by raw rate and act on the top and bottom of that
list, you are — mostly — ranking cities by **how little data you
have on them**, not by anything real about how they're managed.
:::

This is exactly the same problem the multilevel-models chapter raised
for guest satisfaction and, in Chapter 15, for individual managers. It
shows up everywhere in People Analytics because organisational units
are almost never equally sized.

------------------------------------------------------------------------

## The empirical Bayes idea

### Start from Bayes' theorem, as before

For any one city, if we treat each manager as a coin flip with an
unknown "true" concern rate $\theta$, Bayes' theorem says:

$$
\text{posterior rate} \propto \text{prior} \times \text{likelihood (the city's own data)}
$$

The Beta-Binomial model from Chapter 4 gives us a clean way to do this:
if the prior is $\text{Beta}(\alpha, \beta)$ and a city has $n$ managers
with $k$ concerns, the posterior is $\text{Beta}(\alpha + k, \beta + n -
k)$, and its mean is:

$$
\hat\theta = \frac{\alpha + k}{\alpha + \beta + n}
$$

Read that formula carefully: $\alpha$ and $\beta$ act like **extra,
imaginary managers** added to every city — $\alpha$ imaginary concerns
and $\beta$ imaginary clean records. A small city's real data gets
outweighed by these imaginary managers; a large city's real data
swamps them. That's the shrinkage mechanism, in one formula.

### The empirical Bayes shortcut

In Chapter 4 you had to *choose* a prior. Empirical Bayes instead asks:
**what single Beta distribution best describes the spread of raw rates
we actually observed across all our cities?** We estimate $\alpha$ and
$\beta$ directly from the vector of city-level rates, using a simple
method-of-moments fit:

```{r fit-eb-prior}
mu    <- mean(city_rates$raw_rate)
var_r <- var(city_rates$raw_rate)

# Method-of-moments Beta parameters
alpha0 <- mu * ((mu * (1 - mu) / var_r) - 1)
beta0  <- (1 - mu) * ((mu * (1 - mu) / var_r) - 1)

tibble(alpha0, beta0)
```

```{r prior-plot}
tibble(theta = seq(0, 0.6, length.out = 400)) |>
  mutate(density = dbeta(theta, alpha0, beta0)) |>
  ggplot(aes(theta, density)) +
  geom_line(colour = "#122a52", linewidth = 1) +
  geom_vline(xintercept = mu, linetype = "dashed", colour = "#d32f2f") +
  labs(
    title = "The empirically estimated prior",
    subtitle = "Estimated from the spread of raw city rates itself",
    x = expression(theta~"(true concern rate)"), y = "Density"
  )
```

::: callout-note
## A different lens: frequentist statistics has a version of this too

If your background is frequentist statistics, this might feel like it
comes out of nowhere — but it doesn't. In the 1960s, Charles Stein
proved something that shocked the statistics world: if you want to
estimate several unrelated quantities at once (say, the true rate for
each of our cities), you get *better overall accuracy* by pulling every
estimate a little toward the grand average — even when the quantities
have nothing to do with each other. This became known as **Stein's
paradox**, and the **James-Stein estimator** is its practical form.
Empirical Bayes is the natural route to the same destination from a
Bayesian starting point: same shrinkage, same reasoning about small
samples, arrived at via a prior instead of a minimax argument. It's not
a fringe Bayesian trick — it's a well-established result that classical
and Bayesian statistics converge on independently.
:::

::: callout-tip
## For the ML/DS crowd

If you've ever tuned an L1 or L2 regularisation strength via
cross-validation, you've already done something very close to this by
hand. A ridge-regression penalty is mathematically equivalent to
putting a Normal prior on your coefficients; the penalty strength is
equivalent to that prior's precision. Cross-validating the penalty is
choosing the prior's strength empirically by trial and error over held-out
data. Empirical Bayes just estimates that strength directly from the
data's own spread, in closed form, without a validation loop. Same
instinct — "let the data tell you how much to trust extreme values" —
different mechanics.

(One disambiguation while we're here: **empirical Bayes has nothing to
do with a "naive Bayes classifier."** They share a name and a theorem,
not much else — naive Bayes is a classification algorithm that assumes
feature independence; empirical Bayes is a way of estimating priors
from data. Easy to conflate on a first pass.)
:::

::: callout-note
## Why this is a slightly uncomfortable idea, and why it's fine anyway

Using the data to build the prior *and* then using the same data again
in the likelihood sounds circular — full Bayesians will point this out.
In practice it works extremely well as an approximation to a full
hierarchical model (Chapter 15 fits the "proper" version and you'll see
the answers are very close), and it has one big advantage: **no MCMC,
answers in milliseconds.** Treat it as a fast, practical tool — not a
substitute for a full model when the stakes are high enough to justify
one.
:::

### Apply it to every city

```{r apply-eb}
city_eb <- city_rates |>
  mutate(
    eb_rate = (alpha0 + n_concerns) / (alpha0 + beta0 + n_managers)
  ) |>
  arrange(desc(eb_rate))

city_eb
```

```{r eb-shrinkage-plot}
ggplot(city_eb, aes(x = n_managers)) +
  geom_hline(yintercept = mu, linetype = "dashed", colour = "#d32f2f") +
  geom_segment(aes(xend = n_managers, y = raw_rate, yend = eb_rate),
               colour = "grey80") +
  geom_point(aes(y = raw_rate), colour = "grey70", size = 2.5) +
  geom_point(aes(y = eb_rate), colour = "#122a52", size = 2.5) +
  scale_x_log10() +
  labs(
    title = "Empirical Bayes pulls small cities toward the overall rate",
    subtitle = "Grey = raw rate, navy = empirical Bayes estimate, red dashed = overall rate",
    x = "Number of managers (log scale)", y = "Concern rate"
  )
```

Read this the same way you read the shrinkage plot in Chapter 8: small
offices move a long way, large ones barely move, and nobody had to pick
a threshold for "too small to trust."

::: callout-caution
## Try it yourself

Recompute this with `high_hours_flag` instead of `concern_flag`. Does
the ranking of cities change much once you apply empirical Bayes? Which
cities were "extreme" on the raw rate but unremarkable once shrunk?
:::

------------------------------------------------------------------------

## Using external benchmarks as priors

Empirical Bayes needs *enough groups* to estimate a sensible prior. If
your organisation has five offices, estimating a Beta distribution from
five raw rates is itself unreliable — you'd be building a shaky prior
out of shaky ingredients.

A common, entirely legitimate alternative: build your prior from an
**external benchmark** instead of your own groups.

::: callout-note
## A worked (illustrative) example

Suppose your industry body publishes an average annual voluntary
attrition rate of 14%, based on a large cross-industry survey, and you
have no strong reason to think your organisation differs systematically
from that average *before* looking at your own data. You could encode
that as a prior:

```{r benchmark-prior}
# A Beta prior centred on 14%, with moderate confidence
# (equivalent to having already observed ~50 "imaginary" employees
#  at that rate before seeing any of your own data)
benchmark_rate <- 0.14
prior_strength <- 50

alpha_bench <- benchmark_rate * prior_strength
beta_bench  <- (1 - benchmark_rate) * prior_strength

tibble(alpha_bench, beta_bench)
```

Same habit as everywhere else in this book — look at it before using it:

```{r plot-benchmark-prior}
tibble(theta = seq(0, 0.5, length.out = 400)) |>
  mutate(density = dbeta(theta, alpha_bench, beta_bench)) |>
  ggplot(aes(theta, density)) +
  geom_area(fill = "#122a52", alpha = 0.5) +
  geom_vline(xintercept = benchmark_rate, linetype = "dashed", colour = "#d32f2f") +
  labs(
    title = "The benchmark-derived prior: Beta(7, 43)",
    subtitle = "Centred on the industry's 14% rate, moderately confident",
    x = "Attrition rate", y = "Prior density"
  )
```

Now combine it with, say, a 30-person team that has had 2 departures
this year:

```{r benchmark-posterior}
n_team <- 30
k_team <- 2

posterior_mean <- (alpha_bench + k_team) / (alpha_bench + beta_bench + n_team)
raw_rate_team  <- k_team / n_team

tibble(raw_rate_team, posterior_mean, benchmark_rate)
```
:::

Notice what happened: the raw rate (6.7%) looks dramatically better
than the industry benchmark, but with only 30 people that's thin
evidence — the posterior estimate sits between the raw rate and the
benchmark, appropriately hedging until more of your own data arrives.

::: callout-important
## The knob that matters: `prior_strength`

`prior_strength` (the $\alpha + \beta$ total) is doing all the work of
saying *"how much should the benchmark count for, relative to a single
employee's worth of your own data?"* Setting it too high means your own
data can never move the estimate; too low and the benchmark barely
matters. There's no universal right answer — state your choice
explicitly and, ideally, show how sensitive the conclusion is to it (a
prior sensitivity check, exactly as in Chapter 5).
:::

------------------------------------------------------------------------

## Priors from your own past

So far this chapter has built a prior two ways: from the spread across
your own groups, and from an external benchmark. There's a third
source sitting right there in your HRIS, and it's usually the most
persuasive one in a meeting — **your own history**. You have last
year's promotion rates, the last three years of attrition, eight
quarters of offer acceptances. Chapter 4 showed the mechanism already:
today's posterior is tomorrow's prior, so history rolls forward on its
own.

It also came with a warning, which is what this section is about.
Rolling history forward assumes the process that generated it is the
same process running today. Organisations violate that assumption
constantly — a reorganisation, a new promotion policy, a change of
leadership, a hiring freeze. When that happens, evidence from three
years ago describes a process that no longer exists, and the more of
it you have the more confidently wrong it makes you.

So the question isn't whether to use your history. It's **how much of
it still counts.**

### First ask: did it change gradually, or all at once?

This is the whole decision, and it's not a statistical one. You answer
it by knowing your organisation.

Most of the changes People Analytics deals with are **step changes**.
The reorganisation happened in March. The new promotion framework
launched in Q2. The bands were rebuilt on the first of January. There
is a date, people remember it, and the process genuinely is different
either side of it.

::: callout-important
## When there's a date, cut — don't decay

If you can name the month it changed, the honest move is to **use only
the data since then**, and say so in one sentence: *"this uses
promotions from April onward, because the framework changed in
March."*

That's defensible to anyone in the room, it needs no method, and it
survives being challenged. A decay curve applied to a step change is
worse on both counts — it still lets pre-change data influence the
answer, and it obscures a clean editorial decision behind machinery.

The instinct to reach for the sophisticated tool is strong here, and
it's usually wrong. The crude one is more defensible.
:::

The cost of cutting is obvious: you throw data away, and if the change
was recent you may have very little left. That's not a reason to keep
the old data — it's a reason to be honest that you're now in a
small-sample situation, which is what the rest of this chapter is for.

**Gradual drift** is the other case, and it's rarer than it feels.
Nothing changed on a particular Tuesday; the organisation has simply
been getting steadily younger, or the labour market has been slowly
loosening, and evidence from four years ago is less relevant than
evidence from last quarter without there being a line to draw. That's
where decay earns its place.

### Decay weighting, in the same closed form

The mechanic is a small extension of what you've already done. Instead
of counting every historical observation once, count each one by a
weight that falls off with age:

$$
w = \tfrac{1}{2}^{\,\text{age} / h}
$$

where $h$ is a **half-life** — the age at which an observation counts
for half of a fresh one. Then build the prior from the weighted counts
rather than the raw ones.

::: callout-note
## A worked (illustrative) example

Eight quarters of promotion history, oldest first — the kind of table
you'd assemble from an HRIS extract:

```{r decay-history}
history <- tibble(
  quarter    = 1:8,
  eligible   = c(52, 48, 61, 55, 44, 58, 50, 47),
  promoted   = c(19, 17, 21, 17, 12, 15, 12, 11)
) |>
  mutate(quarters_ago = max(quarter) - quarter)

history
```

Pick a half-life and compute the weights. With a four-quarter
half-life, the oldest quarter here counts for about a third of the
most recent one:

```{r decay-weights}
half_life <- 4                                                    # <1>

history <- history |>
  mutate(weight = 0.5 ^ (quarters_ago / half_life))

history |> select(quarters_ago, eligible, promoted, weight)
```

1.  In quarters, matching the unit of `quarters_ago`. This is the one
    genuine judgement call in the section — see the callout below.

```{r decay-weight-plot}
#| code-fold: true
ggplot(history, aes(x = quarters_ago, y = weight)) +
  geom_col(fill = "#122a52", alpha = 0.45, width = 0.7) +
  scale_x_reverse() +
  labs(
    title = "How much each quarter still counts",
    subtitle = "Four-quarter half-life: a year ago counts for half, two years ago a quarter",
    x = "Quarters ago", y = "Weight"
  )
```

Now the prior. The Beta-Binomial update from Chapter 4 said *add your
successes to* $\alpha$ *and your failures to* $\beta$. Here we add the
**weighted** successes and failures instead:

```{r decay-prior}
alpha_hist <- sum(history$weight * history$promoted)
beta_hist  <- sum(history$weight * (history$eligible - history$promoted))

n_raw <- sum(history$eligible)
n_eff <- alpha_hist + beta_hist                                   # <1>

tibble(alpha_hist, beta_hist, n_raw, n_eff)
```

1.  $\alpha + \beta$ is the prior's strength in units of people —
    the same knob as `prior_strength` in the benchmark section, except
    here you didn't choose it directly, the decay chose it for you.

That last number is the one to report. You have `r sum(history$eligible)`
historical observations and roughly `r round(n_eff)` observations'
*worth* of evidence, because you've decided the old ones count for
less. Both numbers are true; only the second one is being used.

```{r decay-prior-plot}
theta <- seq(0, 0.7, length.out = 400)

alpha_flat <- sum(history$promoted)
beta_flat  <- sum(history$eligible - history$promoted)

tibble(
  theta,
  `all history, counted equally` = dbeta(theta, alpha_flat, beta_flat),
  `discounted by age`            = dbeta(theta, alpha_hist, beta_hist)
) |>
  pivot_longer(-theta, names_to = "prior", values_to = "density") |>
  ggplot(aes(theta, density, colour = prior)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("grey70", "#8fabd0")) +
  labs(
    title = "Discounting the past widens the prior — and moves it",
    subtitle = "Promotion rate has been drifting down; the discounted prior follows it",
    x = expression(theta~"(true promotion rate)"), y = "Prior density",
    colour = NULL
  )
```
:::

Two things happen at once there, and both are the point. The
discounted prior is **wider**, because you're honestly claiming less
evidence than you physically hold. And it has **moved**, tracking the
recent quarters rather than averaging across a drift. A prior that
counted all eight quarters equally would be narrower *and* centred in
the wrong place — the worst combination available.

::: callout-important
## Choosing the half-life

There is no data-driven answer, and you should be suspicious of anyone
who offers you one. The half-life is a statement about your
organisation: *how long does it take for the way we promote people to
stop resembling the way we promoted them then?*

Two rules make it defensible. **State it, in plain words** — "evidence
from two years ago counts for a quarter of evidence from this quarter"
is a sentence a stakeholder can push back on, and $h = 4$ isn't. And
**show what it costs you**: refit with a half-life half as long and
twice as long, and report whether the conclusion changes. If it does,
that sensitivity *is* the finding, and the honest answer is that you
don't have enough recent data to settle the question.

Same discipline as the prior sensitivity check in Chapter 5. Same
reason.
:::

If you're modelling rather than counting rates, the same idea arrives
as a `weights =` argument in `brm()` — one weight per row, computed
exactly as above. That's the one place in this book where the `brm()`
skeleton gains an argument, and it's worth the exception: that single
argument is where "how much should the past count?" lives.

### Names worth knowing, and where to stop

Everything above is deliberately the cheap version. There is a
substantial literature underneath it, and the useful thing at this
point isn't a worked example of each — it's knowing what to search
for when your problem outgrows a half-life:

-   **Power priors** — the formal version of what we just did: raise
    the historical likelihood to a power between 0 and 1 to discount
    it. The literature covers how to choose that power, including
    letting the model estimate it.
-   **Change-point models** — for when you're fairly sure something
    shifted but can't name the date. The model finds it, and gives you
    a posterior over *when* rather than making you pick.
-   **State-space and dynamic linear models** — θ becomes something
    that moves, with its own model for how fast it's allowed to move.
    The proper answer to genuine drift, and a different way of working
    rather than an extra argument.
-   **Time-varying effects in `brms`** — `s(time)` or `gp(time)` in a
    model formula, letting a rate or a coefficient bend smoothly over
    time without leaving tools you already know.

::: callout-note
The reason to name these without expanding them is that the hardest
part of the problem is usually knowing that a solution exists and what
it's called. If you've read this far you can recognise the situation —
your history is real but stale, and a single number for "the rate" is
hiding a moving target. That recognition is most of the work. The
search terms are the rest.
:::

------------------------------------------------------------------------

### The workflow steps this chapter skips, and why

Chapter 10's workflow assumes you fitted a model. Most of this chapter
does not, and it is worth being explicit rather than leaving the gap
looking like carelessness.

Empirical Bayes here is **closed-form arithmetic**, not sampling. There
are no chains to diagnose (step 7) because nothing was sampled, and no
posterior predictive check (step 8) in the usual sense because there is
no generative model to simulate from — the Beta prior was estimated from
the group rates by method of moments, and the update is a formula.

What replaces them is the comparison in the next section: fit the full
multilevel model and check the two answers agree. **That is the check.**
If the closed-form shortcut and the properly-fitted model disagree, the
distributional assumption behind the shortcut is wrong, and you have
found out in the only way available.

The steps that do still apply, and are not optional: know what you are
claiming (step 2 — these are descriptions of groups, not effects of
being in one), and temper the result (step 10 — a shrunk rate is still
built on a raw count that may itself be noisy).

------------------------------------------------------------------------

## Empirical Bayes vs. a full multilevel model

| | **Empirical Bayes (this chapter)** | **Multilevel model (Ch. 8, 14)** |
|---|---|---|
| Speed | Instant, closed-form | Seconds to minutes (MCMC) |
| Uncertainty | Point estimate only, by default | Full posterior for every group |
| Assumes | A single, simple distributional shape for the prior | Whatever you specify — more flexible |
| Good for | A quick first look, dashboards, many repeated similar analyses | The version you'd defend to a sceptical stakeholder or publish |

Neither is "more correct" in every situation. A sensible default:
prototype with empirical Bayes because it's instant, then confirm the
conclusion holds with a proper multilevel model before you act on it —
which is exactly what Chapter 15 does, on a closely related problem.

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Almost every People Analytics deliverable involves ranking or comparing
groups of unequal size: offices, recruiting channels, onboarding
cohorts, manager populations. Empirical Bayes is the fastest tool in
this book for stopping a stakeholder from over-reacting to the office
with four people and one bad month. It costs you one extra line of R
and no modelling framework — there's rarely a reason *not* to check a
raw rate this way before it goes in a deck.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Raw rates for small groups are noisy in a *predictable* direction —
    they produce the most extreme values.
2.  Empirical Bayes estimates the prior **from the spread of your own
    groups' rates**, then shrinks each group toward it in closed form.
3.  The shrinkage amount is set automatically by how much groups
    genuinely differ versus how much is just noise.
4.  When you don't have enough groups of your own, an **external
    benchmark** can play the same role as the prior — with an explicit,
    checkable choice of how strongly to weight it.
5.  **Your own history** is the third source of a prior, and the only
    one that needs an expiry date. If you can name the month the
    process changed, cut the data there and say so. If it drifted
    instead, discount older observations by a stated half-life and
    report the effective sample size you're left with.
6.  Empirical Bayes is a fast approximation to the full multilevel
    model in the next chapter — prototype with one, confirm with the
    other.
:::

## Next chapter {.unnumbered}

**Bayesian shrinkage: ranking managers and teams fairly** — we take this
exact idea and apply it properly, with a full Bayesian model, to the
question that probably brought you to this chapter: which managers are
genuinely better or worse, once you account for team size?
