# Hierarchical Models for Pay and Variance

## Welcome back {.unnumbered}

"How much of our pay variation is explained by level, versus function,
versus location, versus... nothing structural at all?" is one of the
most common questions a People Analytics team gets asked, usually
attached to a pay equity review or a compensation redesign. It's also a
question that a simple average-by-group comparison answers badly, for
the same reason Chapter 15 warned you about: small groups produce noisy
averages, and a table of "average pay by department" mixes real
structural variation with pure noise.

This chapter extends the partial-pooling machinery from Chapters 8 and 15 to a new job: not just producing fairer individual estimates, but
**decomposing variance itself** — answering "how much" a grouping
factor matters, as a finding in its own right.

### The request, as it actually arrives {.unnumbered}

The Reward Director has a compensation redesign starting in the autumn
and a table of average pay by department that she doesn't trust:

> *"Before we spend six months rebuilding the grade structure, I need to
> know whether the structure is doing anything. Is pay here mostly
> determined by what job family you're in — or is it mostly down to who
> negotiated well? Because those are different projects."*

Notice what she is not asking for. Not a list of departments ranked by
pay, which is what the table she already has provides. She is asking
**how much** the grouping explains — a single quantity, with enough
honesty about its precision that she can decide whether to commit a
six-month programme to it. That quantity has a name and this chapter
computes it.

::: callout-note
The dataset in this chapter is public labour-market data rather than her
comp file, but the shape of the question and the shape of the answer are
identical, and the last section covers what changes when you point the
same model at a real internal dataset.
:::

## What you'll be able to do by the end {.unnumbered}

1.  Distinguish variation **between** groups from variation **within**
    groups
2.  Fit a variance-components model and compute an **intraclass
    correlation (ICC)**
3.  Interpret "40% of the variance sits at the discipline level" in
    plain language
4.  Extend the idea to multiple, crossed grouping factors (level,
    function, geography) the way a real compensation dataset would need
5.  Know the limits of this kind of analysis for actual pay equity work

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

data("graduates", package = "peopleanalyticsdata")

# Guard against missing values before modelling (Chapter 1 found gaps
# in a similar teaching dataset).
graduates <- graduates |> drop_na(Median_salary, Discipline, Unemployment_rate)
```

`graduates` has US labour-market data for 173 college majors, grouped
into broad disciplines (`Discipline`): `Major`, `Total` (graduates of
working age), `Unemployment_rate`, and `Median_salary`. It's not
People Analytics data in the internal-company sense — it's real
labour-market data (sourced via FiveThirtyEight) — but it gives us
exactly the structure we need: many majors, nested within a smaller
number of disciplines, with genuinely varying pay.

------------------------------------------------------------------------

## The question: where does pay variation actually sit?

### A raw look, discipline by discipline

```{r raw-look}
disc_summary <- graduates |>
  group_by(Discipline) |>
  summarise(
    n_majors = n(),
    mean_salary = mean(Median_salary),
    sd_salary = sd(Median_salary)
  ) |>
  arrange(desc(mean_salary))

disc_summary
```

```{r raw-plot}
graduates |>
  mutate(Discipline = fct_reorder(Discipline, Median_salary, .fun = median)) |>
  ggplot(aes(x = Median_salary, y = Discipline)) +
  geom_boxplot(fill = navy_light, colour = navy, alpha = 0.5,
               outlier.shape = NA) +                                  # <1>
  geom_jitter(height = 0.15, alpha = 0.45, size = 1.4, colour = navy) +
  geom_vline(xintercept = mean(graduates$Median_salary),
             linetype = "dashed", colour = red) +                     # <2>
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Median salary by major, grouped by discipline",
    subtitle = "Each point is one major; boxes summarise its discipline. Red dashed = the grand mean.",
    x = "Median salary (USD)", y = NULL
  )
```

1.  `outlier.shape = NA` because every major is already drawn as a
    point — leaving it on would plot the extreme ones twice.
2.  The convention used throughout this book: navy carries the data, red
    marks a reference value. Two colours doing two different jobs, rather
    than two data series competing.

There's an obvious question sitting in this plot: some of the spread is
*between* disciplines (Engineering majors, as a group, sit higher than
Education majors) and some is *within* a discipline (majors inside
Engineering still differ from each other). **How much of the total
variance is which?**

------------------------------------------------------------------------

## Fitting a variance-components model

This is a job for the same varying-intercepts structure as Chapter 8 —
but this time the quantity we care about most isn't each group's
shrunk estimate, it's the **two variance components themselves**:

New scale, new numbers — worth a look before fitting, as always:

```{r plot-prior-intercept-var}
tibble(x = seq(0, 100000, length.out = 400)) |>
  mutate(density = dnorm(x, 45000, 15000)) |>
  ggplot(aes(x, density)) +
  geom_area(fill = navy, alpha = 0.5) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Intercept: Normal(45000, 15000)",
       subtitle = "Where we think the grand mean sits, before seeing any data",
       x = "Grand mean median salary", y = "Prior density")
```

```{r plot-prior-sd-var}
tibble(x = seq(0, 50000, length.out = 400)) |>
  mutate(density = dexp(x, 0.0001)) |>
  ggplot(aes(x, density)) +
  geom_area(fill = navy, alpha = 0.5) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Variance components: Exponential(0.0001)",
       subtitle = "The same prior is used for both sd(Discipline) and sigma — a mean of $10,000 either way",
       x = "Standard deviation", y = "Prior density")
```

### And what those two together imply about a salary

Plotting each prior on its own says what we believe about each
*parameter*. The prior predictive check asks the more useful question:
put them together, and what salaries does this model think are possible
before it has seen a single major? Chapter 10's step 5 asks for both, and
this is the half that catches mistakes:

```{r prior-predictive-var}
#| code-fold: true
n_sim <- 4000

tibble(
  grand_mean = rnorm(n_sim, 45000, 15000),
  sd_disc    = rexp(n_sim, 0.0001),
  sigma      = rexp(n_sim, 0.0001),
  simulated  = rnorm(n_sim,
                     mean = rnorm(n_sim, grand_mean, sd_disc),        # <1>
                     sd   = sigma)
) |>
  ggplot(aes(simulated)) +
  geom_histogram(bins = 60, fill = navy, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "What these priors think a major's median salary could be",
    subtitle = "Red dashed = zero. A meaningful share of the prior sits below it.",
    x = "Simulated median salary", y = "Simulated draws"
  )
```

1.  The two-stage structure of the model itself, simulated forwards: a
    discipline's mean is drawn around the grand mean, and a major's
    salary is drawn around its discipline's mean.

::: callout-important
## The prior thinks negative salaries are plausible, and they aren't

A good chunk of that distribution is below zero, which is impossible.
This is the standard failure of a Normal prior on a strictly positive
quantity, and it is worth flagging rather than hiding for the same
reason Chapter 22 flags it on the engagement scale: with 173 majors the
likelihood overwhelms the prior immediately, so it changes nothing here.
It would matter with nine data points instead of 173.

The two honest fixes, in increasing order of effort: tighten
`exponential(0.0001)` (a mean of $10,000 for a *between-discipline*
standard deviation is generous — the disciplines plainly don't differ by
that much), or model `log(salary)` instead, which cannot go negative by
construction and is what the crossed-factor model later in this chapter
recommends anyway.

The habit is the point. You cannot see this problem by looking at the two
parameter priors separately — each one is entirely reasonable on its own.
:::

```{r fit-variance}
fit_var <- brm(
  Median_salary ~ 1 + (1 | Discipline),
  data = graduates, family = gaussian(),
  prior = c(prior(normal(45000, 15000), class = Intercept),
            prior(exponential(0.0001), class = sd),
            prior(exponential(0.0001), class = sigma)),
  chains = 4, iter = 2000, seed = 14, refresh = 0
)

summary(fit_var)
```

### Reading the output, line by line

There is less here than in Chapter 12's varying-slopes output, and one
block matters far more than the others:

**Multilevel Hyperparameters — the number this chapter is about.**
`sd(Intercept)` is the **between-discipline** standard deviation: how far
a typical discipline's mean salary sits from the grand mean. This is the
first of the two variance components, and it is a finding in its own
right rather than a nuisance parameter. Note that its credible interval
does not include zero — with sixteen disciplines it comfortably could have,
and that it doesn't is what licenses the rest of the chapter.

**Further Distributional Parameters.** `sigma` is the **within-discipline**
standard deviation: how far a typical major sits from its own
discipline's mean. This is the second component. Read the two side by
side — the ratio between them is, in effect, the answer, and the next
section makes that formal.

**Regression Coefficients.** Just `Intercept` here: the grand mean median
salary across all disciplines. Useful for a sanity check against the raw
plot above, and not much else.

**The two columns on the right, on every line.** `Rhat` should read
`1.00` and `Bulk_ESS` / `Tail_ESS` should be in the hundreds at minimum —
Chapter 10's step 7 and Chapter 11's subject. A model this simple almost
never has trouble, which is exactly why it's worth glancing at now: you
want the habit established before you meet a model that does.

::: callout-tip
Both variance components come with a full credible interval, and that is
the part a classical variance-components analysis struggles to give you.
With only sixteen disciplines, `sd(Intercept)` is genuinely imprecisely
known — and a method that reported it as a single number would hide that.
:::

### Does the model fit?

Before reading anything off it, the posterior predictive check from
Chapter 10's step 7 — does a model this simple actually reproduce the
shape of the salary data?

```{r ppc-var}
pp_check(fit_var, ndraws = 50) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Posterior predictive check: simulated vs actual salaries",
       subtitle = "Thin lines = datasets the model would generate; thick line = the real data",
       x = "Median salary", y = "Density")
```

::: callout-note
## What to look for, and what this one shows

The thin lines are salary distributions the fitted model considers
plausible; the thick line is the data. You want the thick line to look
like one more draw from the same family.

Two things to expect here. The model is symmetric — a Normal around each
discipline's mean — while real pay distributions are **right-skewed**, so
watch whether the simulated lines put more weight in the left tail and
less in the right than the data does. And because it's Normal, some
simulated draws will run below plausible salaries entirely, the same
issue the prior predictive check flagged, now with the data pulling
against it.

Neither invalidates the ICC below — the variance decomposition is fairly
robust to modest misfit in shape — but both are arguments for the
log-scale version recommended later, and both are things you'd want to
have noticed before a compensation committee asks.
:::

### Computing the ICC

The **intraclass correlation (ICC)** answers "what proportion of total
variance sits at the discipline level?" — it's just the between-group
variance divided by the total:

```{r compute-icc}
draws <- as_draws_df(fit_var)

icc_draws <- draws |>
  transmute(
    var_discipline = sd_Discipline__Intercept^2,
    var_residual   = sigma^2,
    icc = var_discipline / (var_discipline + var_residual)
  )

icc_summary <- icc_draws |>
  summarise(
    median_icc = median(icc),
    lower = quantile(icc, 0.025),
    upper = quantile(icc, 0.975)
  )

icc_summary
```

```{r icc-plot}
ggplot(icc_draws, aes(x = icc)) +
  stat_halfeye(fill = navy_light, colour = navy, .width = c(0.5, 0.95)) +
  geom_vline(xintercept = icc_summary$median_icc,
             linetype = "dashed", colour = red) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Posterior distribution of the ICC",
    subtitle = "Proportion of salary variance explained by discipline alone. Red dashed = the median; bars = 50% and 95% intervals.",
    x = "ICC", y = NULL
  )
```

### Saying it out loud

That chart and that table are the finding, so here is the sentence they
support — the "one chart, one sentence" pairing Chapter 10 asked for:

> **About `r scales::percent(icc_summary$median_icc, accuracy = 1)` of
> the variation in median salary sits between disciplines rather than
> within them (95% credible interval
> `r scales::percent(icc_summary$lower, accuracy = 1)` to
> `r scales::percent(icc_summary$upper, accuracy = 1)`). The rest is
> variation between majors inside the same discipline.**

Two things worth noticing about how that sentence is built. It leads with
the quantity, not the model. And the interval is wide — which is the
honest consequence of having only sixteen disciplines, and precisely what a
point-estimate ICC would have concealed. If someone is about to make a
decision on the strength of that number, the width of the interval is as
much a part of the finding as the middle of it.

Which is what the Reward Director needed. A high ICC says the structure
is doing real work and the redesign is a structural project. A low one
says most of the variation is happening *inside* job families, and a new
grade architecture will move less than she hopes — the problem is in how
individual pay decisions get made, which is a different six months of
work. She does not need to understand a variance component to act on
that. She needs one number, its interval, and the sentence above.

::: callout-important
## Reading an ICC

An ICC near 0 says discipline barely matters — majors vary about as
much *within* a discipline as they do *between* disciplines, so
grouping by discipline tells you little. An ICC near 1 says almost all
the variation is structural — once you know the discipline, you know
almost everything about typical pay in that major. Most real
organisational pay data lands somewhere in between, and **that number
is itself the finding** — not a nuisance statistic on the way to
something else.
:::

------------------------------------------------------------------------

## Extending to a real compensation dataset

`graduates` only has one grouping factor. A real internal comp dataset
usually has several **crossed** factors at once — level, function,
geography — none of which is neatly nested inside another the way
`Major` sits inside `Discipline`. The same machinery extends directly:

```r
# Illustrative — not run, no internal comp dataset available here
fit_comp <- brm(
  log(salary) ~ 1 + (1 | level) + (1 | job_family) + (1 | geography),
  data = comp_data, family = gaussian(),
  chains = 4, iter = 2000
)
```

::: callout-tip
`job_family` rather than `function`, deliberately — `function` is a
reserved word in R and cannot be used as a column name without
backticks. A small thing that costs an afternoon the first time an HRIS
extract arrives with a column called exactly that.
:::

Each `(1 | ...)` term gets its own variance component, and you can ask
the same ICC-style question of each one: how much of pay variance sits
at level, versus function, versus geography, versus unexplained
individual variation. (Salary is typically modelled on the
**log** scale in practice — pay differences are usually more
proportional than additive, and log-scale residuals tend to look much
closer to normal.)

::: callout-note
## A different lens: this is "variance components analysis," a classical idea

Decomposing variance into between- and within-group pieces predates
Bayesian statistics by decades — it's the core idea behind the
classical **random-effects ANOVA** and **variance components
analysis** taught in experimental design courses, computed via
expected mean squares rather than a posterior. `lme4::lmer()` plus the
`performance::icc()` helper function will get you a point-estimate ICC
in the frequentist framework, often quickly. The Bayesian version
earns its keep here for the same reason as Chapter 15: a **full
posterior distribution for the ICC itself**, which matters when you
have a modest number of groups (like our 16 disciplines) and don't want
to overstate how precisely you know the answer.
:::

::: callout-tip
## For the ML/DS crowd

If you've done feature engineering for a gradient-boosted model with a
high-cardinality categorical column (postcode, job title, store ID),
you've likely used **target encoding** or **mean encoding** — replacing
a category with the average outcome for that category, usually with
some smoothing toward the global mean for rare categories. That
smoothing is empirical Bayes / partial pooling again, wearing a
different hat, and the ICC computed here is directly related to how
much that smoothing matters: a high ICC means the category is doing
real work and deserves less smoothing; a low ICC means most of what
looks like a category effect is noise, and aggressive smoothing (or
dropping the feature) is the right call.
:::

------------------------------------------------------------------------

## Where this kind of analysis has to be handled carefully

::: callout-important
## This is not, on its own, a pay equity audit

A variance-components model tells you *how much* variance sits at a
given level of structure — it does not, by itself, tell you whether
that structure is fair, legal, or free of bias.

The gap is not a small one, and it is worth being precise about what
would close it. A pay equity analysis controls for legitimate,
defensible factors — role, experience, performance, market rate —
before looking at protected characteristics. **Deciding which factors
are legitimate is a causal question, not a statistical one**, and it is
the hardest part of the whole exercise. Control for something that is
itself a consequence of bias and you will explain the gap away rather
than measure it: if people from one group are systematically placed in
lower-paid roles, then "controlling for role" hides exactly what you
were asked to find.

Chapter 21 is about how to make that decision defensibly, and Chapter 23
about how to make it *with* the people who know the organisation rather
than alone at your desk. Neither turns this chapter into a pay equity
audit on its own, but together they supply the piece that is missing
here.

Two things remain true regardless. This kind of work typically involves
legal counsel and is often conducted under legal privilege. And the
right specialists — legal, compensation, sometimes external counsel —
should be in the room before any model output of this kind goes near a
real pay decision.
:::

::: callout-caution
## Try it yourself

Add `Unemployment_rate` as a fixed-effect predictor alongside
`(1 | Discipline)`. Does the discipline-level variance shrink once
you account for it? What would that tell you about *why* disciplines
differ in pay?
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

"How much does department/level/location actually explain?" comes up
constantly — in comp reviews, in engagement analysis, in attrition
discussions. A variance-components model gives you a defensible,
quantified answer instead of one based on looking at a chart, and the ICC is a genuinely
good one-number summary to put in front of a compensation committee:
it says plainly how much structure there is to explain, before anyone
starts debating why.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Grouped pay (or any continuous) data splits into **between-group**
    and **within-group** variance — and a raw group-average table
    conflates the two.
2.  A varying-intercepts model gives you both variance components
    directly, and the **ICC** turns them into a single interpretable
    number.
3.  The idea extends cleanly to multiple crossed grouping factors —
    level, function, geography — each with its own variance component.
4.  This is a classical idea too (variance components / random-effects
    ANOVA) — the Bayesian version's edge is a full posterior for the
    ICC itself, useful with a modest number of groups.
5.  Variance decomposition is a statistical building block for pay
    equity work, not a substitute for the legal and compensation
    expertise that real pay decisions require.
6.  Two reasonable-looking parameter priors can combine into an
    unreasonable model. The **prior predictive check** is what catches
    it — here, a Normal prior implying negative salaries — and you
    cannot see it by looking at either prior on its own.
:::

## Next chapter {.unnumbered}

**Survival analysis: time-to-event in the employee lifecycle** — from
"how much variance" back to a different shape of question entirely:
not how much, but *when*.
