# Survival Analysis: Time-to-Event in the Employee Lifecycle

## Welcome {.unnumbered}

"Will this employee leave?" is the question most attrition work starts
from. It's the wrong question to *start* with, because it throws away
the most useful part of the data: **when**. Someone who leaves after two
months and someone who leaves after eleven years are both just "a leaver"
to a yes/no model — but they're very different stories, and a model
that only sees yes/no can't tell them apart.

This chapter introduces **survival analysis**: the family of methods
built specifically to model *time until an event*, handling gracefully
the one thing every attrition dataset has and most other methods choke
on — people who **haven't left yet**.

### The request, as it actually arrives {.unnumbered}

The COO has seen an attrition number he doesn't like and has already
decided what to do about it:

> *"We're at 18% and I want it under 12. I've got budget for one thing —
> either we fix onboarding or we fix the two-to-three year mark where I'm
> told people get itchy. Which one?"*

The 18% is a single annual rate, and it cannot answer his question,
because it has averaged away the only thing that distinguishes the two
options: **when** people leave. A model that predicts *whether* someone
leaves can't answer it either — it has thrown the timing away too.

What he needs is the shape of the curve. If departures cluster in the
first nine months, onboarding is the answer and the two-year theory is
folklore. If the curve is flat early and steps down at 30 months, the
opposite. This chapter is how you get that curve, and how you get it
without discarding the majority of the workforce who simply haven't left
yet.

## What you'll be able to do by the end {.unnumbered}

1.  Explain **censoring** and why ignoring it biases your conclusions
2.  Read a **Kaplan-Meier curve** — the standard first look at
    time-to-event data
3.  Fit a **Bayesian parametric survival model** in `brms` with
    covariates
4.  Interpret a survival model's coefficients in plain business language
5.  Add a **frailty term** so groups — industries, sites, managers — get
    their own baseline risk, with Chapter 8's partial pooling doing the
    work
6.  Recognise where the same technique needs adapting for a *positive*
    event (time-to-promotion) that not everyone will ever experience

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(survival)
library(brms)
library(tidybayes)
library(broom)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

# Real, anonymised HR data, originally shared by people analyst Edward
# Babushkin and redistributed on Kaggle under CC BY-NC-SA 4.0 (attribution,
# non-commercial, share-alike). Kaggle has no stable public download URL,
# so — to stay on the right side of that license without redistributing
# someone else's data ourselves — this book doesn't bundle a copy. Download
# it yourself from
# https://www.kaggle.com/datasets/davinwijaya/employee-turnover and save it
# as data/turnover.csv before running this chapter (see data/README.md).
turnover <- read_csv(
  "data/turnover.csv",
  locale = locale(encoding = "latin1"),   # the source file has one non-UTF-8 byte
  show_col_types = FALSE
) |>
  mutate(profession = str_replace(profession, "Finanñe", "Finance"))  # encoding artefact in the source file

# Reference levels chosen for readability: the most common recruitment
# channel and "no coach" become the baseline every coefficient is compared to.
turnover <- turnover |>
  mutate(
    traffic = fct_infreq(traffic),
    coach = factor(coach, levels = c("no", "yes", "my head"))
  )
```

`turnover` follows 1,129 employees at a real company, from their join date
to whichever came first — leaving, or the last time they were observed:
`stag` (months of tenure), `event` (did they leave, 1 = yes), `traffic`
(how they were recruited — job site, referral, agency, and several other
channels), `coach` (whether they had a coach during probation: none, a
dedicated coach, or their own manager acting as coach), plus `industry`,
`profession`, `age`, `gender`, `greywage` (salary fully declared to tax
authorities or not), `way` (commute method), and five Big Five personality
scores. This chapter uses `traffic` and `coach`; the rest are there for the
exercises.

::: callout-note
## Where this data comes from

This is genuine company data, not a simulation — which is exactly the shape
Chapter 1 promised more of: a join-to-departure duration for every
employee, not a fixed observation window. Source: Edward Babushkin,
redistributed on Kaggle, licensed
[CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/) —
attribution required, non-commercial use only, and any shared adaptation
must carry the same license. That's why this chapter asks you to download
your own copy rather than shipping one inside the book.
:::

------------------------------------------------------------------------

## The problem with yes/no

### Censoring: the idea this whole chapter rests on

Take three employees from this data:

| Employee | Months of tenure recorded | Left? | What actually happened |
|---|---|---|---|
| A | 38 | Yes | Left the company after 38 months |
| B | 102 | No | Still employed when the data was pulled, at 102 months' tenure |
| C | 7 | No | Still employed when *they* were last observed, at 7 months |

::: callout-important
B and C are not "successes" in the way a yes/no model would treat
them. We don't know whether they left a month later, or a decade later,
or never. We only know they **hadn't left by the time we stopped
watching**. This is called **right censoring**, and it is the defining
feature of almost all employee lifecycle data — everyone still employed
at the moment you pull the data is censored, not a clean zero.
:::

Throwing censored people away wastes most of your data (in this
dataset, most employees haven't left). Coding them as "0 = didn't
leave" and running ordinary logistic regression quietly assumes you
know they'll never leave, which you don't. Survival analysis is built
to use exactly this kind of incomplete information correctly.

------------------------------------------------------------------------

## A first look: the Kaplan-Meier curve

Before any model, the standard descriptive tool for time-to-event data
is the **Kaplan-Meier estimator** — the survival-analysis equivalent of
a histogram. It answers, at every point in time, "of everyone still at
risk, what fraction are still here?"

```{r km-fit}
km_fit <- survfit(Surv(stag, event) ~ 1, data = turnover)

km_df <- tidy(km_fit)

median_tenure <- km_df |> filter(estimate <= 0.5) |> slice_head(n = 1)  # <1>

ggplot(km_df, aes(x = time, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = navy_light, alpha = 0.30) +
  geom_step(colour = navy, linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = red) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Kaplan-Meier survival curve: probability of still being employed",
    subtitle = "Shaded band = 95% confidence interval. Red dashed = half the workforce.",
    x = "Months since joining", y = "Probability still employed"
  )
```

1.  The first month at which the curve crosses 50% — the **median
    tenure**, and the one number from this chart most stakeholders will
    actually remember.

Read it left to right. Everyone starts employed, so the curve begins at
100%, and each step down is a month in which somebody left. Two features
are worth naming, because they are what a stakeholder will ask about:

**The steep early drop.** The curve falls fastest in the first year or so
— early tenure is when people are most likely to leave, which is why
onboarding and probation get the attention they do. A single "annual
attrition rate" averages this away completely.

**The widening band.** The confidence interval gets wider further right,
because fewer and fewer people have been around long enough to be at
risk at 100 months. The curve's right-hand end is the least reliable part
of it and is routinely over-read.

Where the curve crosses the red line is the **median tenure**: in this
data, about
`r if (nrow(median_tenure) > 0) paste0(round(median_tenure$time), " months") else "beyond the observation window (the curve never reaches 50%)"`.
That's the plain-English version, and it's a better summary than an
attrition percentage because it carries a time unit.

::: callout-note
## A different lens: this is a frequentist tool, and that's fine

The Kaplan-Meier estimator is a **classical, non-parametric** method —
no prior, no posterior, just a direct calculation from the data. It's
worth keeping in your toolkit even in a Bayesian workflow: it's the
fastest possible "what does the raw data look like" check, with no
modelling assumptions to defend, before you commit to any particular
distributional shape.
:::

::: callout-caution
## Try it yourself

Refit the Kaplan-Meier curve separately for each `coach` group
(`Surv(stag, event) ~ coach`) and plot the three curves together. Does
having a coach during probation appear to change how quickly people
leave — and if so, in which part of the tenure range?
:::

------------------------------------------------------------------------

## Adding covariates: a Bayesian parametric model

The Kaplan-Meier curve describes *the whole population*. To ask "does
`traffic` or `coach` change how quickly people leave?" we need a
regression-style model. We'll use a **Weibull accelerated failure time
model** — a standard, fully parametric choice that `brms` fits
natively, including the censoring. Time-to-event might look like an
entirely different kind of problem from everything so far, but notice
the tool doesn't change: still `brm()`, still a formula and priors —
just `family = weibull()` this time, and a `cens()` term in the formula
to handle the censoring.

A Weibull AFT model works on the **log** of time, so — same habit as
every log-scale prior so far — translate before trusting it.

Note what the translation buys, because it's more than readability. On the
log-odds and log-rate scales of earlier chapters, plotting a prior told you
about a *parameter*, and a separate prior predictive check was needed to
find out what it implied about the data. Here the intercept prior
exponentiates directly into **expected months until leaving** — the units
of the outcome itself. So this single chart is both halves of Chapter 10's
step 4 at once: the prior, and the prior predictive check on the intercept.
That's a property of the AFT parameterisation worth knowing about, and one
of the quieter reasons to prefer it over a proportional-hazards
formulation when you have to defend your choices to somebody.

```{r plot-prior-surv}
set.seed(15)

tibble(log_time = rnorm(4000, 0, 2)) |>
  mutate(months = exp(log_time)) |>
  ggplot(aes(months)) +
  geom_histogram(bins = 40, fill = navy, alpha = 0.6) +
  geom_vline(xintercept = max(turnover$stag), linetype = "dashed", colour = red) +
  scale_x_log10(labels = scales::label_number()) +
  labs(title = "What Normal(0, 2) on the log-time scale implies",
       subtitle = "Simulated draws from the intercept prior, translated to expected months until leaving (log10 axis). Red dashed = the longest tenure we actually observe.",
       x = "Implied expected months until leaving", y = "Simulated draws")
```

::: callout-note
A wide spread — from a matter of days out to well over a decade — which
is the point: with censoring doing a lot of the work here, we'd rather
start too vague than accidentally rule out a plausible tenure length
before the data gets a say.

Note also what the log scale is hiding, and why the red line is there.
Plenty of prior mass sits to the *right* of the longest tenure in the
data — the prior is happy to entertain careers of several hundred months.
That's harmless, because the likelihood has 1,129 observations with which
to argue; it's flagged because on a smaller dataset a prior this vague on
a log scale can put real weight on absurd tenures, and the log axis is
exactly what makes that easy to miss.
:::

```{r fit-survival}
turnover <- turnover |>
  mutate(censored = 1 - event)   # brms: 0 = event observed, 1 = right-censored

fit_surv <- brm(
  stag | cens(censored) ~ traffic + coach,
  data = turnover, family = weibull(),
  prior = c(prior(normal(0, 2), class = Intercept),
            prior(normal(0, 1), class = b)),
  chains = 4, iter = 2000, seed = 15, refresh = 0
)

summary(fit_surv)
```

### Reading the output, line by line

This output has one block you haven't met before, and it is the one
people skip:

**Regression Coefficients.** One line per recruitment channel and per
coaching category, each compared against its baseline (the most common
`traffic` channel, and "no coach"). These are on the **log-time** scale,
which is why the next section exponentiates them before saying anything
in business language. A positive coefficient means *longer* expected
tenure than baseline; negative means shorter.

**Further Distributional Parameters — `shape`.** This is the new one, and
it's the Weibull's most useful parameter. It describes whether the risk
of leaving **rises or falls with tenure**:

- `shape` below 1 — the hazard *decreases* over time. Leaving is most
  likely early, and each month survived makes the next one safer. This is
  the usual pattern in employment data, and it's the shape the
  Kaplan-Meier curve above already hinted at.
- `shape` equal to 1 — a constant hazard. The risk of leaving next month
  is the same whether you joined last year or nine years ago. (This
  special case is the exponential distribution.)
- `shape` above 1 — the hazard *increases* with tenure. Risk builds the
  longer someone stays.

Check `shape` against the Kaplan-Meier curve before you trust anything
else in the output. If the model says the hazard is flat and the raw
curve plainly drops off a cliff in year one, the Weibull's shape
assumption is fighting your data and the coefficients inherit that
problem.

**`Rhat` and `Bulk_ESS`, on every line.** Chapter 10's step 6. `Rhat`
should be `1.00`; ESS in the hundreds or better. Censored models
occasionally sample less comfortably than the ones so far, so this is the
first chapter where the glance is more than a formality.

::: callout-note
## Why `cens(censored)`, not `cens(event)`

`brms` expects the censoring indicator to say **"was this a
censoring event"** — 0 for an observed event, 1 for right-censored.
Our `event` column is coded the opposite way (1 = the event we care
about happened), so we flip it. Getting this backwards is a classic,
silent survival-analysis mistake: the model will still run, it'll just
answer the wrong question.
:::

### Reading a Weibull AFT coefficient

Coefficients here don't give you a probability directly — they tell
you how much a predictor **stretches or compresses time**:

```{r interpret-coefs}
coef_table <- fixef(fit_surv) |>
  as_tibble(rownames = "term") |>
  mutate(time_multiplier = exp(Estimate))

coef_table

# One term, pulled out and described in plain language directly from
# the fitted numbers — so the prose below always matches whatever this
# model actually finds, rather than a number typed in by hand.
coach_yes <- coef_table |> filter(term == "coachyes")

coach_summary <- if (coach_yes$Q2.5 > 0) {
  sprintf("a confident, positive effect — about %s%% longer expected tenure than having no coach, with the entire credible interval above zero",
          round((coach_yes$time_multiplier - 1) * 100))
} else if (coach_yes$Q97.5 < 0) {
  sprintf("a confident, negative effect — about %s%% shorter expected tenure than having no coach, with the entire credible interval below zero",
          round((1 - coach_yes$time_multiplier) * 100))
} else {
  sprintf("an estimate that leans %s (time multiplier %s), but the 95%% credible interval (%s to %s) comfortably includes zero — not confident evidence of a real effect either way",
          if_else(coach_yes$Estimate > 0, "positive", "negative"),
          round(coach_yes$time_multiplier, 2),
          round(coach_yes$Q2.5, 2), round(coach_yes$Q97.5, 2))
}
```

Take `coachyes` — employees assigned a dedicated coach during probation,
versus the "no coach" baseline. Exponentiating a coefficient turns it into
a **time multiplier**: above 1 means longer expected tenure, below 1 means
shorter. In this fit, `coachyes` shows `r coach_summary`.

::: callout-important
## Business translation

An accelerated-failure-time coefficient always translates directly into
"longer" or "shorter" expected tenure, not just "more or less likely" —
that's its whole appeal for a business conversation. But translate the
*confidence* honestly too: `r coach_summary` for the coaching effect above.
Scan the rest of the `coef_table` output the same way — for each term,
check whether the interval sits clearly on one side of zero before you
report it as a finding, not just the point estimate.
:::

### Does the model fit? Checking it against the raw curve

The posterior predictive check for a survival model is a purpose-built
one: simulate datasets from the fitted model, compute a Kaplan-Meier
curve for each, and lay them over the Kaplan-Meier curve of the real
data. If the model's shape assumption is wrong, this is where it shows.

```{r ppc-surv}
pp_check(fit_surv, type = "km_overlay", ndraws = 50) +                # <1>
  labs(title = "Posterior predictive check: simulated vs actual survival curves",
       subtitle = "Thin lines = tenure curves the model would generate; thick line = the real data",
       x = "Months since joining", y = "Probability still employed")
```

1.  Note what we do **not** pass. The underlying `bayesplot` function
    needs a `status_y` argument saying which observations were events and
    which were censored — otherwise the observed curve would treat every
    censored tenure as a departure. `brms` supplies it automatically,
    because you already told the model in the `cens(censored)` term, and
    passing it again by hand is an error rather than a belt-and-braces
    precaution. Declare censoring once, in the formula, and everything
    downstream inherits it.

::: callout-tip
## For the ML/DS crowd

If you've built "will this customer/employee churn in the next 30
days" as a classification problem, you've been solving a cruder version
of this. Two well-known bridges back to more familiar ML territory:
**discrete-time survival models are literally logistic regression**,
fitted on data reshaped into one row per person-per-time-period, with
time itself as a predictor — no new machinery required, just a
different data shape. And gradient-boosted survival models (e.g.
XGBoost's Cox objective, or DeepSurv-style neural survival models) are
solving exactly this problem at scale — modelling a hazard rather than
a class label, so that "hasn't happened yet" is used as information
instead of being dropped or mislabelled as "no."
:::

::: callout-important
## What a Weibull typically gets wrong here, and why it still earns its place

Watch the **early months** in particular. A Weibull has one shape
parameter to describe the whole hazard curve, and real employment hazards
often aren't one shape: a spike of early leavers in the first few months,
then a long, much flatter tail. If the simulated curves fall too slowly
at the start and too quickly later, that's the single-shape assumption
straining, and you're seeing exactly the trade-off the Cox comparison
below describes.

That's a reason to caveat the "expected months" numbers, not a reason to
abandon the model. The *coefficients* — the relative comparison between
coached and uncoached — are considerably more robust to shape misfit than
the absolute predicted times are. Which is a useful general rule for
parametric survival models: trust the comparison further than the
forecast.
:::

### Predicted survival curves by group

The coefficient answers "longer or shorter." A stakeholder deciding
whether to fund probation coaching wants to see *how much longer, and by
when* — which means a curve, not a number:

```{r predicted-curves}
newdata <- tibble(
  traffic = levels(turnover$traffic)[1],   # the most common recruitment channel, held fixed
  coach = c("no", "yes")
) |>
  mutate(
    traffic = factor(traffic, levels = levels(turnover$traffic)),
    coach = factor(coach, levels = levels(turnover$coach)),
    profile = c("No coach", "Had a coach")
  )

pred_draws <- newdata |>
  add_predicted_draws(fit_surv, ndraws = 2000)

pred_draws |>
  group_by(profile) |>
  summarise(
    median_months = median(.prediction),
    q25 = quantile(.prediction, 0.25),
    q75 = quantile(.prediction, 0.75),
    .groups = "drop"
  )
```

```{r predicted-curves-plot}
#| code-fold: true
horizon <- seq(0, 100, by = 2)

surv_curves <- pred_draws |>
  ungroup() |>
  select(profile, .prediction) |>
  cross_join(tibble(month = horizon)) |>                              # <1>
  group_by(profile, month) |>
  summarise(still_here = mean(.prediction > month), .groups = "drop")

ggplot(surv_curves, aes(month, still_here, colour = profile)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("No coach" = navy_light,
                                 "Had a coach" = navy)) +
  labs(
    title = "Predicted survival curves: coached vs uncoached during probation",
    subtitle = "Share of a comparable cohort still employed, from the fitted model",
    x = "Months since joining", y = "Predicted probability still employed",
    colour = NULL
  ) +
  theme(legend.position = "top")
```

1.  The survival curve straight from the predictive draws: at each month,
    what fraction of simulated employees are predicted to still be there.
    No distributional algebra needed — just counting draws, which is the
    same trick as Chapter 20's `P(A beats B)`.

Two employees who differ only in whether they had a coach during probation
have different predicted tenures above — with a full posterior distribution
behind every one of those numbers, not just a point estimate. Compare the
gap here with what `coach_summary` said about the underlying coefficient:
a real difference in typical predicted values can still coexist with an
uncertain coefficient, because the predicted-values summary folds in all
of the model's other residual variation too — another reason to read the
coefficient's own interval, not just look at predicted numbers.

::: callout-important
## The chart, and the sentence that goes with it

This is the deliverable Chapter 10 asked for. The chart shows two cohorts
diverging over time; the sentence is the coefficient's honest reading:
`r coach_summary`.

Note what the pairing prevents. The chart alone invites someone to read
the gap between two lines as settled fact. The sentence alone throws away
the *timing* — where in the tenure curve the gap opens up, which is what
determines whether a coaching intervention would pay for itself. Neither
is sufficient; together they're a recommendation somebody can act on.
:::

::: callout-note
## A different lens: how this compares to a Cox model

The most common survival model in classical statistics is the **Cox
proportional hazards model** (`survival::coxph()`). Cox is
*semi-parametric* — it makes no assumption about the shape of the
baseline hazard, which is why it's so popular. The trade-off: it's
built to compare *relative risk* between groups, and getting an actual
predicted time-to-event out of it takes extra steps. The Weibull AFT
model used here is fully parametric (it assumes a specific hazard
shape), which is a stronger assumption — but it directly hands you a
predictive distribution for "when," which is usually exactly what a
business stakeholder is asking for. Fitting both and comparing is a
reasonable habit; `brms` can also fit a Bayesian Cox model
(`family = cox()`) if you want the semi-parametric version with
posterior uncertainty.
:::

------------------------------------------------------------------------

## Frailty: survival with a grouping factor

### The question this chapter has been building toward

Everything so far has asked what makes people leave *sooner or later*.
There is a version of that question People Analytics teams care about
more, and it has been waiting since Chapter 8:

> **Do people leave faster under some managers than others — beyond what
> the makeup of their teams already explains?**

That is a multilevel question inside a survival model, and it has its own
name. A **frailty model** gives each group its own multiplier on the risk
of the event happening — its own "frailty". Groups with high frailty lose
people faster than their covariates predict; groups with low frailty hold
on to them longer.

::: callout-note
The name comes from medical statistics, where the groups were patients
and the frailty was literal. Nothing about it is specific to health, and
in an HR setting the vocabulary is unhelpful — you are modelling
*something about the group that you did not measure*, which is precisely
what Chapter 8's varying intercepts did for averages.
:::

### The formula is the one you already know

The grouping term is identical to Chapter 8's, dropped into the survival
model:

```{r}
#| eval: false
stag | cens(censored) ~ traffic + coach                    # what we fitted
stag | cens(censored) ~ traffic + coach + (1 | industry)   # with frailty
```

```{r fit-frailty}
fit_frailty <- brm(
  stag | cens(censored) ~ traffic + coach + (1 | industry),
  data = turnover, family = weibull(),
  prior = c(prior(normal(0, 2),   class = Intercept),
            prior(normal(0, 1),   class = b),
            prior(exponential(2), class = sd)),           # <1>
  chains = 4, iter = 2000, seed = 16, refresh = 0
)

summary(fit_frailty)
```

1.  The one new prior. `sd` is how much industries differ in their
    baseline risk, on the log-time scale — `exponential(2)` keeps it
    modest without ruling out a real difference. Same class of prior,
    and the same reasoning, as Chapter 8's `sd(Intercept)`.

### Reading it

`sd(Intercept)` under **Multilevel Hyperparameters** is the finding.
It says how much industries differ in tenure once recruitment channel
and coaching are accounted for. Near zero means the differences you can
see between industries are explained by who they hire and how they
support them. Clearly above zero means something unmeasured about the
industry itself is shortening or lengthening careers.

::: callout-important
This is the survival chapter's version of Chapter 8's argument, and it
closes an arc that started there.

Chapter 8 showed that a league table of **averages** across unequal
groups is mostly a league table of group sizes. The same is true of a
league table of **attrition rates** — and it is worse, because attrition
counts are usually small, the observation windows differ, and censoring
means the raw number is not even measuring what people think it is.

Swap `industry` for `manager_id` and this model answers the question
every HR director eventually asks: *which managers actually lose people
faster?* Partial pooling handles the manager with four reports the same
way Chapter 15 handled the small office — it pulls the estimate toward
the average by exactly as much as the thin evidence warrants.
:::

::: callout-warning
## When you can skip this

**Skip it when the groups are large and similar in size.** Frailty earns
its place when some groups are small, which is exactly when raw
comparisons mislead. With forty roughly equal business units, the
grouped and ungrouped answers will be close.

**Skip it when the grouping is a nuisance rather than the question.** If
you are asking what recruitment channel does and industry is just
something to account for, adding `(1 | industry)` widens your intervals
appropriately but changes little else. Worth doing, not worth a section
in the deck.

**Don't skip it when the group is going to be named.** The moment an
estimate is attached to an individual manager or site — in a review, a
scorecard, a conversation about someone's job — the small-group problem
stops being a statistical nicety. That is the case where the raw number
is not merely imprecise but predictably unfair to whoever has the
smallest team.
:::

------------------------------------------------------------------------

## The other direction: time-to-promotion

Survival analysis isn't only for negative events. The `promotion`
dataset frames the identical technique around a positive one:

```{r promotion-example}
data("promotion", package = "peopleanalyticsdata")
promotion <- promotion |> drop_na(year, promoted, diverse, flexible, store)

promotion |>
  count(promoted) |>
  mutate(pct = n / sum(n))
```

Here, `year` is the number of years until promotion (or until last
observed, for those not yet promoted), and `promoted` is the event
indicator — the same *shape* of data as the attrition model above. But
there's a real difference underneath that shape, worth stopping on
before fitting anything.

### Why "not yet" doesn't mean the same thing here

Every survival model in this chapter so far rests on an assumption you
haven't had to notice, because attrition happens to satisfy it:
**given long enough, everyone eventually experiences the event.**
Everybody leaves a company eventually — resignation, retirement,
redundancy, the company itself closing down. So for attrition, "hasn't
left yet" always means exactly that: hasn't left *yet*. Every censored
employee is a genuine still-waiting case, and a model built on the
assumption that the survival curve eventually decays to zero is
describing something true.

Promotion doesn't share that guarantee. Some people will be promoted
eventually, just later than others — but some people will *never* be
promoted, not because we stopped watching too soon, but because it was
never going to happen: a capped individual-contributor role, someone
who's opted out of the track, a team with no room above them.
Statistically, that's a **cured** (or "immune") fraction sitting inside
the censored group — language borrowed from medical survival analysis,
where some patients are cured and will never relapse, and others just
haven't relapsed *yet*. Our data can't tell the two kinds of censored
case apart for any one still-unpromoted employee; they look identical,
right up until (or unless) one of them gets promoted.

A Weibull AFT model — like every standard survival model — implicitly
assumes there's no cured fraction: whatever the shape, it expects the
survival curve to keep decaying toward zero if you wait long enough.
Fit it to data with a genuinely large cured fraction, and it has no way
to say "some of these people will just never be promoted" — it can only
stretch one distribution to somehow cover both the people who'll be
promoted eventually *and* the people who never will, which usually
means an inflated tail and an "expected time to promotion" that
doesn't correspond to anything real.

The tell is in the shape of the Kaplan-Meier curve:

```{r km-promotion}
km_promo <- survfit(Surv(year, promoted) ~ 1, data = promotion)

tidy(km_promo) |>
  ggplot(aes(x = time, y = estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = navy_light, alpha = 0.30) +
  geom_step(colour = navy, linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = red) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Kaplan-Meier curve: probability not yet promoted",
    subtitle = "Still heading toward the red line by the end of the window, or already levelling off well above it?",
    x = "Years", y = "Probability not yet promoted"
  )
```

::: callout-important
If that curve is still visibly falling by the end of the observed
window, a standard survival model is a reasonable choice — most of the
remaining un-promoted cases are plausibly still waiting. If it's already
flattened out well above zero instead, that's a cured fraction showing
itself directly in the data: a substantial share of the sample was
never going to be promoted, however much longer you watched. With only
`r scales::percent(mean(promotion$promoted), accuracy = 0.1)` of this
sample promoted at all, look carefully at the shape above before
trusting a Weibull AFT's "expected time to promotion" here.
:::

Two practical ways to handle a genuine cured fraction, rather than
quietly modelling as if it isn't there:

1.  **Reframe the question with a bounded horizon.** Instead of an
    open-ended "how long until promoted" — which only really makes
    sense if everyone eventually gets there — ask a bounded one:
    "promoted within 3 years, yes or no?" That's back to ordinary
    logistic regression (Chapter 9's toolkit), and it sidesteps the
    cured-fraction problem entirely by never assuming an infinite
    horizon.
2.  **Fit a genuine cure model.** The formal fix: a two-part model that
    first estimates *who's even eligible to ever be promoted* (a
    logistic sub-model for the cured fraction), then fits a survival
    distribution for *time to promotion, given eligibility*. Not a
    family built into `brms` — `flexsurvcure` is the standard R
    package for the frequentist version — but conceptually it's still
    one recipe built from two linked pieces, and an ambitious reader
    could construct the Bayesian version as a custom `brms` mixture
    likelihood.

This isn't unique to promotion — the same question is worth asking of
any "good" event you model this way (first big sale, first internal
move, first stretch assignment): is this something everyone will get
to eventually, or does your population genuinely include people who
never will?

::: callout-caution
## Try it yourself

First, look at the Kaplan-Meier curve above and judge for yourself:
does this look like a "some people will never be promoted" situation?
Then fit `year | cens(1 - promoted) ~ diverse + flexible + store`,
family `weibull()`, exactly as before — but, knowing what you now know,
treat its "expected time to promotion" number with real scepticism
rather than reporting it at face value. What would you tell a
stakeholder instead: a coefficient's *direction* (raises or lowers the
chances of earlier promotion, without claiming a trustworthy "expected
time"), or a reframed bounded-horizon question instead?
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Attrition, promotion, time-to-productivity for new hires, time until a
first internal move — almost every "how long until X" question in the
employee lifecycle is a survival question, and almost all of that data
is censored (anyone still in the state you're measuring is a censored
observation, not a clean zero). Reaching for a plain classifier on this
kind of data is one of the most common avoidable mistakes in People
Analytics — it silently discards timing information and mishandles
everyone still "in progress."
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  **Censoring** — not knowing the outcome for people still "in
    progress" — is the defining feature of employee lifecycle data, and
    needs to be modelled, not ignored or coded as zero.
2.  The **Kaplan-Meier curve** is the fast, assumption-light first look
    at time-to-event data.
3.  A **Weibull accelerated failure time model** in `brms` adds
    covariates while handling censoring correctly, and gives
    coefficients you can read as "longer" or "shorter" expected time.
4.  The same machinery extends to positive events (time-to-promotion),
    not just negative ones (time-to-attrition) — but check first
    whether everyone in your population will eventually experience the
    event. Attrition guarantees that; many positive events, like
    promotion, don't.
5.  The Weibull's **`shape`** parameter says whether the risk of leaving
    rises or falls with tenure, and checking it against the raw
    Kaplan-Meier curve — along with a **`km_overlay`** posterior
    predictive check — is how you find out whether one shape is enough to
    describe your hazard. When it isn't, trust the comparison between
    groups further than the absolute predicted times.
6.  A Kaplan-Meier curve that **plateaus well above zero** is the
    signal of a **cured fraction** — people who will never experience
    the event, not just people you haven't waited long enough for.
    Standard survival models assume no such fraction exists; when one
    does, reframe as a bounded-horizon logistic question or reach for
    a formal cure model instead of trusting the expected-time output.
:::

## Next chapter {.unnumbered}

**Analysing Likert and survey data the Bayesian way** — from "when"
back to "how much," but for outcomes that are ordered categories, not
continuous numbers — the other data shape every People Analytics
practitioner meets constantly.
