# Analysing Likert and Survey Data the Bayesian Way

## Welcome {.unnumbered}

If you work in People Analytics, you work with Likert scales — 1–4,
1–5, 1–7, "strongly disagree" to "strongly agree." They're everywhere:
engagement surveys, exit interviews, pulse checks, interview scorecards.
And they're routinely analysed in a way that quietly produces the wrong
answer: treated as if they were ordinary continuous numbers, averaged,
t-tested, put into a linear regression.

This chapter is about the gap between "1, 2, 3, 4 are numbers" and "1,
2, 3, 4 are *ordered categories*, not equally-spaced quantities" — and
the Bayesian tool built specifically for the second, more accurate,
description: **ordinal regression**.

### The request, as it actually arrives {.unnumbered}

The engagement survey has closed and the CHRO wants the story, not the
scores:

> *"Management came out at 3.1 and benefits at 3.4, so the deck says
> benefits are in better shape. Fine. But I have money for one programme
> next year. Which of these actually moves how people feel about working
> here?"*

Two different questions are tangled up in that, and the usual analysis
answers neither. Averaging the sub-scales and ranking them answers "which
scores highest," which is not the same as "which matters most." And a
linear regression on the averaged scores answers a version of "which
matters most" in units — *happiness points per management point* — that
can't be turned back into anything she can say to a board.

What she can use is a probability: how much of the workforce moves from
the unhappy categories into the happy ones. That's an ordinal model's
native output, and this chapter's closing chart is exactly that.

## What you'll be able to do by the end {.unnumbered}

1.  Explain why treating a Likert scale as continuous can mislead —
    and when it's a harmless shortcut
2.  Recognise the **proportional odds** idea behind ordinal regression
3.  Fit a Bayesian ordinal model with `brms::cumulative()`
4.  Read and plot **predicted category probabilities**, not just a
    single coefficient
5.  Know where to go for the classical (frequentist) and deeper
    treatments of this topic

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

# An *ordered* four-step ramp for the four Likert categories. Ordered data
# needs a palette that gets monotonically darker — a categorical palette
# (or a red somewhere in the middle) tells the eye the categories are
# unrelated, which is the exact opposite of what an ordinal scale means.
likert_pal <- c("#dce6f2", "#8fabd0", "#3d68a8", "#122a52")

data("employee_survey", package = "peopleanalyticsdata")

# Guard against missing values before modelling (Chapter 1 found gaps
# in a similar teaching dataset).
employee_survey <- employee_survey |>
  drop_na(Happiness, Ben1, Ben2, Ben3, Work1, Work2, Work3,
          Man1, Man2, Man3, Car1, Car2, Car3, Car4)
```

`employee_survey` has 2,833 responses to an engagement survey on a
**4-point scale**, increasingly positive: an overall `Happiness`
rating, three benefits questions (`Ben1-3`), three work-environment
questions (`Work1-3`), three management questions (`Man1-3`), and four
career-prospects questions (`Car1-4`).

```{r composites}
employee_survey <- employee_survey |>
  mutate(
    Man_avg  = (Man1 + Man2 + Man3) / 3,
    Work_avg = (Work1 + Work2 + Work3) / 3,
    Ben_avg  = (Ben1 + Ben2 + Ben3) / 3,
    Car_avg  = (Car1 + Car2 + Car3 + Car4) / 4,
    Happiness = factor(Happiness, ordered = TRUE)
  )

glimpse(employee_survey)
```

We'll average each sub-scale's items into a single composite predictor
to keep the model readable, and predict `Happiness` — kept as an
**ordered factor**, which matters for everything that follows.

------------------------------------------------------------------------

## The problem: why "just average it" can mislead

### A quick look at the shape

```{r happiness-bar}
ggplot(employee_survey, aes(x = Happiness, fill = Happiness)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = likert_pal) +                            # <1>
  scale_x_discrete(labels = c("1\nvery\nunhappy", "2\nsomewhat\nunhappy",
                              "3\nsomewhat\nhappy", "4\nvery\nhappy")) +
  labs(title = "Distribution of overall happiness",
       subtitle = "2,833 responses on a 4-point scale",
       x = NULL, y = "Number of responses")
```

1.  The same ordered ramp used for every Likert chart in this chapter, so
    "darker means more positive" means the same thing on every one of
    them.

Note the shape, because it decides how much of this chapter you need. The
responses are **bunched toward the positive end** and there are only
**four** categories — which, as the note below spells out, is exactly the
combination where treating the scale as continuous stops being a harmless
shortcut.

### What "treat it as continuous" assumes

A plain linear regression of `Happiness` (as a number) on `Man_avg`
assumes the *gap* between 1 and 2 means the same thing, in the same
units, as the gap between 3 and 4. There's no reason to believe that:
the psychological distance between "strongly disagree" and "disagree"
is not guaranteed to equal the distance between "agree" and "strongly
agree."

```{r naive-lm}
naive_lm <- lm(as.numeric(Happiness) ~ Man_avg, data = employee_survey)
summary(naive_lm)$coefficients
```

Two numbers to read off that, in the `Estimate` column. The
`(Intercept)` is the predicted happiness score for someone rating their
management a hypothetical zero — already outside the scale, which is the
first hint. `Man_avg` is the slope: how many happiness points a one-point
improvement in management perception buys, and the model insists it's the
*same* number wherever on the scale you start.

That second assumption is the one to test, and the way to test it is to
ask the model what it predicts at the ends:

```{r naive-lm-problem}
tibble(Man_avg = c(1, 4)) |>
  mutate(predicted = predict(naive_lm, newdata = tibble(Man_avg = Man_avg)))
```

::: callout-important
Watch what a linear model is willing to predict at the edges of the
scale: values that don't correspond to any real response category, and
a straight-line relationship that treats every one-point move as
identical regardless of where on the scale it happens. With only four
categories and responses piled up near the top (common in engagement
data), this isn't a small rounding issue — it changes the conclusion.

The deeper problem is that "3.6" is not an answer anybody gave. There is
no response category between "somewhat happy" and "very happy", so a
predicted 3.6 cannot be translated into a statement about people. The
ordinal model's output — a probability for each *actual* category — can.
:::

::: callout-note
## To be fair to the shortcut

Treating a Likert item as continuous isn't always wrong in practice —
with enough categories (5–7+), a roughly symmetric distribution, and a
large sample, results from a linear model and an ordinal model often
point the same direction. The risk rises exactly where engagement data
usually sits: few categories, heavily skewed toward the positive end,
and the deltas between adjacent categories are unlikely to be equal.
Check your own data's shape before assuming either way.
:::

------------------------------------------------------------------------

## The idea behind ordinal regression

::: callout-tip
## For the ML/DS crowd

If you've trained a model on ordinal survey targets before, there's a
good chance you did one of two things that quietly lose information:
treated it as a **regression** target (implicitly assuming equal
spacing, as above) or as **multiclass classification** with a
categorical cross-entropy loss (which treats predicting "2" instead of
the true "4" as no worse than predicting "3" instead of "4" — throwing
away the fact that categories are ordered). **Ordinal regression** is
the dedicated middle ground, and it shows up under different names
across tooling: "ordered logit"/"ordered probit" in econometrics, cumulative
link models in statistics, and libraries like Python's `mord` in more
ML-flavoured toolchains. Same underlying idea as this chapter's model,
whichever name you meet it under.
:::

### A latent variable underneath

Ordinal regression imagines a continuous, unobserved "true happiness"
for every respondent, which gets chopped into the four categories we
observe by three **thresholds**:

```
very unhappy | threshold 1 | somewhat unhappy | threshold 2 | somewhat happy | threshold 3 | very happy
```

The model estimates where those thresholds sit, and how predictors
shift the *whole underlying distribution* left or right. This is
exactly the same latent-threshold idea Keith McNulty's book develops in
depth in **Chapter 7, "Proportional Odds Logistic Regression for
Ordered Category Outcomes"** — if you want the classical derivation and
more worked examples of this model, that's the place to go deeper. This
chapter gives you the Bayesian fitting and interpretation of the same
model family — and, true to form throughout this book, fitting it is
still `brm()` with a formula and priors, just `family = cumulative("logit")`
in place of whatever came before.

### Fitting it

The same weakly-informative shape used for the log-odds slopes back in
Chapter 9 — worth a quick look again, since it's a new model family
even if the prior itself isn't new:

```{r plot-prior-ordinal}
tibble(b = seq(-4, 4, length.out = 400)) |>
  mutate(density = dnorm(b, 0, 1)) |>
  ggplot(aes(b, density)) +
  geom_area(fill = navy, alpha = 0.5) +
  geom_vline(xintercept = 0, colour = red, linetype = "dashed") +
  labs(title = "Prior for each sub-scale's effect: Normal(0, 1)",
       subtitle = "Red dashed = no effect. Centred there, and wide enough to find one.",
       x = "Effect on cumulative log-odds", y = "Prior density")
```

::: callout-note
## Why there's no separate prior predictive check here

Chapter 10's step 5 asks for a prior predictive check as well as a prior
plot, and this chapter does the plot and not the simulation. That's a
deliberate choice rather than an omission, and the reason is worth
knowing.

An ordinal model's outcome is **bounded by construction**. Whatever the
priors, the model can only ever predict a probability distribution across
the four categories that exist — it cannot produce a negative salary
(Chapter 16) or an engagement score of 140 (Chapter 22), because there is
nothing for it to predict except "1, 2, 3 or 4". The absurd-implications
failure mode that a prior predictive check exists to catch is largely
unavailable to it.

What remains checkable is whether the *implied distribution across
categories* is sensible before seeing data, and the honest version of that
check is the posterior predictive `type = "bars"` chart two sections down.
The general rule: prior predictive checks earn their keep in proportion to
how far the outcome scale extends beyond the plausible. On a four-category
scale, that's not very far.
:::

```{r fit-ordinal}
fit_ord <- brm(
  Happiness ~ Man_avg + Work_avg + Ben_avg + Car_avg,
  data = employee_survey, family = cumulative("logit"),
  prior = c(prior(normal(0, 1), class = b)),
  chains = 4, iter = 2000, seed = 16, refresh = 0
)

summary(fit_ord)
```

### Reading the output, line by line

This output looks like every other `brms` summary and one block means
something quite different from usual. That block is worth slowing down
on, because it's where the thresholds we just drew as a diagram actually
live:

**Regression Coefficients — but read the Intercepts separately.** There
are two kinds of row here, mixed together:

- `Intercept[1]`, `Intercept[2]`, `Intercept[3]` are **the thresholds** —
  the three cut-points from the diagram above, on the cumulative log-odds
  scale. Three of them for four categories, always one fewer than the
  number of categories. They should come out in **increasing order**
  (`Intercept[1]` < `Intercept[2]` < `Intercept[3]`), because they're
  positions along a line; `brms` enforces this, so if they ever look out
  of order you've misread which rows you're looking at. They are not
  "the intercept" in the Chapter 6 sense, and there's rarely much reason
  to interpret their values directly.
- `Man_avg`, `Work_avg`, `Ben_avg`, `Car_avg` are the actual predictor
  effects — how far each one **shifts the whole latent distribution**
  relative to those fixed thresholds. Positive means shifting people
  toward the happier end.

**No `sigma`.** Worth noticing its absence. A Normal model needs a
residual standard deviation; here the latent scale's spread is fixed at 1
by convention, because the thresholds and the coefficients are only ever
identified relative to it. Nothing has gone wrong.

**`Rhat` and `Bulk_ESS`.** Chapter 10's step 7, as always. Ordinal models
with several correlated predictors — and these four sub-scales *are*
correlated, since happy people tend to rate everything highly — can
sample less crisply than the models so far. This is a reasonable place to
actually look rather than assume.

::: callout-tip
The thing to take from that block: **the thresholds are not the finding
and the coefficients are not the deliverable either.** The coefficients
are on a cumulative log-odds scale that nobody can interpret out loud.
Two sections below turns them into predicted category probabilities,
which is the form a stakeholder can use.
:::

::: callout-note
## "Proportional odds" — the assumption in the name

The model assumes each predictor's effect is the *same* across every
threshold — moving from "very unhappy" to "somewhat unhappy" and
moving from "somewhat happy" to "very happy" are shifted by the same
amount for a one-point change in, say, `Man_avg`. This is a real
assumption, not a technicality — it's checkable (compare to a model
that relaxes it) and worth stating explicitly when you report results.
:::

### Reading the coefficients

```{r ordinal-coefs}
fixef(fit_ord) |>
  as_tibble(rownames = "term") |>
  filter(!str_detect(term, "Intercept"))
```

These are log-odds on the *cumulative* scale — not intuitive to read
directly. Before turning them into something usable, the posterior
predictive check Chapter 10 asks for. For a categorical outcome there's a
purpose-built version: compare the predicted count in each response
category against the observed count.

```{r ppc-ordinal}
pp_check(fit_ord, type = "bars", ndraws = 200) +
  labs(title = "Posterior predictive check: predicted vs actual response counts",
       subtitle = "Bars and intervals = what the model predicts; points = what people actually answered",
       x = "Happiness rating", y = "Number of responses")
```

::: callout-note
## What this catches that a coefficient can't

One bar per response category, and you want each observed point sitting
inside its interval. Read it as four separate questions: does the model
reproduce how many people said "very unhappy"? "Somewhat unhappy"? And so
on.

The specific failure to look for in engagement data is the **top
category**. If the model under-predicts the number of 4s — the point
sitting above its bar — that's a sign the proportional-odds assumption is
straining at the positive end, where all the responses are piled up. That
matters because the top category is usually the one being reported to the
business ("X% of staff are very happy"), so misfit there lands directly in
the headline number.

This check is also the reason to prefer the ordinal model over the linear
one on evidence rather than principle. A linear model can't be checked
this way at all — it doesn't predict categories, so there's nothing to
compare.
:::

The far more useful output is the predicted probability of
each response category:

```{r predicted-probs}
newdata <- tibble(
  Man_avg = seq(1, 4, length.out = 30),
  Work_avg = mean(employee_survey$Work_avg),
  Ben_avg  = mean(employee_survey$Ben_avg),
  Car_avg  = mean(employee_survey$Car_avg)
)

pred_probs <- newdata |>
  add_epred_draws(fit_ord) |>
  group_by(Man_avg, .category) |>
  summarise(mean_prob = mean(.epred), .groups = "drop")

ggplot(pred_probs, aes(x = Man_avg, y = mean_prob, fill = .category)) +
  geom_area(position = "stack") +
  scale_fill_manual(
    values = likert_pal,                                              # <1>
    labels = c("1 very unhappy", "2 somewhat unhappy",
               "3 somewhat happy", "4 very happy")
  ) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    title = "Predicted happiness category, across management perception",
    subtitle = "Each band's height is the predicted probability of that response",
    x = "Management sub-scale average (1–4)", y = "Predicted probability",
    fill = "Happiness"
  )
```

1.  The ordered ramp again, and this is the chart where it matters most.
    An earlier draft of this figure used red for category 3 — which reads
    as a *highlight* in the middle of an ordered sequence, telling the eye
    that "somewhat happy" is a different kind of thing from its
    neighbours. On an ordinal scale, colour has to be monotonic or it
    contradicts the model it's illustrating.

```{r headline-shift}
#| code-fold: true
# The headline number, computed rather than typed, so the sentence below
# always matches the fitted model.
very_happy <- pred_probs |>
  filter(.category == "4") |>
  filter(Man_avg == min(Man_avg) | Man_avg == max(Man_avg)) |>
  arrange(Man_avg)

shift_low  <- very_happy$mean_prob[1]
shift_high <- very_happy$mean_prob[2]
```

::: callout-important
## Business translation

This chart says something a single coefficient can't: as management
perception improves, the whole distribution of happiness responses
shifts — the "very unhappy" and "somewhat unhappy" bands shrink, and
"very happy" grows. That's a much more useful (and honest) statement
than "the coefficient was significant."

The sentence to put next to it, in Chapter 10's one-chart-one-sentence
form:

> **Among people who rate their management at the bottom of the scale,
> about `r scales::percent(shift_low, accuracy = 1)` are predicted to say
> they're very happy. At the top of the management scale that rises to
> about `r scales::percent(shift_high, accuracy = 1)` — holding benefits,
> work environment and career perceptions at their average.**

Both numbers are proportions of people in categories that people actually
chose. Nobody has to be told what a cumulative log-odds is, and nothing
has been rounded into a 3.6 that means nothing.
:::

::: callout-note
## A different lens: the classical version, and where it strains

The frequentist proportional-odds model (`MASS::polr()` or
`ordinal::clm()` in R) fits the same model and will usually give you
very similar point estimates to the Bayesian version above — this
isn't an area where the two philosophies typically disagree much on the
answer. Where the Bayesian version earns its keep: a full posterior for
every predicted probability (not just a point estimate with a
Wald-style standard error), and a natural extension to add a
multilevel structure (e.g. responses nested within department or team,
exactly as in Chapters 8 and 15) without leaving the same modelling
framework.
:::

------------------------------------------------------------------------

## Aside: multiple raters on the same Likert scale

`recruiting` — another `peopleanalyticsdata` dataset, on hiring
decisions — has three interviewers each rating candidates 1–5 (`int1`,
`int2` from line managers, `int3` from HR). The same ordinal framework
applies to asking whether raters use the scale consistently:

```{r recruiting-aside}
data("recruiting", package = "peopleanalyticsdata")
recruiting <- recruiting |> drop_na(int1, int2, int3)

recruiting |>
  summarise(
    mean_int1 = mean(int1), mean_int2 = mean(int2), mean_int3 = mean(int3),
    cor_12 = cor(int1, int2), cor_13 = cor(int1, int3)
  )
```

::: callout-caution
## Try it yourself

Fit `int3 ~ int1 + int2` with `family = cumulative("logit")`. Does the
HR interviewer's rating track the line managers' ratings closely, or
does it look like a different scale is being applied? This is a
lightweight, model-based way to ask an inter-rater reliability question
you'd otherwise reach for Cohen's kappa or an ICC to answer.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Every engagement survey, exit interview scorecard, interview rating,
and pulse check in your organisation is ordinal data. The habit of
averaging Likert items and running a t-test or linear regression is
extremely common and usually harmless when categories are numerous and
responses spread out — and genuinely misleading exactly when survey
data tends to look in practice: few categories, bunched toward the
positive end. Checking your data's shape before choosing, rather than
defaulting to the familiar tool, is the actual skill here.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Likert scales are **ordered categories**, not guaranteed to be
    equally spaced — treating them as continuous is a real assumption,
    not a neutral default.
2.  **Ordinal regression** models a latent continuous scale cut by
    thresholds into the categories you observe.
3.  The **proportional odds** assumption says a predictor's effect is
    the same across every threshold — checkable, and worth stating.
4.  `brms::cumulative("logit")` fits this model with full posteriors;
    plotting **predicted category probabilities** communicates far more
    than a single coefficient.
5.  The frequentist version (`polr()`/`clm()`) usually agrees closely on
    the answer — the Bayesian version's edge is full uncertainty and an
    easy path to adding multilevel structure.
6.  `pp_check(type = "bars")` is the check that matters here: it asks
    whether the model reproduces how many people actually chose each
    category, and misfit in the **top** category is the failure that lands
    in your headline number.
7.  Colour on an ordinal chart has to be **monotonic**. A sequential ramp
    says "these categories are ordered"; a categorical palette, or a
    highlight colour in the middle of the sequence, contradicts the model
    you just fitted.
:::

## Next chapter {.unnumbered}

**Modelling what you can't see** — this chapter took the survey
responses at face value. The next one asks what happens when the number
in the column is not quite the quantity it claims to be, or is not there
at all.
