# Modelling What You Can't See: Measurement Error, Missing Data & Latent Variables

## Welcome {.unnumbered}

Every model in this book so far has quietly assumed something that is
almost never true: that the numbers in the columns are the quantities
they claim to be. That an engagement score of 3.4 *is* the person's
engagement. That a performance rating of 4 *is* their performance. That
one salesperson's 26 months of tenure *is* the same 26 months as
somebody else's.

They aren't. Every one of those numbers is a measurement of something
you can't observe directly, taken with an instrument of limited quality,
and the gap between the two is not small. The meta-analytic literature
puts the inter-rater reliability of supervisor performance ratings at
around 0.52 — meaning that when two managers rate the same person,
roughly **half the variance in the rating has nothing to do with the
person being rated.**

That's not a caveat. Half your data being noise changes what your models
can and can't find, and in a specific and predictable direction.

This chapter is about what to do when you don't observe your variables
exactly. It has two halves that look unrelated and are the same problem:
**measurement error** (you observe the variable, but wrongly) and
**missing data** (you don't observe it at all). The Bayesian treatment of
both is identical in spirit — if you don't know a value, put a
distribution on it and let it stay uncertain — and it's one of the places
where this framework does something the alternatives genuinely struggle
to.

::: callout-important
The core idea in one sentence: **an unobserved quantity is just another
parameter — give it a model rather than a guess.**
:::

## What you'll be able to do by the end {.unnumbered}

1.  Explain what a reliability coefficient means in units you can use
2.  Predict the direction and size of **attenuation** — why measurement
    error in a predictor drags its coefficient toward zero
3.  Fit a model that accounts for a known measurement error using
    `brms::me()`, and read the difference
4.  Distinguish **noise** from **bias**, and explain why calibration
    sessions address the smaller of the two problems
5.  Name the three missingness mechanisms and say which one your data
    is plausibly in
6.  Impute missing values *inside* the model with `brms::mi()`, so the
    uncertainty propagates instead of disappearing
7.  Recognise when the thing you want to measure is latent, and know
    what to read next

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

data("salespeople", package = "peopleanalyticsdata")
data("employee_survey", package = "peopleanalyticsdata")
```

Note what we haven't done: `drop_na()`. Every previous chapter opened by
quietly discarding incomplete rows. This one is about what that costs.

------------------------------------------------------------------------

# Part 1 — Measurement error

## Not every wrong number is wrong in the same way

Those three examples fail in three different ways, and they do not have
the same fix. Separating them now saves a great deal of confusion later.

**A number can be noisy.** Ask two managers to rate the same person and
you get two answers. Ask the same manager twice and you may still get two
answers. What is recorded is the real value plus something random, and
the something random has no particular direction. Engagement scores and
performance ratings are both this. **It is the only one of the three with
a statistical fix, and it is what most of this chapter is about.**

**A number can mean something other than what you assume.** Tenure is
recorded exactly — there is no noise in a start date. But is it
continuous service, service in role, or service with the group? Does an
acquisition reset it? A career break? Does everyone hired before the 2015
system migration share a start date, because that is when the field was
created? These errors are systematic rather than random, so no amount of
modelling averages them away. You fix the definition, or you flag the
affected rows and say so.

**A number can cover different amounts of the world.** Two people each
have 26 months of tenure, and those 26 months contain different numbers
of working days. Not by much, but by more than you would guess. Simulate
it — everybody on exactly 26 months, only the start date varying — and
the working days run from about **547 to 562**, a fifteen-day spread with
a standard deviation of around three and a half days. Weekday alignment
accounts for only two of those; the rest is public holidays, because a
26-month window catches either two or three of each annual holiday
depending on which month you started.

The clearer version of the same problem is monthly reporting. Comparing
February's sales against March's compares 28 days with 31 — around 10%
— before anybody has done anything differently. The number is exact. The
**unit** is not constant.

### Does that matter? The arithmetic says it depends

This is worth a moment, because the answer is not "yes" or "no" and the
thing that decides it is the same ratio the next section is about: how
much of the variation you see is real, and how much is the calendar.

| What you are comparing | Reliability of "months" | Effect on a coefficient |
|---|---|---|
| A whole workforce, tenure 1 month to 20 years | 0.9999 | 0.01% — ignore it |
| One cohort, everybody around 26 months | 0.96 | about 4% |
| February against March | ≈ 0 | the calendar *is* the finding |

Across a real workforce, genuine differences in tenure are so much larger
than three days that the jitter disappears. Narrow the comparison and it
grows. Compare one month with another and there is nothing left but
calendar.

::: callout-important
Two things make this different from the noise in a rating, and both point
to the same conclusion.

**You can work out the true value.** Nobody knows a person's real
engagement, which is why the rest of this chapter models it. But working
days are *computable* from a start date. When the truth is recoverable,
recover it — do not model the error in a number you could simply
calculate correctly.

**It is systematic, not random.** Two people who started in the same
month share the same distortion. That is harmless if start dates are
scattered, and dangerous if they are not: in an organisation that hires
seasonally, start month is tangled up with role, contract type and
department. At that point the calendar stops being noise that shrinks
your estimate a little and becomes bias that pushes it in a direction.
:::

So the rule is **use the right unit**, and that means one of two things.
Sometimes it is an offset, as in Chapter 15 — a manager with thirty
reports has more opportunity than one with five, so transfer counts were
divided through by team size before anyone compared managers. Sometimes
it is simply computing working days instead of months. Either way you are
removing the problem rather than modelling it, which is always the better
trade when it is available.

------------------------------------------------------------------------

## What reliability actually means

The standard model is deliberately simple. An observed score is a true
score plus an error:

$$
x_{\text{observed}} = x_{\text{true}} + \varepsilon
$$

and **reliability** is the share of the observed variance that comes from
the true score:

$$
r = \frac{\mathrm{Var}(x_{\text{true}})}{\mathrm{Var}(x_{\text{observed}})}
$$

That definition is more useful than it looks, because it converts a
reliability coefficient into a standard deviation you can reason about.
If true performance had a standard deviation of 15 points and ratings are
0.52 reliable, then the observed variance is $225 / 0.52 \approx 433$,
the error variance is the difference, and the error has a standard
deviation of about **14.4 points**:

```{r reliability-to-sd}
sd_true     <- 15
reliability <- 0.52

sd_observed <- sd_true / sqrt(reliability)
sd_error    <- sqrt(sd_observed^2 - sd_true^2)

tibble(sd_true, reliability, sd_observed, sd_error)
```

The noise in a single performance rating is almost exactly as large as
the real differences between people. Put another way: a one-rating-point
gap between two employees is close to a coin flip.

::: callout-important
This is the number to keep in your head for the rest of the chapter.
Not "ratings are imperfect" — everyone agrees with that and then models
as if they weren't. **The error is the same size as the signal.**
:::

::: callout-tip
## For the ML/DS crowd

The closest familiar idea is label noise, with one important difference:
noise in the *labels* mostly costs you accuracy, while noise in the
*features* costs you interpretability in a directional way. A model
trained on noisy features can still predict well — it just can't tell you
how much each feature matters, because every coefficient has been shrunk
by that feature's reliability.

This has a practical consequence for feature-importance work. If you rank
features by SHAP value or permutation importance and one feature is
measured with a much noisier instrument than the others, you have ranked
your measurement quality as much as your signal. Well-instrumented
features (system-logged: tenure, absence days, headcount) will
systematically outrank survey- or judgement-derived ones (engagement,
performance, potential) regardless of which actually matters — because
the first group has no noise to attenuate them, whatever else may be
wrong with it.
:::

## Attenuation: where your effect went

Measurement error in a *predictor* doesn't just add noise. It biases the
coefficient, systematically, toward zero. This is the single most
practically important fact in the chapter and it's easy to demonstrate.

```{r attenuation-sim}
n <- 4000

true_engagement <- rnorm(n)                                          # <1>
outcome         <- 0.5 * true_engagement + rnorm(n)                  # <2>

sd_err   <- sqrt((1 - reliability) / reliability)                    # <3>
measured <- true_engagement + rnorm(n, sd = sd_err)

attenuation <- tibble(true_engagement, measured, outcome)
```

1.  Standardised, so the coefficient below is directly readable.
2.  The true effect is 0.5. Half a standard deviation of outcome per
    standard deviation of engagement — a large effect by People
    Analytics standards.
3.  With `sd(true) = 1`, this is the error SD implied by a reliability of
    0.52 — the same arithmetic as the block above, rearranged.

```{r slope-helper}
#| include: false
# Pull one named coefficient out of an lm() into a one-row tibble, so the
# comparison below reads as a table rather than two blocks of model output.
slope_of <- function(fit, term, label) {
  s <- summary(fit)$coefficients[term, ]
  tibble(model = label, estimate = s[["Estimate"]], se = s[["Std. Error"]])
}
```

```{r attenuation-fits}
bind_rows(
  slope_of(lm(outcome ~ true_engagement, data = attenuation),
           "true_engagement", "if we could measure it perfectly"),
  slope_of(lm(outcome ~ measured, data = attenuation),
           "measured", "using the score we actually have")
)
```

The estimate falls from 0.5 to roughly 0.26 — and there's nothing
approximate about that number. The attenuated slope is the true slope
multiplied by the reliability, and 0.5 × 0.52 = 0.26. Measurement error
doesn't blur the answer; it shrinks it by a factor you can calculate.

```{r attenuation-curve}
#| code-fold: true
sweep <- tibble(r = seq(0.2, 1, by = 0.02)) |>
  mutate(
    estimate = map_dbl(r, \(rel) {
      m <- true_engagement + rnorm(n, sd = sqrt((1 - rel) / rel))
      coef(lm(outcome ~ m))[["m"]]
    })
  )

ggplot(sweep, aes(r, estimate)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = red) +
  geom_line(colour = navy, linewidth = 1) +
  geom_vline(xintercept = 0.52, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 0.52, y = 0.12, hjust = -0.08, size = 3.2,
           colour = "grey40", label = "typical performance rating") +
  labs(
    title = "How much of your effect survives your measurement",
    subtitle = "Red dashed = the true effect of 0.5",
    x = "Reliability of the predictor", y = "Estimated coefficient"
  )
```

::: callout-important
## The consequence nobody draws

If your engagement survey has a reliability of 0.7 and your performance
rating has a reliability of 0.52, and you regress one on the other, you
are not measuring a weak relationship between engagement and performance.
You are measuring a strong relationship between two noisy proxies, and
reporting the product.

"Engagement explains surprisingly little of performance" is one of the
most-repeated findings in the field. Some of it is real. A good deal of
it is arithmetic.
:::

::: callout-note
## A different lens: this problem has two older names

Psychometrics has had a fix since 1904: Spearman's **correction for
attenuation** divides the observed correlation by the square root of the
product of the two reliabilities. Econometrics knows the same problem as
**errors-in-variables** or **classical measurement error**, and reaches
for instrumental variables when the error is in a regressor.

Both work, and both are worth knowing. What neither gives you is
uncertainty that accounts for the correction itself. Spearman's formula
returns a point estimate corrected by another point estimate; the
resulting number has a confidence interval that is quietly wrong, because
the reliability coefficient was itself estimated. The Bayesian version
below treats the true score as an unknown parameter, which means the
uncertainty about what people's real scores are shows up where it belongs
— in a wider posterior — rather than being corrected away.
:::

## The Bayesian version

`brms` handles this directly. If you can supply a standard error for each
observation of a noisy predictor, `me()` treats the true value as a
parameter to be estimated rather than a number to be trusted.

::: callout-note
## A worked (illustrative) example

`salespeople` gives us one customer rating per person with no indication
of how it was produced. Suppose — and this is the kind of thing you'd
establish from the survey vendor, not invent — that each rating is the
mean of a handful of customer responses, so each carries a standard error
we can estimate:

```{r construct-se}
salespeople_me <- salespeople |>
  drop_na(sales, customer_rate, performance) |>
  mutate(
    customer_rate_c  = customer_rate - mean(customer_rate),
    customer_rate_se = 0.45                                          # <1>
  )
```

1.  Illustrative, and deliberately a single number rather than a column —
    in real work this varies by person, because some people have more
    responses behind their score than others. The people with the fewest
    responses are exactly the ones whose scores need the most
    discounting, which is the Chapter 14 argument arriving by a different
    route.

The model is the Chapter 6 skeleton with one term wrapped. We fit the
naive version alongside it, because the comparison is the whole point:

```{r fit-me}
priors <- c(
  prior(normal(400, 200),   class = Intercept),                      # <1>
  prior(normal(0, 100),     class = b),
  prior(exponential(0.005), class = sigma)
)

fit_naive <- brm(
  sales ~ customer_rate_c,
  data = salespeople_me, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 18, refresh = 0
)

fit_me <- brm(
  sales ~ me(customer_rate_c, customer_rate_se),                     # <2>
  data = salespeople_me, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 18, refresh = 0,
  save_pars = save_pars(latent = TRUE)                               # <3>
)
```

1.  Chapter 6's priors, unchanged. The only difference between these two
    models is `me()`.
2.  The one wrapped term. Everything else is identical to the line above.
3.  Keeps the estimated true scores in the fitted object. You don't need
    them for the coefficient, but they're worth looking at once — each
    person now has a posterior for what their rating "really" was.

Before reading anything off it, the two checks the workflow asks for.
`me()` adds a latent true score for every observation, so the parameter
count jumps from three to several hundred — which makes step 7 less of a
formality than usual:

```{r ppc-me}
pp_check(fit_me, ndraws = 100) +
  labs(title = "Posterior predictive check: the measurement-error model",
       subtitle = "Light lines = datasets the model would generate; dark = the real data",
       x = "Sales (thousands of dollars)", y = "Density")
```

::: callout-note
Watch `Rhat` and `ESS` on this one in particular. A latent variable per
observation is a large, weakly-identified posterior, and it is common for
the true-score parameters to mix poorly even when the coefficient you
care about is fine. Chapter 11's remedy applies: more iterations first,
then `adapt_delta`.
:::

```{r read-me}
me_comparison <- bind_rows(
  fit_naive |> gather_draws(b_customer_rate_c) |>
    median_qi(.width = 0.95) |> mutate(model = "taking the rating at face value"),
  fit_me |> gather_draws(`bsp_.*`, regex = TRUE) |>
    median_qi(.width = 0.95) |> mutate(model = "allowing for measurement error")
) |>
  mutate(width = .upper - .lower)

me_comparison |>
  select(model, estimate = .value, .lower, .upper, width)
```
:::

Two things to notice, and the second is the one people miss. The
coefficient is **larger** than the naive version, because we've undone
some of the attenuation. And the interval is **wider**, because we've
stopped pretending we know each person's rating exactly. Both directions
are correct. A method that made the estimate bigger *and* more precise
would be selling you something.

Both of those are much easier to see than to read:

```{r me-comparison-plot}
#| code-fold: true
ggplot(me_comparison, aes(x = .value, y = fct_rev(factor(model)))) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_linerange(aes(xmin = .lower, xmax = .upper),
                 colour = navy_light, linewidth = 3) +
  geom_point(colour = navy, size = 3) +
  labs(
    title = "What allowing for measurement error does to the estimate",
    subtitle = "Point = posterior median; bar = 95% credible interval",
    x = "Sales per point of customer rating", y = NULL
  )
```

This is the shape to commit to memory, because it is what an honest
correction looks like: the point moves **out**, away from zero, and the bar
gets **longer**. Any technique that claims to remove bias should be
expected to cost you precision, and one that doesn't is worth being
suspicious of.

It's also the reason this comparison belongs on a chart rather than in the
table above. Two rows of numbers invite the reader to look at the estimates
and skip the interval widths — which inverts the lesson, because the widths
are the half that's easy to forget.

::: callout-important
## Check the sampler here more carefully than usual

Every model in this chapter is harder to fit than anything before it, and
this is the one place in the book where Chapter 10's step 6 is more than a
formality.

The reason is structural. `me()` adds **one latent parameter per
observation** — 350-odd unknown "true" customer ratings, each informed by a
single noisy measurement — and `mi()` does the same for every missing
value. Models with a parameter per row are inherently harder to explore
than models with five parameters, and they can look fine while quietly
sampling badly.

```{r diagnose-me}
#| eval: false
summary(fit_me)                 # scan Rhat and Bulk_ESS on the bsp_ row
plot(fit_me, variable = "bsp_mecustomer_rate_ccustomer_rate_se")
```

Two things to watch, and neither shows up in the coefficient table:

- **Divergent transitions.** `brms` warns about these on fitting, and the
  warning is easy to scroll past in a rendered document. Divergences on a
  measurement-error model usually mean the stated standard error is too
  small relative to the spread in the data — the model is being asked to
  believe the ratings more precisely than the likelihood can support.
- **Low `Bulk_ESS` on the `bsp_` coefficient specifically.** The latent
  scores and the coefficient trade off against each other, which is the
  same ridge-walking problem Chapter 22 diagnoses in its event-study model.
  More iterations is the first response; a tighter prior on `sigma` is the
  second.

The uncomfortable implication for the comparison above: if the `me()` model
sampled poorly, its wider interval might be the sampler's uncertainty
rather than the measurement error's. Those look identical in the chart, and
only the diagnostics distinguish them.
:::

### Noise and bias are not the same problem, and HR works on the wrong one

Almost all organisational effort on rating quality targets **bias** —
systematic, directional error. The lenient manager who rates everyone
highly, the severe one, the halo effect, recency. Calibration sessions
exist to align distributions across managers, and they work: bias is
systematic, so it can be identified and corrected.

**Noise** is the unsystematic part — the same manager rating the same
person differently on a different day, two managers disagreeing for
reasons that have nothing to do with the employee. Kahneman, Sibony and
Sunstein's *Noise* documents across many professional judgement domains
that noise is typically the larger of the two, and calibration does
essentially nothing about it. The only real remedies are structural:
multiple independent raters, averaged; or a model that shrinks extreme
ratings toward the population (Chapters 14 and 15).

The practical version: **the bottom 5% of a rating distribution is not
the bottom 5% of performers.** At a reliability of 0.52 it's a mixture of
genuinely weak performers and people who had a bad year, a new manager,
or a hard project. Managing out on that basis is not selecting against
poor performance so much as partly selecting against bad luck.

::: callout-note
## This connects backwards to Chapter 15

Shrinkage and measurement-error correction are the same operation applied
at different ends of the model. Chapter 15 shrank noisy *outcome*
estimates toward the population mean, in proportion to how little data
each manager had. `me()` does the same thing to a noisy *predictor*, in
proportion to its stated error.

Both answer "how much should I trust this number?" with a quantity rather
than a judgement call. If Chapter 15 made sense, this chapter is the same
idea pointed in the other direction.
:::

------------------------------------------------------------------------

# Part 2 — Missing data

## What `drop_na()` has been doing

Every chapter so far has opened with a line like this:

```{r the-habit}
#| eval: false
salespeople <- salespeople |> drop_na()
```

In Chapter 1 that was defensible — three missing values in 351 rows. It
is also the single most common way analyses go quietly wrong, because the
defensibility depends entirely on *why* the values are missing, and
`drop_na()` never asks.

Complete-case analysis is unbiased under one condition and biased under
the other two. So it's worth knowing which one you're in.

## Three mechanisms, with the PA versions

**MCAR — missing completely at random.** The missingness has nothing to
do with anything. A file transfer dropped rows; a survey page failed to
load for a random subset. Dropping these rows costs you precision and
nothing else. This is rare, and it is the assumption `drop_na()` makes.

**MAR — missing at random.** The missingness depends on things you
*have* observed. Contractors don't get performance ratings; a business
unit adopted the engagement survey a year late; exit interview data
exists only for voluntary leavers. Given the observed variables, the
missingness carries no extra information. Dropping these rows **is**
biased — you're systematically deleting whole categories — but the
information needed to fix it is in your data.

**MNAR — missing not at random.** The missingness depends on the missing
value itself. The people most disengaged didn't fill in the engagement
survey. The employees with the worst manager relationships skipped the
manager questions. Nothing in your data can fix this, and no method
recovers it without an assumption you can't check.

::: callout-important
Survey non-response is very rarely MCAR, and treating it that way is
probably the most consequential unexamined assumption in People
Analytics. The employees who don't respond are not a random sample of
employees — that's the entire reason response rate is reported alongside
the results.
:::

## Imputing inside the model

The instinct is to fill the gaps first and model afterwards — mean
imputation, or a quick predictive fill. Both are worse than they look,
for the same reason: they manufacture certainty. Once the value is filled
in, every downstream model treats it as data, and your intervals come out
too narrow.

The Bayesian move is to refuse to fill it in. A missing value is an
unknown quantity, exactly like a coefficient, so it gets a posterior and
the uncertainty flows through to everything downstream.

::: callout-important
## First, a subtlety that decides whether any of this matters

MCAR / MAR / MNAR is the standard framing and it is not quite enough to
tell you whether `drop_na()` is safe, because it describes the
missingness without reference to the model you're about to fit.

The sharper rule, for a regression coefficient specifically:

> **Complete-case analysis is unbiased if missingness depends only on
> variables already in your model — even when it depends on them
> strongly.** It becomes biased when missingness depends on the
> *outcome*, or on something you haven't measured.

That's a genuinely useful thing to know, because it says the common
case — "contractors have no performance rating, and job type is in the
model anyway" — costs you precision and nothing else. Dropping those rows
is fine.

The case that hurts is the one where whether a value is missing depends
on how the person turned out. Exit surveys returned mostly by the
disgruntled. Customer feedback that arrives for the accounts that went
well. Manager assessments completed for the people the manager wanted to
promote. That's the version worth demonstrating, so it's the one we'll
build.
:::

### A worked example, start to finish

Start with the complete data, and fit the model we're trying to recover.
Because we're about to create the missingness ourselves, we have
something no real analysis ever has — a known right answer:

```{r fit-full}
sp_full <- salespeople |>
  drop_na() |>
  mutate(customer_rate_c = customer_rate - mean(customer_rate))

fit_full <- brm(
  sales ~ customer_rate_c + performance,
  data = sp_full, family = gaussian(),
  prior = priors,
  chains = 4, iter = 2000, seed = 18, refresh = 0
)
```

Now knock holes in `customer_rate`, with the probability depending on
`sales` — the outcome:

```{r make-mar}
sp_mar <- sp_full |>
  mutate(
    p_missing       = if_else(sales < median(sales), 0.40, 0.05),     # <1>
    customer_rate_c = if_else(runif(n()) < p_missing,
                              NA_real_, customer_rate_c)
  )

sp_mar |> summarise(missing = sum(is.na(customer_rate_c)), n = n())
```

1.  Customer ratings come back eight times less often for the
    lower-selling half of the salesforce. This is still **MAR** — the
    missingness depends on `sales`, which we observe — but it depends on
    the outcome, which is the case the callout above flagged as the
    damaging one.

The model becomes two linked formulas: one for the outcome, using
`mi(customer_rate_c)` where the predictor would go, and one describing
how `customer_rate_c` itself is generated.

```{r fit-mi}
fit_mi <- brm(
  bf(sales ~ mi(customer_rate_c) + performance) +                     # <1>
    bf(customer_rate_c | mi() ~ performance) +                        # <2>
    set_rescor(FALSE),                                                # <3>
  data = sp_mar, family = gaussian(),
  prior = c(                                                          # <4>
    prior(normal(400, 200),   class = Intercept, resp = "sales"),
    prior(normal(0, 100),     class = b,         resp = "sales"),
    prior(exponential(0.005), class = sigma,     resp = "sales")
  ),
  chains = 4, iter = 2000, seed = 18, refresh = 0
)
```

1.  The model we actually care about. `mi(customer_rate_c)` says: use the
    imputed values, and keep their uncertainty. `sales` has no gaps, so
    it needs no `| mi()` of its own — only the variable being imputed
    declares one.
2.  The imputation model. Note what is **not** on the right-hand side:
    `sales`, even though that's what the missingness depends on. It
    doesn't need to be, and that's the part worth understanding — see
    below.
3.  We're not modelling residual correlation between the two outcomes;
    the second formula exists only to serve the first.
4.  The same priors as the other two fits, but with `resp = "sales"`
    attached to each. This is a multivariate model now — two formulas, so
    two response variables — and an unqualified `prior(...)` would be
    ambiguous about which one it applies to. Leaving `prior` off entirely
    is the easy mistake here: the model runs perfectly happily on `brms`
    defaults, and the three-way comparison below stops being like-for-like
    without anything warning you. The `customer_rate_c` submodel keeps
    default priors deliberately — it's a nuisance model, and it's on a
    centred scale we've said nothing informative about.

And the complete-case model — the analysis every previous chapter in this
book would have run without comment:

```{r fit-cc}
fit_cc <- brm(
  sales ~ customer_rate_c + performance,
  data = sp_mar, family = gaussian(),                                 # <1>
  prior = priors,
  chains = 4, iter = 2000, seed = 18, refresh = 0
)
```

1.  Same data frame, gaps and all. `brm()` drops the incomplete rows and
    tells you it has — which is exactly the behaviour under examination.

```{r compare-mi}
mi_comparison <- bind_rows(
  fit_full |> gather_draws(b_customer_rate_c) |> median_qi(.width = 0.95) |>
    mutate(approach = "the answer we're trying to recover"),
  fit_cc |> gather_draws(b_customer_rate_c) |> median_qi(.width = 0.95) |>
    mutate(approach = "complete cases only"),
  fit_mi |> gather_draws(`bsp_sales_.*`, regex = TRUE) |>              # <1>
    median_qi(.width = 0.95) |>
    mutate(approach = "imputed in-model")
) |>
  mutate(approach = factor(approach, levels = c(                       # <2>
    "the answer we're trying to recover",
    "complete cases only",
    "imputed in-model"
  )))

mi_comparison |>
  select(approach, estimate = .value, .lower, .upper)
```

1.  Matched by regex rather than by name — see the trap below.
2.  Fixed order, so the truth sits at the top of the chart below and the
    two candidate methods are read against it rather than against each
    other.

```{r mi-comparison-plot}
#| code-fold: true
truth <- mi_comparison |>
  filter(approach == "the answer we're trying to recover") |>
  pull(.value)

ggplot(mi_comparison, aes(x = .value, y = fct_rev(approach))) +
  geom_vline(xintercept = truth, linetype = "dashed", colour = red) +
  geom_linerange(aes(xmin = .lower, xmax = .upper),
                 colour = navy_light, linewidth = 3) +
  geom_point(colour = navy, size = 3) +
  labs(
    title = "Three answers to the same question, one of which is right",
    subtitle = "Red dashed = the estimate from the complete data, which we only know because we made the holes ourselves",
    x = "Sales per point of customer rating", y = NULL
  )
```

Read it against the red line, which is the only reason this example is
worth running: we know the answer. Two things to look for.

**Where the complete-case interval sits relative to the line.** Its
midpoint is pulled off the truth — not by a rounding amount, and in a
direction determined by how the missingness was generated rather than by
chance. That is bias, and running the analysis again on more data of the
same kind would not fix it.

**How wide the in-model interval is.** Wider than complete cases, and that
is correct behaviour rather than a cost. The imputed model holds its
uncertainty about the missing ratings open; the complete-case model bought
its narrower interval by discarding the rows it wasn't sure about, which is
a form of confidence it hasn't earned.

::: callout-note
## A naming trap worth knowing about

`mi()` and `me()` are what `brms` calls *special* terms, and their
coefficients don't get the `b_` prefix you'd expect from every other
model in this book. They get `bsp_` — `sp` for "special" — so the
coefficient above is `bsp_sales_micustomer_rate_c`, not
`b_sales_micustomer_rate_c`. Asking `tidybayes` for the latter fails with
`No variables found matching spec`, which is a confusing error message
for what is really a spelling problem.

The general habit, worth using any time a model gains an unfamiliar term:

```{r list-variables}
variables(fit_mi) |> head(10)
```

Read the names off the fitted object rather than guessing them. Matching
with `regex = TRUE`, as above, then survives the next naming surprise
too. The same `bsp_` prefix appears in the `me()` model earlier in this
chapter, and in monotonic effects (`mo()`) if you ever reach for those on
ordinal predictors.
:::

The complete-case estimate is computed on a sample that over-represents
high sellers, because theirs are the ratings that came back. Nobody chose
that sample, and nothing in the model output announces it.

::: callout-important
## Why the imputation model doesn't need `sales` in it

This is the part that looks like it shouldn't work.

The missingness depends on `sales`. Standard advice for multiple
imputation is that the imputation model must contain everything the
missingness depends on, *including the outcome* — advice that strikes most
people as backwards the first time they hear it, and is therefore
routinely forgotten. Yet the formula above imputes `customer_rate_c` from
`performance` alone.

It works because the two submodels are fitted **jointly**. The missing
ratings are unknown parameters inside a single posterior, and that
posterior is conditioned on all the observed data — including each
person's sales. A salesperson with high sales and a missing rating gets
imputed values that lean high, because the outcome submodel says high
sellers tend to have high ratings. Nobody wrote that rule down; it falls
out of fitting everything at once.

Which is the practical argument for imputing inside the model rather than
before it. The two-step workflow relies on the analyst remembering a
counter-intuitive rule. The joint model can't forget it.
:::

::: callout-tip
## For the ML/DS crowd

This is why `sklearn`'s `SimpleImputer` in a pipeline, or filling with
the median before a gradient-boosting fit, is fine for prediction and
misleading for inference. The imputation is a point estimate, so the
model downstream cannot distinguish an observed value from a guessed one
and reports the same confidence in both.

Multiple imputation (`mice`) is the classical fix and a good one — it
generates several completed datasets, fits the model to each, and pools.
`mi()` does the same thing in a single pass, with the imputation and the
analysis sharing one posterior instead of being stitched together
afterwards.
:::

------------------------------------------------------------------------

# Part 3 — When the thing itself is unobservable {#sec-latent-variables}

Measurement error assumes there's a true score your instrument is
missing. Sometimes the construct has no direct measurement at all, and
what you have is several imperfect indicators of it.

`employee_survey` is built this way. The items come in blocks:

```{r survey-blocks}
employee_survey |> names()
```

`Ben1`–`Ben3` ask about benefits, `Work1`–`Work3` about the work itself,
`Man1`–`Man3` about the manager, `Car1`–`Car4` about career. Nobody
cares about the answer to `Man2` specifically. The question of interest
is whether someone feels well managed — a **latent variable**, indicated
by three items, none of which is the thing itself.

Two moves are available, and the difference matters:

- **Average the block into an index.** What most teams do. It's fast, and
  it treats the index as though it were measured exactly — bringing you
  straight back to Part 1 of this chapter, now with an unstated
  reliability.
- **Model the latent variable.** Let the construct be a parameter that
  the three items are noisy indicators of, and estimate the relationship
  between *constructs* rather than between indices.

::: callout-note
## Where to go, and an honest note on scope

The second route is standard practice in psychometrics and in the
psychology-adjacent parts of People Analytics, under the names
confirmatory factor analysis and structural equation modelling. It is
rarer in the economics tradition this book mostly draws on, and I don't
think it would be useful for me to write a shallow chapter on a technique
I use less than the ones in the rest of the book.

So this is a signpost rather than a treatment. If your survey has item
blocks and you want to model constructs rather than indices, the two
things worth knowing are that the measurement model (do these items hang
together?) and the structural model (how do the constructs relate?) are
separate steps and should be assessed separately, and that Keith
McNulty's *Handbook of Regression Modeling in People Analytics* covers
both properly with worked code. `blavaan` is the Bayesian implementation;
`brms` can express simpler measurement models directly.

What you should take from Part 1 regardless: **if you're going to use an
index, find out its reliability and act as though it matters — because
it does, by a factor of exactly that size.**
:::

------------------------------------------------------------------------

## The thread running through all three

These look like three chapters stapled together and they're one idea in
three costumes.

| | What you don't observe | What the model does |
|---|---|---|
| Measurement error | The true value behind a noisy score | Estimates it, given a stated error |
| Missing data | The value, at all | Estimates it, given the other variables |
| Latent variables | The construct behind several indicators | Estimates it, given the indicators |

In every case the classical instinct is to produce a best guess and carry
on, and in every case that guess becomes indistinguishable from data one
line later. The Bayesian instinct is to leave the unknown quantity
uncertain and let it stay uncertain all the way to the interval you
report.

::: callout-important
## And one more, which belongs to the next chapters

There's a fourth member of this family: **the people you don't observe at
all.** Your data contains only those the organisation hired and who
haven't yet left — and both of those were determined partly by the
variables you're studying.

That's not a missing-data problem in the sense of Part 2, because the
rows aren't in your file with gaps in them; they were never in your file.
It's a selection problem, and Chapter 21 shows why it's structurally the
same as adjusting for a collider. It's also why a validated selection
test can appear to have no relationship with performance among your
employees: the range has been restricted by the very process that used
it.
:::

::: callout-caution
## Try it yourself

1.  Rerun the attenuation simulation with a reliability of 0.9 — a
    well-built cognitive test rather than a performance rating. How much
    of the effect survives? What does that say about which of your
    variables you should worry about first?
2.  Take the `me()` model and let the standard error vary by person
    (say, larger for salespeople with lower `sales`, on the grounds that
    fewer customers responded). Does the coefficient move? Which
    individuals' estimated true scores move most?
3.  Change the missingness in `sp_mar` from MAR to MNAR — make it depend
    on `customer_rate_c` itself rather than on `sales` — and refit.
    The model will run happily and give you a confident, wrong answer.
    That is the point of the exercise.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Two sentences from this chapter will change more analyses than any model
in it.

The first: **"What's the reliability of this measure?"** Ask it about
every survey scale and every rating before you model with it. Half the
time nobody knows, which is itself the finding. When you do know, you can
say how much of an effect your instrument is capable of detecting before
you run anything.

The second: **"Why is this missing?"** Not how much — why. The answer
determines whether dropping those rows is free, biased-but-fixable, or
fatal, and it takes one conversation with whoever owns the system.

Both are questions rather than techniques, which is why they're worth
more than the code above. A stakeholder who has been told that half the
variance in performance ratings is noise will interpret every league
table you show them for the rest of the year differently — and better.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Reliability converts into an error standard deviation you can reason
    about. At 0.52, the noise in a performance rating is about as large
    as the real differences between people.
2.  Measurement error in a predictor **attenuates** its coefficient
    toward zero by a factor equal to the reliability. Weak findings are
    sometimes weak instruments.
3.  `brms::me()` treats the true score as a parameter: the coefficient
    grows and the interval widens, and both are correct.
4.  **Noise is not bias.** Calibration fixes the systematic part and
    leaves the larger, random part untouched. The bottom of a rating
    distribution is partly bad luck.
5.  `drop_na()` is only safe under MCAR, which survey data almost never
    is. MAR is fixable with information you already have; MNAR isn't
    fixable at all.
6.  `brms::mi()` imputes inside the model, so missingness costs you
    interval width instead of silently costing you accuracy.
7.  Measurement error, missing data and latent constructs are one
    problem: **an unobserved quantity should stay uncertain.**
8.  Both of this chapter's headline findings are **shapes, not numbers** —
    a point moving outward with a lengthening interval (`me()`), and a
    biased midpoint against a known truth (`mi()`). Plot the intervals
    rather than tabling them, or readers will compare the estimates and
    skip the widths, which inverts the lesson.
:::

## Next chapter {.unnumbered}

**Bayesian A/B testing for people experiments** — having spent this
chapter on what your data can't tell you, the next one is about the
cleanest case where it can: a comparison where you controlled how people
were assigned.
