# Bayesian A/B Testing for People Experiments

## Welcome back {.unnumbered}

People Analytics runs experiments constantly, even when nobody calls
them that: did the new interview process change the hire rate, did the
flexible-work pilot change retention, does one group's promotion rate
genuinely differ from another's. Unlike a web A/B test with millions of
visitors overnight, these experiments usually have **dozens or
hundreds** of people in each group, and the answer is needed for a
decision, not a p-value.

This chapter builds a Bayesian A/B test from the same Beta-Binomial
machinery you met in Chapter 4, and shows why its output — "there's an
87% chance A beats B, and if it does, it's probably by 2–9 points" — is
usually a better fit for a People Analytics decision than a
significance test.

## What you'll be able to do by the end {.unnumbered}

1.  Set up a Bayesian A/B test for two groups' rates using Beta
    posteriors
2.  Compute **P(A beats B)** directly from posterior samples
3.  Report the size of a likely difference, not just its direction
4.  Compare this to a classical proportions test and to Bayes-Factor
    hypothesis testing, and know when each is the better tool
5.  Apply appropriate care when the "test" involves a protected
    characteristic

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). Navy and light
# navy carry the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

data("promotion", package = "peopleanalyticsdata")

# Guard against missing values before grouping and modelling (Chapter 1
# found gaps in a similar teaching dataset).
promotion <- promotion |> drop_na(diverse, flexible, store, promoted)
```

`promotion` has 1,134 employee records from a retail company: `diverse`
(membership of a diversity group at the company), `flexible` (worked
part-time for 6+ months), `store` (joined working in retail stores),
and `promoted` (the outcome).

------------------------------------------------------------------------

## The question

Promotion-rate comparisons between groups are one of the most common —
and most consequential — comparisons a People Analytics team is asked
to make, often as part of routine equity monitoring. We'll use it as
our worked example:

```{r raw-rates}
# `diverse` may be coded as Y/N or 1/0 depending on data version — normalise
# to an explicit, readable label so every downstream step is unambiguous.
promotion <- promotion |>
  mutate(diverse_grp = if_else(diverse %in% c("Y", "1", 1, TRUE),
                                "diverse_group", "comparison_group"))

group_rates <- promotion |>
  group_by(diverse_grp) |>
  summarise(
    n = n(),
    promoted_n = sum(promoted == 1 | promoted == "Y", na.rm = TRUE),
    raw_rate = promoted_n / n
  ) |>
  mutate(label = if_else(diverse_grp == "diverse_group",
                         "Diversity group", "Comparison group"))       # <1>

group_rates
```

1.  A presentation-ready label alongside the raw coding. Legends and axis
    labels in this book never show a raw variable value — `diverse_grp`
    is for the code, `label` is for the reader.

Describe it, then plot it — Chapter 10's step 2, and worth doing even when
the data is only four numbers, because the plot is where the sample sizes
become visible:

```{r raw-rates-plot}
ggplot(group_rates, aes(x = raw_rate, y = fct_rev(label))) +
  geom_col(fill = navy_light, colour = navy, width = 0.55) +
  geom_vline(xintercept = weighted.mean(group_rates$raw_rate, group_rates$n),
             linetype = "dashed", colour = red) +
  geom_text(aes(label = paste0(promoted_n, " of ", n)),
            hjust = -0.15, size = 3.4, colour = navy) +                # <1>
  scale_x_continuous(labels = scales::percent,
                     expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Raw promotion rate by group",
    subtitle = "Red dashed = the overall rate. Counts shown because the denominators do the arguing.",
    x = "Promotion rate", y = NULL
  )
```

1.  The counts on the bars deliberately. Two rates of 8% and 12% mean
    something completely different at n = 40 than at n = 400, and a bar
    chart of rates alone hides which situation you're in — the single most
    common way a rate comparison misleads a stakeholder.

::: callout-important
## This example is not a randomised experiment, and the distinction matters

Everything below is a valid Bayesian comparison of two rates. It is not
an A/B *test*, because nobody assigned anyone to a group: membership of a
diversity group is a characteristic people arrived with, not a treatment
that was applied. Every confounder Chapter 21 is about — differences in
role, tenure, function, manager — is still fully in play, and the
Beta-Binomial machinery does nothing whatever about them.

The reason to use it as the worked example anyway is that the *arithmetic*
of comparing two rates is identical either way, and this is the version a
People Analytics team is asked for far more often than a true experiment.
The reason to say so out loud is that the method's output — "an 87% chance
one rate is higher" — sounds equally authoritative in both cases, and only
one of them licenses a causal reading.

When you do have random assignment (a training pilot with a randomised
control group, a randomised change to an interview process), the same code
below gives you a genuine causal answer. When you don't, it gives you an
accurate description of a difference whose cause is still open.
:::

::: callout-important
## Handle this kind of comparison with care

A raw or modelled promotion-rate gap between groups is a **starting
signal**, not a conclusion. Real equity analysis needs to account for
role, tenure, performance, and other legitimate factors before drawing
any conclusion about a specific group difference, and — exactly as in
Chapter 16 — usually belongs alongside legal and compensation
expertise, sometimes under legal privilege, before it goes anywhere
near a decision. Everything below is the *statistical method*, not a
complete equity analysis on its own.
:::

------------------------------------------------------------------------

## The Bayesian A/B test

### One posterior per group

Exactly as in Chapter 4, each group's promotion rate gets a Beta
posterior: start from a weakly-informative `Beta(2, 2)` prior (centred
on 50%, easy to move), and update with each group's promotions and
non-promotions. Same habit as always — look at the prior before
combining it with any data:

```{r plot-prior-ab}
tibble(theta = seq(0, 1, length.out = 400)) |>
  mutate(density = dbeta(theta, 2, 2)) |>
  ggplot(aes(theta, density)) +
  geom_area(fill = navy, alpha = 0.5) +
  geom_vline(xintercept = weighted.mean(group_rates$raw_rate, group_rates$n),
             linetype = "dashed", colour = red) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Prior for each group's promotion rate: Beta(2, 2)",
       subtitle = "A gentle hump centred on 50%. Red dashed = the rate the data will actually report.",
       x = "Promotion rate", y = "Prior density")
```

::: callout-note
## A prior centred a long way from the answer

That red line is a long way from the middle of the prior, and it's worth
being explicit about why that's fine here and when it wouldn't be.

`Beta(2, 2)` is worth about **two prior observations per group** — a
single notional promoted and a single notional not-promoted, plus the
uniform baseline. Against denominators in the hundreds, it is
overwhelmed immediately and the posterior sits essentially where the data
puts it. If you knew the organisation's overall promotion rate runs around
10%, a prior like `Beta(2, 18)` would encode that honestly and cost
nothing — and with 30 people per group instead of 500, it would change the
answer materially. Chapter 14 is the version of this argument where the
prior is estimated from your own data rather than picked.

The habit worth keeping: look at where your prior sits relative to the
answer *before* you combine them, so that if the two disagree you find out
now rather than in the meeting.
:::

```{r beta-posteriors}
prior_a <- 2; prior_b <- 2

posteriors <- group_rates |>
  mutate(
    post_alpha = prior_a + promoted_n,
    post_beta  = prior_b + (n - promoted_n)
  )

posteriors
```

```{r posterior-plot}
theta_grid <- seq(0, 0.6, length.out = 400)

curves <- posteriors |>
  reframe(
    theta = theta_grid,
    density = dbeta(theta_grid, post_alpha, post_beta),
    .by = label
  )

prior_curve <- tibble(                                                 # <1>
  theta = theta_grid,
  density = dbeta(theta_grid, prior_a, prior_b)
)

ggplot(curves, aes(x = theta, y = density)) +
  geom_area(data = prior_curve, fill = "grey85", colour = "grey60",
            linetype = "dotted") +
  geom_area(aes(fill = label), alpha = 0.25, position = "identity") +
  geom_line(aes(colour = label), linewidth = 1) +
  scale_colour_manual(values = c("Comparison group" = navy_light,
                                 "Diversity group" = navy)) +
  scale_fill_manual(values = c("Comparison group" = navy_light,
                               "Diversity group" = navy)) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Prior and both posteriors, on one chart",
    subtitle = "Grey dotted = the Beta(2, 2) prior. Coloured curves = each group after seeing its data.",
    x = "Promotion rate", y = "Density", colour = NULL, fill = NULL
  ) +
  theme(legend.position = "top")
```

1.  The prior drawn on the same axes as the posteriors, deliberately —
    Chapter 5's habit, and the clearest single picture of what Bayesian
    updating actually did here. The flat grey hump is what we believed
    beforehand; the two narrow spikes are what 1,100 records did to it.

The gap between the grey and the coloured curves is the whole point of the
exercise: almost all of the information in those posteriors came from the
data, which is exactly what you want a weakly-informative prior to
produce. The gap between the two *coloured* curves is the finding, and how
much they overlap is the honest answer to "is this real."

### The question a p-value doesn't answer directly: P(A beats B)

This is Chapter 5's technique — subtract the draws — applied to two
Beta posteriors instead of two Normal ones. Draw many samples from
each posterior, subtract them element by element, and count how often
the difference lands above zero. Simple, and exactly what most
stakeholders actually want to know:

```{r p-a-beats-b}
n_draws <- 100000

samples <- posteriors |>
  reframe(
    draw = 1:n_draws,
    sample = rbeta(n_draws, post_alpha, post_beta),
    .by = diverse_grp
  ) |>
  pivot_wider(names_from = diverse_grp, values_from = sample, id_cols = draw)


samples <- samples |>
  mutate(diff = diverse_group - comparison_group)

p_higher <- mean(samples$diff > 0)
p_higher
```

```{r diff-plot}
ggplot(samples, aes(x = diff)) +
  geom_histogram(bins = 60, fill = navy_light, colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Posterior distribution of the difference in promotion rates",
    subtitle = paste0("Red dashed = no difference. P(difference > 0) = ",
                      scales::percent(p_higher, accuracy = 0.1)),
    x = "Difference in promotion rate (diversity group − comparison group)",
    y = "Posterior draws"
  )
```

```{r credible-interval-diff}
diff_ci <- quantile(samples$diff, c(0.025, 0.5, 0.975))

diff_ci
```

Those three numbers are the finding, so here they are in the form you'd
actually say them — the difference in **percentage points**, which is the
unit a stakeholder thinks in:

> **The diversity group's promotion rate is
> `r scales::percent(diff_ci[[2]], accuracy = 0.1)` different from the
> comparison group's, with a 95% credible interval running from
> `r scales::percent(diff_ci[[1]], accuracy = 0.1)` to
> `r scales::percent(diff_ci[[3]], accuracy = 0.1)`. The probability the
> diversity group's underlying rate is genuinely higher is
> `r scales::percent(p_higher, accuracy = 0.1)`.**

Read the interval before the probability. If it straddles zero, the
probability statement is telling you which side of the fence the bulk of
the posterior sits on — not that a difference has been established.

::: callout-note
## Two workflow steps that legitimately don't apply here

Chapter 10's eleven steps include diagnosing the sampler (step 7) and a
posterior predictive check (step 8), and this chapter does neither. That's
correct, and worth being explicit about so it doesn't read as a shortcut.

**There is no sampler to diagnose.** A Beta prior updated by binomial data
gives a Beta posterior *exactly* — that's what "conjugate" means. `rbeta()`
above is drawing from a distribution we already know in closed form, not
exploring an unknown posterior by MCMC. There is no `Rhat`, because there
are no chains that could fail to agree.

**The posterior predictive check is nearly trivial.** With two groups, two
counts and two denominators, the model has as many parameters as it has
data points to reproduce. It will match the observed promotion counts
essentially by construction, so a check that it does tells you nothing.

The general lesson, which arrives properly in Chapter 21: the workflow is a
checklist, not a ritual. Skipping a step because it cannot inform you is
different from skipping it because you didn't get round to it — and the
difference is that you can say which one you did.
:::

::: callout-important
## Reporting this properly

Report **three things together**, not a single number: the probability
that the difference runs in the direction you asked about (here
`r scales::percent(p_higher, accuracy = 0.1)` that the diversity group's
rate is the *higher* of the two), the likely *size* of the difference
(the credible interval), and
whether that size is big enough to matter for a real decision. A
99%-confident but tiny difference and an 80%-confident but large one
often call for different actions — collapsing both into "significant /
not significant" throws that distinction away.
:::

------------------------------------------------------------------------

## Is the difference big enough to act on?

A **region of practical equivalence (ROPE)** makes this explicit: pick
the smallest difference in promotion rate you'd actually consider
meaningful (say, 3 percentage points), and ask how much of the
posterior falls outside it.

```{r rope}
rope_width <- 0.03

rope_result <- samples |>
  summarise(
    p_practically_higher = mean(diff > rope_width),
    p_practically_lower  = mean(diff < -rope_width),
    p_practically_equivalent = mean(abs(diff) <= rope_width)
  )

rope_result
```

```{r rope-plot}
#| code-fold: true
ggplot(samples, aes(x = diff)) +
  geom_histogram(bins = 60, fill = navy_light, colour = "white") +
  annotate("rect", xmin = -rope_width, xmax = rope_width,
           ymin = -Inf, ymax = Inf, fill = "grey50", alpha = 0.18) +   # <1>
  geom_vline(xintercept = c(-rope_width, rope_width),
             linetype = "dashed", colour = red) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "The same posterior, with a region of practical equivalence",
    subtitle = paste0(
      "Shaded band = differences too small to act on (±3pp). ",
      scales::percent(rope_result$p_practically_equivalent, accuracy = 0.1),
      " of the posterior falls inside it."),
    x = "Difference in promotion rate", y = "Posterior draws"
  )
```

1.  The ROPE drawn as a band rather than described in a table. This is the
    chart to show a stakeholder, because it converts "is there a
    difference" into "is the difference in the zone we said we'd care
    about" — a question they set the terms of, not you.

Read the three numbers as a set that has to add to one. The two outer
columns are the probability the difference is **big enough to matter** in
each direction; the middle column is the probability it is real but too
small to act on, or absent altogether.

> **`r scales::percent(rope_result$p_practically_equivalent, accuracy = 1)`
> of the posterior falls inside ±3 percentage points — the range we agreed
> in advance was too small to change anything. The probability the gap is
> larger than that in either direction is
> `r scales::percent(rope_result$p_practically_higher + rope_result$p_practically_lower, accuracy = 1)`.**

This reframes "is there a difference" as "is there a difference big
enough to be worth acting on" — usually the actual question behind the
request. The critical procedural point: **pick the ROPE width before you
see the posterior.** Choosing it afterwards, once you know where the
distribution sits, is p-hacking with different vocabulary.

------------------------------------------------------------------------

## How this compares to other approaches

::: callout-note
## A different lens: the classical two-proportion test

`prop.test()` or a chi-square test asks: *if the two groups truly had
the same promotion rate, how unusual would data this different be?*
— producing a p-value. The Bayesian version above asks the question
most stakeholders actually mean when they say "is this real": *given
the data we've seen, how likely is it that the rates genuinely
differ, and by how much?* With the sample sizes typical of People
Analytics work, the two methods usually agree on direction; the
Bayesian version's edge is answering the second, size-and-confidence
question directly, in one step, without a separate power calculation.
:::

```{r classical-comparison}
prop.test(x = group_rates$promoted_n, n = group_rates$n)
```

Worth reading against the Bayesian output above, line by line, because the
two are closer than the philosophical argument suggests:

- **`p-value`** — the probability of seeing a gap at least this large *if
  the two rates were truly identical*. Note the direction of that
  sentence. It is not the probability that the rates are the same, and it
  is not the probability the gap is real, which is what
  `r scales::percent(p_higher, accuracy = 0.1)` above actually gives you.
- **`95 percent confidence interval`** — compare it to the credible
  interval from `quantile(samples$diff, ...)`, but mind the direction of
  the subtraction. `prop.test()` works down the rows of `group_rates` and
  reports comparison group *minus* diversity group; `samples$diff` runs
  the other way round. The two intervals are mirror images, so flip the
  sign on one before you line them up. Once you do: with hundreds of
  observations per group and a weak prior, the two land in nearly the
  same place, which is the honest version of "these methods usually
  agree." What differs is what you're allowed to say about the interval:
  the credible interval is a statement about where the difference probably
  is, the confidence interval is a statement about the procedure that
  produced it.
- **`sample estimates`** — the two raw rates, unchanged from
  `group_rates`. No shrinkage, no prior, no pooling.

The one thing `prop.test()` does not give you at all is the ROPE question.
There is no way to ask it "how likely is the gap bigger than three
percentage points," because it only ever tests against exactly zero.

::: callout-note
## Why this book uses estimation, not Bayes Factors, for this kind of test

Keith McNulty's book covers **Bayesian hypothesis testing** using
**Bayes Factors** — comparing the evidence for "the rates are equal"
against "the rates differ" as a ratio, the Bayesian analogue of a
p-value-producing test. It's a legitimate and well-developed approach,
useful in particular when the actual question is *"is there truly a
non-zero effect at all?"* — common in scientific replication contexts,
and covered in his Chapter 12. This book leans instead on the
estimation approach above (posteriors, credible intervals, and a ROPE),
because the questions People Analytics usually needs answered are
"how big" and "are we confident enough to act," not "is the effect
exactly zero" — and estimation answers those directly, in the same
units as the decision. Bayes Factors can also be sensitive to exactly
how you specify the alternative hypothesis, which adds a layer of
judgment estimation avoids. Neither is "the correct Bayesian method" —
pick the one that matches the actual question.
:::

::: callout-tip
## For the ML/DS crowd

If you've heard of **Thompson sampling** or **multi-armed bandits** —
used in live experimentation platforms and recommender systems to
decide which variant to show next — you've seen this exact method
running continuously. Thompson sampling literally draws one sample
from each arm's Beta posterior and picks whichever arm's sample is
highest, repeating as data arrives; it's the sequential, automated
sibling of the one-shot comparison in this chapter. The maths is
identical — sampling from Beta posteriors and comparing — just wired
into a live decision loop instead of a one-time analysis.
:::

::: callout-caution
## Try it yourself

Rerun this comparison for `flexible` vs. not, and separately for
`store` vs. not. Do either of those produce a clearer signal than
`diverse` did? Report your answer the way you would to a stakeholder —
probability of a real difference, likely size, and whether it clears
your ROPE.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Most People Analytics "did X work" questions are exactly this shape —
two groups, a rate, and a decision to make with a modest sample. This
method scales down gracefully to small groups (unlike a chi-square test,
which gets unreliable with few observations per cell), gives you a
direct, presentable probability statement, and — via the ROPE — keeps
you honest about the difference between "statistically distinguishable"
and "big enough to matter."
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  A Bayesian A/B test puts a Beta posterior on each group's rate and
    compares them **directly by sampling**, rather than via a
    significance test.
2.  **P(A beats B)** and the **credible interval on the difference**
    together answer "is it real" and "how big," in one step.
3.  A **ROPE** turns "is there a difference" into the more useful "is
    the difference big enough to act on."
4.  This book favours estimation over Bayes-Factor hypothesis testing
    for exactly this kind of question — see the note above for when
    you'd reach for the other approach instead.
5.  Comparisons involving protected characteristics need the same care
    as the pay analysis in Chapter 16 — a statistical signal, not a
    complete equity conclusion on its own.
6.  The **arithmetic is the same whether or not anyone was randomised, and
    the licence to interpret it isn't.** With random assignment this is a
    causal answer; without it — as in this chapter's worked example — it
    is an accurate description of a difference whose cause is still open,
    and Chapters 21 and 22 are what you reach for next.
7.  Set the **ROPE width before you look at the posterior**. Choosing it
    afterwards is p-hacking in different clothes.
:::

## Next chapter {.unnumbered}

**Causal structure: drawing the graph** — an A/B test is the easy case,
because you controlled who went into which group. The next two chapters
are about the far more common situation where you didn't, and what can
still be said.
