# The Elicitation Workflow: Building Priors With Your Stakeholders

## Welcome back {.unnumbered}

Every chapter so far has built a prior one of two ways: a weakly
informative default (Chapter 5), or one estimated from data — your own
groups (Chapter 14) or an external benchmark (also Chapter 14). There's
a third source, often better than both, that this chapter is built
around: **asking the people who actually know**.

This isn't just a statistical technique. Done well, it's an
organisational one — the difference between presenting a model to
stakeholders at the end, and having them recognise their own judgment
inside it from the start.

## What you'll be able to do by the end {.unnumbered}

1.  Run a structured **elicitation** session using SHELF's quartile
    questions, which a non-statistician can actually answer
2.  Fit a formal probability distribution to elicited judgements with
    the `SHELF` package
3.  Combine several experts' judgements into one pooled prior
4.  Run a **prior sensitivity analysis** across experts, to see whether
    the data makes the conclusion robust to whose judgement you used
5.  Feed an elicited prior directly into a `brms` model
6.  Explain why involving stakeholders in the prior changes how they
    receive the posterior
7.  Use a prior's width as a quick, principled way to prioritise which
    analyses are worth commissioning next

------------------------------------------------------------------------

## Setup

```{r setup}
#| eval: false
install.packages("SHELF")
```

```{r setup-real}
library(tidyverse)
library(SHELF)
library(brms)
library(tidybayes)

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). The navy ramp
# carries the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_mid   <- "#3d68a8"
navy_light <- "#8fabd0"
red        <- "#d32f2f"
```

This chapter uses the **`SHELF`** package (the Sheffield Elicitation
Framework, built by Jeremy Oakley and Anthony O'Hagan) — a set of
tools for turning expert judgement into a fitted probability
distribution, widely used in health economics and risk analysis. It's
a genuinely different tool from anything else in this book: the input
isn't a dataset, it's a conversation.

A distinction worth drawing at the start, because it is easy to miss.
SHELF is two things: a **protocol** for running the conversation, and
an **R package** that fits distributions to whatever numbers the
conversation produced. This chapter uses both. The questions in the
next section are the ones SHELF prescribes, and I recommend them over
the alternatives — including the version I used myself for years,
which gets its own callout further down.

------------------------------------------------------------------------

## The problem elicitation solves

You're about to model the expected effect of a new manager-training
programme on first-year attrition. There's no historical data on *this*
programme yet — but your Head of L&D has run three similar
interventions before and has a real, informed opinion about what to
expect. A default weakly-informative prior throws that opinion away. A
prior "estimated from data" doesn't apply, because there isn't any yet.

::: callout-important
The Head of L&D's judgement **is** data — it's just data in someone's
head instead of a spreadsheet. Elicitation is the structured process
for getting it out in a form a model can use.
:::

------------------------------------------------------------------------

## Asking the right questions

Nobody can answer "what's your prior probability distribution for the
effect size?" — that's not how people think. They *can* answer
calibrated questions about specific values:

::: callout-note
## The quartile method — ask in this order

1.  **"What's the lowest the true reduction could plausibly be? And
    the highest?"** — the **plausible limits**, asked first and
    deliberately as a range you'd be astonished to fall outside.
2.  **"Given that range, what value would you say the truth is equally
    likely to fall above or below?"** — the **median**.
3.  **"Now think only about the lower half. What value splits *that*
    in two?"** — the **lower quartile**. Then the same for the upper
    half — the **upper quartile**.

Two things make this work, and both are deliberate.

**The limits come before the best guess.** Ask for a central estimate
first and everything afterwards gets anchored to it. Ask for the range
first and the expert has to think about how wrong they could be before
they commit to being right.

**Each question splits a range in half.** People are noticeably better
at "is it above or below this?" than at "how likely is this?", and the
quartile questions are just that judgement applied twice.

SHELF also offers the **tertile** method (thirds instead of quarters)
and the **roulette** method, where the expert distributes a fixed
number of chips across bins to draw their own distribution. Use
roulette with a group and a whiteboard; it is the easiest of the three
for people who dislike numbers.
:::

::: callout-warning
## Why I no longer ask for the 10th and 90th percentiles

For years I used a three-number version — 10th percentile, median,
90th percentile — which I adapted from Douglas Hubbard's *How to
Measure Anything*. Hubbard's version is wider still: a **90% confidence
interval**, which is the 5th and 95th percentiles. If you have read
that book, or seen the 10/90 version in something of mine, this is the
correction.

The problem is the same for both, and it is the reason SHELF **refuses
to ask for tail quantiles at all**. Judging the far tail of your own
uncertainty is the specific thing people are worst at, and the further
out you ask, the more exposed the answer is. The quartiles are the
least extreme of the three formats. Hubbard's 5th and 95th are the
most.

The evidence is blunt. Soll and Klayman found subjective intervals
sometimes only **40% as wide as they needed to be** — and, the finding
worth remembering, three formats that look almost identical produced
roughly **40 percentage points of difference in overconfidence**. *How*
you ask moves the answer by more than *who* you ask.

**In fairness to Hubbard, he has an answer to this and it is a good
one.** His method is not "ask for a 90% interval"; it is "train people
until their 90% intervals actually contain the truth 90% of the time",
using equivalent-bet tests and repeated scored feedback. Calibration
training demonstrably works. If you are willing to run it, the format
matters much less. Most organisations are not going to run it, and an
untrained 90% interval is the worst of the three options rather than
the best.

So: **ask for quartiles.** If you inherit numbers already collected as
10/50/90, don't rerun the session — `fitdist()` takes whatever `probs`
you give it. Just don't design a new one that way.
:::

Suppose your Head of L&D says the reduction in first-year attrition
could plausibly be anywhere from nothing at all up to about 25
percentage points; that 5 points splits that range in two; and that
the halves either side of 5 split at **2.5** and **10**.

Three numbers to fit, then, plus a lower limit: 2.5 (lower quartile),
5 (median), 10 (upper quartile).

### Fitting a distribution to those three numbers

```{r fit-single-expert}
judgements <- fitdist(
  vals  = c(2.5, 5, 10),
  probs = c(0.25, 0.5, 0.75),   # <1>
  lower = 0     # a reduction can't be negative in this framing
)

judgements$Normal
judgements$best.fitting
```

1.  The quartiles, per the protocol above. `fitdist()` will accept any
    probabilities you hand it — the argument for using these ones is
    about the conversation, not the code.

`fitdist()` fits several candidate distributions (Normal, log-normal,
gamma, beta, Student-t) to the three elicited points and reports which
fits most closely.

Two outputs, and they answer different questions:

- **`judgements$Normal`** gives the `mean` and `sd` of the best-fitting
  Normal. This is the pair you'll hand to `brms` later, because
  `prior(normal(...))` is what a `brms` prior wants. It says the Head of
  L&D's belief, forced into a Normal shape, is centred around
  `r round(judgements$Normal$mean, 1)` points with a standard deviation of
  about `r round(judgements$Normal$sd, 1)`.
- **`judgements$best.fitting`** names the distribution family that
  reproduces her three numbers most faithfully. With an asymmetric set of
  judgements like 2.5 / 5 / 10 — the gap upward being twice the gap
  downward — this is often *not* the Normal, because a Normal can't be
  skewed. That mismatch is information: she thinks the upside is longer
  than the downside, and a Normal prior discards that.

::: callout-note
## Which one should you actually use?

The honest answer is that it depends on how much the skew matters, and the
way to find out is to check rather than assume.

Using `judgements$Normal` is the pragmatic default — it's what plugs
straight into `brms`, and with any real quantity of data the difference
between a Normal and a skew-Normal prior is invisible in the posterior.
Using the best-fitting family means either finding a `brms` prior that
matches it or writing a custom one, which is real work.

The check that settles it: fit the model both ways and compare the
posteriors. That's the same sensitivity logic this chapter applies to
*whose* prior you use, applied instead to *which shape* you fitted it
with — and if the answer moves, you've learned that the skew was carrying
weight and is worth the effort.
:::

Plot it to check the shape makes sense — and, more
importantly, to **show the expert**:

```{r plot-fit}
plotfit(judgements, d = "best", lp = TRUE,
        showPlot = FALSE, returnPlot = TRUE) +      # <1>
  geom_vline(xintercept = judgements$limits$lower,
             linetype = "solid", colour = "grey40") +
  geom_vline(xintercept = judgements$Normal$mean,
             linetype = "dashed", colour = red) +
  theme_minimal(base_size = 13)

theme_set(theme_minimal(base_size = 13))     # <2>
```

1.  The dashed red line is the **Normal** fit's mean, drawn against the
    *best-fitting* curve. With only three elicited points the best fit
    is often skew-normal, and a skewed distribution's mean does not sit
    at its peak — so the two deliberately disagree. That gap is the
    skew, made visible.
2.  `plotfit()` resets ggplot's global theme as a side effect. Put it
    back, or every later chart in the chapter changes appearance.

```{r feedback}
feedback(judgements, quantiles = c(0.1, 0.25, 0.5, 0.75, 0.9))
```

That table is the single most useful thing to put in front of the expert,
and it repays a moment on how to read it. Each **column** is a fitted
distribution family; each **row** is a quantile. The numbers are what that
fitted distribution says the value at that quantile is.

So find the `0.25`, `0.5` and `0.75` rows and compare them back to the
2.5, 5 and 10 she actually said. A good fit reproduces those three almost
exactly — that's what "fitted" means. The rows that earn their keep are the
ones she *didn't* give you: `0.1` and `0.9`. Those are the fit's own
extrapolation out into the tails, a claim she never made, and they're the
natural thing to test with her: *"this says there's a one-in-ten chance the
reduction comes in above [the 0.9 row] — does that sound right to you?"*

If she balks at that number, the three elicited values need revisiting.
That is the loop working, not a problem with the tool.

::: callout-important
## The feedback loop is the whole point

Show the fitted curve back to the expert and ask: *"does this look
like what you meant?"* Experts are often surprised by how fat or thin
the tails of the fitted distribution are compared to what they
intended — and adjusting the elicited numbers until the fitted curve
actually matches their belief is where the real value of this process
lives. Skipping this step and just taking the first fitted distribution
defeats the purpose.
:::

::: callout-tip
## For the ML/DS crowd: LLMs as a mechanical fitting shortcut

Turning three elicited numbers into distribution parameters is a
solved, mechanical curve-fitting problem — which makes it a
reasonable task to hand to an LLM if you're prototyping and don't
have `SHELF` to hand: describe the elicited quartiles and median
and ask for the Beta(α, β) or Gamma(α, β) that reproduces
them. It lowers the technical bar without changing anything about the
process itself — you should still verify the distribution you get
back actually reproduces your three numbers, and still show the
fitted curve back to the expert. For real work, `SHELF::fitdist()`
remains the better choice: it fits and compares several candidate
families at once and was purpose-built for exactly this job.
:::

------------------------------------------------------------------------

## Multiple experts, one pooled prior

Real decisions rarely rest on one person's judgement. The fastest
version of this is to elicit from several experts **independently**
(critically: before they discuss it with each other, to avoid one
confident voice anchoring everyone else), then pool:

```{r multiple-experts}
# Rows = the three elicited points; columns = one expert each
vals_matrix <- matrix(
  c(2.5, 5, 10,    # Head of L&D
    2.8, 4,  6.5,  # a line manager who ran a pilot
    0.5, 3,  9),   # an external L&D consultant, more uncertain
  nrow = 3, ncol = 3
)

probs <- c(0.25, 0.5, 0.75)

group_fit <- fitdist(vals = vals_matrix, probs = probs, lower = 0)

plotfit(group_fit, d = "normal", lp = TRUE, xu = 20,   # <1>
        showPlot = FALSE, returnPlot = TRUE) +
  geom_vline(xintercept = group_fit$limits$lower,
             linetype = "solid", colour = "grey40") +
  geom_vline(xintercept = group_fit$Normal$mean,
             linetype = "dashed", colour = red) +
  theme_minimal(base_size = 13)

theme_set(theme_minimal(base_size = 13))
```

1.  `xu` sets the right-hand edge of the axis by hand. We gave
    `fitdist()` a lower limit but no upper one, so the fitted curves run
    to infinity and `plotfit()`'s own choice of axis limit fails on some
    SHELF versions. Any finite value a little above the largest elicited
    number works; there is nothing statistical in the choice.

::: callout-important
## A practical trap: don't let an elicited value sit exactly on the bound

Notice the third expert's low estimate above is `0.5`, not `0`, even
though `lower = 0` here. If an elicited value is *exactly equal* to the
bound you've set, several of the candidate distributions `fitdist()`
tries (log-normal, gamma) need values strictly greater than that bound
— they work with the log of the gap between the value and the bound,
and `log(0)` is `-Inf`. The optimiser then fails with a rather
unhelpful `non-finite value supplied by optim` error that gives no hint
the real cause was a boundary value.

In practice, experts often say "zero" as shorthand for "as low as it
could plausibly go," not literally zero. If you're given that answer,
swap in a small non-zero value instead (0.5, or something like 1% of
the scale you're working in) — it keeps the same intent without
sitting exactly on a boundary the fit can't handle.
:::

::: callout-note
## The fuller version: a Delphi-style round

Eliciting once and pooling, as above, is fine for a fast, low-stakes
prior. For a decision with real stakes, the classic **Delphi method**
adds two more steps worth the extra hour:

1.  **Elicit individually** — each expert's limits, median and
    quartiles, before anyone sees anyone else's numbers.
2.  **Share the spread anonymously** — everyone sees the range of
    answers, not who gave which one.
3.  **Discuss the outliers out loud.** Someone at the extreme often
    holds private or local information the others don't — understanding
    *why* they differ is frequently as valuable as the pooled number
    itself.
4.  **Re-elicit a second round** in light of that discussion, then
    pool.

I run it towards **calibration, not consensus** — if experts are still
spread out after a second round, that's telling you something true
about how much the organisation actually knows, not a process failure.

That is my departure from Delphi, not a description of it, and it is
worth being straight about. The original method was built at RAND to
produce *reliable consensus*, and treating persistent spread as a
finding rather than a failure inverts its purpose. Be straight about
the evidence too: Sackman's 1974 RAND critique called Delphi
unreliable and unvalidated, and Rowe and Wright's later tally gives it
twelve wins to two against simply averaging a single round — better
than nothing, with no consistent advantage over other structured
procedures. Use it because structure beats a meeting, not because the
method has been shown to be the best one.
:::

::: callout-tip
## Disagreement is information, not a problem to hide

If your three experts' ranges barely overlap, that's a real finding —
it says the organisation doesn't actually have a shared view of what to
expect, which is worth surfacing *before* the training programme
launches, not after it underperforms someone's private expectation.
Sometimes the disagreement isn't really about the number at all: two
experts can give different answers because they're implicitly reasoning
from different causal models of what drives attrition, and the
elicitation session is what surfaces that.
:::

::: callout-warning
## Averaging confident disagreement produces a prior nobody holds

The obvious thing to do with three curves is average them, and it has
a failure mode worth seeing before you rely on it.

Take two experts who are each **confident** and who **disagree** — one
says the reduction is around 3 points, the other around 12, and both
give narrow ranges. Average their distributions and you get a
**bimodal** prior: two humps with a valley between them. Read as a
statement of belief, it says the truth is likely to be near 3 or near
12 and *unlikely* to be near 7 — which is a claim neither expert made
and nobody in the room would sign.

The fix is not a cleverer weighting. It is to build **one distribution
together** rather than averaging separate ones, which is what SHELF's
"Rational Impartial Observer" device exists to do: the group is asked
what a reasonable neutral person would believe having heard all the
arguments, and that single distribution is what gets fitted.

The reassuring counterweight, from a study that compared them
directly: every aggregation method beat every individual expert under
every scoring rule. **That you aggregated matters more than how.** Just
look at the pooled curve before you use it, and if it has two humps,
go back to the room.
:::

------------------------------------------------------------------------

## An alternative to pooling: does whose prior you use change the answer?

Pooling turns several experts' judgements into one prior. A different,
often more useful question is: **if I'd used each expert's prior on
its own, would we end up recommending something different?** Instead
of running the model once with a pooled prior, run it once *per
expert's* prior and compare the results. That turns disagreement
between experts into a direct answer to two questions a decision-maker
cares about far more than the prior itself:

1.  Does the pilot data dominate, so the conclusion barely depends on
    whose prior you started from? If so, that's reassuring, and worth
    saying explicitly: *"the result is robust to who you believed
    beforehand."*
2.  Or does the choice of prior still meaningfully swing the
    conclusion, even after seeing the data? If so, that's not a
    problem to hide — it's telling you the decision genuinely still
    rests on whose judgement you trust, and stakeholders deserve to
    see that openly rather than have it buried inside a single pooled
    number.

Recall the running example from earlier in this chapter: no data yet
exists for a training programme that hasn't launched. To actually run
the sensitivity check below rather than just describe it, here's a
small **simulated** stand-in for the pilot that would eventually
produce that data — clearly not real, but enough to show the mechanics
properly. The next section's model reuses this same simulated pilot.

```{r simulate-pilot}
set.seed(2026)

training_pilot_data <- tibble(
  condition = factor(rep(c("Control", "Trained"), each = 20),
                      levels = c("Control", "Trained"))
) |>
  mutate(
    attrition_reduction = if_else(
      condition == "Trained",
      rnorm(n(), mean = 4, sd = 5),
      rnorm(n(), mean = 0, sd = 5)
    )
  )
```

```{r sensitivity-analysis}
# One brms fit per expert's individually elicited prior, instead of
# one fit using a pooled prior.
sensitivity_fits <- group_fit$Normal |>
  as_tibble(rownames = "expert") |>
  pmap(function(expert, mean, sd) {
    fit <- brm(
      attrition_reduction ~ 1 + condition,
      data = training_pilot_data,
      prior = prior_string(
        paste0("normal(", mean, ",", sd, ")"),
        class = "b", coef = "conditionTrained"
      ),
      chains = 4, iter = 2000, refresh = 0
    )
    fixef(fit)["conditionTrained", ] |>
      as_tibble_row() |>
      mutate(expert = expert, .before = 1)
  }) |>
  list_rbind()

sensitivity_fits
```

::: callout-tip
## Reading this table

Look at the spread of `Estimate` across experts, and how much their
`Q2.5`–`Q97.5` intervals overlap — not any single row. If all three
posteriors land in roughly the same place despite starting from
noticeably different priors, that's the finding: the pilot data was
informative enough to make the conclusion robust to whose starting
judgement you used. If they still disagree meaningfully, the prior is
still doing real work — the choice of whose judgement to trust (or how
to pool it) is still a live decision, not a formality.
:::

### The chart that makes the point

Three rows of numbers is the wrong format for a comparison whose whole
content is *how much things moved*. Put each expert's prior next to the
posterior it produced, and the answer is immediate:

```{r sensitivity-plot}
#| code-fold: true
expert_labels <- c("Head of L&D", "Line manager", "External consultant")

prior_intervals <- group_fit$Normal |>
  as_tibble(rownames = "expert") |>
  transmute(
    expert = expert_labels[row_number()],
    stage = "Prior (elicited)",
    estimate = mean,
    lower = qnorm(0.025, mean, sd),
    upper = qnorm(0.975, mean, sd)
  )

posterior_intervals <- sensitivity_fits |>
  transmute(
    expert = expert_labels[row_number()],
    stage = "Posterior (after the pilot)",
    estimate = Estimate,
    lower = Q2.5,
    upper = Q97.5
  )

bind_rows(prior_intervals, posterior_intervals) |>
  mutate(stage = factor(stage, levels = c("Prior (elicited)",
                                          "Posterior (after the pilot)"))) |>
  ggplot(aes(x = estimate, y = fct_rev(factor(expert)), colour = stage)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  geom_linerange(aes(xmin = lower, xmax = upper),
                 linewidth = 2, alpha = 0.85,
                 position = position_dodge(width = 0.55)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.55)) +
  scale_colour_manual(values = c("Prior (elicited)" = navy_light,
                                 "Posterior (after the pilot)" = navy)) +
  labs(
    title = "Three experts' priors, and where the pilot data took each of them",
    subtitle = "Red dashed = no effect. Light bars = what each expert believed; dark bars = what the model concluded.",
    x = "Reduction in first-year attrition (percentage points)",
    y = NULL, colour = NULL
  ) +
  theme(legend.position = "top")
```

::: callout-important
## What to look for, in this order

**How far each dark bar sits from its light one.** That distance is what
the data did. Large movement means the pilot was informative; barely any
movement means the prior is still driving the answer, and with 40
observations in this simulated pilot you should expect somewhere in
between.

**How much the three dark bars overlap each other.** This is the actual
deliverable of a sensitivity analysis. Three posteriors sitting on top of
one another means the conclusion is robust to whose judgement you started
from — say that out loud, because it's the strongest thing you can say
about a result built on elicited priors. Three that remain visibly apart
means the recommendation genuinely still depends on whose view you trust,
and that belongs in front of the decision-maker rather than averaged away.

**Whether any dark bar crosses the red line while others don't.** The
uncomfortable case, and the one worth checking for explicitly: the
programme looks worth funding on one expert's prior and not on another's.
That is not a modelling failure. It is a finding about how much the
organisation actually knows, and it is far better to discover it here than
in the meeting where the budget gets signed.
:::

::: callout-important
## Why this is worth showing back to your experts

This is one of the most direct ways to make elicitation land with a
sceptical stakeholder. Show the Head of L&D specifically: *"here's
what the model concludes using your prior alone, and here's what it
concludes using the line manager's, or the consultant's."* If their
own number would have led to a materially different recommendation,
that's a concrete, honest answer to "did my input actually matter?" If
it wouldn't have — if the data swamps the disagreement — that's just
as useful to know, and it reassures everyone that the conclusion
doesn't quietly hinge on one person's opinion.
:::

::: callout-note
## A different lens: this is a tipping-point sensitivity analysis

Rerunning the same model under different assumptions and watching
whether the conclusion changes isn't a Bayesian-only idea — it's the
same instinct behind a one-way sensitivity analysis or tornado diagram
in classical decision analysis, or behind checking whether a
regression result survives a different choice of controls. What's
specifically Bayesian here is that the "assumption being varied" is a
formally elicited prior rather than an arbitrary modelling choice — so
this sensitivity check and the elicitation process earlier in this
chapter directly reinforce each other.
:::

------------------------------------------------------------------------

## Putting the elicited prior to work

The whole point is to use this in an actual model — take the fitted
parameters straight into `brms`:

```{r use-in-model}
# Reuses the simulated pilot from the previous section - this is the
# single-expert version (judgements), fed straight into brms, of the
# same idea the sensitivity check just ran once per expert.
fitted_normal <- judgements$Normal

fit_training <- brm(
  attrition_reduction ~ 1 + condition,
  data = training_pilot_data,
  prior = prior_string(
    paste0("normal(", fitted_normal$mean, ",", fitted_normal$sd, ")"),
    class = "b", coef = "conditionTrained"
  ),
  chains = 4, iter = 2000, refresh = 0
)

fixef(fit_training)
```

### Check the elicited prior before you trust it

An elicited prior gets no exemption from Chapter 10's step 5. If anything
it needs the check more than a weakly-informative default does, because an
informative prior is by definition capable of doing damage. So: what does
this prior, combined with the rest of the model, think the pilot data
could look like?

```{r prior-predictive-elicited}
#| code-fold: true
n_sim <- 4000

tibble(
  baseline = rnorm(n_sim, 0, 10),                                     # <1>
  effect   = rnorm(n_sim, fitted_normal$mean, fitted_normal$sd),      # <2>
  sigma    = rexp(n_sim, 0.2),
  trained  = rnorm(n_sim, baseline + effect, sigma)
) |>
  ggplot(aes(trained)) +
  geom_histogram(bins = 60, fill = navy, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  labs(
    title = "What the elicited prior thinks a trained group's outcome could be",
    subtitle = "Red dashed = no change. Simulated before seeing any pilot data.",
    x = "Simulated reduction in first-year attrition (percentage points)",
    y = "Simulated draws"
  )
```

1.  `brms`' default intercept prior, roughly — the control group's
    baseline, about which we elicited nothing.
2.  The elicited prior, the one part of this that came from a person.

::: callout-important
## Two things this catches, and one of them is political

The statistical check first: does the simulated distribution stay inside
the range of outcomes the measure can actually take? A reduction in
first-year attrition of 60 percentage points is not possible in an
organisation with 20% first-year attrition, and if the prior thinks it is,
either the elicited numbers or the scale got misunderstood somewhere.

The second is the one worth the extra ten minutes. **Show this chart back
to the expert too.** She answered three questions about where a range
splits; this
is what those three answers imply about the actual pilot result, which is
not a translation anyone does in their head. If she looks at it and says
"no, I'd be astonished by anything above eight points," the elicitation
isn't finished — and you have found that out before the prior is doing work
inside a model with her name attached to it.

That's the same feedback loop as `plotfit()` and `feedback()` earlier,
extended one step further: from *does this curve match your belief about
the parameter* to *does this match what you'd expect to see happen*.
:::

### Reading the result

```{r read-training}
fixef(fit_training) |>
  as_tibble(rownames = "term")
```

Two rows. `Intercept` is the control group's mean — no training, and
nothing much expected to happen, so it should sit near zero.
`conditionTrained` is the finding: the estimated additional reduction in
first-year attrition for the trained group, with `Q2.5` and `Q97.5` giving
the 95% credible interval around it.

Compare that `Estimate` against the elicited prior mean of about
`r round(fitted_normal$mean, 1)` points. Where it lands between the prior
and what the pilot alone would have said is the whole mechanic of Bayesian
updating made visible in one number — and it is exactly the comparison the
Head of L&D will make first, because it answers "did what I said count for
anything?"

This directly closes the loop from Chapter 5: instead of picking a
weakly-informative default because you didn't have anything better,
you now have a prior that's genuinely informed — and that a named
person in the business would recognise and stand behind.

------------------------------------------------------------------------

## What a prior's width tells you about what to analyse next

There's a second, less obvious payoff to running this process
properly: an elicited prior doesn't just feed a model — its **width**
tells you something about whether the analysis is worth commissioning
at all.

This section is the informal version of something Chapter 24 does
arithmetically. Here you compare prior widths by eye to decide what to
work on next; there, the same question is answered in pounds, computed
from the posterior draws themselves. If the idea below appeals, that is
where it is finished.

::: callout-tip
## The logic, in plain terms

If your experts' pooled prior is already narrow — everyone roughly
agrees what the answer is — a full analysis has little room to change
anyone's mind, however rigorously it's done. If the pooled prior is
wide — real disagreement, real stakes — that's exactly the situation
where spending analyst time to narrow it down pays off.

The parameter everyone argues about is, mechanically, the one worth
investigating. That is the whole idea, and you can act on it with
nothing more than the elicited widths in front of you.
:::

You don't need to compute anything formally to get that benefit.
Simply comparing pooled prior widths across a list of competing project
requests gives a People Analytics team a principled way to answer
"which of these should we analyse first?" — a tightly clustered prior
means the organisation basically already knows the answer; a wide one,
especially on a high-stakes question, is where a proper analysis has
the most to add.

::: callout-note
## The four quantities, and which one answers which question

Decision analysts have formalised this, and the names are easy to
confuse — I have confused them myself. They are four different
quantities answering four different questions, and they are used in
this order:

| | Question it answers |
|---|---|
| **EVPI** | Is *any* further evidence worth anything at all? Use it to screen. |
| **EVPPI** | *Which* uncertain input carries the decision? Use it to triage. |
| **EVSI** | What is *this specific study*, with this design and this sample size, worth? Use it to price. |
| **ENBS** | Population EVSI minus what the study costs. Use it to decide. |

The one people reach for when they mean "what is this piece of
research worth" is **EVSI**, not EVPPI. EVPPI tells you which
*parameter* to go after; EVSI tells you what a particular attempt to
go after it would be worth.

Chapter 24 computes the first two directly from posterior draws — both
turn out to be a couple of lines of R — and shows why the answer often
points at a cost input rather than at the effect the analyst wanted to
improve.
:::

::: callout-note
## A different lens: this is a decision-analysis idea wearing Bayesian clothes

Value of information comes from decision analysis and health
economics, not from Bayesian statistics specifically — Raiffa and
Schlaifer set out EVPI and EVSI in 1961, long before `brms` or modern
computing. But it depends entirely on having a prior to measure the
width of, which is exactly what this chapter's elicitation process
hands you for free. It's a good example of how a useful idea travels
between fields once you have the right building block in hand.

It also travels less far than it should. Health economics invented
this machinery and still does not use it routinely. The obstacle is
not technical.
:::

------------------------------------------------------------------------

## Why do this at all: the organisational case

::: callout-important
## Elicitation is also a change-management tool

A stakeholder who contributed their own judgement to the prior has a
fundamentally different relationship to the posterior than one seeing
the model for the first time in a results deck. They can see exactly
where their input mattered, ask informed questions about what moved
and why, and — because they were part of building it — are far more
likely to trust and act on the result. Running the elicitation session
*is* stakeholder management, not a detour from it.
:::

::: callout-note
## A different lens: this isn't unique to Bayesian statistics

Structured expert-judgement elicitation predates Bayesian statistics
in practice — the **Delphi method** (anonymous, iterative rounds of
expert forecasting, developed at RAND in the 1950s) is a
philosophy-agnostic ancestor of the same instinct: judgement, gathered
systematically, beats judgement gathered informally. What Bayesian
elicitation adds is a formal destination for that judgement — a
probability distribution that plugs directly into a model — rather
than a converged point forecast.

Delphi's serious rival is worth knowing about even though this book
doesn't use it. **Roger Cooke's classical model** scores experts on
*seed questions* whose answers are already known, then weights each
expert by how well calibrated and how informative they proved to be —
with poor calibrators weighted all the way to zero. It is the
uncomfortable idea in this field, because it means the most senior
voice in the room can be formally worth nothing. Carry the counterweight
honestly: it performs less well out of sample than it does in, and the
aggregation finding above suggests the choice of method matters less
than doing it at all.
:::

::: callout-tip
## For the ML/DS crowd

Formal elicitation is close cousin to a few things you may already
do informally: setting an informed **search space and starting point**
for Bayesian hyperparameter optimisation based on past experience
rather than a blind uniform range, or **warm-starting** a model with
domain-informed initial parameters. The difference here is rigor — a
documented, feedback-checked distribution instead of "I'll just set the
range to something that feels about right." In safety-critical Bayesian
ML applications (clinical trial design, risk modelling), formal
elicitation exactly like this is often a required, audited step, not an
optional nicety.
:::

------------------------------------------------------------------------

## The other thing you can elicit

Everything so far has elicited a **quantity**: how high might this rate
be, how confident are you, where are the edges. That is the obvious
thing to ask an expert for, and it is not the only thing they know.

Go back to Chapter 21. Before that chapter could estimate anything, it
had to state which variables belonged in the model — and it made the
point that no fit statistic can decide this, because it is a question
about cause and effect. Somebody has to say what they believe causes
what.

That somebody is the same person sitting in this workshop.

::: callout-important
A DAG is an elicited object.

It is not a statistical result. It is a written record of what a group
of people believe about how their organisation works, drawn before the
data is touched — which is exactly what a prior is, pointed at
*structure* instead of at *magnitude*.

The two halves of Part V are therefore the same move, done twice.
Chapter 21 elicits the shape of the problem; this chapter elicits the
numbers in it. Both are ways of getting judgement onto the page where it
can be argued with, instead of leaving it in someone's head where it
silently sets the answer anyway.
:::

### Running the structural version

The same session, the same people, often the same hour. The mechanics
are lighter, because you are drawing rather than quantifying:

1.  **Name the outcome and the thing you might change.** *Attrition, and
    the manager-coaching programme.*
2.  **Ask what else affects the outcome.** Write each on the board. Do
    not filter yet.
3.  **Ask what affects those.** This is where disagreement surfaces, and
    where the useful conversation is.
4.  **Draw the arrows, and read them back.** "So you're saying tenure
    affects both whether someone gets coaching *and* whether they leave
    anyway?" — that sentence is what turns a vague belief into a
    checkable claim.
5.  **Ask what is missing and unmeasured.** Motivation, manager support,
    team context. Draw those too, marked as unobserved. They are the
    reason your estimate will need a caveat, and it is better to know
    which caveat before you write it.
6.  **Take the picture away and run `adjustmentSets()` on it.**

Step 4 is where the value is, in the same way the feedback loop earlier
in this chapter is where the value is for numbers. Two HR business
partners who both nod along to "engagement drives retention" will draw
different arrows the moment you ask whether manager quality acts
*through* engagement or alongside it. That disagreement is worth more
than either answer, and it never surfaces if you draw the graph alone at
your desk.

::: callout-note
There is a practical bonus, and it is the reason this converts sceptics.
The variable list in your model stops being a methodological choice you
have to defend and becomes **a decision the business made**, on a
whiteboard, with witnesses. When somebody later asks why you controlled
for tenure but not for performance rating, the answer is a picture they
helped draw.
:::

------------------------------------------------------------------------

### Where this sits in the workflow

This chapter inverts Chapter 10's step 5. Everywhere else in the book the
prior is an input you choose quickly and check; here it is the **output**,
and the whole chapter is the checking.

That inversion explains the two steps that look missing. There is no
posterior predictive check, because the elicitation produces a prior
rather than a fitted model — the check that belongs here is the **prior
predictive check**, which the chapter runs and shows back to the expert,
and which matters more than usual precisely because an informative prior
is capable of doing damage a weak one cannot. And there are no
convergence diagnostics, because nothing was sampled.

Once the elicited prior goes into a real model, every step of the
workflow applies again in full. The elicited prior does not earn an
exemption; if anything it raises the bar, because you now have to defend
where the numbers came from as well as what came out.

------------------------------------------------------------------------

## A short checklist for running one of these sessions

::: callout-tip
## Before you next need a prior nobody has data for

1.  **Ask the plausible limits before the best guess**, then the
    median, then the quartiles — the order is what reduces anchoring.
    Don't ask for tail percentiles.
2.  **Elicit from each expert independently first**, share anonymously,
    and discuss the outliers before a second round — pool only at the
    end.
3.  **Always show the fitted curve back and ask if it looks right** —
    the feedback loop, not the first fit, is the point.
4.  **Document the elicited values and each expert's stated reasoning**
    — a useful audit trail if the model's conclusions are challenged
    later.
5.  **Treat wide disagreement as a finding**, not noise to average
    away — and check the pooled curve isn't bimodal before you use it.
6.  **Rerun the model per expert's prior before you pool**, and check
    whether the conclusion actually depends on whose judgement you
    used — it's a better answer to "does my opinion matter?" than the
    pooled number alone.
7.  **Compare prior widths across competing analysis requests** before
    committing analyst time — a wide, high-stakes prior is where an
    analysis earns its keep.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

Every technique in this book — empirical Bayes, shrinkage, hierarchical
pay models, survival analysis, ordinal survey models, Bayesian A/B
tests — needs a prior somewhere. Most of the time a sensible default or
a data-driven estimate is exactly right. But the moments that matter
most in People Analytics — a genuinely new intervention, a decision
with real stakes and no precedent — are usually exactly the moments
where a data-driven prior doesn't exist and a default feels arbitrary.
That's when this chapter's tools earn their place, and when doing the
work *with* the people who'll act on the answer is worth the extra
hour it takes.

There is one more thing you can elicit, and the next chapter is built
on it: what each possible outcome is actually worth.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  Expert judgement is a legitimate source for a prior — **elicitation**
    is the structured process for extracting it properly.
2.  Ask calibrated questions, not "what's your probability
    distribution." Use **SHELF's quartile method** — limits, then
    median, then quartiles — and avoid tail percentiles such as the
    10/90 or Hubbard's 90% interval, because the format you choose
    moves the answer more than the choice of expert does.
3.  `SHELF::fitdist()` turns those judgements into a formal
    distribution; `plotfit()` and `feedback()` close the loop by
    checking it against what the expert actually meant.
4.  Multiple experts' elicited judgements can be **pooled** — and
    disagreement between them is itself a useful finding, especially
    once discussed in a Delphi-style second round. Look at the pooled
    curve before using it: averaging confident disagreement produces a
    **bimodal** prior that nobody in the room actually holds.
5.  Instead of pooling, you can rerun the model **once per expert's
    prior** and compare the results — a direct, honest answer to
    whether the data makes the conclusion robust to whose judgement you
    used, and a compelling thing to show back to the experts themselves.
6.  Running this process *with* stakeholders is as much about buy-in
    and shared ownership of the result as it is about getting a better
    number.
7.  A wide elicited prior signals a question worth analysing further; a
    narrow one signals the organisation already agrees on the answer —
    a simple, practical way to prioritise competing analysis requests.
    Formally these are four distinct quantities: **EVPI** screens,
    **EVPPI** triages, **EVSI** prices a specific study, **ENBS**
    decides.
8.  An elicited prior needs a **prior predictive check** like any other —
    more so, because an informative prior is capable of doing real damage.
    Show that check back to the expert as well: three answers about
    where a range splits imply a distribution of *outcomes* that nobody
    translates in their head.
:::

## Next chapter {.unnumbered}

**From posterior to decision.** You have spent this chapter putting a
distribution on something nobody had data for. The last chapter does it
once more — on what each outcome is *worth* — and then spends the
posterior.
