# From Posterior to Decision: What an Analysis Is Worth

## Welcome back {.unnumbered}

Every chapter so far has ended in the same place. A posterior, an
interval, and a sentence you can say out loud without misleading
anybody.

Nobody acts on a posterior.

The decision at the end of a People Analytics project is almost always
binary and almost always financial: do we fund this, or not. Between
the posterior and that decision sits a step this book has not taken
yet, and which most statistics books never take at all — working out
what each possible outcome is **worth**, and spending the posterior
against it.

### The complaint this chapter is about {.unnumbered}

People Analytics has spent fifteen years being asked to demonstrate
impact, and has mostly answered in one of two unsatisfying ways.

The first is a correlation dressed as a consequence: *engaged teams
sell 21% more*, with no account of what would happen if you
intervened. Chapter 13 dealt with why that sentence usually cannot
support the weight put on it.

The second is worse, and more common in board packs: a single
confident number. *This programme delivered a 3.2x return.* It is
arrived at by multiplying point estimates together, it carries no
uncertainty, and everybody in the room knows it is decorative. The
finance function knows it best of all, which is why the number is
politely received and quietly discounted.

The bridge is not more rigour in the estimate. It is a **loss
function** — a statement of what each outcome costs — evaluated over
the posterior you already have. What comes out is still a return, but
it comes out with an interval, and an interval is the thing a finance
director will actually argue with rather than ignore.

::: callout-important
This chapter is the one I would most like a reader who *commissions*
analysis to read. Almost none of it is statistics. It is about what you have to write down before an analysis can change anything, and most of that writing down happens before the model is fitted.
:::

## What you'll be able to do by the end {.unnumbered}

1.  Explain why a posterior is an input to a decision, not a decision
2.  Put an honest **distribution** on the cost of an outcome nobody has
    priced, and get it vouched for
3.  Run a **decision tree on posterior draws**, and read the
    distribution of net benefit rather than a single expected value
4.  Derive an action **threshold** from the relative cost of the two
    mistakes, instead of from convention
5.  Compute **EVPI** and **EVPPI** directly from draws, and say which
    uncertainty is worth paying to reduce
6.  Recognise when a **wide** posterior is an asset rather than a
    problem

------------------------------------------------------------------------

## Setup

```{r setup}
library(tidyverse)
library(peopleanalyticsdata)
library(brms)
library(tidybayes)
library(SHELF)          # Chapter 23's elicitation tool, used again here

theme_set(theme_minimal(base_size = 13))
set.seed(2026)

# Shared colour tokens (matching theme/academicdesign*.scss). The navy ramp
# carries the data; red is reserved for reference lines and annotations,
# never for a second data series.
navy       <- "#122a52"
navy_mid   <- "#3d68a8"
navy_light <- "#8fabd0"
red        <- "#d32f2f"

data("managers", package = "peopleanalyticsdata")

managers <- managers |>
  drop_na(transfers, group_size, employee_id, city, performance_group)
```

We return to Chapter 15's managers, and to Chapter 15's model — the
partially pooled transfer rate, with `group_size` as the exposure:

```{r refit-shrink}
fit_shrink <- brm(
  transfers ~ 1 + (1 | employee_id) + offset(log(group_size)),
  data = managers, family = poisson(),
  prior = c(prior(normal(-2.5, 1), class = Intercept),
            prior(exponential(2), class = sd)),
  chains = 4, iter = 2000, seed = 13, refresh = 0
)
```

Chapter 15 stopped where most such analyses stop: a fair ranking, and
a warning about publishing it. The question nobody asked there is the
one the HR Director actually has.

> *"We can afford to put about thirty managers through the coaching
> programme. Which thirty, and is it worth doing at all?"*

------------------------------------------------------------------------

## Part 1 — The multiplication nobody does

A posterior tells you what is plausible. A decision needs one more
thing: what each outcome is worth. Multiply them together, average
over the posterior, and you have the expected value of an action.

That is the whole of decision theory in a sentence, and the reason it
is not routine in People Analytics is not that the mathematics is
hard. It is one line of R. The reason is that the second input — the
cost — is the one nobody will write down.

::: callout-note
## A different lens: this predates all of it

Choosing the action that minimises expected loss is not a Bayesian
invention. Decision analysis was assembled at Stanford and Harvard in
the 1960s — Raiffa and Schlaifer's *Applied Statistical Decision
Theory* in 1961, Howard's work on decision analysis from 1966 — and
decision trees were being drawn for business audiences in the *Harvard
Business Review* by 1964.

What is genuinely Bayesian is the *input*. Expected loss is an average
over a probability distribution for the unknown quantity. A
frequentist analysis does not produce one — a confidence interval is a
statement about the procedure, not a distribution over the parameter,
as Chapter 5 laboured. So the classical route has to substitute a point
estimate and lose the uncertainty at exactly the moment it matters
most.

This is the strongest practical argument in the book for working the
Bayesian way, and it has taken twenty-three chapters to be able to
make it.
:::

There is also an uncomfortable piece of the field's own history here.
HR very nearly got there first: Cronbach and Gleser reframed selection
testing as statistical decision theory in the 1950s, and the utility
formula that came out of it is still taught. The field kept the
formula and dropped the framework — which is why utility analysis is
remembered as an arithmetic exercise about assessment methods, rather
than as the decision-theoretic apparatus it was drawn from.

------------------------------------------------------------------------

## Part 2 — Pricing an outcome nobody has priced

Here is the objection, and it is a fair one:

> *"You cannot put a number on somebody resigning."*

You cannot put **a** number on it. That was never the requirement.
Everything else in this book is a quantity we are uncertain about,
handled by putting a distribution on it and carrying the uncertainty
through. A cost is no different. The demand for a single defensible
figure is what makes the exercise feel impossible, and it is a demand
nobody made.

So elicit it, exactly as Chapter 23 elicited a prior.

### Three questions, asked of the right person

The right person is not you. It is whoever owns the budget that
absorbs the cost — usually a finance business partner, sometimes the
function head.

What we want from them is the **net** cost of a departure: what it
costs to replace someone, less whatever we save by no longer employing
them. That framing matters, because the net figure is not always
positive. If we were paying somebody more than they were delivering,
their leaving is worth something, and a loss function is perfectly
capable of carrying a negative loss.

> *When someone in this population leaves and we replace them, what
> does it cost us net of what we stop paying? Start with the range
> you'd be astonished to fall outside. Now split it: what figure is it
> equally likely to be above or below? And what splits each half
> again?*

Those are Chapter 23's quartile questions, asked about a cost instead
of an effect. Say the answers come back at £10,000, £15,000 and
£23,000 — recruitment fees, the hiring manager's time, the ramp-up
period where the replacement is paid full salary at partial
productivity — inside a plausible range running from **minus** £20,000
to £90,000. The negative end is the case where we were carrying
somebody.

### Fitting it, with the limits doing real work

Chapter 23's tool does this, and here the limits are not decoration:

```{r cost-distribution}
n_draws <- 4000

net_cost_fit <- fitdist(
  vals  = c(10000, 15000, 23000),
  probs = c(0.25, 0.50, 0.75),
  lower = -20000,                      # <1>
  upper =  90000
)

net_cost_fit$Beta                      # <2>
round(net_cost_fit$ssq, 5)             # <3>
```

1.  The plausible limits, asked for first and now used. Supplying both
    is what lets `fitdist()` offer a **Beta**, the only family in its
    list bounded at *both* ends.
2.  The two shape parameters of the fitted Beta, rescaled onto our
    interval rather than sitting on 0–1.
3.  The sum of squared error for every family it tried. Read it before
    choosing — it is the evidence for the callout below.

```{r cost-draws}
net_cost_draws <- sampleFit(net_cost_fit, n = n_draws)[, "beta"]   # <1>

quantile(net_cost_draws, c(0.25, 0.50, 0.75))                      # <2>
range(net_cost_draws)
```

1.  `sampleFit()` returns a matrix with one column per fitted family —
    and, a trap worth knowing, its column names are **not** the names
    in the fitted object. `Log.normal` becomes `lognormal`,
    `Student.t` becomes `t`. Run `colnames()` on it once rather than
    assuming; a misspelt column name here returns `NULL` rather than
    an error.
2.  Check the fit reproduces roughly what you were given, and that the
    range respects the limits. Both should hold, and the draws come
    back in pounds rather than on a 0–1 scale.

And, per the habit this book has kept since Chapter 5, look at it
before using it:

```{r plot-cost-prior}
tibble(net_cost = net_cost_draws) |>
  ggplot(aes(x = net_cost)) +
  geom_density(fill = navy, colour = navy, alpha = 0.30, linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +   # <1>
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "The elicited net cost of one departure",
    subtitle = "Red dashed = break even. Mass to its left: departures that saved us money.",
    x = "Net cost per departure (£)", y = "Density"
  )
```

1.  Zero is the value this distribution is argued against, so it takes
    the reference colour. The mass to its left is finance's honest
    admission that some departures do us a favour.

::: callout-important
## Why the Beta, when it isn't the best fit

`net_cost_fit$ssq` does not name the Beta as the closest fit — a
skew-normal fits the three points several times better, and for the
programme effect below a log-t fits better still. Taking the Beta is a
deliberate override, and it is worth saying why out loud.

Sum of squared error measures how closely a curve passes through the
three points you elicited. It says **nothing about what the curve does
beyond them**. The skew-normal and the log-t are unbounded, and the
log-t in particular has a very heavy right tail. Fitted to a proportion
that cannot exceed 1, it will happily put probability above 1 — and
every one of those draws enters the decision as a programme that more
than eliminates the problem.

So we trade a little accuracy at three points for a family that cannot
produce impossible values anywhere. That is usually the right trade,
and it is the same argument Chapter 23 made for taking `$Normal` over
`$best.fitting`: pick the family for the property you need, then check
what it cost you.

Two conditions on that. Look at the `ssq` rather than assuming the
penalty is small — if the Beta is *far* worse, the expert's judgement
may not be Beta-shaped and you should go back to them. And say what you
did, because "we used the bounded fit rather than the closest one" is
the sort of choice a reviewer is entitled to see.
:::

### The promise

The elicited distribution is only worth something if the organisation
will stand behind it. The move that makes this work in practice is
small and political rather than statistical:

::: callout-important
## Get finance to agree to the range, not to certify the number

You are not asking the finance business partner to agree that a
resignation costs £15,000. You are asking a much weaker and much more
answerable question: *would you accept this range in a business case?*

That is a question a finance partner can say yes to, because it
commits them to nothing they don't already believe, and it converts
your analysis from "HR's estimate" into "the number finance already
signed off". Every subsequent argument is then about the decision
rather than about whether you were allowed to use the figure.

Do it in writing, do it before you see the result, and record who
said it.
:::

That last point is not bureaucratic caution. A cost distribution
agreed *after* the effect is known will be negotiated, unconsciously
and in good faith, into whatever makes the answer come out the way
the room wants. This is Chapter 13's pre-commitment discipline
applied to the other input.

::: callout-warning
## Uncertainty is not the same thing as disagreement about values

A distribution says *we don't know this number*. It does not say *we
value this differently*, and the two get confused constantly.

If the Head of Sales and the Head of HR give different figures for the
cost of a resignation because one is including lost pipeline and the
other isn't, that is uncertainty about a shared quantity, and pooling
is the right response.

If they differ because one thinks the only cost that counts is cash
out of the door this year and the other thinks the cost is the
capability you no longer have, that is a **disagreement about values**,
and averaging it produces a loss function neither of them holds. The
right move there is not to pool. It is to run the decision twice, once
under each loss function, and show both. If the action is the same
either way, the disagreement was never decision-relevant and you can
say so. If it isn't, you have found the real argument, and it was
never a statistical one.
:::

### The second elicited quantity

The programme's effect needs the same treatment. Nobody has run it
yet, so there is no data — this is Chapter 23's situation exactly. The
Head of L&D gives a proportional reduction in transfer requests with a
median of 20% and quartiles at 12% and 35%, inside a range running from
no effect at all to eliminating the problem entirely:

```{r effect-distribution}
effect_fit <- fitdist(
  vals  = c(0.12, 0.20, 0.35),
  probs = c(0.25, 0.50, 0.75),
  lower = 0,                       # <1>
  upper = 1
)

effect_draws <- sampleFit(effect_fit, n = n_draws)[, "beta"]

quantile(effect_draws, c(0.25, 0.50, 0.75))
range(effect_draws)                # <2>
```

1.  A proportional reduction lives on [0, 1] by definition. Handing
    those limits to `fitdist()` is the whole fix — the Beta cannot
    produce a value outside them, so nothing downstream needs guarding.
2.  Confirm it. Every draw should sit inside the limits.

::: callout-caution
## What this replaced, and why it is worth showing

An earlier draft of this chapter fitted a lognormal to those same three
numbers by hand, then clipped anything above 1 back down to 1. It
looked harmless. It was not, and the failure is instructive enough to
keep.

A lognormal has two parameters, so fixing the median and the
inter-quartile range fixes the tail as well — you do not get to choose
it. Fitted to 0.12 / 0.20 / 0.35, it put **2.1% of draws above 1.0**:
85 in 4,000, each one clipped to exactly 1, and each one therefore
carrying the largest effect the model allows. On a density plot they
show up as a spike at the boundary that no expert ever described, and
because they sit at the maximum they dragged about 8% of the mean
effect with them. Multiplied by an unbounded cost, that produced a long
right tail on the net-benefit chart which was entirely an artefact of
the fitting choice.

The lesson is not "avoid the lognormal". It is that **a guard like
`pmin()` is a sign the family is wrong**. Clipping does not remove the
misfit, it piles it up on the boundary — where it does more damage than
leaving it alone would have done.
:::

And the programme has a price, which is the one number that genuinely
is known: say £4,000 per manager, invoiced.

------------------------------------------------------------------------

## Part 3 — The decision tree, run on draws

A decision tree is the structure of the decision: what you choose,
what happens by chance afterwards, and what each ending is worth. Drawn
for one manager, ours is small.

![The coaching decision as a classical decision tree. Squares are
choices, circles are uncertainties resolving, triangles are endings with
a value attached — notation that has barely changed since it appeared in
the *Harvard Business Review* in 1964.](../figures/decision-tree-classical.svg){#fig-tree-classical}

Read @fig-tree-classical left to right. You choose; the world then does
something you don't control; and each path ends somewhere with a number
on it. Multiply each ending by the probability of reaching it, add them
up, and the branch with the higher total is the one to take.

That is a complete method, and for sixty years it has been how this is
done. It also has a weakness this book has spent twenty-three chapters
sharpening your eye for.

::: callout-warning
## Three branches is not what you believe

The probabilities in @fig-tree-classical came from the quartiles the
Head of L&D gave us: a quarter chance of the top band, a half of the
middle, a quarter of the bottom. That is a fair summary of her
judgement, and it is still a **summary**. She does not think the
reduction is exactly 12%, 20% or 35% and nothing else. Those were the
points where her belief splits into quarters, not the only values it
can take.

The same is true of the other two nodes. Chapter 15's model gave us a
posterior for the manager's rate, not one number. The elicited cost is
a Beta spread across a range that reaches below zero, not three prices.

Collapsing each of them to a handful of branches is a choice made for
arithmetic that stopped being hard decades ago.
:::

So put a draw at each node instead, and do it four thousand times.

```{r manager-rate-draws}
manager_draws <- managers |>
  mutate(group_size_original = group_size, group_size = 1) |>
  add_epred_draws(fit_shrink, re_formula = NULL, ndraws = n_draws) |>
  ungroup() |>
  select(employee_id, group_size_original, .draw, rate = .epred)   # <1>
```

1.  `.epred` here is each manager's expected transfers **per person
    managed**, exactly as in Chapter 15 — `group_size = 1` makes the
    offset drop out. We keep the real team size separately, because the
    number of transfers a coaching programme could avoid depends on how
    many people the manager actually has.

::: callout-caution
## This frame is large, and that is a real constraint

571 managers times 4,000 draws is a little over two million rows, which
is fine on a laptop and will not be fine on ten times the managers.

If you need to cut it, reduce `ndraws` — but note that `.draw` keeps its
original numbering when `add_epred_draws()` subsamples, so
`net_cost_draws[.draw]` below would index past the end of a shorter cost
vector. Either keep the two the same length, as here, or join on `.draw`
explicitly rather than indexing by it. It is the kind of mismatch that
fails silently by recycling rather than loudly with an error.
:::

Now the multiplication. One line per node, and the whole decision
collapses into it:

```{r net-benefit-draws}
decision_draws <- manager_draws |>
  mutate(
    net_cost          = net_cost_draws[.draw],                      # <1>
    effect            = effect_draws[.draw],
    transfers_avoided = rate * group_size_original * effect,        # <2>
    benefit           = transfers_avoided * net_cost,               # <3>
    net_benefit       = benefit - 4000                              # <4>
  )
```

1.  Indexing by `.draw` pairs each posterior draw with one cost draw
    and one effect draw. This is Chapter 5's rule again: combine draw
    by draw, never summary by summary. Here the three quantities are
    independent, so any consistent pairing is valid — but doing it by
    draw index is the habit that stays correct when they aren't.
2.  Expected transfers avoided per year for this manager: their rate,
    times their team size, times the proportional reduction.
3.  What those avoided departures were worth. Note this can come out
    **negative** on draws where the net cost was negative — a departure
    we would have been better off allowing. That is the loss function
    doing its job, not a bug.
4.  Net of the programme fee. Everything is now in pounds, which is
    the unit the decision is made in.

![The same tree, with a draw at every node instead of a branch. The
skeleton is unchanged — what changes is that each path now carries one
draw from each of the three distributions, and the ending is itself a
distribution.](../figures/decision-tree-draws.svg){#fig-tree-draws}

@fig-tree-draws is @fig-tree-classical with one substitution. Same
choice, same structure, same £4,000 fee. The chance node stops being a
fork with three labelled exits and becomes four thousand paths, and the
triangle at the end of them stops being a number and becomes a shape.

### What comes out is a distribution, not a number

Take one manager with a reasonably large team and look at the whole
thing:

```{r one-manager-plot}
example_id <- managers |>
  filter(group_size >= 20) |>
  slice_max(transfers / group_size, n = 1, with_ties = FALSE) |>
  pull(employee_id)

one_manager <- decision_draws |> filter(employee_id == example_id)

x_lo <- quantile(one_manager$net_benefit, 0.005)               # <1>
x_hi <- quantile(one_manager$net_benefit, 0.99)
off_screen <- mean(one_manager$net_benefit < x_lo |
                   one_manager$net_benefit > x_hi)

ggplot(one_manager, aes(x = net_benefit)) +
  geom_histogram(bins = 90, fill = navy_light, colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = red) +
  geom_vline(xintercept = mean(one_manager$net_benefit),
             linetype = "dotted", colour = navy) +
  coord_cartesian(xlim = c(x_lo, x_hi)) +                          # <2>
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net benefit of coaching one manager",
    subtitle = paste0(
      "Red dashed = break even. Navy dotted = the expected value.\n",
      "Axis clipped; ", scales::percent(off_screen, accuracy = 0.1),
      " of draws sit outside it, nearly all in the right tail."),
    x = "Net benefit (£ per year)", y = "Draws"
  )
```

1.  This distribution has a long right tail and a thin left one, and
    left alone they squash the part you need to read into a narrow band
    in the middle. Clip both ends, not just the ugly one.
2.  `coord_cartesian()` zooms rather than filters — the bars and the
    mean are still computed on every draw, so the dotted line stays in
    the right place. Using `xlim()` instead would silently *drop* the
    tail and move the mean, which is a different chart making a
    different claim.

::: callout-warning
## Never clip an axis without saying so

The subtitle reports what fell off the edge, and that is not
politeness. A clipped axis with no note is indistinguishable from a
distribution that genuinely stops there, and the reader has no way to
tell which they are looking at.

Chapter 5 made the same point about prior predictive plots. The rule is
the same wherever it comes up: clip for legibility, then say what you
clipped.
:::

```{r one-manager-summary}
one_manager |>
  summarise(
    expected_value = mean(net_benefit),
    p_worth_it     = mean(net_benefit > 0),
    lower          = quantile(net_benefit, 0.025),
    upper          = quantile(net_benefit, 0.975)
  )
```

::: callout-important
## Say it in plain English

*"Coaching this manager is worth about [expected value] a year on our
central estimate, with a 95% range from [lower] to [upper]. There is a
[p_worth_it] probability it pays for itself."*

The single expected value is the number you would have reported
without any of this. It is not wrong. It is one summary of the shape
above, and on its own it hides the two things the decision-maker
needs: how often this loses money, and how badly.
:::

This is what health economists call **probabilistic sensitivity
analysis**, and it has been standard practice in drug reimbursement
decisions for twenty years. There is no methodological innovation here
for us to claim. The only novelty is doing it to an HR decision.

------------------------------------------------------------------------

## Part 4 — Which thirty, and where the threshold comes from

The HR Director asked for thirty managers. Rank by expected net
benefit and take the top thirty:

```{r rank-managers}
by_manager <- decision_draws |>
  group_by(employee_id, group_size_original) |>
  summarise(
    expected_net = mean(net_benefit),
    p_positive   = mean(net_benefit > 0),
    .groups = "drop"
  ) |>
  arrange(desc(expected_net))

by_manager |> head(10)
```

```{r selected-set}
selected <- by_manager |> slice_head(n = 30)

selected |>
  summarise(
    total_expected_net = sum(expected_net),
    worst_p_positive   = min(p_positive),
    expected_wrong     = sum(1 - p_positive)     # <1>
  )
```

1.  The expected number of managers in the chosen thirty for whom this
    will turn out to have been money wasted. It is the sum of the
    per-manager probabilities of being wrong, and it is the number
    nobody ever puts on the slide.

::: callout-caution
## The number you will be asked about later

`expected_wrong` is uncomfortable and worth reporting anyway. Here it
comes out near **fourteen of thirty** — so on our own numbers, close to
half the programme is expected to be money that did not come back. When
two or three of those surface in a year's time, you want to have said
so in advance rather than to be explaining it afterwards.

Selecting the top of a ranking guarantees some of this. The units that
rank highest are disproportionately the ones whose estimates ran lucky,
and Chapter 15's shrinkage reduces that effect without removing it. A
selected set is always worse than it looked at selection time. Saying
by how much, up front, is the difference between a forecast and an
excuse.
:::

### So when is fourteen too many?

It is the obvious next question and it has a slightly surprising
answer: **on its own, never.** The count of wrong calls is not a
decision quantity, and there is no threshold on it.

Look at where it came from. The previous section derived an action
threshold from the ratio of two costs, and with a cheap intervention
and an expensive miss it landed well below 50%. If you have decided to
act whenever the probability of benefit exceeds — say — a third, then
by construction you will be wrong on a good share of the set. That is
not the rule misfiring. It is the rule working.

::: callout-important
A high `expected_wrong` and a low action threshold are **the same fact
stated twice**. Being alarmed by the first while having deliberately
chosen the second is being alarmed by your own loss function.
:::

The quantity that *does* decide how many to fund is the marginal
manager. Keep adding while the next one still has positive expected
net benefit, and stop when they don't:

```{r marginal-rule}
by_manager |>
  mutate(rank = row_number(),                       # <1>
         cumulative_net = cumsum(expected_net)) |>
  filter(rank %in% c(10, 20, 30, 40, 60, 100)) |>
  select(rank, expected_net, p_positive, cumulative_net)

sum(by_manager$expected_net > 0)                    # <2>
```

1.  `by_manager` is already sorted by expected net benefit, so the row
    number is the rank.
2.  How many managers are worth coaching *at all*. If this is larger
    than thirty, the constraint is the budget rather than the quality
    of the candidates — which is a much better problem, and a
    much better slide.

Read the two columns together. As long as `expected_net` is still
positive at rank 30, the thirtieth manager is a positive-value
purchase and the programme is budget-constrained, not quality-
constrained. The misses are the price of the hits, and
`cumulative_net` is what the portfolio bought.

::: callout-warning
## Two things that should stop you, and neither is the count

**Concentration.** A portfolio whose total rests on two or three
managers is fragile in a way the headline number hides. Check it:

```{r concentration}
selected |>
  summarise(
    top_3_share = sum(sort(expected_net, decreasing = TRUE)[1:3]) /
                  sum(expected_net)
  )
```

If three of the thirty carry most of the value, you do not have a
programme, you have three bets and twenty-seven passengers — and the
honest recommendation may be to fund the three properly and think again
about the rest.

**A cost you left out of the loss function.** Fourteen visible misses
can be survivable financially and fatal politically. A programme widely
believed to have failed does not get renewed, and the next one is
harder to fund — which is a real cost that our loss function says
nothing about, because we never elicited it.

That is a legitimate reason to fund fewer, higher-confidence managers
than the arithmetic recommends. It is not a legitimate reason to
pretend the arithmetic said something else. If reputational risk is
carrying the decision, elicit *that* as a cost and put it in, or state
plainly that you are overriding the model and why. Chapter 13's point,
one layer up: the decision you can defend is the one where you can say
which considerations were priced and which were not.
:::

### Does the tail choose who gets coached?

That net-benefit distribution is strongly right-skewed, and the mean of
a skewed distribution sits above its middle. Since we are ranking on
the mean, it is worth asking whether the ranking is being set by the
tail rather than by the bulk — whether we are funding the managers with
the widest posteriors instead of the highest rates.

There is a direct test. Rank on the median instead, which the tail
barely moves, and see how much the chosen set changes:

```{r rank-robustness}
alt <- decision_draws |>
  group_by(employee_id) |>
  summarise(mean_net   = mean(net_benefit),
            median_net = median(net_benefit),
            .groups = "drop")

length(intersect(
  alt |> slice_max(mean_net,   n = 30) |> pull(employee_id),
  alt |> slice_max(median_net, n = 30) |> pull(employee_id)
))
```

Twenty-eight of the thirty are the same people. The tail is ugly and it
is not deciding anything.

::: callout-important
## The right question about an alarming number

Summarising every manager against every draw gives a maximum transfer
rate above 1 — more than one request per person managed. That looks
wrong, and chasing it is the natural instinct.

Two things make it less alarming than it appears. It is the largest of
**two and a bit million** values, so it is an extreme order statistic
rather than a claim about any particular manager; and it is *not
impossible*, because the denominator is average headcount over a period
and a person can ask to move more than once. Unlike the effect in Part
2, which genuinely could not exceed 1, there is nothing here to bound.

The extreme draws belong to managers with five to thirteen reports —
which is Chapter 15's argument arriving at the decision layer. Few
reports, wide posterior.

But none of that is the point. The question is never *"is this number
uncomfortable?"*, it is **"does it change what we do?"** — and the
check above answers it in five lines. Twenty-eight of thirty. It
doesn't.

Had the overlap come out at twenty, the conclusion would have been the
opposite and the fix would have been a decision, not a repair: rank on
a lower quantile of net benefit, and say openly that you would rather
fund managers you are confident about than managers you are unsure
about. That is a legitimate risk-averse loss function, chosen
deliberately.

One caution on that, since it is easy to over-learn. Expected value is
what decision theory tells you to maximise, so the **mean is the right
quantity** and the median is a robustness check, not a replacement.
Switch to a quantile only when you have decided you are risk-averse and
are prepared to say so.
:::



### Where the threshold actually comes from

"Act when the expected net benefit is positive" is one rule, and it is
the right one when the two mistakes cost the same. They rarely do.

Suppose the coaching budget is uncommitted and could be returned; the
cost of coaching a manager who didn't need it is then just the £4,000.
But the cost of *not* coaching a manager who did need it is a year of
avoidable transfers — which the model says is often several times
larger. The two errors are not symmetric, and the break-even rule
quietly assumes they are.

```{r threshold}
cost_of_acting_wrongly    <- 4000
cost_of_not_acting_wrongly <- mean(one_manager$benefit)

threshold <- cost_of_acting_wrongly /
  (cost_of_acting_wrongly + cost_of_not_acting_wrongly)

threshold
```

Act when the probability of benefit exceeds that threshold. With a
cheap intervention and an expensive miss, it comes out well below
50% — which says you should coach managers who are quite unlikely to
need it, because the occasional hit pays for the many misses.

::: callout-important
That is the whole of the "should we set the cut at 0.5?" question, and
the answer has nothing to do with statistics. The threshold is a ratio
of two costs. Anybody who sets it at 0.5, or at 0.05, without writing
those costs down has made a financial decision by accident.
:::

::: callout-tip
## When you can skip this

Most analyses do not need any of Part 3 or Part 4, and it is worth
being able to say why rather than skipping it from habit.

**Skip it when the action is already decided.** If the programme is
running whatever you find, an expected-value calculation is theatre.
Say so and spend the time elsewhere.

**Skip it when the decision is not reversible in the relevant sense.**
Some decisions are made once, on values rather than returns — whether
to publish a pay gap, for instance. Pricing them is a category error.

**Don't skip it when the money is contested**, which in practice means
whenever another function wants the same budget. This is the only
form in which a People Analytics result competes on equal terms with a
capital request, and declining to put it in that form is choosing to
lose the argument politely.

And if you skip it, say what you skipped. "We have not attempted to
value the outcome, so this shows effect size rather than return"
costs one sentence and prevents someone else supplying the missing
multiplication badly.
:::

------------------------------------------------------------------------

## Part 5 — What is it worth to know more?

The decision above rests on three uncertain inputs: the manager's
underlying rate, the cost of a transfer, and the programme's effect.
Each could be pinned down further, at a price. A better cost figure
means a week of finance's time. A better effect estimate means
running a pilot, which means Chapter 22.

Which one is worth buying?

This has a precise answer, and — the pleasant surprise — it computes
directly from the draws you already have.

### EVPI: is any further evidence worth anything?

Today you must choose without knowing how things will turn out, so
you choose the action with the best average. If instead you knew the
truth before choosing, you would pick the better action every time.
The difference between those two is the **expected value of perfect
information**.

```{r evpi}
per_draw <- decision_draws |>
  filter(employee_id == example_id) |>
  pull(net_benefit)

decide_now      <- max(mean(per_draw), 0)      # <1>
decide_knowing  <- mean(pmax(per_draw, 0))     # <2>

evpi <- decide_knowing - decide_now
evpi
```

1.  Choose once, using the average. Coaching if its mean net benefit
    beats doing nothing, which is worth zero.
2.  Choose per draw, as if you already knew which world you were in.
    You would coach only in the draws where it pays.

::: callout-note
EVPI is an **upper bound**, and that is what makes it useful as a
screen. It is what you would pay to eliminate *all* the uncertainty,
perfectly and instantly. No real study delivers that. So if EVPI comes
out below the cost of the cheapest study you could run, stop — no
amount of further evidence can pay for itself, and you should act on
what you have.
:::

### EVPPI: which uncertainty is carrying the decision?

EVPI says whether to buy information. It doesn't say *what* to buy.
For that, ask what it would be worth to resolve one input completely
while the others stay uncertain — the **expected value of partial
perfect information**.

```{r evppi}
evppi_for <- function(net, parameter, n_bins = 30) {
  binned <- tibble(net = net, parameter = parameter) |>
    mutate(bin = ntile(parameter, n_bins)) |>          # <1>
    group_by(bin) |>
    summarise(best_in_bin = max(mean(net), 0),         # <2>
              .groups = "drop")

  mean(binned$best_in_bin) - max(mean(net), 0)         # <3>
}

evppi_table <- tibble(
  input = c("Net cost per departure", "Programme effect", "Manager's rate"),
  evppi = c(
    evppi_for(one_manager$net_benefit, one_manager$net_cost),
    evppi_for(one_manager$net_benefit, one_manager$effect),
    evppi_for(one_manager$net_benefit, one_manager$rate)
  )
) |>
  arrange(desc(evppi))

evppi_table
```

1.  Group the draws by the input we're imagining we could learn. Within
    a bin that input is approximately known, so the bin mean is what
    you'd expect having learned it.
2.  Having learned it, choose the better action for that bin.
3.  The value of deciding *after* learning this input, minus the value
    of deciding now. Same shape as the EVPI calculation above — only
    the thing being learned has changed.

::: callout-warning
The binning above is the simple version, and it is an approximation
that gets noisy when the number of draws is modest or the bins are
many. The published method regresses net benefit on the parameter
instead of binning it, which is both more stable and more work.

Use the binned version to see the shape of the answer — which input
dominates, and by roughly how much. Don't quote its third significant
figure at anyone.
:::

Plot the three side by side and you have the artefact worth taking to
a steering group:

```{r evppi-tornado}
evppi_table |>
  ggplot(aes(x = evppi, y = fct_reorder(input, evppi))) +
  geom_col(fill = navy) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "What it would be worth to resolve each uncertainty",
    subtitle = "Per manager, per year. The top bar is what to go and find out.",
    x = "Expected value of partial perfect information (£)", y = NULL
  )
```

::: callout-important
Read it for the gaps, not the order. The programme effect comes top,
which is the answer an analyst would have guessed — but the **net cost
is barely behind it**, and the manager's own rate, the only one of the
three we have a fitted model for, is worth markedly less to resolve
than either.

That second place is the useful finding. Resolving the effect means
running a pilot: months, a budget, and Chapter 22's design work.
Resolving the cost means a better afternoon with finance. Almost the
same value, at a fraction of the cost — and it is the one nobody puts
on a research plan.

So the question the chart really answers is not *which uncertainty is
largest* but **which is largest per pound of effort to reduce**. A
People Analytics function that asked that before commissioning work
would reorder its own queue.
:::

### The two you'll need names for

EVPI and EVPPI both imagine *perfect* information. Real studies give
you noisy information from a finite sample, and pricing a specific
study — this design, this many participants — is **EVSI**, the
expected value of sample information. Subtract what the study costs,
across everyone it would affect, and you have **ENBS**, the expected
net benefit of sampling, which is the quantity that actually says run
it or don't.

::: callout-note
## Four names, four questions, one order

| | Question it answers |
|---|---|
| **EVPI** | Is any further evidence worth anything? *Screen.* |
| **EVPPI** | Which input carries the decision? *Triage.* |
| **EVSI** | What is this specific study worth? *Price.* |
| **ENBS** | Does it beat what it costs? *Decide.* |

They are easy to mix up, and the commonest slip — one I have made — is
to say EVPPI when you mean EVSI. EVPPI points at a *parameter*; EVSI
prices an *attempt to learn it*.

This book computes the first two, because they need nothing but the
draws in front of you. EVSI needs you to simulate the study you would
run, fit the model you would fit, and see how much the posterior
narrows — which is Chapter 22's design analysis with a price attached.
:::

------------------------------------------------------------------------

## Part 6 — When a wide posterior is an asset

Every chapter so far has treated uncertainty as a cost. Wider
interval, weaker claim, less you can say. Under a decision framing
that stops being reliably true, and the exception is worth ending on.

Consider two candidates for a role. One is a known quantity who will
be solidly average. The other might be excellent or might be poor,
with equal odds and the same average. For a single fixed period the
firm is indifferent between them — same mean, and the mean is what
you get.

Now let the firm act on what it learns. If the second candidate turns
out well you keep them; if they turn out badly you part company. The
upside is retained and the downside is truncated, so the risky
candidate is worth **more** than the safe one, by an amount that grows
with the spread. Variance has become an asset.

Hiring is where this argument is usually made, and probation is the
mechanism that makes it work. But nothing in it is about hiring.

::: callout-important
## The catch, which is the interesting half

Option value only exists if you can act on the downside. The more it
costs to correct the mistake — long notice periods, strong employment
protection, a culture that never exits anyone, a manager who will not
admit the hire was wrong — the smaller the option is worth, until at
some point the safe candidate is genuinely the better bet.

So an organisation's appetite for high-variance talent is determined
by its ability to correct mistakes, not by how bold its hiring
language is. "We hire for potential" is a claim about your exit
process, and it is usually being made by firms whose exit process
cannot support it.
:::

### The same shape, on the decision we have been making

Our coaching programme has the identical structure, with a pilot
playing the part probation played above. Committing the whole budget
now buys you `max(mean(net), 0)`. Running a pilot first, and keeping
the right to stop, buys you `mean(pmax(net, 0))` — you only continue
in the worlds where continuing pays.

Those are the two quantities from the EVPI calculation, in the same
order.

![A staged version of the coaching decision. The second square — the
choice you still get to make once the pilot has told you something — is
where the option value lives.](../figures/decision-tree-staged.svg){#fig-tree-staged}

Compare @fig-tree-staged with @fig-tree-classical. The lower branch is
the decision we have been making all chapter: choose once, live with
it. The upper branch defers, and the extra square is the whole of the
difference. Take that square away — no pilot, no ability to stop the
rollout, no willingness to admit the programme isn't working — and the
two branches become the same decision.

The hiring version of @fig-tree-staged is the same drawing with
different words on it: *hire* in place of *pilot*, *probation period*
in place of *what the pilot shows*, and *confirm or exit* in the second
square.

::: callout-note
I remember a trading desk manager making just this decision when I worked as the European Graduate Recruitment Manager for the Markets division of a large, American investment bank. We had got through the whole of the recruitment process and the manager genuinely couldn't determine whether the remaining candidate was going to be brilliant or a disaster. In the end he told me to extend an offer. His logic was if the candidate was brilliant he didn't want to risk his competitors getting him. If he was poor he could always get rid of him. What made my internal client a brilliant trader was exactly the ability to price risk.
:::

::: callout-important
**The value of being able to change your mind is the value of
information.** They are the same number, arrived at from opposite
directions — one asks what you'd pay to learn, the other what you'd
pay to stay flexible. Which means a staged rollout is not caution, and
it is not indecision. It is a purchase, it has a price, and Part 5
tells you what it is worth.
:::

------------------------------------------------------------------------

## Where this chapter sits in the workflow

::: callout-note
## Chapter 10's steps, and the two that change shape here

Nothing in Chapter 10's eleven steps is skipped — the model in this
chapter is Chapter 15's, diagnosed and checked there. Two steps do
change shape, and both changes are the point of the chapter.

**Step 2, the claim, becomes financial.** Everywhere else the claim
was about a quantity — a rate, a difference, an effect. Here it is
about an action and its price, and it has to be written down before
anything is fitted, because both cost inputs must be agreed before the
result is known.

**Step 11, communicate, changes units.** The posterior arrives in
transfers per person and leaves in pounds per year. That translation
is where the elicited distributions enter, and it is the step that
makes the analysis legible to people who will never look at an
interval.
:::

------------------------------------------------------------------------

## On the job {.unnumbered}

::: callout-important
## Why this matters day to day

People Analytics is asked to demonstrate impact and answers with
either a correlation or an invented multiple. This chapter is the
third option, and its components are unremarkable individually: a
posterior you already have, two elicited distributions, one
multiplication.

What makes it work in a room is not the arithmetic. It is that the
cost distribution was agreed with finance in advance, and that the
answer arrives with an interval instead of a decimal point. A number
with an honest range invites an argument about the decision. A number
without one invites an argument about the number.

The single most useful habit here is also the cheapest: before the
analysis, ask which of the two mistakes is worse and roughly by how
much. You can do that in a meeting, without any of the machinery
above, and it will change more analyses than the machinery will.
:::

::: callout-caution
## The honest close

None of this is new, and it is not being withheld from People
Analytics by anybody. Health economics built this apparatus, made it
mandatory for reimbursement decisions, and *still* does not use it
routinely. Food safety has formal guidance for it. Conservation
biology teaches it with spreadsheets.

The obstacle is not technical, and you should not expect adopting it
to be a matter of showing people the chart. Expect instead that the
first time you ask a finance partner for three numbers, the interesting
part will be the conversation, not the distribution you fit to it.
:::

------------------------------------------------------------------------

## Summary {.unnumbered}

::: callout-note
## Today you learned

1.  A posterior is an **input to a decision, not a decision**. The
    missing step is what each outcome is worth.
2.  You can price outcomes nobody has priced — not with a number, but
    with an **elicited distribution**, using Chapter 23's machinery.
3.  Get finance to **vouch for the range** rather than certify the
    number, in writing, before the result is known.
4.  Uncertainty about a cost and **disagreement about a value** are
    different things. Pool the first; run the decision twice for the
    second.
5.  Running a decision tree on **posterior draws** gives a distribution
    of net benefit rather than a single expected value. Health
    economics calls this probabilistic sensitivity analysis.
6.  An action **threshold is a ratio of two costs**, not a convention.
    Setting it at 0.5 without writing the costs down is a financial
    decision made by accident.
7.  Report the **expected number of wrong calls** in a selected set — a
    selected set is always worse than it looked at selection. But a high
    miss count is not a reason to stop: it is **the same fact as a low
    action threshold**. What decides how many to fund is the marginal
    unit's expected net benefit, not the miss rate.
8.  When a number looks alarming, the test is not whether it is
    comfortable but **whether it changes the decision**. Re-rank on a
    summary the tail doesn't move, and compare the chosen sets.
9.  **EVPI** screens whether evidence is worth anything, **EVPPI**
    says which input to go after, **EVSI** prices a specific study and
    **ENBS** decides. The first two compute directly from draws.
10. Under a decision framing a **wide posterior can be an asset** —
    but only to the extent you can act on the downside. The value of
    staying flexible and the value of information are the same number.
:::

## And that's the book {.unnumbered}

You started by asking what a plausible range for a single number was,
and finished by pricing what it would be worth to narrow one. In
between, every model was the same `brm()` call with different
arguments, and every answer was a distribution you could ask any
question of.

If one habit survives, make it the one this chapter ends on: **before
you analyse, ask what you would do differently depending on the
answer, and what each wrong turn would cost.** Ask it out loud, in the
meeting, before anyone opens R. Most of what matters follows from
that, and the arithmetic is the easy part.

------------------------------------------------------------------------

## Before you go {.unnumbered}

### This one is still being written

You have reached the end of a book that isn't finished. Chapters get
revised as I teach them and as I write about the same ideas elsewhere,
so what you have just read is a version rather than a final text.

That makes corrections genuinely useful rather than merely polite. If
something is wrong, unclear, or simply doesn't work when you run it,
I would rather hear about it. There are three ways, in ascending order
of effort:

-   **Edit this page.** Every page has an *Edit this page* link in the
    right-hand margin, which takes you to the source on GitHub.
-   **Open an issue**, using the *Report an issue* link beside it — best
    for "this is wrong" or "this didn't run", where a fix isn't obvious.
-   **Email me**, at
    [andrew@andrewmarritt.ch](mailto:andrew@andrewmarritt.ch).

Typos are welcome. Arguments are more welcome. If you think a chapter
reaches the wrong conclusion, that is the most useful message I can
receive, and it will not be the first time.

### Working Ideas

Much of the thinking here started as writing in my newsletter,
**Working Ideas**, at
[andrewmarritt.substack.com](https://andrewmarritt.substack.com). It
takes a deliberately multi-disciplinary view of work — personnel
economics, psychology, decision analysis, and statistics tend to study
the same problems in isolation from each other, and a lot of what looks
novel in People Analytics turns out to be well understood somewhere
else.

Several chapters here have a newsletter essay behind them, and the two
will keep feeding each other. If this book was useful, that is the
place to carry on.

### Teaching it

I teach this material, and I do it in two settings. As an **academic
course** — the first version of this book was the course my students
sat through, and their questions shaped most of it. And **inside
organisations**, with People Analytics and HR teams who want to work
this way but have nobody to get them started.

The company version is usually the more interesting one, because we can
run it on your problems rather than on the `salespeople` dataset. If
either would be useful to you, do get in touch.

### Free, and how to support it if you want to {#supporting-the-book}

The online version of this book is free, and will stay free. There is
no paywalled edition and no upsell — I would rather it were read.

If it saved you time and you would like to put something behind it, the
simplest way is a **paid subscription to Working Ideas**. You get the
newsletter, and it funds the time that goes into projects like this one.
There is no obligation whatsoever, and nothing in the book is withheld
from anyone who doesn't.

### Getting in touch

For teaching enquiries, corrections, disagreements, or anything else:

> **Andrew Marritt**\
> [andrew@andrewmarritt.ch](mailto:andrew@andrewmarritt.ch)\
> [andrewmarritt.substack.com](https://andrewmarritt.substack.com)

I read everything, and I reply to most of it.
